---
title: "Unit 3.1: Theory of Random Number Generation"
author: "Dr. Manoj C. Patil"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
    code_folding: hide
    theme: cosmo
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    echo = TRUE,
    warning = FALSE,
    message = FALSE,
    fig.align = "center",
    fig.width = 7,
    fig.height = 4.5
)

set.seed(123)
library(tidyverse)
library(ggplot2)
library(dplyr)
library(knitr)
library(kableExtra)
library(gridExtra)
library(patchwork)
```

# Learning Objectives

By the end of this unit, a student should be able to

- explain what is meant by a pseudo-random number, and why it differs
  fundamentally from a truly random number,
- list and justify the statistical and computational properties that a
  good random number generator must possess,
- derive the general linear congruential recurrence and distinguish
  its multiplicative and mixed special cases,
- state and prove the Hull-Dobell theorem, and use it to certify
  whether a given linear congruential generator attains its maximum
  possible period,
- state and prove the maximal-period result for multiplicative
  congruential generators with prime modulus,
- implement congruential generators in R, diagnose their structural
  weaknesses through simulation, and compare them against R's built-in
  generator,
- apply a defensible set of criteria when selecting or evaluating a
  random number generator for a statistical computing task.

# Prerequisites

A reader should already be comfortable with

- elementary modular arithmetic, including the notions of congruence,
  greatest common divisor, and prime factorization,
- the definition of the Uniform$(0,1)$ distribution and its role as
  the basic building block of stochastic simulation,
- basic group theory at the level of cyclic groups and the order of an
  element (used only in one proof; a first-principles explanation is
  provided),
- elementary R programming: writing functions, `for` loops, and vector
  operations.

# Introduction

Every stochastic simulation, from a simple Monte Carlo estimate of a
definite integral to a large-scale Markov chain Monte Carlo posterior
sampler, rests on a single computational primitive: the ability to
produce, on demand, a long sequence of numbers that behave as if they
were independent draws from the Uniform$(0,1)$ distribution. Every
other random variate used in simulation — Normal, Exponential,
Binomial, Dirichlet, or the trajectory of a queueing network — is
built, directly or indirectly, from this one primitive. Understanding
how it is produced is therefore not a peripheral topic in statistical
computing; it is the foundation on which the rest of the subject
stands.

Historically, statisticians did not generate random numbers by
algorithm at all. Physical devices were used instead. The RAND
Corporation's 1955 volume *A Million Random Digits with 100,000 Normal
Deviates* was compiled from the output of an electronic roulette
wheel, and earlier still, Tippett's 1927 table of random sampling
numbers was built from digits appearing in census records. Physical
generation has an appealing purity — the numbers really are the
product of a stochastic physical process — but it is slow, expensive
to produce in bulk, and, most damagingly for scientific use,
non-reproducible. If a Monte Carlo experiment produces an anomalous
result, a researcher must be able to replay the exact same sequence of
"random" inputs in order to isolate the cause. A physical table cannot
be regenerated on demand; a table lookup exhausts itself once
consumed.

The modern resolution to this tension is the pseudo-random number
generator: a wholly deterministic algorithm, implemented in software,
whose output nonetheless passes an extensive battery of statistical
tests for uniformity and independence (the subject of Unit 3.2). The
first algorithmic proposals date to the 1940s. John von Neumann's
"middle-square method" (1946) squared a seed number and extracted the
middle digits as the next value; it was recognized almost immediately
to degenerate into short cycles for many seeds, but it established the
principle that a recursive arithmetic rule could stand in for a
physical randomizer. D. H. Lehmer's 1949 proposal — multiplying the
previous value by a constant and reducing modulo a large number — is
the direct ancestor of the generators studied in this unit, and
remains, in refined form, one of the most widely deployed ideas in
computational statistics.

The practical importance of this material is threefold. First,
correctness: a simulation built on a flawed generator can produce
answers that are confidently wrong, a failure mode far more dangerous
than an outright crash, because nothing in the output signals that an
error has occurred. Second, reproducibility: setting a seed makes an
entire analysis, however elaborate, exactly repeatable, which is
essential for debugging, for peer review, and for regulatory
submissions in fields such as clinical trial simulation. Third,
efficiency: modern statistical computing routinely calls for
generating billions of variates, so the arithmetic simplicity and
speed of congruential methods remain relevant even though more
sophisticated generators (such as the Mersenne Twister used by
default in R) have since superseded them for general use.

# Intuition

Deterministic arithmetic can look unpredictable if the rule is chosen
carefully. Consider a mechanical odometer with $m$ positions, numbered
$0, 1, \dots, m-1$, connected to a gear that advances the dial by a
fixed number of teeth, $a$, every time it turns, and which is also
nudged forward by a fixed additional increment $c$ at every step. If
the dial currently reads $X_n$, the reading after the next turn is

$$
X_{n+1} = \bigl(a X_n + c\bigr) \bmod m,
$$

that is, multiply the current reading by $a$, add $c$, and keep only
the remainder after dividing by $m$. Because the rule involves both a
multiplication and a modular "wrap-around," the dial does not sweep
through its positions in the obvious ascending order; instead, for a
suitable choice of $a$, $c$, and $m$, it visits every one of the $m$
positions exactly once before returning to its starting position. To
someone who does not know the rule, the sequence of readings looks
erratic and unpredictable, even though it is generated by an entirely
deterministic three-parameter formula.

This is the core idea behind every generator studied in this unit.
"Randomness," from the machine's point of view, is not the absence of
a rule; it is the presence of a rule complicated enough, in its
interaction with modular arithmetic, that short-range structure is
destroyed while every possible value is still reached exactly once per
cycle. The mathematical problem that follows is precise and
answerable: for which choices of $a$, $c$, and $m$ does the dial visit
every position before repeating, and, separately, for which choices
does the resulting sequence also look statistically like an
independent uniform sample rather than merely a permutation of the
integers $0$ through $m-1$? The first question is answered completely
by number theory (the Hull-Dobell theorem, proved below); the second
question requires the empirical tests developed in Unit 3.2.

# Formal Definitions

## Pseudo-Random Numbers

**Definition.** A **pseudo-random number generator** (PRNG) is a
deterministic algorithm, specified by

- a finite state space $S$,
- a deterministic transition function $g: S \to S$,
- an output function $h: S \to [0,1]$,
- an initial state $x_0 \in S$, called the **seed**,

which produces a sequence $x_0, x_1, x_2, \dots$ by iterating
$x_{n+1} = g(x_n)$, and reports the observable sequence
$u_n = h(x_n)$ as the generated numbers, intended to imitate
independent draws from the Uniform$(0,1)$ distribution.

where

- $S$ denotes the (finite) set of internal states the algorithm can
  occupy; for the generators of this unit, $S = \{0, 1, \dots, m-1\}$,
- $g$ denotes the recurrence relation advancing the state,
- $h$ denotes the normalization map that converts an integer state
  into a number in $[0,1]$,
- $x_0$ denotes the seed, the only genuinely arbitrary input, usually
  chosen from the system clock or supplied explicitly for
  reproducibility.

Because $S$ is finite and $g$ is deterministic, the sequence of states
must eventually repeat: once a state recurs, the entire subsequent
trajectory recurs identically. This observation motivates the next
definition.

**Definition.** The **period** of a pseudo-random number generator,
started from seed $x_0$, is the smallest positive integer $P$ such
that $x_{n+P} = x_n$ for all sufficiently large $n$. If $P$ equals the
maximum possible value $|S|$, the size of the state space, the
generator is said to have **full period** from that seed.

A generator with a short period is unsafe for simulation, because the
"random" sequence begins repeating itself well before a realistic
simulation sample size is exhausted, silently reintroducing exact
dependence into what is meant to be an independent sample.

## Desirable Properties of a Random Number Generator

A pseudo-random number generator intended for statistical work should
satisfy the following properties.

- **Uniformity.** The generated values should be statistically
  indistinguishable from a genuine Uniform$(0,1)$ sample; no region of
  $[0,1]$ should be systematically over- or under-represented.
- **Independence.** No subsequence of outputs should be predictable,
  even approximately, from earlier outputs; in particular, consecutive
  values should show no serial correlation.
- **Long period.** The period should comfortably exceed the largest
  sample size the generator will ever be asked to produce in a single
  program run.
- **Reproducibility.** Given the same seed, the generator must produce
  the same sequence on any machine and at any time, so that
  simulations can be exactly replicated for debugging and review.
- **Efficiency.** Generating a single value should require only a
  handful of arithmetic operations, since simulations often demand
  billions of draws.
- **Portability.** The output should not depend on machine-specific
  floating point or integer arithmetic; the same seed should give the
  same sequence across hardware and software platforms.
- **Absence of lattice structure.** Tuples of consecutive outputs,
  viewed as points in $k$-dimensional space, should fill the unit
  hypercube densely rather than falling on a small number of
  parallel hyperplanes (a defect examined in detail below for the
  RANDU generator).

# Theorems

## Theorem 1.1: Hull-Dobell Theorem

**Statement.** Consider the linear congruential generator

$$
X_{n+1} = \bigl(a X_n + c\bigr) \bmod m, \qquad n = 0, 1, 2, \dots,
$$

with integer parameters $0 \le a < m$, $0 \le c < m$, modulus $m \ge
2$, and arbitrary integer seed $X_0$ with $0 \le X_0 < m$. This
generator has period exactly $m$, for every choice of seed $X_0$, if
and only if the following three conditions hold simultaneously.

1. $\gcd(c, m) = 1$.
2. $a \equiv 1 \pmod{p}$ for every prime $p$ that divides $m$.
3. $a \equiv 1 \pmod{4}$ if $4$ divides $m$.

**Assumptions.** The recurrence is exactly linear (no terms beyond
$aX_n + c$), and $m$, $a$, $c$ are fixed integers known in advance,
so that the divisibility conditions can be checked directly from the
prime factorization of $m$.

**Proof.** Unrolling the recurrence by induction on $n$ gives the
closed-form expression

$$
X_n \;=\; \Bigl(a^n X_0 \;+\; c \, S(n)\Bigr) \bmod m,
\qquad
S(n) = \sum_{i=0}^{n-1} a^i,
\tag{1.1}
$$

which is verified by substitution: $X_1 = (aX_0 + c) \bmod m$ matches
$n=1$ since $S(1)=1$, and if (1.1) holds for $n$ then
$X_{n+1} = (aX_n + c) \bmod m = \bigl(a(a^nX_0 + cS(n)) + c\bigr) \bmod
m = \bigl(a^{n+1}X_0 + c(aS(n)+1)\bigr) \bmod m$, and $aS(n)+1 =
S(n+1)$ by direct expansion of the geometric sum, completing the
induction.

The generator has full period $m$ from every seed precisely when the
map $f(x) = (ax+c) \bmod m$ is a single cyclic permutation of
$\{0,1,\dots,m-1\}$, which happens if and only if $X_m = X_0$ for
every $X_0$, and no smaller positive $n<m$ satisfies $X_n = X_0$ for
every $X_0$. Using (1.1), the requirement $X_n \equiv X_0 \pmod m$ for
every $X_0$ is equivalent to the pair of conditions

$$
a^n \equiv 1 \pmod m
\qquad\text{and}\qquad
c\,S(n) \equiv 0 \pmod m,
\tag{1.2}
$$

obtained by comparing coefficients of $X_0$ (take $X_0 = 1$ and $X_0 =
0$ in turn). Full period is therefore equivalent to $n=m$ being the
smallest positive integer satisfying (1.2).

*Sufficiency.* Suppose conditions 1-3 hold. Write $b = a - 1$; by
condition 2, $p \mid b$ for every prime $p \mid m$, and by condition
3, $4 \mid b$ whenever $4 \mid m$. A standard number-theoretic lemma
(a special case of the *Lifting-the-Exponent* result; see Knuth, 1997,
Section 3.2.1.2) states that under exactly these two divisibility
conditions on $b = a-1$, the smallest $n$ with $a^n \equiv 1 \pmod{m}$
is $n = m / \gcd(c,m) \cdot \gcd(c,m)$-independent, and in fact, for
every prime power $p^{e}$ exactly dividing $m$, $p^{e} \mid S(n)$
whenever $p^{e} \mid n$ under these same conditions on $b$. Combining
this fact prime-power by prime-power via the Chinese Remainder
Theorem shows that $a^m \equiv 1 \pmod m$ and $m \mid S(m)$
automatically, so $n=m$ satisfies (1.2). Because $\gcd(c,m)=1$
(condition 1), the congruence $c\,S(n) \equiv 0 \pmod m$ forces $m
\mid S(n)$ regardless of $c$, and the same lemma shows no smaller
positive $n$ can satisfy both parts of (1.2) simultaneously across
every prime power factor of $m$. Hence the smallest such $n$ is
exactly $m$, giving full period.

*Necessity.* Suppose the generator has full period $m$ from every
seed. Taking $X_0 = 0$ gives $X_1 = c \bmod m$, and since the orbit of
$0$ must visit all $m$ states before returning to $0$, in particular
it must return to $0$ only after $m$ steps; a short computation using
(1.2) with $n$ ranging over proper divisors of $m$ shows this forces
$\gcd(c,m)=1$ (condition 1), for otherwise the orbit of $0$ would be
confined to the proper subgroup of $\mathbb{Z}_m$ generated by
$\gcd(c,m)$, of size strictly less than $m$. Conditions 2 and 3 follow
by considering the action of $f$ restricted modulo each prime power
factor $p^{e}$ of $m$: if $a \not\equiv 1 \pmod p$ for some prime
$p \mid m$, or if $4 \mid m$ but $a \not\equiv 1 \pmod 4$, the induced
map on $\mathbb{Z}_{p^e}$ fails to be a single $p^e$-cycle, which
propagates (via the Chinese Remainder Theorem decomposition
$\mathbb{Z}_m \cong \prod \mathbb{Z}_{p^e}$) to a cycle structure on
$\mathbb{Z}_m$ shorter than $m$, contradicting full period. $\blacksquare$

**Remarks.** The proof reduces the problem to prime-power moduli and
recombines the result using the Chinese Remainder Theorem, which is
the standard strategy in the number theory of congruential generators
(Knuth, 1997). Condition 3 is easy to overlook and is the most common
source of an incorrectly "full period" generator design: a modulus
divisible by 4 with $a \equiv 3 \pmod 4$ will fail to achieve full
period even though conditions 1 and 2 may hold.

**Examples.** Worked verification of these conditions for specific
parameter sets is given in the Numerical Examples section below.

**Applications.** The theorem is the standard certificate used to
validate any newly proposed linear or mixed congruential generator
before it is deployed in simulation software, and historically it was
used to diagnose the short-period failures of several early generator
implementations.

## Theorem 1.2: Maximal Period of a Multiplicative Congruential Generator

**Statement.** Consider the multiplicative congruential generator

$$
X_{n+1} = \bigl(a X_n\bigr) \bmod m,
$$

with $m$ prime and seed $X_0 \not\equiv 0 \pmod m$. The period of the
generator, from this seed, equals the multiplicative order of $a$
modulo $m$, and this period attains its maximum possible value, $m-1$,
if and only if $a$ is a **primitive root** modulo $m$.

**Assumptions.** $m$ is prime, so the nonzero residues
$\{1, 2, \dots, m-1\}$ form a group $(\mathbb{Z}/m\mathbb{Z})^{*}$
under multiplication modulo $m$; $X_0$ is taken nonzero, since the
orbit of $0$ under multiplication is the fixed point $\{0\}$ and is
excluded from consideration.

**Proof.** Since $m$ is prime, $(\mathbb{Z}/m\mathbb{Z})^{*}$ is a
finite group of order $m-1$ under multiplication. For $X_0 \ne 0$,
the sequence $X_n = a^n X_0 \bmod m$ lies entirely in this group, and
$X_n = X_0$ if and only if $a^n \equiv 1 \pmod m$, that is, if and
only if $n$ is a multiple of $\operatorname{ord}(a)$, the
multiplicative order of $a$ modulo $m$ (the smallest positive integer
$k$ with $a^k \equiv 1 \pmod m$; such $k$ exists and divides $m-1$ by
Lagrange's theorem applied to the cyclic group
$(\mathbb{Z}/m\mathbb{Z})^{*}$). Hence the period of the orbit of
$X_0$ is exactly $\operatorname{ord}(a)$, independent of which nonzero
$X_0$ is chosen, since the map $x \mapsto a^n x$ is a bijection of the
group for every $n$.

Because $\operatorname{ord}(a)$ divides $|(\mathbb{Z}/m\mathbb{Z})^{*}|
= m-1$, the largest value the period can possibly take is $m-1$
itself, and by definition $a$ achieves $\operatorname{ord}(a) = m-1$
precisely when $a$ generates the entire group, that is, when $a$ is a
primitive root modulo $m$. $\blacksquare$

**Remarks.** Unlike the Hull-Dobell theorem, which certifies full
period $m$ (including the absorbing state $0$), a multiplicative
generator with $c=0$ can never itself reach period $m$, because $X_n
=0$ is a fixed point that, once entered, never leaves; hence the best
attainable period for an MCG is $m-1$, achieved by choosing $m$ prime
and $a$ a primitive root modulo $m$.

**Examples.** A full numerical illustration is given below.

**Applications.** This theorem is the design principle behind the
classical Lehmer generator and the Park-Miller "minimal standard"
generator, both discussed later in this unit.

# Mathematical Derivations

## The General Linear Congruential Recurrence

The starting point for every generator in this unit is the recurrence

$$
X_{n+1} = \bigl(aX_n + c\bigr) \bmod m, \qquad n = 0, 1, 2, \dots,
\tag{1.3}
$$

with parameters: modulus $m$ (a positive integer, the size of the
state space), multiplier $a$ (an integer, $0 \le a < m$), increment
$c$ (an integer, $0 \le c < m$), and seed $X_0$ (an integer, $0 \le
X_0 < m$). The output stream reported to the user is the normalized
sequence

$$
U_n = \frac{X_n}{m} \in [0,1),
\tag{1.4}
$$

intended to approximate independent Uniform$(0,1)$ variates. Three
named special cases of (1.3) are distinguished by the value of $c$.

- **Linear Congruential Generator (LCG).** The umbrella term for
  recurrence (1.3) in full generality, for any $a$ and $c$.
- **Multiplicative Congruential Generator (MCG).** The special case
  $c = 0$, so that $X_{n+1} = (aX_n) \bmod m$. This is historically
  the oldest variant (Lehmer, 1949) and is attractive because it
  requires one fewer arithmetic operation per step.
- **Mixed Congruential Generator.** The special case $c \ne 0$, in
  which both a multiplicative term $aX_n$ and an additive term $c$
  are present; the name reflects the "mixing" of multiplication and
  addition in the recurrence. (Some texts use "LCG" loosely as a
  synonym for the mixed case and reserve "multiplicative congruential
  generator" purely for the $c=0$ case, as this unit does; the
  underlying recurrence is identical, and only the value of $c$
  distinguishes the two.)

## Why Modular Reduction Produces Bounded, Deterministic Output

Without the modulo operation, iterating $X_{n+1} = aX_n + c$ would
produce a sequence of unboundedly growing (or shrinking) integers, of
no direct use as a $[0,1]$-valued output. The reduction modulo $m$
confines the state to the finite set $\{0, 1, \dots, m-1\}$, at the
cost of making the map $X_n \mapsto X_{n+1}$ many-to-one on the
integers but a well-defined function on this finite set. Because a
function from a finite set to itself must eventually revisit a state
(by the pigeonhole principle, among any $m+1$ successive states, two
must coincide), the sequence $X_n$ is guaranteed to be eventually
periodic; the constructions above show precisely when the period is as
long as it can possibly be.

## Expected Behaviour Under Full Period

When a generator attains full period $m$ from seed $X_0$, the
sequence $X_0, X_1, \dots, X_{m-1}$ is a permutation of $\{0, 1,
\dots, m-1\}$, visiting each state exactly once per cycle. Consequently,
over one full cycle, the empirical distribution of $U_n = X_n/m$ is
*exactly* uniform on the discrete grid $\{0, 1/m, \dots, (m-1)/m\}$,
with sample mean and variance

$$
\frac{1}{m}\sum_{n=0}^{m-1} U_n = \frac{m-1}{2m}
\;\longrightarrow\; \frac12,
\qquad
\frac{1}{m}\sum_{n=0}^{m-1}\Bigl(U_n - \tfrac12\Bigr)^2
\;\longrightarrow\; \frac{1}{12},
\tag{1.5}
$$

as $m \to \infty$, which are exactly the mean and variance of a true
Uniform$(0,1)$ random variable. This calculation shows that full
period is a *necessary* condition for good uniformity over one cycle,
but it says nothing about the *sequential* (independence) structure of
the output within the cycle — a generator can achieve perfect
marginal uniformity over its full cycle while still producing
badly correlated successive values, as the RANDU example below
demonstrates.

# Algorithms

## Algorithm 1.1: Linear Congruential Generator

**Purpose.** Generate a sequence of $n$ pseudo-random numbers in
$[0,1)$ using the general (mixed) linear congruential recurrence.

**Input.** Seed $X_0$; multiplier $a$; increment $c$; modulus $m$;
required sample size $n$.

**Output.** A vector $U_1, \dots, U_n$ of pseudo-random numbers in
$[0,1)$.

**Step-by-step procedure.**

1. Validate parameters: $0 \le X_0 < m$, $0 \le a < m$, $0 \le c < m$.
2. Set the current state to $X_0$.
3. For $i = 1, \dots, n$: compute $X_i = (aX_{i-1}+c) \bmod m$, and
   report $U_i = X_i/m$.
4. Return the vector $(U_1, \dots, U_n)$.

**Pseudo-code.**

```
Algorithm: Linear Congruential Generator
Input:  seed x0, multiplier a, increment c, modulus m, size n
Output: vector u of length n

1. x <- x0
2. For i = 1 to n:
       x <- (a * x + c) mod m
       u[i] <- x / m
3. Return u
```

**Flowchart description.** Start; initialize state to the seed;
repeat n times: multiply the state by $a$, add $c$, reduce modulo
$m$, store the normalized value; after the loop, return the
collected vector; end.

**Complexity.** Each step performs one multiplication, one addition,
and one modulo reduction, all $O(1)$ under fixed-precision arithmetic,
so generating $n$ values costs $O(n)$ time and $O(1)$ auxiliary space
beyond the output vector itself.

**Practical considerations.** The product $aX_n$ can exceed the range
of standard 32-bit signed integers even when $a$ and $m$ individually
fit within that range; production implementations either use 64-bit
(or wider) integer arithmetic for the intermediate product, or apply
Schrage's decomposition, which computes $(aX_n) \bmod m$ using only
quantities that stay within the original bit width.

## Algorithm 1.2: Multiplicative Congruential Generator

**Purpose.** Generate pseudo-random numbers using the special case
$c=0$, historically the first congruential method to be proposed.

**Input.** Seed $X_0 \ne 0$; multiplier $a$; prime modulus $m$;
required sample size $n$.

**Output.** A vector $U_1, \dots, U_n$ of pseudo-random numbers in
$(0,1)$.

**Step-by-step procedure.**

1. Validate $0 < X_0 < m$ and that $m$ is prime.
2. Set the current state to $X_0$.
3. For $i = 1, \dots, n$: compute $X_i = (aX_{i-1}) \bmod m$, and
   report $U_i = X_i/m$.
4. Return the vector $(U_1, \dots, U_n)$.

**Pseudo-code.**

```
Algorithm: Multiplicative Congruential Generator
Input:  seed x0 (nonzero), multiplier a, prime modulus m, size n
Output: vector u of length n

1. x <- x0
2. For i = 1 to n:
       x <- (a * x) mod m
       u[i] <- x / m
3. Return u
```

**Flowchart description.** Identical to Algorithm 1.1 with the
addition step omitted; the state is simply multiplied by $a$ and
reduced modulo $m$ at every iteration.

**Complexity.** $O(n)$ time, $O(1)$ auxiliary space, marginally
cheaper per step than the general LCG because no addition is
performed.

**Practical considerations.** Because $X_n = 0$ is an absorbing state,
a seed accidentally set to $0$ produces a degenerate constant-zero
sequence; the seed must always be validated as strictly between $0$
and $m$.

## Algorithm 1.3: Mixed Congruential Generator

**Purpose.** Generate pseudo-random numbers using the recurrence with
$c \ne 0$, chosen so that, by the Hull-Dobell theorem, full period $m$
can be certified for any modulus, not only a prime one.

**Input.** Seed $X_0$; multiplier $a$; increment $c \ne 0$; modulus
$m$; required sample size $n$.

**Output.** A vector $U_1, \dots, U_n$ of pseudo-random numbers in
$[0,1)$.

**Step-by-step procedure.** Identical to Algorithm 1.1, with the
additional pre-check (before generation begins) that the Hull-Dobell
conditions hold for the chosen $(a,c,m)$, to guarantee full period.

**Pseudo-code.**

```
Algorithm: Mixed Congruential Generator
Input:  seed x0, multiplier a, increment c (nonzero), modulus m, size n
Output: vector u of length n

1. Verify Hull-Dobell conditions for (a, c, m)
2. x <- x0
3. For i = 1 to n:
       x <- (a * x + c) mod m
       u[i] <- x / m
4. Return u
```

**Flowchart description.** Start; verify the Hull-Dobell conditions
once, before iteration begins; if satisfied, proceed exactly as in
Algorithm 1.1; if not satisfied, warn that the requested period is not
guaranteed; end.

**Complexity.** $O(n)$ time for generation, plus a one-time $O(\log
m)$ cost to factorize $m$ for the parameter check; $O(1)$ auxiliary
space beyond the output.

**Practical considerations.** Choosing $m$ as a power of $2$ (e.g.
$m = 2^{31}$) makes the modulo operation a fast bitmask in hardware,
which is why early mixed congruential generators, including the one
used in early versions of the C standard library, favoured this
choice; the trade-off is that the low-order bits of $X_n$ then have
much shorter periods than the high-order bits (the $k$-th bit from
the bottom has period dividing $2^k$), so applications that consume
only a few low bits of each output can be badly non-random even when
the full-width sequence has full period $m$.

# R Programming

```{r generator-functions}
set.seed(123)

#' Generate pseudo-random numbers from a linear congruential generator
#'
#' Purpose: implements the general recurrence
#'   X[n+1] = (a * X[n] + c) mod m
#' and returns the normalized output U[n] = X[n] / m.
#'
#' Arguments:
#'   seed          integer seed X0, with 0 <= seed < modulus
#'   multiplier    integer multiplier a
#'   increment     integer increment c (use 0 for a multiplicative generator)
#'   modulus       integer modulus m
#'   sample_size   number of values to generate
#'
#' Returns:
#'   a numeric vector of length sample_size, with entries in [0, 1)
#'
#' Example:
#'   generate_lcg(seed = 7, multiplier = 5, increment = 3,
#'                modulus = 16, sample_size = 5)
generate_lcg <- function(seed, multiplier, increment, modulus, sample_size) {
  state <- numeric(sample_size)
  current_value <- seed

  for (index in seq_len(sample_size)) {
    current_value <- (multiplier * current_value + increment) %% modulus
    state[index] <- current_value
  }

  state / modulus
}

# A multiplicative congruential generator is the special case c = 0
generate_mcg <- function(seed, multiplier, modulus, sample_size) {
  generate_lcg(
    seed = seed, multiplier = multiplier, increment = 0,
    modulus = modulus, sample_size = sample_size
  )
}

#' Check the Hull-Dobell conditions for full period m
#'
#' Purpose: verifies, for a proposed (multiplier, increment, modulus)
#' triple, whether the mixed congruential generator built from them
#' attains full period m for every seed.
#'
#' Arguments:
#'   multiplier    integer multiplier a
#'   increment     integer increment c
#'   modulus       integer modulus m
#'
#' Returns:
#'   a named logical vector with the three individual conditions and
#'   an overall TRUE/FALSE verdict
#'
#' Example:
#'   check_hull_dobell(multiplier = 21, increment = 3, modulus = 20)
check_hull_dobell <- function(multiplier, increment, modulus) {
  primes <- prime_factors_of(modulus)

  condition_gcd <- gcd_two_integers(increment, modulus) == 1
  condition_multiplier <- all((multiplier - 1) %% primes == 0)
  condition_mod4 <- if (modulus %% 4 == 0) {
    (multiplier - 1) %% 4 == 0
  } else {
    TRUE
  }

  c(
    gcd_condition = condition_gcd,
    prime_condition = condition_multiplier,
    mod4_condition = condition_mod4,
    full_period = condition_gcd && condition_multiplier && condition_mod4
  )
}

# Helper: greatest common divisor via the Euclidean algorithm
gcd_two_integers <- function(first_value, second_value) {
  while (second_value != 0) {
    remainder <- first_value %% second_value
    first_value <- second_value
    second_value <- remainder
  }
  abs(first_value)
}

# Helper: distinct prime factors of a positive integer (trial division;
# adequate for the moderate moduli used for teaching in this unit)
prime_factors_of <- function(integer_value) {
  factors <- integer(0)
  remaining <- integer_value
  divisor <- 2

  while (divisor * divisor <= remaining) {
    if (remaining %% divisor == 0) {
      factors <- c(factors, divisor)
      while (remaining %% divisor == 0) remaining <- remaining / divisor
    }
    divisor <- divisor + 1
  }
  if (remaining > 1) factors <- c(factors, remaining)

  factors
}
```

Both `generate_lcg()` and `check_hull_dobell()` are pure functions of
their arguments, use `<-` for assignment throughout, and avoid any
deprecated base-R random number interfaces such as direct
`.Random.seed` manipulation.

# Simulation Studies

## Empirical Period of a Small Mixed Congruential Generator

```{r period-demo}
set.seed(123)

# A small modulus makes the full cycle visible in one table
demo_sequence <- generate_lcg(
  seed = 7, multiplier = 5, increment = 3,
  modulus = 16, sample_size = 20
) * 16

tibble(
  step = 0:19,
  state = as.integer(demo_sequence)
) %>%
  kable(caption = "First 20 states of a mixed LCG with a=5, c=3, m=16") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

The Hull-Dobell check confirms the theoretical prediction before the
table is even inspected.

```{r period-demo-check}
check_hull_dobell(multiplier = 5, increment = 3, modulus = 16)
```

With `full_period = TRUE`, the theorem guarantees the cycle length is
exactly $16$; the table above confirms this directly, since the state
sequence visits every integer from $0$ to $15$ exactly once in the
first sixteen steps before repeating.

## Diagnosing RANDU: A Historically Important Failure

RANDU is a multiplicative congruential generator, once distributed by
IBM in the 1960s and 1970s, with $a = 65539$, $c=0$, $m = 2^{31}$.
Although RANDU passes simple one-dimensional uniformity tests, its
successive *triples* are catastrophically non-random. Because $a =
2^{16}+3$, squaring gives $a^2 = 2^{32} + 6\cdot 2^{16} + 9$, and since
$2^{32} \equiv 0 \pmod{2^{31}}$,

$$
a^2 \equiv 6a - 9 \pmod{2^{31}}
\tag{1.6}
$$

(direct arithmetic: $6a - 9 = 393{,}234 - 9 = 393{,}225$, which agrees
with $a^2 \bmod 2^{31} = 393{,}225$). Because the generator is
multiplicative, $X_{n+2} = a^2 X_n \bmod 2^{31}$, so substituting
(1.6) gives the exact linear identity

$$
X_{n+2} \equiv 6X_{n+1} - 9X_n \pmod{2^{31}},
\tag{1.7}
$$

which means every consecutive triple $(X_n, X_{n+1}, X_{n+2})$ lies
exactly on one of a small number of parallel planes in
three-dimensional space, rather than filling the unit cube (Marsaglia,
1968).

```{r randu-demo, fig.cap="Successive triples from RANDU collapse onto a small number of parallel planes; a good generator (right) shows no such structure."}
set.seed(123)

randu_generate <- function(seed, sample_size) {
  generate_mcg(seed = seed, multiplier = 65539, modulus = 2^31, sample_size = sample_size)
}

n_points <- 3000
randu_stream <- randu_generate(seed = 1, sample_size = n_points + 2)
good_stream <- runif(n_points + 2)

combine_plane <- function(stream) {
  6 * stream[2:(n_points + 1)] - 9 * stream[1:n_points]
}

randu_df <- tibble(
  x = randu_stream[1:n_points],
  y = randu_stream[2:(n_points + 1)],
  z = randu_stream[3:(n_points + 2)],
  plane_value = combine_plane(randu_stream) %% 1
)

good_df <- tibble(
  x = good_stream[1:n_points],
  y = good_stream[2:(n_points + 1)],
  plane_value = combine_plane(good_stream) %% 1
)

p_randu <- ggplot(randu_df, aes(x = x, y = plane_value)) +
  geom_point(alpha = 0.35, size = 0.6, colour = "firebrick") +
  labs(
    title = "RANDU: 6X[n+1] - 9X[n] (mod 1) against X[n]",
    x = expression(U[n]), y = "planar combination (mod 1)"
  ) +
  theme_minimal(base_size = 11)

p_good <- ggplot(good_df, aes(x = x, y = plane_value)) +
  geom_point(alpha = 0.35, size = 0.6, colour = "steelblue") +
  labs(
    title = "runif(): same combination, no structure",
    x = expression(U[n]), y = "planar combination (mod 1)"
  ) +
  theme_minimal(base_size = 11)

p_randu + p_good
```

The RANDU panel collapses onto a small number of near-horizontal
bands, a direct visual signature of identity (1.7): because
$6U_{n+1}-9U_n \pmod 1$ is (almost) a function of $U_{n+2}$ alone, the
combination clusters tightly rather than filling the plot. The
`runif()` panel, generated by R's Mersenne-Twister algorithm, shows no
such banding. This is the classical demonstration, due to Marsaglia
(1968), that passing univariate uniformity tests is not sufficient for
a generator to be fit for multivariate simulation; the lattice
structure of the underlying recurrence must also be examined.

## Comparison with R's Built-In Generator

```{r comparison-summary}
set.seed(123)

sample_size <- 5000

lcg_sample <- generate_lcg(
  seed = 12345, multiplier = 16807, increment = 0,
  modulus = 2^31 - 1, sample_size = sample_size
)

builtin_sample <- runif(sample_size)

comparison_table <- tibble(
  Generator = c("Park-Miller MCG", "R runif (Mersenne-Twister)"),
  Mean = c(mean(lcg_sample), mean(builtin_sample)),
  Variance = c(var(lcg_sample), var(builtin_sample)),
  Theoretical_Mean = 0.5,
  Theoretical_Variance = 1 / 12
)

kable(comparison_table, digits = 5,
      caption = "Sample moments compared to the Uniform(0,1) target") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

Both generators reproduce the theoretical mean of $0.5$ and variance
of $1/12$ closely, illustrating that first-moment agreement alone
cannot distinguish a well-designed congruential generator such as
Park-Miller from R's default; higher-dimensional and sequential tests,
covered in Unit 3.2, are required to tell them apart from a
substantially weaker design such as RANDU.

# Graphs

```{r histogram-comparison, fig.cap="Histograms of 5000 draws: a well-parameterized MCG (left) closely matches the Uniform(0,1) target, shown against R's default generator (right)."}
set.seed(123)

p_lcg_hist <- ggplot(tibble(u = lcg_sample), aes(x = u)) +
  geom_histogram(bins = 30, fill = "steelblue", colour = "white") +
  labs(title = "Park-Miller MCG", x = expression(U[n]), y = "Count") +
  theme_minimal(base_size = 11)

p_builtin_hist <- ggplot(tibble(u = builtin_sample), aes(x = u)) +
  geom_histogram(bins = 30, fill = "darkgreen", colour = "white") +
  labs(title = "R runif()", x = expression(U[n]), y = "Count") +
  theme_minimal(base_size = 11)

p_lcg_hist + p_builtin_hist
```

```{r lag-scatter, fig.cap="Lag-1 scatterplots: successive pairs (U[n], U[n+1]) for the Park-Miller generator (left) and RANDU (right)."}
set.seed(123)

lag_df_good <- tibble(
  u_now = lcg_sample[1:(sample_size - 1)],
  u_next = lcg_sample[2:sample_size]
)

randu_full <- randu_generate(seed = 1, sample_size = sample_size)
lag_df_randu <- tibble(
  u_now = randu_full[1:(sample_size - 1)],
  u_next = randu_full[2:sample_size]
)

p_lag_good <- ggplot(lag_df_good, aes(x = u_now, y = u_next)) +
  geom_point(alpha = 0.25, size = 0.5, colour = "steelblue") +
  labs(title = "Park-Miller MCG", x = expression(U[n]), y = expression(U[n+1])) +
  theme_minimal(base_size = 11)

p_lag_randu <- ggplot(lag_df_randu, aes(x = u_now, y = u_next)) +
  geom_point(alpha = 0.25, size = 0.5, colour = "firebrick") +
  labs(title = "RANDU", x = expression(U[n]), y = expression(U[n+1])) +
  theme_minimal(base_size = 11)

p_lag_good + p_lag_randu
```

The lag-1 scatterplot for the Park-Miller generator fills the unit
square with no visible pattern, while RANDU already shows faint
diagonal banding even in two dimensions, consistent with the more
severe three-dimensional collapse documented above.

# Numerical Examples

**Example 1: Manual iteration of a mixed congruential generator.**

*Problem.* Given $X_0 = 7$, $a=5$, $c=3$, $m=16$, compute $X_1$
through $X_5$ by hand and report the corresponding $U_n$ values.

*Solution.*

$$
\begin{aligned}
X_1 &= (5 \times 7 + 3) \bmod 16 = 38 \bmod 16 = 6, & U_1 &= 6/16 = 0.375\\
X_2 &= (5 \times 6 + 3) \bmod 16 = 33 \bmod 16 = 1, & U_2 &= 1/16 = 0.0625\\
X_3 &= (5 \times 1 + 3) \bmod 16 = 8 \bmod 16 = 8, & U_3 &= 8/16 = 0.5\\
X_4 &= (5 \times 8 + 3) \bmod 16 = 43 \bmod 16 = 11, & U_4 &= 11/16 = 0.6875\\
X_5 &= (5 \times 11 + 3) \bmod 16 = 58 \bmod 16 = 10, & U_5 &= 10/16 = 0.625
\end{aligned}
$$

*Interpretation.* The state sequence $7, 6, 1, 8, 11, 10, \dots$
shows no obvious pattern to a naive observer, illustrating the
intuition of Section 4; the R simulation above confirms that this
particular parameter set has full period $16$.

**Example 2: Verifying the Hull-Dobell conditions.**

*Problem.* Determine whether $a=21$, $c=3$, $m=20$ gives a full-period
mixed congruential generator.

*Solution.* The prime factors of $m=20 = 2^2 \times 5$ are $2$ and
$5$. Check each condition:

$$
\begin{aligned}
\text{Condition 1:}\quad & \gcd(3, 20) = 1. \quad \checkmark\\
\text{Condition 2:}\quad & a - 1 = 20, \quad 20 \equiv 0 \pmod 2, \quad 20 \equiv 0 \pmod 5. \quad \checkmark\\
\text{Condition 3:}\quad & 4 \mid 20, \text{ and } 20 \equiv 0 \pmod 4. \quad \checkmark
\end{aligned}
$$

All three conditions hold, so this generator has full period $20$ for
every seed.

*Interpretation.* Note that $a-1=20$ is itself divisible by $m$, so
$a \equiv 1 \pmod m$; this is the simplest reliable recipe for
choosing a multiplier that is guaranteed to satisfy conditions 2 and
3 simultaneously, at the cost of $a$ being uncomfortably close to $1$,
which can weaken the mixing of digits in $aX_n$. Better choices
satisfy the same congruences modulo the prime factors of $m$ without
being congruent to $1$ modulo $m$ itself; see the Practical Tips
section.

**Example 3: Maximal period of a multiplicative congruential
generator.**

*Problem.* With prime modulus $m=31$, verify that $a=3$ is a primitive
root modulo $31$, and hence that the resulting MCG has period $30$.

*Solution.* The order of $3$ modulo $31$ must divide $m-1=30$. Check
the proper divisors of $30$ (namely $1,2,3,5,6,10,15$) to make sure
none of them satisfies $3^k \equiv 1 \pmod{31}$:

$$
3^1=3,\ 3^2=9,\ 3^3=27,\ 3^5=243\equiv 26,\
3^6\equiv 3\times 26=78\equiv 16,\ 3^{10}\equiv 16^2\times 3^{-2}\dots
$$

more directly, computing successive powers modulo $31$ gives $3, 9,
27, 19, 26, 16, 17, 20, 29, 25, 13, 8, 24, 10, 30, 28, 22, 4, 12, 5,
15, 14, 11, 2, 6, 18, 23, 7, 21, 1$ for exponents $1$ through $30$,
and $1$ first appears only at exponent $30$. Hence
$\operatorname{ord}(3) = 30 = m-1$, confirming that $3$ is a
primitive root modulo $31$ and that the MCG with $a=3, m=31$ has
maximal period $30$.

*Interpretation.* By Theorem 1.2, every nonzero seed produces a cycle
visiting all $30$ nonzero residues modulo $31$ before repeating; this
is verified computationally below.

```{r example3-verify}
set.seed(123)

mcg_period_31 <- generate_mcg(seed = 1, multiplier = 3, modulus = 31, sample_size = 30) * 31

tibble(step = 1:30, state = as.integer(mcg_period_31)) %>%
  summarise(distinct_states = n_distinct(state), full_period_expected = 30) %>%
  kable(caption = "Distinct states visited in one cycle of the m=31 MCG")
```

# Applications

- **Statistics.** Every classical resampling procedure (bootstrap,
  permutation tests, jackknife) and every simulation-based power
  calculation begins by generating a uniform stream from a
  congruential or similar generator.
- **Machine Learning.** Weight initialization, dropout masks,
  mini-batch shuffling, and train-test splitting all consume uniform
  or derived random numbers, and reproducible seeds are essential for
  comparing model variants fairly.
- **Data Science.** Random sampling of large datasets, cross-
  validation fold assignment, and synthetic data generation for
  testing pipelines all rely on the same primitive.
- **Finance.** Monte Carlo option pricing and value-at-risk estimation
  simulate thousands of price paths, each path consuming a long
  uniform stream converted into asset return innovations.
- **Reliability Engineering.** Simulating component failure times
  under a fitted lifetime distribution (Exponential, Weibull) to
  estimate system reliability functions begins with a uniform
  generator combined with the inverse transformation method of Unit
  3.3.
- **Monte Carlo Methods.** Numerical integration by simple or
  importance-sampled Monte Carlo is, at its core, an average of a
  function evaluated at uniform (or uniform-derived) sample points.
- **Bayesian Statistics.** Markov chain Monte Carlo samplers
  (Metropolis-Hastings, Gibbs sampling) require a uniform stream at
  every acceptance step, in addition to whatever proposal-generation
  mechanism is used.
- **Operations Research.** Discrete-event simulation of queueing
  systems, inventory policies, and network reliability all generate
  interarrival and service times from an underlying uniform stream.
- **Engineering.** Monte Carlo tolerance analysis in manufacturing and
  structural reliability analysis under random loading both depend on
  a trustworthy uniform generator as their first building block.

# Advantages

- Congruential generators are extremely fast, requiring only one or
  two arithmetic operations per output value.
- They are simple to implement correctly in any programming language,
  with no special hardware or library support required.
- Their theoretical properties (period, lattice structure) are fully
  understood via the number theory developed above, unlike some
  more elaborate generators whose properties are known only
  empirically.
- They are exactly reproducible from a stored seed, which is essential
  for scientific replicability.

# Limitations

- The maximum period of a congruential generator is bounded by the
  modulus $m$; for a modulus that fits in 32 bits, the period is far
  too short for the largest modern simulation studies, which is why
  contemporary software defaults to generators with astronomically
  larger periods, such as the Mersenne Twister (period $2^{19937}-1$).
- Even at full period, congruential generators are well known to
  exhibit lattice structure in moderate dimensions: $k$-tuples of
  consecutive outputs fall on a limited number of parallel
  hyperplanes, as demonstrated for RANDU above; this defect worsens as
  $k$ increases (Marsaglia, 1968).
- Poorly chosen parameters can silently produce a much shorter period
  than the modulus suggests, with no error or warning at run time.
- Congruential generators are entirely unsuitable for cryptographic
  applications, since observing a short run of outputs is often enough
  to reconstruct the multiplier and reconstruct all future values.

# Practical Tips

- **Never reseed from the system clock inside a loop.** Doing so
  destroys reproducibility and can, on fast machines, produce
  identical seeds across successive calls if the clock resolution is
  coarser than the loop's execution time.
- **Always report the seed used for a simulation study.** A single
  call to `set.seed()` at the start of a script or analysis, recorded
  in the write-up, is sufficient to make the entire study exactly
  reproducible.
- **Prefer a multiplier that is not congruent to $1 \pmod m$ by an
  excessively small margin.** Choosing $a = 1 + m$ trivially satisfies
  the Hull-Dobell conditions but produces very poor mixing in the
  high-order bits; well-tested multipliers (such as those tabulated by
  L'Ecuyer, 1999, for various moduli) should be preferred over
  ad-hoc choices.
- **Beware of integer overflow.** Computing $aX_n$ directly can exceed
  the range of the machine's native integer type well before the
  modulo operation is applied; use a wider integer type, floating
  point with adequate mantissa precision, or Schrage's algorithm to
  avoid silently wrapping into negative or incorrect values.
- **Do not rely on the low-order bits of a power-of-two-modulus
  generator.** As noted in Algorithm 1.3, the $k$ lowest bits of $X_n$
  cycle with period dividing $2^k$, so extracting, say, a single low
  bit to simulate a fair coin flip from such a generator can be far
  from independent draws.
- **Never use RANDU, or any similarly flawed historical generator, in
  new work.** It is retained in the curriculum purely as a
  cautionary example of the gap between passing simple tests and
  being fit for genuine multivariate simulation.

# Summary

- A pseudo-random number generator is a deterministic algorithm
  producing a sequence that imitates independent Uniform$(0,1)$ draws.
- Good generators combine uniformity, independence, long period,
  reproducibility, efficiency, portability, and freedom from lattice
  structure.
- The linear congruential recurrence $X_{n+1} = (aX_n+c) \bmod m$
  specializes to the multiplicative case ($c=0$) and the mixed case
  ($c \ne 0$).
- The Hull-Dobell theorem gives exact, checkable conditions on $a$,
  $c$, and $m$ for the mixed generator to achieve full period $m$.
- A multiplicative generator with prime modulus $m$ achieves its
  maximal period $m-1$ exactly when the multiplier is a primitive
  root modulo $m$.
- Full period guarantees perfect marginal uniformity over one cycle,
  but says nothing about multivariate independence; the RANDU example
  shows that a full-period generator can still have severely flawed
  higher-dimensional structure.
- Modern software (including R's default generator) uses far more
  sophisticated algorithms with astronomically longer periods, but
  congruential generators remain the clearest vehicle for
  understanding the theory of pseudo-randomness.

# Key Formulae

- General linear congruential recurrence:
  $X_{n+1} = (aX_n + c) \bmod m$.
- Normalized output: $U_n = X_n / m$.
- Hull-Dobell full-period conditions: $\gcd(c,m)=1$;
  $a \equiv 1 \pmod p$ for every prime $p \mid m$; $a \equiv 1 \pmod
  4$ if $4 \mid m$.
- Maximal MCG period: $\operatorname{ord}(a) \mid m-1$ for $m$ prime,
  with equality when $a$ is a primitive root modulo $m$.
- RANDU planar identity: $X_{n+2} \equiv 6X_{n+1} - 9X_n \pmod{2^{31}}$.
- Full-cycle moments: mean $\to 1/2$, variance $\to 1/12$ as $m \to
  \infty$.

# Key Terms

- **Congruential generator.** A generator whose recurrence is defined
  by modular arithmetic on a linear (or affine) function of the
  previous state.
- **Full period.** The property that a generator's period equals the
  size of its state space.
- **Hull-Dobell theorem.** The number-theoretic characterization of
  full-period mixed congruential generators.
- **Lattice structure.** The tendency of consecutive outputs of a
  congruential generator, viewed as points in several dimensions, to
  fall on a limited number of parallel hyperplanes.
- **Modulus ($m$).** The integer defining the size of the state space
  of a congruential generator.
- **Multiplier ($a$).** The integer by which the state is multiplied
  at each step of a congruential generator.
- **Increment ($c$).** The integer added at each step of a mixed
  congruential generator; zero for a purely multiplicative generator.
- **Period.** The number of steps after which a generator's output
  sequence begins to repeat.
- **Primitive root.** An integer $a$ whose powers modulo a prime $m$
  generate every nonzero residue class, i.e. $\operatorname{ord}(a) =
  m-1$.
- **Pseudo-random number generator (PRNG).** A deterministic algorithm
  producing output intended to imitate a genuinely random sequence.
- **Seed.** The initial state supplied to a pseudo-random number
  generator, which entirely determines its subsequent output.

# Exercises

## Multiple Choice Questions

1. A pseudo-random number generator is best described as
   (a) a genuinely random physical process
   (b) a deterministic algorithm imitating randomness
   (c) an undefined mathematical object
   (d) a lookup table of pre-recorded random digits

2. The period of a generator with state space of size $m$ can be at
   most
   (a) $m^2$
   (b) $2m$
   (c) $m$
   (d) unbounded

3. In the recurrence $X_{n+1} = (aX_n+c) \bmod m$, setting $c=0$ gives
   (a) a mixed congruential generator
   (b) a multiplicative congruential generator
   (c) an additive generator
   (d) a Mersenne Twister generator

4. The Hull-Dobell theorem provides conditions for
   (a) uniformity of the output values
   (b) full period of a mixed congruential generator
   (c) independence of successive outputs
   (d) convergence of Monte Carlo estimates

5. If $4 \mid m$, the Hull-Dobell theorem additionally requires
   (a) $a \equiv 1 \pmod 2$
   (b) $a \equiv 3 \pmod 4$
   (c) $a \equiv 1 \pmod 4$
   (d) $c \equiv 0 \pmod 4$

6. A multiplicative congruential generator with prime modulus $m$
   attains its maximal period $m-1$ when
   (a) $a=1$
   (b) $a$ is a primitive root modulo $m$
   (c) $c=1$
   (d) $m$ is a power of two

7. RANDU is a well-known example of
   (a) a generator with excellent multidimensional uniformity
   (b) a generator with full period but severe lattice structure
   (c) a purely additive generator
   (d) a cryptographically secure generator

8. The low-order bits of a congruential generator with modulus $m =
   2^k$ tend to
   (a) have longer periods than the high-order bits
   (b) have shorter periods than the high-order bits
   (c) be exactly independent of the high-order bits
   (d) be uniformly distributed regardless of parameter choice

9. Schrage's algorithm is used to
   (a) test uniformity of generated numbers
   (b) avoid integer overflow when computing $(aX_n) \bmod m$
   (c) compute the period of a generator
   (d) generate Normal random variates

10. Which of the following is NOT a desirable property of a random
    number generator?
    (a) Long period
    (b) Reproducibility
    (c) Machine-dependent output
    (d) Statistical independence of outputs

**Answers.** 1(b), 2(c), 3(b), 4(b), 5(c), 6(b), 7(b), 8(b), 9(b),
10(c).

## Short Answer Questions

1. Distinguish a pseudo-random number generator from a physical
   random number source, and explain why reproducibility matters for
   simulation studies.
2. Define the period of a random number generator and explain why
   the pigeonhole principle guarantees that every congruential
   generator is eventually periodic.
3. State the three conditions of the Hull-Dobell theorem in your own
   words.
4. Why can a multiplicative congruential generator with $c=0$ never
   attain full period $m$?
5. Explain what a primitive root modulo a prime is, and why it is
   relevant to the design of a multiplicative congruential generator.
6. What is meant by "lattice structure" in a random number generator,
   and why is it a more serious flaw than failing a simple univariate
   uniformity test?
7. Explain briefly why RANDU passes simple tests of uniformity despite
   being a poor generator overall.
8. Why is choosing the modulus as a power of two computationally
   convenient, and what disadvantage does this choice introduce?
9. Explain why integer overflow is a practical concern in
   implementing congruential generators, and describe one remedy.
10. Give two applications, from different fields, that rely on the
    availability of a fast, reproducible pseudo-random number
    generator.

## Long Answer Questions

1. State and prove the Hull-Dobell theorem, explaining clearly how
   the Chinese Remainder Theorem is used to combine prime-power
   conditions into a statement about a general modulus $m$.
2. State and prove the theorem on the maximal period of a
   multiplicative congruential generator with prime modulus, and
   explain the role of the multiplicative group
   $(\mathbb{Z}/m\mathbb{Z})^{*}$ in the proof.
3. Derive the closed-form expression for $X_n$ in terms of $X_0$, $a$,
   $c$, $m$, and $n$, and use it to explain why full period is
   equivalent to the two congruence conditions given in the unit.
4. Derive the identity $X_{n+2} \equiv 6X_{n+1}-9X_n \pmod{2^{31}}$
   for the RANDU generator, and explain its geometric consequence for
   the distribution of consecutive triples.
5. Compare and contrast the linear congruential, multiplicative
   congruential, and mixed congruential generators in terms of
   computational cost, achievable period, and parameter selection
   criteria.
6. Discuss, with reference to at least two historical examples, why
   passing univariate statistical tests is not sufficient evidence
   that a generator is fit for use in a Monte Carlo study.
7. Explain the practical numerical issues (overflow, precision) that
   arise when implementing a congruential generator with a large
   modulus, and describe how Schrage's algorithm addresses them.
8. Discuss the criteria a statistician should apply when selecting a
   random number generator for a large-scale simulation study,
   relating each criterion back to the theory developed in this unit.
9. Explain, using the full-cycle moment calculation of Section 6, why
   full period is necessary but not sufficient for a congruential
   generator to be considered "good."
10. Trace the historical development of pseudo-random number
    generation from von Neumann's middle-square method through Lehmer's
    multiplicative congruential generator to the design principles
    embodied in the Hull-Dobell theorem.

## Programming Exercises

1. Implement `generate_lcg()` from first principles (already provided
   in this unit) and use it to reproduce the twenty-step table of
   Example 1 exactly.
2. Write an R function that, given $a$, $c$, and $m$, empirically
   determines the observed period of the generator from a fixed seed
   by iterating until the state repeats, and compare the result to the
   prediction of the Hull-Dobell theorem.
3. Implement `check_hull_dobell()` from first principles (already
   provided) and use it to test at least five different $(a,c,m)$
   combinations of your own choosing, tabulating the results.
4. Reproduce the RANDU lattice-structure plot of this unit using a
   different pair of small integer coefficients in the planar
   combination, and comment on whether the banding is still visible.
5. Implement the Park-Miller minimal standard generator
   ($a=16807$, $m=2^{31}-1$, $c=0$) and verify empirically, for a
   sample of size 100,000, that its sample mean and variance are close
   to the theoretical Uniform$(0,1)$ values.
6. Write an R function that searches, for a fixed prime modulus $m$
   between $10$ and $100$, for all primitive roots modulo $m$, and
   verify your answer for $m=31$ against Example 3 of this unit.
7. Using `system.time()`, benchmark your `generate_lcg()` function
   against R's built-in `runif()` for sample sizes of $10^4$, $10^5$,
   and $10^6$, and discuss the results.
8. Write a function that implements Schrage's algorithm for computing
   $(aX) \bmod m$ without intermediate overflow, and verify it against
   direct computation (using R's arbitrary-precision handling via
   `as.numeric()`, since R's doubles safely represent integers well
   beyond the moduli used in this unit) for a range of test inputs.
9. Simulate a mixed congruential generator with a modulus that is a
   power of two, extract only the lowest bit of each output, and
   demonstrate empirically that the resulting bit sequence has a much
   shorter period than the full generator.
10. Write an R function that takes a candidate multiplier, increment,
    and modulus, verifies the Hull-Dobell conditions, and, if they
    hold, generates and returns a full period's worth of uniform
    values, raising an informative error otherwise.

## Mini Projects

1. **A parameter atlas for mixed congruential generators.** Using
   `check_hull_dobell()`, systematically search moduli $m \in
   \{2^{10}, 2^{16}, 10^4, 10^5\}$ for multipliers $a$ that (i)
   satisfy the Hull-Dobell conditions and (ii) are not trivially close
   to $1 \pmod m$ (say, within $0.001m$). For each modulus, report the
   number of valid multipliers found, and, for one representative
   choice per modulus, produce a lag-1 scatterplot analogous to those
   in this unit. Summarize which moduli offer the richest set of
   well-behaved multipliers.

2. **Reproducing the RANDU controversy.** Implement RANDU and at least
   one alternative multiplicative congruential generator with a prime
   modulus of comparable size (e.g. $m = 2^{31}-1$). For both
   generators, produce three-way lag plots analogous to the two-
   dimensional plots in this unit (consider projecting the third
   dimension using colour or facets), run the diagnostic identity of
   Equation (1.7) numerically to detect planar structure, and write a
   short report explaining, in your own words, why RANDU's flaw
   remained undetected for years despite its widespread use.

**Difficulty labels.** Multiple choice, short answer: **Easy**. Long
answer questions 3-7, programming exercises 1-6: **Medium**. Long
answer questions 1, 2, 8-10, programming exercises 7-10: **Medium to
Advanced**. Mini projects: **Advanced**.

# References

Hull, T. E., & Dobell, A. R. (1962). Random number generators. *SIAM
Review*, 4(3), 230-254.

Knuth, D. E. (1997). *The art of computer programming, Volume 2:
Seminumerical algorithms* (3rd ed.). Addison-Wesley.

L'Ecuyer, P. (1999). Tables of linear congruential generators of
different sizes and good lattice structure. *Mathematics of
Computation*, 68(225), 249-260.

Marsaglia, G. (1968). Random numbers fall mainly in the planes.
*Proceedings of the National Academy of Sciences*, 61(1), 25-28.

Park, S. K., & Miller, K. W. (1988). Random number generators: Good
ones are hard to find. *Communications of the ACM*, 31(10), 1192-1201.

Ripley, B. D. (1987). *Stochastic simulation*. Wiley.

Rizzo, M. L. (2019). *Statistical computing with R* (2nd ed.). CRC
Press.

Ross, S. M. (2020). *Simulation* (6th ed.). Academic Press.
