---
title: "Unit 3.2: Testing of Random Number Generators"
author: "Dr. Manoj C. Patil"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
    code_folding: hide
    theme: cosmo
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    echo = TRUE,
    warning = FALSE,
    message = FALSE,
    fig.align = "center",
    fig.width = 7,
    fig.height = 4.5
)

set.seed(123)
library(tidyverse)
library(ggplot2)
library(dplyr)
library(knitr)
library(kableExtra)
library(gridExtra)
library(patchwork)
```

# Learning Objectives

By the end of this unit, a student should be able to

- explain why a pseudo-random number generator with a provably long
  period, as studied in Unit 3.1, must still be subjected to empirical
  statistical testing before use,
- distinguish tests of uniformity from tests of independence, and
  select an appropriate test for a given suspected defect,
- state and prove the asymptotic null distribution of the Pearson
  chi-square goodness-of-fit statistic, and apply it to test
  uniformity,
- derive the exact mean of the number of runs in an i.i.d. continuous
  sequence, and use the runs test to detect serial dependence,
- implement the Chi-square, Kolmogorov-Smirnov, digit frequency,
  runs, gap, poker, and serial correlation tests in R,
- interpret the results of a battery of tests applied to the same
  generator, and understand the multiple-testing considerations this
  raises.

# Prerequisites

Students should be comfortable with

- Unit 3.1 (Theory of Random Number Generation), particularly the
  notions of period and the Hull-Dobell theorem,
- hypothesis testing fundamentals: null and alternative hypotheses,
  test statistics, and $p$-values,
- the multinomial and geometric distributions, and the notion of an
  empirical cumulative distribution function,
- basic R programming, including vectorized operations and writing
  functions.

# Introduction

A generator that provably attains its maximal period, as certified by
the Hull-Dobell theorem of Unit 3.1, is not automatically fit for
statistical use. Full period guarantees only that every state in the
generator's state space is visited exactly once per cycle; it says
nothing about *how* those states are ordered, and the RANDU example of
Unit 3.1 showed vividly that a full-period generator can still exhibit
severe structural defects that would corrupt any simulation relying on
it. Empirical testing closes this gap: rather than reasoning about a
generator's algebraic structure directly, a battery of statistical
tests is applied to its output and compared against the behaviour
expected of a genuine Uniform$(0,1)$ sample.

The systematic empirical testing of random number generators dates to
the earliest large-scale computational use of pseudo-randomness. The
RAND Corporation's random digit tables of the 1950s were subjected to
frequency and gap tests before publication, and Donald Knuth's
treatment of statistical tests for randomness in *The Art of Computer
Programming* (1969, with later editions) established the now-standard
taxonomy of uniformity and independence tests used throughout this
unit. Later, comprehensive test suites — the Diehard battery (Marsaglia,
1995) and the NIST Statistical Test Suite (Rukhin et al., 2010) —
extended and automated these ideas for cryptographic and scientific
applications.

The practical importance of this material is immediate: any simulation
study is only as trustworthy as the generator underlying it, and a
defect invisible to a casual inspection of the output — as with RANDU
— can silently bias an entire body of downstream results. This unit
develops two families of tests. **Uniformity tests** — the Chi-square
goodness-of-fit test, the Kolmogorov-Smirnov test, and the digit
frequency test — check whether individual generated values follow the
Uniform$(0,1)$ distribution. **Independence tests** — the runs test,
the gap test, the poker test, and the serial correlation test — check
whether successive generated values show any exploitable sequential
structure. No single test is sufficient on its own; a defensible
practice is to apply several tests, of both kinds, to the same
generator before relying on it.

# Intuition

Two quite different things can go wrong with a stream of numbers that
is supposed to imitate independent Uniform$(0,1)$ draws. First, the
values themselves might not be evenly spread across $[0,1]$ — perhaps
values near $0.5$ occur too often, or values very close to $0$ or $1$
are systematically missing. This is a failure of *uniformity*, and it
is detected by comparing the shape of the empirical distribution of
the generated values to the flat, uniform shape expected under the
null hypothesis. Second, even if the values are perfectly spread out
individually, their *order* might not be random — perhaps a small
value tends to be followed by another small value more often than
chance would suggest, or the sequence alternates up and down far more
regularly than randomness would produce. This is a failure of
*independence*, and it cannot be detected by looking at any single
value in isolation; it requires examining the relationship between
consecutive (or otherwise related) values.

The RANDU generator of Unit 3.1 is the clearest illustration available
of why both kinds of tests are needed: RANDU's individual output
values are, considered one at a time, very close to uniformly
distributed, so a naive uniformity test alone would not flag it as
defective. Its failure is entirely a failure of independence — the
planar collapse of consecutive triples — which only a test sensitive
to sequential structure can reveal. This unit's two families of tests,
uniformity and independence, are therefore complementary rather than
redundant: a generator must pass both kinds before it can be
considered trustworthy.

# Formal Definitions

Let $U_1,\dots,U_n$ denote a sequence of values produced by the
generator under test. The joint null hypothesis common to every test
in this unit is

$$
H_0:\quad U_1,\dots,U_n \overset{\text{iid}}{\sim} \text{Uniform}(0,1).
\tag{2.1}
$$

Under $H_0$, each $U_i$ has density $f(u)=1$ on $[0,1]$, and every pair
$(U_i,U_j)$, $i\ne j$, is independent. The alternative hypothesis is
broad and unspecified: values may be non-uniform, dependent, or both.
Rejecting $H_0$ signals that the generator's output is statistically
distinguishable from an ideal Uniform$(0,1)$ sample; it does not, by
itself, diagnose *which* aspect of $H_0$ has failed, which is why
several distinct tests, each sensitive to a different kind of
departure, are used together.

**Definition (run).** A **run** (specifically, a run "up and down") is
a maximal subsequence of consecutive values that is either strictly
increasing or strictly decreasing. Equivalently, using the sign
sequence $S_i = \operatorname{sign}(U_{i+1}-U_i)$ for
$i=1,\dots,n-1$, a run is a maximal block of consecutive equal signs.

**Definition (gap).** For a fixed sub-interval $[\alpha,\beta]\subset
[0,1]$, a **gap** is the number of consecutive generated values lying
outside $[\alpha,\beta]$ between two successive values that do lie
inside it.

**Definition (poker hand).** For a fixed integer $m$, the **poker
hand** associated with a generated value is the pattern of repeated
digits among its first $m$ decimal digits (for example, "all
different," "one pair," "three of a kind"), by direct analogy with a
hand of playing cards.

**Definition (serial correlation).** The **lag-$k$ serial
correlation** of the sequence $U_1,\dots,U_n$ is the sample
autocorrelation

$$
r_k = \frac{\sum_{i=1}^{n-k}(U_i-\bar U)(U_{i+k}-\bar U)}
{\sum_{i=1}^{n}(U_i-\bar U)^2}, \qquad \bar U = \frac1n\sum_{i=1}^n U_i.
\tag{2.2}
$$

# Theorems

## Theorem 2.1: Asymptotic Null Distribution of the Pearson Chi-square Statistic

**Statement.** Partition $[0,1]$ into $k$ equal bins, and let
$O_1,\dots,O_k$ be the observed counts of $U_1,\dots,U_n$ falling into
each bin. Under $H_0$, $(O_1,\dots,O_k)$ is Multinomial$(n; 1/k,
\dots,1/k)$, and the Pearson statistic

$$
\chi^2 = \sum_{i=1}^{k} \frac{(O_i - n/k)^2}{n/k}
\tag{2.3}
$$

converges in distribution, as $n\to\infty$ with $k$ fixed, to a
$\chi^2_{k-1}$ distribution.

**Assumptions.** The bins are fixed and non-random, each with equal
probability $1/k$ under $H_0$; $n\to\infty$ with $k$ held fixed.

**Proof.** By the multivariate central limit theorem applied to the
multinomial counts, writing $p_i=1/k$,

$$
\frac{1}{\sqrt n}\bigl(O_1-np_1,\dots,O_k-np_k\bigr)
\;\xrightarrow{d}\; N_k(\mathbf{0}, \boldsymbol\Sigma), \qquad
\boldsymbol\Sigma = \operatorname{diag}(\mathbf p) - \mathbf p\mathbf p',
$$

where $\mathbf p = (p_1,\dots,p_k)'$. The matrix $\boldsymbol\Sigma$ is
singular, with rank exactly $k-1$: it annihilates the vector
$\mathbf 1 = (1,\dots,1)'$, since $\boldsymbol\Sigma\mathbf1 =
\mathbf p - \mathbf p(\mathbf p'\mathbf 1) = \mathbf p - \mathbf p =
\mathbf 0$ (using $\mathbf p'\mathbf 1 = \sum_i p_i = 1$), reflecting
the linear constraint $\sum_i (O_i-np_i) = n-n=0$ that the counts must
always satisfy. The statistic (2.3) can be written as the quadratic
form

$$
\chi^2 = \frac1n \bigl(O-n\mathbf p\bigr)' D^{-1} \bigl(O-n\mathbf p\bigr),
\qquad D = \operatorname{diag}(\mathbf p),
$$

and it is a standard result for quadratic forms in a (possibly
singular) Normal vector (see Rao, 1973, §3b.4) that if
$V\sim N_k(\mathbf 0,\boldsymbol\Sigma)$ and $D^{-1}\boldsymbol\Sigma$
is idempotent of rank $r$, then $V'D^{-1}V \sim \chi^2_r$. Direct
computation confirms $D^{-1}\boldsymbol\Sigma = I_k - \mathbf 1\mathbf
p'$ is idempotent — $(I_k-\mathbf1\mathbf p')^2 = I_k - 2\mathbf1\mathbf
p' + \mathbf1(\mathbf p'\mathbf1)\mathbf p' = I_k-\mathbf1\mathbf p'$,
using $\mathbf p'\mathbf1=1$ — with rank equal to its trace,
$\operatorname{tr}(I_k-\mathbf1\mathbf p') = k - \mathbf p'\mathbf1 =
k-1$. Applying this result to the limiting vector
$V = \boldsymbol\Sigma^{1/2}Z$ ($Z$ standard Normal in $k$ dimensions)
shows the limiting distribution of $\chi^2$ is $\chi^2_{k-1}$.
$\blacksquare$

**Remarks.** The "loss" of exactly one degree of freedom, relative to
the naive guess of $k$, is a direct consequence of the single linear
constraint $\sum_i O_i = n$ that the multinomial counts must satisfy;
this is the same mechanism, generalized, that removes one degree of
freedom whenever a sample mean is estimated from the same data used to
compute a sum of squares elsewhere in statistics.

**Examples.** A full worked numerical example is given in Section 12.

**Applications.** Beyond testing uniformity of a random number
generator, the identical asymptotic argument underlies the Chi-square
goodness-of-fit test for any fully specified discrete or discretized
distribution, and the digit frequency and poker tests of this unit are
both direct applications of Theorem 2.1 to transformed data.

## Theorem 2.2: Mean Number of Runs in an I.I.D. Continuous Sequence

**Statement.** Let $U_1,\dots,U_n$ be i.i.d. with a continuous
distribution (so that ties occur with probability zero), and let $R$
denote the number of runs up and down in the sequence. Under $H_0$,

$$
E(R) = \frac{2n-1}{3}, \qquad
\operatorname{Var}(R) = \frac{16n-29}{90}.
\tag{2.4}
$$

**Assumptions.** The $U_i$ are i.i.d. from a continuous distribution
(only continuity is needed for $E(R)$; the specific distribution is
irrelevant, since runs depend only on the relative ranks of the
values).

**Proof (mean).** A run boundary occurs at every interior "turning
point": an index $i \in \{2,\dots,n-1\}$ at which $U_i$ is either a
local maximum ($U_{i-1}<U_i>U_{i+1}$) or a local minimum
($U_{i-1}>U_i<U_{i+1}$) among its two neighbours. Let $Y_i$ be the
indicator of a turning point at position $i$. Among the three values
$U_{i-1},U_i,U_{i+1}$, all $3! = 6$ orderings are equally likely by
continuity and exchangeability, and exactly $4$ of these $6$ orderings
place the middle-indexed value $U_i$ at either the largest or the
smallest of the three (the orderings in which $U_i$ ranks first or
third among the three, in whichever direction); the remaining $2$
orderings place $U_i$ in the middle rank. Hence
$P(Y_i=1) = 4/6 = 2/3$ for each of the $n-2$ interior positions
$i=2,\dots,n-1$. Writing $T = \sum_{i=2}^{n-1} Y_i$ for the total
number of turning points,

$$
E(T) = (n-2)\cdot\frac23 = \frac{2n-4}{3}.
$$

The number of runs and the number of turning points are related by the
elementary combinatorial identity $R = T+1$ (each turning point marks
the start of a new run, and there is always one more run than there
are turning points, since the sequence begins in some initial run
before the first turning point). Therefore

$$
E(R) = E(T)+1 = \frac{2n-4}{3}+1 = \frac{2n-1}{3},
$$

as claimed. $\blacksquare$

**Proof (variance, outline).** The variance requires
$\operatorname{Var}(T) = \sum_i \operatorname{Var}(Y_i) +
2\sum_{i<j}\operatorname{Cov}(Y_i,Y_j)$. Since $Y_i$ is a Bernoulli$(2/3)$
indicator, $\operatorname{Var}(Y_i) = \tfrac23\cdot\tfrac13 = \tfrac29$
for each of the $n-2$ terms. Turning-point indicators $Y_i$ and $Y_j$
with $|i-j|\ge 3$ depend on entirely disjoint triples of observations
and are therefore independent, contributing zero covariance; only the
lag-$1$ and lag-$2$ neighbouring pairs, which share one or two
observations, contribute non-zero covariance terms, each computable by
the same equally-likely-orderings technique used above but applied to
the $4$ (respectively $5$) jointly relevant observations. Carrying out
this enumeration in full (Knuth, 1997, §3.3.2, Theorem A) gives
$\operatorname{Cov}(Y_i,Y_{i+1}) = -\tfrac{7}{180}$ and
$\operatorname{Cov}(Y_i,Y_{i+2}) = \tfrac{1}{90}$ for interior
positions, and summing all contributions yields
$\operatorname{Var}(R) = \operatorname{Var}(T) = (16n-29)/90$. The
full accounting is a lengthy but elementary enumeration and is not
reproduced here; see the cited source for the complete derivation.

**Remarks.** For large $n$, a central limit theorem for sums of
locally dependent indicators (the dependence here extends only to
neighbours within lag $2$) gives the standardized statistic

$$
Z = \frac{R-E(R)}{\sqrt{\operatorname{Var}(R)}} \;\xrightarrow{d}\; N(0,1),
\tag{2.5}
$$

which is the basis of the runs test used in Section 8. The exact
formulas (2.4) are validated by direct Monte Carlo simulation in
Section 10.

**Examples.** A full worked numerical example is given in Section 12.

**Applications.** The runs test, built on this theorem, is one of the
most widely used independence diagnostics in random number generator
testing, and the same turning-point construction is used in
nonparametric trend-detection tests elsewhere in statistics (the
Cox-Stuart test and related procedures).

# Mathematical Derivations

## The Kolmogorov-Smirnov Statistic

For the ordered sample $U_{(1)} \le \cdots \le U_{(n)}$, the empirical
distribution function is $F_n(x) = n^{-1}\sum_{i=1}^n \mathbf
1\{U_i\le x\}$, a non-decreasing step function jumping by $1/n$ at
each order statistic. The one-sample Kolmogorov-Smirnov statistic

$$
D_n = \sup_{0\le x\le 1} |F_n(x)-x|
\tag{2.6}
$$

measures the largest vertical gap between $F_n$ and the Uniform$(0,1)$
cdf $F(x)=x$. Because $F_n$ is a step function and $F$ is continuous
and increasing, the supremum is attained either just before or exactly
at one of the jump points, giving the computable form

$$
D_n = \max_{1\le i\le n} \max\left\{ \frac{i}{n} - U_{(i)},\;
U_{(i)} - \frac{i-1}{n} \right\},
\tag{2.7}
$$

since between consecutive order statistics $F_n$ is constant while $F$
increases, so the maximum deviation on each such interval occurs at
one of its two endpoints. Kolmogorov (1933) showed that under $H_0$,
$\sqrt n\, D_n$ converges in distribution to a limit with cdf
$K(t) = 1-2\sum_{j=1}^\infty (-1)^{j-1}e^{-2j^2t^2}$; the proof
requires weak convergence of the empirical process to a Brownian
bridge and is beyond the scope of this course (proof omitted; see
Billingsley, 1968, Ch. 3, for the complete argument).

## The Digit Frequency Statistic

Writing each generated value in decimal form,
$U_i = 0.d_{i1}d_{i2}d_{i3}\dots$, and fixing a digit position $j$, the
digits $d_{1j},\dots,d_{nj}$ are, under $H_0$, i.i.d. discrete
Uniform on $\{0,1,\dots,9\}$ (a direct consequence of the Uniform$(0,1)$
density being flat, so every decimal digit at every position is
equally likely to be any of the ten values). Letting $O_d$ denote the
observed count of digit $d$, the statistic

$$
\chi^2_{\text{digit}} = \sum_{d=0}^{9} \frac{(O_d - n/10)^2}{n/10}
\tag{2.8}
$$

is exactly the Theorem 2.1 statistic with $k=10$ equiprobable
categories, so it is asymptotically $\chi^2_9$ under $H_0$.

## The Poker Test Probabilities

For $m=5$ decimal digits per hand (digits $0$-$9$, drawn with
repetition, $10^5$ equally likely digit strings), the probability of
each hand type is computed by direct enumeration. For **all
different**, the first digit can be any of $10$, the second any of the
remaining $9$, and so on:

$$
P(\text{all different}) = \frac{10\cdot9\cdot8\cdot7\cdot6}{10^5}
= \frac{30240}{100000} = 0.3024.
\tag{2.9}
$$

For **one pair**, choose which $2$ of the $5$ positions form the pair
($\binom{5}{2}=10$ ways), choose the pair's digit ($10$ ways), and fill
the remaining $3$ positions with distinct digits different from the
pair's digit, in order ($9\times8\times7=504$ ways):

$$
P(\text{one pair}) = \frac{\binom{5}{2}\cdot 10 \cdot 9\cdot8\cdot7}{10^5}
= \frac{50400}{100000} = 0.5040.
\tag{2.10}
$$

The remaining five categories are obtained by the identical
position-counting-times-digit-counting technique (Table 1), and the
seven probabilities sum to exactly $1$, as required of a complete
partition of the outcome space.

```{r poker-probability-table}
poker_hand_probabilities <- tibble(
  Hand_Type = c(
    "All different", "One pair", "Two pairs", "Three of a kind",
    "Full house", "Four of a kind", "Five of a kind"
  ),
  Probability = c(
    prod(10:6) / 10^5,
    choose(5, 2) * 10 * 9 * 8 * 7 / 10^5,
    (choose(5, 2) * choose(3, 2) / 2) * 10 * 9 * 8 / 10^5,
    choose(5, 3) * 10 * 9 * 8 / 10^5,
    choose(5, 3) * 10 * 9 / 10^5,
    choose(5, 4) * 10 * 9 / 10^5,
    10 / 10^5
  )
)

kable(poker_hand_probabilities, digits = 4,
      caption = "Table 1: Theoretical poker-hand probabilities for m=5 digits") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))

cat("Sum of probabilities:", sum(poker_hand_probabilities$Probability), "\n")
```

# Algorithms

## Algorithm 2.1: Chi-square Goodness-of-Fit Test for Uniformity

**Purpose.** Test whether a generated sample is consistent with the
Uniform$(0,1)$ distribution by comparing observed and expected bin
counts.

**Input.** Sample $U_1,\dots,U_n$; number of bins $k$.

**Output.** Test statistic $\chi^2$ and $p$-value.

**Step-by-step procedure.**

1. Partition $[0,1]$ into $k$ equal-width bins.
2. Compute the observed count $O_i$ in each bin.
3. Compute $\chi^2 = \sum_i (O_i-n/k)^2/(n/k)$.
4. Compute $p = 1-F_{\chi^2_{k-1}}(\chi^2)$; reject $H_0$ if $p$ is
   small.

**Pseudo-code.**

```
Algorithm: Chi-square Goodness-of-Fit Test
Input:  u[1..n], k
Output: chi-square statistic, p-value

1. bins <- k equal-width intervals partitioning [0,1]
2. O[1..k] <- observed counts of u in each bin
3. E <- n / k
4. chi_sq <- sum((O - E)^2 / E)
5. p_value <- 1 - pchisq(chi_sq, df = k - 1)
6. Return chi_sq, p_value
```

**Flowchart description.** Start; bin the sample; compute observed
counts; compute expected counts under uniformity; compute the
Chi-square statistic; compare to the $\chi^2_{k-1}$ distribution to
obtain a $p$-value; end.

**Complexity.** $O(n+k)$: binning the sample is $O(n)$, and computing
the statistic from the counts is $O(k)$.

**Practical considerations.** Choose $k$ so that every expected count
$n/k$ is at least $5$, the conventional threshold for the asymptotic
$\chi^2$ approximation of Theorem 2.1 to be adequate; merge adjacent
bins if this threshold is violated.

## Algorithm 2.2: Kolmogorov-Smirnov Test for Uniformity

**Purpose.** Test uniformity by comparing the empirical distribution
function directly to the Uniform$(0,1)$ cdf, without binning.

**Input.** Sample $U_1,\dots,U_n$.

**Output.** Test statistic $D_n$ and (asymptotic) $p$-value.

**Step-by-step procedure.**

1. Sort the sample to obtain order statistics $U_{(1)}\le\cdots\le
   U_{(i)}$.
2. Compute $D_n$ using the closed form (2.7).
3. Obtain the $p$-value from the Kolmogorov limiting distribution (or
   an exact finite-sample table for small $n$).

**Pseudo-code.**

```
Algorithm: Kolmogorov-Smirnov Test
Input:  u[1..n]
Output: D_n, p-value

1. Sort u to get u_(1) <= ... <= u_(n)
2. For i = 1 to n:
       d_plus[i]  <- i/n - u_(i)
       d_minus[i] <- u_(i) - (i-1)/n
3. D_n <- max(d_plus, d_minus)
4. p_value <- P(sup |F_n - F| >= D_n) under H0
5. Return D_n, p_value
```

**Flowchart description.** Start; sort the sample; compute, at each
order statistic, the two one-sided deviations from the target cdf;
take the overall maximum; compare to the Kolmogorov distribution;
end.

**Complexity.** $O(n\log n)$, dominated by sorting.

**Practical considerations.** Unlike the Chi-square test, the K-S test
requires no arbitrary binning choice and remains valid for any sample
size, including small $n$, using exact tables; it is generally more
powerful than the Chi-square test against smooth departures from
uniformity, but less naturally suited to discrete or heavily
tied data.

## Algorithm 2.3: Runs Test for Independence

**Purpose.** Test whether the ordering of a sequence shows evidence of
serial dependence, using the number of up/down runs.

**Input.** Sample $U_1,\dots,U_n$.

**Output.** Standardized statistic $Z$ and (asymptotic) $p$-value.

**Step-by-step procedure.**

1. Compute the sign sequence $S_i = \operatorname{sign}(U_{i+1}-U_i)$
   for $i=1,\dots,n-1$.
2. Count the number of runs $R$ (maximal blocks of consecutive equal
   signs).
3. Compute $Z = (R-E(R))/\sqrt{\operatorname{Var}(R)}$ using (2.4).
4. Compute $p = 2(1-\Phi(|Z|))$.

**Pseudo-code.**

```
Algorithm: Runs Test
Input:  u[1..n]
Output: Z, p-value

1. s[1..n-1] <- sign(u[i+1] - u[i])
2. R <- 1 + count of i such that s[i] != s[i-1]
3. E_R <- (2n - 1) / 3
4. V_R <- (16n - 29) / 90
5. Z <- (R - E_R) / sqrt(V_R)
6. p_value <- 2 * (1 - Phi(abs(Z)))
7. Return Z, p_value
```

**Flowchart description.** Start; compute the sign of each consecutive
difference; count runs of consecutive equal signs; standardize using
the exact mean and variance of Theorem 2.2; compare to the standard
Normal distribution; end.

**Complexity.** $O(n)$.

**Practical considerations.** The runs test detects *both* excessive
regularity (too few runs, indicating positive dependence, values
tending to stay high or low for stretches) and excessive alternation
(too many runs, indicating negative dependence, values tending to
zig-zag more than chance allows); a two-sided test is therefore always
appropriate.

# R Programming

```{r prng-test-functions}
set.seed(123)

#' Chi-square goodness-of-fit test for Uniform(0,1)
#'
#' Purpose: implements Algorithm 2.1.
#'
#' Arguments:
#'   sample_values   numeric vector in [0, 1]
#'   number_of_bins  number of equal-width bins (default 10)
#'
#' Returns:
#'   a one-row tibble with the test name, statistic, and p-value
#'
#' Example:
#'   chi_square_uniformity_test(runif(1000))
chi_square_uniformity_test <- function(sample_values, number_of_bins = 10) {
  sample_size <- length(sample_values)
  bin_assignment <- cut(
    sample_values,
    breaks = seq(0, 1, length.out = number_of_bins + 1),
    include.lowest = TRUE
  )
  observed_counts <- as.numeric(table(bin_assignment))
  expected_counts <- rep(sample_size / number_of_bins, number_of_bins)

  chi_square_statistic <- sum((observed_counts - expected_counts)^2 / expected_counts)
  p_value <- 1 - pchisq(chi_square_statistic, df = number_of_bins - 1)

  tibble(Test = "Chi-square", Statistic = chi_square_statistic, P_Value = p_value)
}

#' Kolmogorov-Smirnov test for Uniform(0,1)
#'
#' Purpose: implements Algorithm 2.2 using R's built-in ks.test().
#'
#' Arguments:
#'   sample_values   numeric vector in [0, 1]
#'
#' Returns:
#'   a one-row tibble with the test name, statistic, and p-value
#'
#' Example:
#'   kolmogorov_smirnov_uniformity_test(runif(1000))
kolmogorov_smirnov_uniformity_test <- function(sample_values) {
  test_result <- ks.test(sample_values, "punif", min = 0, max = 1)
  tibble(
    Test = "Kolmogorov-Smirnov",
    Statistic = as.numeric(test_result$statistic),
    P_Value = test_result$p.value
  )
}

#' Digit frequency chi-square test
#'
#' Purpose: implements the test of Section 7.2, applied to a chosen
#' decimal digit position.
#'
#' Arguments:
#'   sample_values     numeric vector in [0, 1]
#'   digit_position    which decimal digit to test (1 = first digit)
#'
#' Returns:
#'   a one-row tibble with the test name, statistic, and p-value
#'
#' Example:
#'   digit_frequency_test(runif(1000), digit_position = 1)
digit_frequency_test <- function(sample_values, digit_position = 1) {
  sample_size <- length(sample_values)
  extracted_digit <- floor((sample_values * 10^digit_position) %% 10)
  observed_counts <- as.numeric(table(factor(extracted_digit, levels = 0:9)))
  expected_counts <- rep(sample_size / 10, 10)

  chi_square_statistic <- sum((observed_counts - expected_counts)^2 / expected_counts)
  p_value <- 1 - pchisq(chi_square_statistic, df = 9)

  tibble(
    Test = paste0("Digit Frequency (position ", digit_position, ")"),
    Statistic = chi_square_statistic, P_Value = p_value
  )
}

#' Runs (up and down) test for independence
#'
#' Purpose: implements Algorithm 2.3.
#'
#' Arguments:
#'   sample_values   numeric vector in [0, 1]
#'
#' Returns:
#'   a one-row tibble with the test name, statistic, and p-value
#'
#' Example:
#'   runs_test(runif(1000))
runs_test <- function(sample_values) {
  sample_size <- length(sample_values)
  sign_sequence <- sign(diff(sample_values))
  sign_sequence <- sign_sequence[sign_sequence != 0]

  number_of_runs <- 1 + sum(sign_sequence[-1] != sign_sequence[-length(sign_sequence)])

  expected_runs <- (2 * sample_size - 1) / 3
  variance_runs <- (16 * sample_size - 29) / 90
  z_statistic <- (number_of_runs - expected_runs) / sqrt(variance_runs)
  p_value <- 2 * (1 - pnorm(abs(z_statistic)))

  tibble(Test = "Runs (Up/Down)", Statistic = z_statistic, P_Value = p_value)
}

#' Chi-square helper with automatic bin merging for low expected counts
#'
#' Purpose: merges adjacent categories with expected count below 5
#' before computing the Chi-square statistic, following the practical
#' consideration noted for Algorithm 2.1.
#'
#' Arguments:
#'   observed_counts    numeric vector of observed counts
#'   expected_counts    numeric vector of expected counts (same length)
#'
#' Returns:
#'   a list with the chi-square statistic and its degrees of freedom
merge_and_compute_chi_square <- function(observed_counts, expected_counts) {
  index <- 1
  while (index <= length(expected_counts)) {
    if (expected_counts[index] < 5 && length(expected_counts) > 1) {
      merge_target <- if (index < length(expected_counts)) index + 1 else index - 1
      expected_counts[merge_target] <- expected_counts[merge_target] + expected_counts[index]
      observed_counts[merge_target] <- observed_counts[merge_target] + observed_counts[index]
      expected_counts <- expected_counts[-index]
      observed_counts <- observed_counts[-index]
    } else {
      index <- index + 1
    }
  }

  degrees_of_freedom <- length(observed_counts) - 1
  chi_square_statistic <- sum((observed_counts - expected_counts)^2 / expected_counts)

  list(statistic = chi_square_statistic, df = degrees_of_freedom)
}

#' Gap test for independence
#'
#' Purpose: tests whether the lengths of gaps between successive
#' values falling in a fixed sub-interval follow the geometric
#' distribution implied by independence.
#'
#' Arguments:
#'   sample_values   numeric vector in [0, 1]
#'   lower_bound     lower endpoint of the target interval
#'   upper_bound     upper endpoint of the target interval
#'
#' Returns:
#'   a one-row tibble with the test name, statistic, and p-value
#'
#' Example:
#'   gap_test(runif(2000), lower_bound = 0, upper_bound = 0.5)
gap_test <- function(sample_values, lower_bound = 0, upper_bound = 0.5) {
  interval_probability <- upper_bound - lower_bound
  hit_positions <- which(sample_values >= lower_bound & sample_values < upper_bound)

  if (length(hit_positions) < 2) {
    return(tibble(Test = "Gap Test", Statistic = NA_real_, P_Value = NA_real_))
  }

  gap_lengths <- diff(hit_positions) - 1
  max_gap <- max(gap_lengths)

  observed_counts <- as.numeric(table(factor(gap_lengths, levels = 0:max_gap)))
  expected_probabilities <- interval_probability * (1 - interval_probability)^(0:max_gap)
  expected_counts <- sum(observed_counts) * expected_probabilities

  merged_result <- merge_and_compute_chi_square(observed_counts, expected_counts)
  p_value <- 1 - pchisq(merged_result$statistic, df = merged_result$df)

  tibble(Test = "Gap Test", Statistic = merged_result$statistic, P_Value = p_value)
}

#' Poker test for independence
#'
#' Purpose: classifies groups of m decimal digits into poker-hand
#' categories and compares observed frequencies to the theoretical
#' probabilities derived in Table 1.
#'
#' Arguments:
#'   sample_values     numeric vector in [0, 1]
#'   digits_per_hand   number of decimal digits per hand (default 5)
#'
#' Returns:
#'   a one-row tibble with the test name, statistic, and p-value
#'
#' Example:
#'   poker_test(runif(2000))
poker_test <- function(sample_values, digits_per_hand = 5) {
  sample_size <- length(sample_values)
  five_digit_integers <- floor(sample_values * 10^digits_per_hand)

  classify_poker_hand <- function(integer_value) {
    digit_string <- formatC(integer_value, width = digits_per_hand, flag = "0")
    digit_counts <- table(strsplit(digit_string, "")[[1]])

    if (any(digit_counts == 5)) return("Five of a kind")
    if (any(digit_counts == 4)) return("Four of a kind")
    if (any(digit_counts == 3) && any(digit_counts == 2)) return("Full house")
    if (any(digit_counts == 3)) return("Three of a kind")
    if (sum(digit_counts == 2) == 2) return("Two pairs")
    if (any(digit_counts == 2)) return("One pair")
    "All different"
  }

  hand_labels <- c(
    "All different", "One pair", "Two pairs", "Three of a kind",
    "Full house", "Four of a kind", "Five of a kind"
  )
  observed_hands <- vapply(five_digit_integers, classify_poker_hand, character(1))
  observed_counts <- as.numeric(table(factor(observed_hands, levels = hand_labels)))

  theoretical_probabilities <- c(
    prod(10:6) / 10^5,
    choose(5, 2) * 10 * 9 * 8 * 7 / 10^5,
    (choose(5, 2) * choose(3, 2) / 2) * 10 * 9 * 8 / 10^5,
    choose(5, 3) * 10 * 9 * 8 / 10^5,
    choose(5, 3) * 10 * 9 / 10^5,
    choose(5, 4) * 10 * 9 / 10^5,
    10 / 10^5
  )
  expected_counts <- sample_size * theoretical_probabilities

  merged_result <- merge_and_compute_chi_square(observed_counts, expected_counts)
  p_value <- 1 - pchisq(merged_result$statistic, df = merged_result$df)

  tibble(Test = "Poker Test", Statistic = merged_result$statistic, P_Value = p_value)
}

#' Serial correlation test for independence
#'
#' Purpose: tests the lag-k sample autocorrelation against its
#' asymptotic Normal null distribution.
#'
#' Arguments:
#'   sample_values   numeric vector in [0, 1]
#'   lag             the lag k (default 1)
#'
#' Returns:
#'   a one-row tibble with the test name, statistic, and p-value
#'
#' Example:
#'   serial_correlation_test(runif(1000), lag = 1)
serial_correlation_test <- function(sample_values, lag = 1) {
  sample_size <- length(sample_values)
  sample_mean <- mean(sample_values)

  numerator <- sum(
    (sample_values[1:(sample_size - lag)] - sample_mean) *
      (sample_values[(1 + lag):sample_size] - sample_mean)
  ) / (sample_size - lag)
  denominator <- sum((sample_values - sample_mean)^2) / sample_size

  lag_k_correlation <- numerator / denominator
  z_statistic <- lag_k_correlation * sqrt(sample_size - lag)
  p_value <- 2 * (1 - pnorm(abs(z_statistic)))

  tibble(
    Test = paste0("Serial Correlation (lag ", lag, ")"),
    Statistic = z_statistic, P_Value = p_value
  )
}

#' Apply the full battery of uniformity and independence tests
#'
#' Purpose: applies all seven tests of this unit to a single sample,
#' using the same sample for every test so that comparisons across
#' generators are consistent.
#'
#' Arguments:
#'   sample_values   numeric vector in [0, 1]
#'
#' Returns:
#'   a tibble with one row per test
#'
#' Example:
#'   evaluate_generator(runif(2000))
evaluate_generator <- function(sample_values) {
  bind_rows(
    chi_square_uniformity_test(sample_values),
    kolmogorov_smirnov_uniformity_test(sample_values),
    digit_frequency_test(sample_values, digit_position = 1),
    runs_test(sample_values),
    gap_test(sample_values, lower_bound = 0, upper_bound = 0.5),
    poker_test(sample_values),
    serial_correlation_test(sample_values, lag = 1)
  )
}
```

# Simulation Studies

## Testing a Good Generator Against a Deliberately Poor One

This study compares R's default generator (`runif()`, Mersenne
Twister) against the small-modulus mixed congruential generator with
$a=5$, $c=3$, $m=16$ studied in Unit 3.1 (verified there, via the
Hull-Dobell theorem, to have full period exactly $16$, far too short
for a sample of $2000$), applying the full battery of Section 8-9 to
each.

```{r generator-comparison}
set.seed(123)

sample_size <- 2000
good_generator_sample <- runif(sample_size)

poor_lcg_generator <- function(seed, multiplier, increment, modulus, size) {
  state <- numeric(size)
  current_value <- seed
  for (index in seq_len(size)) {
    current_value <- (multiplier * current_value + increment) %% modulus
    state[index] <- current_value
  }
  state / modulus
}
poor_generator_sample <- poor_lcg_generator(
  seed = 7, multiplier = 5, increment = 3, modulus = 16, size = sample_size
)

results_good <- evaluate_generator(good_generator_sample) %>% mutate(Generator = "Good (runif)")
results_poor <- evaluate_generator(poor_generator_sample) %>% mutate(Generator = "Poor (LCG, m=16)")

conclusion_label <- function(p_val) {
  if (is.na(p_val)) return("Insufficient data")
  if (p_val < 0.05) "Reject H0 (evidence against uniformity/independence)" else "Fail to reject H0"
}

all_results <- bind_rows(results_good, results_poor) %>%
  mutate(
    Conclusion = vapply(P_Value, conclusion_label, character(1)),
    Statistic = round(Statistic, 4),
    P_Value = round(P_Value, 4)
  ) %>%
  select(Generator, Test, Statistic, P_Value, Conclusion)

kable(all_results, caption = "Full test battery applied to a good and a deliberately poor generator") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

As expected, the deliberately poor generator, whose period of $16$ is
vastly shorter than the sample size of $2000$, fails essentially every
test (the sample necessarily repeats the same $16$ values $125$ times
over, producing extreme non-uniformity and extreme serial dependence),
while the good generator fails none of them.

## Empirical Validation of the Runs Test Moments (Theorem 2.2)

```{r runs-moment-validation}
set.seed(123)

validate_n <- 50
number_of_replications <- 5000

count_runs <- function(sample_values) {
  sign_sequence <- sign(diff(sample_values))
  sign_sequence <- sign_sequence[sign_sequence != 0]
  1 + sum(sign_sequence[-1] != sign_sequence[-length(sign_sequence)])
}

simulated_run_counts <- replicate(
  number_of_replications, count_runs(runif(validate_n))
)

tibble(
  Quantity = c("Theoretical E(R)", "Empirical mean R",
               "Theoretical Var(R)", "Empirical variance R"),
  Value = c(
    (2 * validate_n - 1) / 3, mean(simulated_run_counts),
    (16 * validate_n - 29) / 90, var(simulated_run_counts)
  )
) %>%
  kable(digits = 4, caption = "Theorem 2.2 moments vs. Monte Carlo simulation, n = 50")
```

The close agreement between the theoretical and empirical moments
validates the formulas of Theorem 2.2, including the variance formula
whose full derivation was only outlined above.

# Graphs

```{r lag-plot-comparison, fig.cap="Lag-1 scatterplots: the good generator (left) fills the unit square, while the poor generator (right), with period only 16, visits just 16 distinct points."}
lag_df_good <- tibble(
  u_now = good_generator_sample[1:(sample_size - 1)],
  u_next = good_generator_sample[2:sample_size]
)
lag_df_poor <- tibble(
  u_now = poor_generator_sample[1:(sample_size - 1)],
  u_next = poor_generator_sample[2:sample_size]
)

p_good <- ggplot(lag_df_good, aes(x = u_now, y = u_next)) +
  geom_point(alpha = 0.25, size = 0.5, colour = "steelblue") +
  labs(title = "Good generator (runif)", x = expression(U[n]), y = expression(U[n+1])) +
  theme_minimal(base_size = 11)

p_poor <- ggplot(lag_df_poor, aes(x = u_now, y = u_next)) +
  geom_point(alpha = 0.5, size = 1.2, colour = "firebrick") +
  labs(title = "Poor generator (LCG, m=16)", x = expression(U[n]), y = expression(U[n+1])) +
  theme_minimal(base_size = 11)

p_good + p_poor
```

```{r runs-histogram, fig.cap="Simulated distribution of the number of runs (n = 50) against the Normal approximation implied by Theorem 2.2."}
ggplot(tibble(runs = simulated_run_counts), aes(x = runs)) +
  geom_histogram(aes(y = after_stat(density)), binwidth = 1, fill = "darkgreen", colour = "white") +
  stat_function(
    fun = dnorm,
    args = list(mean = (2 * validate_n - 1) / 3, sd = sqrt((16 * validate_n - 29) / 90)),
    colour = "darkred", linewidth = 1
  ) +
  labs(title = "Distribution of the number of runs (n = 50)", x = "Number of runs R", y = "Density") +
  theme_minimal(base_size = 11)
```

# Numerical Examples

**Example 1: Chi-square goodness-of-fit test.**

*Problem.* A sample of $n=40$ values is binned into $k=4$ equal bins
with observed counts $O = (8, 12, 11, 9)$. Test uniformity at the
$5\%$ level.

*Solution.* Expected count per bin: $E = 40/4 = 10$.

$$
\chi^2 = \frac{(8-10)^2}{10} + \frac{(12-10)^2}{10} + \frac{(11-10)^2}{10} + \frac{(9-10)^2}{10}
= \frac{4+4+1+1}{10} = 1.0.
$$

With $df=k-1=3$, the critical value at $5\%$ is $\chi^2_{3,0.05}=
7.815$; since $1.0 < 7.815$, fail to reject $H_0$.

*Interpretation.* The observed counts are close to their expected
value of $10$ in every bin, so there is no evidence against
uniformity; this is expected of a well-behaved generator.

**Example 2: Kolmogorov-Smirnov test.**

*Problem.* For the ordered sample $u_{(1)},\dots,u_{(5)} = 0.10,
0.35, 0.42, 0.68, 0.90$, compute $D_5$.

*Solution.* Using (2.7) with $n=5$:

| $i$ | $u_{(i)}$ | $i/n - u_{(i)}$ | $u_{(i)} - (i-1)/n$ |
|---|---|---|---|
| 1 | 0.10 | 0.10 | 0.10 |
| 2 | 0.35 | 0.05 | 0.15 |
| 3 | 0.42 | 0.18 | 0.02 |
| 4 | 0.68 | 0.12 | 0.08 |
| 5 | 0.90 | 0.10 | 0.10 |

The largest entry in either column is $0.18$, so $D_5 = 0.18$.

*Interpretation.* The largest deviation occurs at $i=3$, where the
third order statistic ($0.42$) falls well short of its expected
position ($3/5=0.6$); with only $n=5$ observations, a deviation of
this size is not unusual, and exact small-sample K-S tables (rather
than the asymptotic Kolmogorov distribution) would be used for a
formal decision at this sample size.

**Example 3: Runs test.**

*Problem.* For the sequence
$0.51, 0.62, 0.40, 0.85, 0.33, 0.71, 0.29, 0.58, 0.44, 0.77$
($n=10$), compute the runs test statistic.

*Solution.* The signs of consecutive differences are
$+,-,+,-,+,-,+,-,+$ (the sequence alternates at every single step), so
every sign change creates a new run: $R=9$. From (2.4) with $n=10$,
$E(R) = 19/3 = 6.333$ and $\operatorname{Var}(R) = 131/90 = 1.4556$,
giving

$$
Z = \frac{9 - 6.333}{\sqrt{1.4556}} = \frac{2.667}{1.2065} = 2.211.
$$

Since $|Z|=2.211 > 1.96$, reject $H_0$ at the $5\%$ level.

*Interpretation.* This hypothetical sequence alternates far more
regularly than randomness would produce (an i.i.d. sequence has a
$2/3$ chance of a turning point at each interior position, not
certainty), and the runs test correctly flags this excessive
alternation as evidence of negative serial dependence.

**Example 4: Serial correlation test.**

*Problem.* For the sequence $0.2, 0.5, 0.3, 0.8, 0.4, 0.6, 0.1, 0.9$
($n=8$), compute the lag-1 serial correlation statistic.

*Solution.* $\bar u = 3.8/8 = 0.475$. Computing (2.2) term by term
gives numerator sum $-0.30813$ (divided by $n-k=7$: $-0.04402$) and
denominator sum $0.555$ (divided by $n=8$: $0.06938$), so
$r_1 = -0.04402/0.06938 = -0.6345$, and
$Z = r_1\sqrt{n-k} = -0.6345\sqrt7 = -1.679$.

*Interpretation.* Since $|Z|=1.679 < 1.96$, this particular short
sequence does not provide significant evidence of lag-1 dependence at
the $5\%$ level, even though the sample correlation itself
($-0.63$) looks numerically large — a reminder that with only $n=8$
observations, sample correlations are highly variable under $H_0$ and
formal testing, not eyeballing the coefficient, is required.

# Applications

- **Statistics.** Certifying the pseudo-random number generator
  underlying any statistical software package before it is used for
  simulation-based inference, bootstrap resampling, or power analysis.
- **Machine Learning.** Verifying that random train-test splits,
  dropout masks, and weight initializations are drawn from a generator
  free of the sequential dependence that would bias cross-validation
  estimates.
- **Data Science.** Auditing third-party or legacy random number
  generators embedded in data pipelines before trusting synthetic data
  or sampling procedures built on them.
- **Finance.** Validating the Monte Carlo engines used for derivative
  pricing and risk simulation, where an undetected generator defect
  can silently bias a valuation.
- **Reliability Engineering.** Certifying the generator used in
  Monte Carlo reliability and Weibull-analysis software before
  deployment in safety-critical design work.
- **Monte Carlo Methods.** Testing is a mandatory pre-requisite,
  not an optional check, before any large-scale Monte Carlo study,
  precisely because an undetected flaw of the kind exhibited by RANDU
  can corrupt every downstream result without any visible error.
- **Bayesian Statistics.** Verifying the pseudo-random substrate
  underlying a Markov chain Monte Carlo sampler, since the validity of
  MCMC convergence diagnostics presumes a trustworthy underlying
  uniform stream.
- **Operations Research.** Validating simulation software used for
  queueing, inventory, and network reliability studies.
- **Engineering.** Certifying random number generators used in
  Monte Carlo tolerance analysis and structural reliability
  simulations.

# Advantages

- The tests in this unit require no knowledge of the generator's
  internal algorithm; they operate purely on its output, so they apply
  equally to congruential generators, the Mersenne Twister, or any
  other generator.
- Uniformity and independence tests are complementary and together
  provide broad diagnostic coverage, as illustrated by the RANDU case,
  which independence tests catch but naive uniformity tests alone
  would miss.
- The tests are computationally cheap and can be run routinely as a
  sanity check before any simulation study.

# Limitations

- No finite battery of tests can *prove* a generator is fit for use;
  it can only fail to detect a defect that the chosen tests happen not
  to be sensitive to. A generator that passes every test in this unit
  may still fail a test not included here.
- Applying many tests to the same sample inflates the overall false
  positive rate: at a nominal $5\%$ level per test, running all seven
  tests in this unit's battery gives a substantially higher than $5\%$
  chance of at least one false rejection under $H_0$, purely by
  chance, if the tests are treated as independent.
- Passing statistical tests of the kind in this unit is necessary but
  not sufficient for cryptographic applications, which require
  unpredictability against an adversary with knowledge of prior
  outputs — a much stronger requirement than the tests here are
  designed to detect.

# Practical Tips

- **Apply both uniformity and independence tests.** The RANDU example
  demonstrates that either family alone can miss a serious defect that
  the other family catches immediately.
- **Account for multiple testing when running a full battery.** A
  Bonferroni-adjusted significance level ($\alpha/m$ for $m$ tests) or
  a holistic reading of the pattern of $p$-values across tests is more
  defensible than treating each test's $5\%$ threshold in isolation.
- **Use a sample size appropriate to the test.** The Chi-square and
  digit frequency tests require expected bin counts of at least $5$
  for the asymptotic approximation to be trustworthy; the runs test
  and serial correlation test rely on asymptotic normality that is
  poor for very small $n$, as Example 4 illustrates.
- **Do not over-interpret a single rejected test as proof of a fatal
  flaw.** At the $5\%$ level, one test in twenty will reject a
  perfectly good generator purely by chance; a pattern of rejections
  across several independent tests, or a very extreme $p$-value, is
  much stronger evidence than an isolated borderline rejection.
- **Re-verify any hand-rolled combinatorial formula (as in the poker
  test) against a computational check before trusting it.** The
  poker-hand probabilities derived in Section 7.3 were verified to sum
  to exactly $1$ before use; an error in such a formula silently
  invalidates the corresponding test's null distribution.

# Summary

- Full period alone (Unit 3.1) does not certify a generator for
  statistical use; empirical testing of the output is also required.
- Uniformity tests (Chi-square, Kolmogorov-Smirnov, digit frequency)
  check whether individual values follow Uniform$(0,1)$; independence
  tests (runs, gap, poker, serial correlation) check whether the
  ordering of values shows exploitable structure.
- Theorem 2.1 establishes the asymptotic $\chi^2_{k-1}$ null
  distribution of the Pearson statistic; Theorem 2.2 gives the exact
  mean and variance of the number of runs, validated here both
  analytically (for the mean) and by simulation (for the variance).
- The digit frequency and poker tests are direct applications of the
  Chi-square framework to transformed (digit-based) data.
- No battery of tests can prove a generator is defect-free, and
  running many tests on the same sample raises multiple-testing
  considerations that must be accounted for when interpreting results.

# Key Formulae

- Chi-square statistic: $\chi^2=\sum_i (O_i-E_i)^2/E_i \sim \chi^2_{k-1}$.
- K-S statistic: $D_n = \max_i \max\{i/n-U_{(i)},\, U_{(i)}-(i-1)/n\}$.
- Runs test moments: $E(R)=(2n-1)/3$, $\operatorname{Var}(R)=(16n-29)/90$.
- Serial correlation: $r_k = \dfrac{\sum_{i=1}^{n-k}(U_i-\bar U)(U_{i+k}-\bar U)}{\sum_{i=1}^n (U_i-\bar U)^2}$,
  $Z=r_k\sqrt{n-k}$.
- Poker-hand probabilities (m=5 digits): all different $0.3024$; one
  pair $0.5040$; two pairs $0.1080$; three of a kind $0.0720$; full
  house $0.0090$; four of a kind $0.0045$; five of a kind $0.0001$.

# Key Terms

- **Gap test.** An independence test based on the lengths of gaps
  between successive values falling inside a fixed interval.
- **Kolmogorov-Smirnov test.** A uniformity test comparing the
  empirical distribution function to the target cdf directly.
- **Multiple testing.** The inflation of the overall false-positive
  rate that results from applying several hypothesis tests to the same
  data.
- **Pearson chi-square statistic.** A goodness-of-fit statistic
  comparing observed and expected category counts.
- **Poker test.** An independence test classifying groups of digits by
  their repeated-digit pattern.
- **Run.** A maximal subsequence of consecutive strictly increasing or
  strictly decreasing values.
- **Serial correlation.** The sample autocorrelation between values
  separated by a fixed lag.
- **Turning point.** An interior position at which a sequence changes
  from increasing to decreasing or vice versa.

# Exercises

## Multiple Choice Questions

1. The joint null hypothesis tested throughout this unit is that the
   generated values are
   (a) independent but not necessarily uniform
   (b) uniform but not necessarily independent
   (c) i.i.d. Uniform(0,1)
   (d) Normally distributed

2. Under $H_0$, the Pearson chi-square statistic (2.3) is
   asymptotically distributed as
   (a) $\chi^2_k$
   (b) $\chi^2_{k-1}$
   (c) $N(0,1)$
   (d) $t_{k-1}$

3. The Kolmogorov-Smirnov statistic $D_n$ measures
   (a) the largest vertical gap between the empirical and target cdf
   (b) the sample correlation at lag 1
   (c) the number of runs in the sequence
   (d) the chi-square statistic on digit counts

4. The runs test is sensitive to
   (a) non-uniformity only
   (b) both excessive regularity and excessive alternation
   (c) excessive alternation only
   (d) tied values only

5. Under $H_0$, the mean number of runs for a sequence of length $n$
   is
   (a) $n/2$
   (b) $(2n-1)/3$
   (c) $(16n-29)/90$
   (d) $n-1$

6. In the proof of Theorem 2.2, the probability that an interior point
   is a turning point is
   (a) $1/2$
   (b) $1/3$
   (c) $2/3$
   (d) $3/4$

7. The RANDU generator of Unit 3.1 is an example of a generator that
   (a) fails simple uniformity tests but passes independence tests
   (b) passes simple uniformity tests but fails independence tests
   (c) fails both uniformity and independence tests
   (d) passes all known statistical tests

8. Running seven tests, each at the 5% level, on the same sample
   (a) has no effect on the overall false positive rate
   (b) decreases the overall false positive rate
   (c) increases the overall false positive rate
   (d) makes the tests exact rather than asymptotic

9. In the poker test with $m=5$ digits, the probability of "all
   different" is
   (a) $0.5040$
   (b) $0.3024$
   (c) $0.1080$
   (d) $0.0720$

10. The digit frequency test is asymptotically distributed as
    (a) $\chi^2_9$
    (b) $\chi^2_{10}$
    (c) $N(0,1)$
    (d) $\chi^2_1$

**Answers.** 1(c), 2(b), 3(a), 4(b), 5(b), 6(c), 7(b), 8(c), 9(b),
10(a).

## Short Answer Questions

1. Explain why full period, as certified by the Hull-Dobell theorem,
   does not by itself guarantee a generator is fit for statistical
   use.
2. State the joint null hypothesis common to every test in this unit.
3. Distinguish uniformity tests from independence tests, giving one
   example of each.
4. Explain why the Pearson chi-square statistic loses exactly one
   degree of freedom relative to the number of categories.
5. Define a "run" and explain the identity $R=T+1$ relating the
   number of runs to the number of turning points.
6. Explain why the digit frequency test can be viewed as a special
   case of the Chi-square goodness-of-fit test.
7. Why does the K-S test require no arbitrary binning choice, unlike
   the Chi-square test?
8. Explain, using the RANDU example, why both uniformity and
   independence tests are needed.
9. What practical adjustment should be made when applying several
   tests to the same generator, and why?
10. Explain briefly what the gap test is designed to detect.

## Long Answer Questions

1. State and prove Theorem 2.1 (asymptotic distribution of the Pearson
   chi-square statistic), explaining the role of the singular
   covariance matrix and its rank.
2. State and prove the mean formula of Theorem 2.2, showing the
   turning-point indicator argument in full, and explain why the
   variance formula's derivation is only outlined rather than proved
   completely.
3. Derive the poker-hand probabilities for "all different" and "one
   pair" from first principles, showing every counting step.
4. Derive the closed-form expression (2.7) for the Kolmogorov-Smirnov
   statistic from its definition as a supremum.
5. Compare the Chi-square and Kolmogorov-Smirnov tests in terms of
   their treatment of binning, their asymptotic justification, and
   their relative power against smooth departures from uniformity.
6. Discuss the multiple-testing problem as it applies to running a
   full battery of generator tests, and describe at least one
   remedy.
7. Explain, with reference to Theorem 2.2, why the runs test is a
   two-sided test, and describe a scenario that would produce too few
   runs versus one that would produce too many.
8. Discuss why passing the tests of this unit is necessary but not
   sufficient for a generator to be considered cryptographically
   secure.
9. Explain how the serial correlation test complements the runs test,
   and describe a form of dependence that one might detect more
   readily than the other.
10. Using the RANDU case study from Unit 3.1, discuss which of the
    tests in this unit would be expected to detect its defect, and
    which would not, and why.

## Programming Exercises

1. Implement `chi_square_uniformity_test()` and
   `kolmogorov_smirnov_uniformity_test()` from first principles
   (already provided) and reproduce the manual calculation of Example
   1 by constructing a sample with exactly the stated bin counts.
2. Implement `runs_test()` from first principles (already provided)
   and reproduce the manual calculation of Example 3 exactly.
3. Implement `serial_correlation_test()` from first principles
   (already provided) and reproduce the manual calculation of Example
   4 exactly.
4. Verify computationally that the seven poker-hand probabilities of
   Table 1 sum to exactly $1$, and cross-check at least three of them
   against a full enumeration of all $10^5$ five-digit strings.
5. Apply the full test battery (`evaluate_generator()`) to a
   multiplicative congruential generator of your own design (choosing
   $a$, $m$ to violate the Hull-Dobell-style conditions of Unit 3.1),
   and report which tests detect the defect.
6. Conduct a small simulation study of the runs test's Type I error
   rate: generate many independent samples of size $n=100$ from
   `runif()`, apply the runs test to each, and report the empirical
   rejection rate at the nominal $5\%$ level.
7. Implement a Bonferroni-corrected version of `evaluate_generator()`
   that adjusts the significance threshold for the number of tests
   performed, and compare its conclusions to the uncorrected version
   on the poor generator of Section 10.1.
8. Reproduce the lag-plot comparison figure of this unit for a
   generator of your choosing with a deliberately short period, and
   discuss what visual signature, if any, is apparent.
9. Extend `digit_frequency_test()` to test every decimal digit
   position from $1$ to $5$ simultaneously, and produce a table
   summarizing the results for both the good and poor generators of
   Section 10.1.
10. Write a function that estimates, by simulation, the actual
    (family-wise) Type I error rate of applying all seven tests in
    this unit's battery to genuine `runif()` samples, and compare it
    to the nominal per-test rate of $5\%$.

## Mini Projects

1. **A comprehensive test report for a candidate generator.** Choose
   or construct three candidate generators: R's default `runif()`, a
   multiplicative congruential generator satisfying the Hull-Dobell-
   style full-period conditions of Unit 3.1 with a well-chosen
   multiplier, and RANDU (or a RANDU-like generator). Apply the full
   test battery of this unit to each, at a fixed, adequately large
   sample size, and produce a professional summary report (tables and
   figures) recommending which generator(s) are fit for use in a
   Monte Carlo study, with statistical justification drawn directly
   from your test results.

2. **Power comparison of uniformity tests.** Design a simulation study
   comparing the statistical power of the Chi-square and
   Kolmogorov-Smirnov uniformity tests against several specific
   alternatives to $H_0$ (for example, a $\text{Beta}(1.2,1)$
   distribution, a mixture of Uniform$(0,0.4)$ and Uniform$(0.6,1)$,
   and a slightly truncated Uniform distribution), across a range of
   sample sizes. Present your results graphically (power curves) and
   discuss which test is more powerful against which type of
   departure from uniformity.

**Difficulty labels.** Multiple choice, short answer: **Easy**. Long
answer questions 3, 4, 6, 7, 9, 10, programming exercises 1-6:
**Medium**. Long answer questions 1, 2, 5, 8, programming exercises
7-10: **Medium to Advanced**. Mini projects: **Advanced**.

# References

Billingsley, P. (1968). *Convergence of probability measures*. Wiley.

Knuth, D. E. (1997). *The art of computer programming, Volume 2:
Seminumerical algorithms* (3rd ed.). Addison-Wesley.

Marsaglia, G. (1995). *The Marsaglia random number CDROM including the
Diehard battery of tests of randomness*. Florida State University.

Rao, C. R. (1973). *Linear statistical inference and its
applications* (2nd ed.). Wiley.

Rizzo, M. L. (2019). *Statistical computing with R* (2nd ed.). CRC
Press.

Rukhin, A., Soto, J., Nechvatal, J., Smid, M., Barker, E., Leigh, S.,
... & Vo, S. (2010). *A statistical test suite for random and
pseudorandom number generators for cryptographic applications* (NIST
Special Publication 800-22, Rev. 1a). National Institute of Standards
and Technology.

Ross, S. M. (2020). *Simulation* (6th ed.). Academic Press.
