---
title: "Unit 3.4: The Acceptance-Rejection Method"
author: "Dr. Manoj C. Patil"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: false
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
    code_folding: hide
    theme: cosmo
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    echo = TRUE,
    warning = FALSE,
    message = FALSE,
    fig.align = "center",
    fig.width = 7,
    fig.height = 4.5
)

set.seed(123)
library(tidyverse)
library(ggplot2)
library(dplyr)
library(knitr)
library(kableExtra)
library(gridExtra)
library(patchwork)
```

# Learning Objectives

By the end of this unit, a student should be able to

- explain why the inverse transformation method of Unit 3.3 is
  insufficient whenever a quantile function is unavailable, and how
  the acceptance-rejection method fills this gap,
- state and prove that the acceptance-rejection algorithm produces a
  variate with exactly the target density, using the conditional
  distribution of the accepted candidate,
- derive the bounding constant $M$ that governs the algorithm and
  explain why the tightest valid $M$ minimizes computational cost,
- define exponential tilting and use it to construct an efficient
  envelope for simulating extreme-tail events,
- implement the acceptance-rejection algorithm in R for both a
  bounded and an unbounded target density, and empirically verify the
  theoretical acceptance probability,
- evaluate when the method is advantageous relative to alternative
  generation techniques, and diagnose common implementation errors.

# Prerequisites

A reader should already be comfortable with

- the inverse transformation method and the notion of a quantile
  function (Unit 3.3),
- basic probability density manipulations, including conditioning on
  an event and computing conditional densities,
- elementary calculus, in particular locating the maximum of a smooth
  function by differentiation,
- the moment generating function of a random variable,
- generating Uniform$(0,1)$ and Exponential random variables in R.

# Introduction

The inverse transformation method reduces the problem of simulating
any random variable to a single recipe: draw a uniform variate and
apply the quantile function. The recipe is elegant, but it has a
practical weak point. Many distributions that arise routinely in
statistics — the Gamma distribution for non-integer shape parameter,
the standard Normal distribution, most posterior distributions
encountered in Bayesian inference — do not admit a quantile function
in closed form, and numerical inversion of the cumulative distribution
function at every draw can be slow or numerically delicate. The
acceptance-rejection method, sometimes called rejection sampling,
addresses exactly this situation. It requires no quantile function at
all; it requires only the ability to evaluate the target density
(possibly up to an unknown multiplicative constant) and the ability to
simulate from some other, more convenient distribution.

The idea originates with John von Neumann, who described it in an
unpublished 1947 note in connection with Monte Carlo calculations for
the Manhattan Project, and it was popularized in the open literature
through the 1950s and 1960s as electronic computing made large-scale
simulation practical. The method's genius lies in converting a
sampling problem into a geometric one: if a target density can be
enclosed beneath a scaled version of a density that is easy to sample,
then sampling uniformly from the region beneath the easy density and
discarding points that fall above the target curve leaves behind
points that are exactly distributed according to the target. No
inversion, and no closed-form cumulative distribution function, is
required anywhere in the procedure.

The practical importance of the method extends well beyond
distributions lacking a quantile function. In Bayesian statistics, a
posterior density is typically known only up to an unknown
normalizing constant — the method, remarkably, still works without
ever needing that constant, which makes it a natural exact-sampling
benchmark against which approximate Markov chain Monte Carlo methods
are compared. In reliability and insurance mathematics, a refinement
of the method known as exponential tilting makes it possible to
simulate efficiently from the extreme tail of a distribution, a
setting in which naive methods waste nearly all of their computational
effort. This unit develops the theory, the choice of the bounding
constant $M$, exponential tilting, and the algorithms in full.

# Intuition

Picture the graph of a target probability density $f$, and suppose it
is possible to draw, at will, a random point that lands uniformly
somewhere on or under the graph of some *other* curve $Mg(x)$, where
$g$ is itself a probability density that is easy to sample from and
$M \ge 1$ is a constant large enough that $Mg(x)$ lies everywhere on
or above $f(x)$. If a large number of such uniformly scattered points
are generated under the curve $Mg$, then the subset of those points
that happen to fall *under* the curve $f$ is, geometrically, uniformly
scattered under $f$ — and a point uniformly scattered under a density
curve has, as its horizontal coordinate, exactly a draw from that
density. Points falling above $f$ but still under $Mg$ are simply
discarded ("rejected"); the surviving ("accepted") points supply an
exact sample from $f$.

Turning this geometric picture into an algorithm requires only a way
to generate a point uniformly under $Mg$. This is done in two stages:
first draw the horizontal coordinate $Y$ from the density $g$ (which
is assumed to be easy), and then, conditional on $Y=y$, draw the
vertical coordinate uniformly between $0$ and $Mg(y)$. The point is
accepted if this vertical coordinate falls at or below $f(y)$. Because
comparing a Uniform$(0,1)$ draw $U$ against the ratio $f(y)/(Mg(y))$
is equivalent to comparing a uniform draw on $(0, Mg(y))$ against
$f(y)$, the acceptance rule reduces to the single check
$U \le f(y)/(Mg(y))$, with no reference to areas or integrals needed
at run time.

```{r intuition-figure, echo=FALSE, fig.cap="Geometric idea of acceptance-rejection: candidate points are generated uniformly under the envelope Mg(x); the shaded points falling under f(x) are accepted, and the outline points are rejected."}
x_grid <- seq(0, 1, length.out = 300)
target_density <- 6 * x_grid * (1 - x_grid)
envelope_height <- 1.5

set.seed(123)
n_demo_points <- 400
y_candidates <- runif(n_demo_points)
v_candidates <- runif(n_demo_points, min = 0, max = envelope_height)
f_at_y <- 6 * y_candidates * (1 - y_candidates)
accepted <- v_candidates <= f_at_y

demo_df <- tibble(y = y_candidates, v = v_candidates, accepted = accepted)

ggplot() +
  geom_line(
    data = tibble(x = x_grid, f = target_density),
    aes(x = x, y = f), colour = "steelblue", linewidth = 1
  ) +
  geom_hline(yintercept = envelope_height, colour = "darkorange", linewidth = 1) +
  geom_point(
    data = demo_df, aes(x = y, y = v, colour = accepted), size = 1.2, alpha = 0.7
  ) +
  scale_colour_manual(values = c("FALSE" = "grey70", "TRUE" = "firebrick")) +
  labs(
    title = "Points uniform under Mg(x) = 1.5; accepted points lie under f(x)",
    x = "y", y = "height", colour = "Accepted"
  ) +
  theme_minimal(base_size = 12)
```

The picture also makes the efficiency of the method transparent: the
fraction of candidate points that are accepted is exactly the ratio of
the area under $f$ (which is $1$, since $f$ is a density) to the area
under $Mg$ (which is $M$, since $g$ is a density). An envelope that
hugs the target closely, with $M$ close to its minimum possible value,
wastes little effort on rejected points; a loose envelope with large
$M$ wastes most of its draws.

# Formal Definitions

## The Acceptance-Rejection Setup

**Definition.** Let $f$ be a target probability density on a support
$\mathcal{X}$, and let $g$ be a **proposal density** (also called an
**envelope density** or **instrumental density**) on the same support,
from which random variates can be readily generated. Suppose there
exists a finite constant $M \ge 1$, called the **bounding constant**,
such that

$$
f(x) \le M\,g(x) \qquad \text{for every } x \in \mathcal{X}.
\tag{4.1}
$$

The **acceptance-rejection method** generates a candidate
$Y \sim g$ together with an independent $U \sim \text{Uniform}(0,1)$,
and accepts $X = Y$ if

$$
U \le \frac{f(Y)}{M\,g(Y)},
\tag{4.2}
$$

repeating with a fresh $(Y,U)$ pair whenever the candidate is
rejected.

where

- $f$ denotes the density from which a sample is ultimately desired,
- $g$ denotes the density actually used to generate candidates,
- $M$ denotes the constant bounding the ratio $f/g$ over the entire
  support, so that the right-hand side of (4.2) never exceeds $1$,
- $U$ denotes the auxiliary uniform variate used to decide acceptance.

## Conditional Distribution of the Accepted Variate

The central claim of the method, proved formally as Theorem 4.1 below,
is that the variate reported by the algorithm — the value of $Y$ on
the trial that is finally accepted — has *exactly* the conditional
distribution

$$
X \;\overset{d}{=}\; \bigl(Y \mid U \le f(Y)/(Mg(Y))\bigr),
\tag{4.3}
$$

and that this conditional distribution coincides exactly with the
target density $f$, not merely approximately. This is the feature that
elevates the method from a heuristic to an exact simulation technique:
however loose or tight the envelope $Mg$ is chosen (subject to $(4.1)$
holding), the output is never biased; only the *speed* of generation
depends on the tightness of the envelope.

# Theorems

## Theorem 4.1: Correctness of the Acceptance-Rejection Algorithm

**Statement.** Let $f$ and $g$ be densities on a common support
$\mathcal{X}$, and let $M \ge 1$ satisfy $f(x) \le Mg(x)$ for all $x
\in \mathcal{X}$. If $Y \sim g$ and $U \sim \text{Uniform}(0,1)$
independently, and $X$ denotes the value of $Y$ conditional on the
event $\{U \le f(Y)/(Mg(Y))\}$, then $X$ has density exactly $f$.
Moreover, the probability that any given trial is accepted equals
$1/M$.

**Assumptions.** $g(x) > 0$ wherever $f(x) > 0$ (so that the ratio
$f(x)/g(x)$ is well defined on the support of $f$), and $M$ is finite.

**Proof.** For a fixed candidate $Y=y$, the conditional acceptance
probability is

$$
P\bigl(U \le f(y)/(Mg(y)) \mid Y=y\bigr) = \frac{f(y)}{Mg(y)},
$$

since $U$ is Uniform$(0,1)$ and the right-hand side of (4.2) lies in
$[0,1]$ by the bounding condition (4.1). Averaging over the
distribution of $Y$ gives the unconditional acceptance probability

$$
P(\text{accept}) = \int_{\mathcal{X}} \frac{f(y)}{Mg(y)}\, g(y)\, dy
= \frac{1}{M}\int_{\mathcal{X}} f(y)\, dy = \frac{1}{M},
\tag{4.4}
$$

using $\int f = 1$ since $f$ is a density. This proves the second
claim. For the first claim, fix $x \in \mathcal{X}$ and compute the
joint probability that $Y \le x$ *and* the candidate is accepted:

$$
P(Y \le x,\ \text{accept})
= \int_{-\infty}^{x} \frac{f(y)}{Mg(y)}\, g(y)\, dy
= \frac{1}{M}\int_{-\infty}^{x} f(y)\, dy
= \frac{F(x)}{M},
\tag{4.5}
$$

where $F$ is the cumulative distribution function of $f$. By the
definition of conditional probability and using (4.4) and (4.5),

$$
P(X \le x) = P(Y \le x \mid \text{accept})
= \frac{P(Y\le x,\ \text{accept})}{P(\text{accept})}
= \frac{F(x)/M}{1/M} = F(x).
$$

Since $P(X \le x) = F(x)$ for every $x$, the accepted variate $X$ has
cumulative distribution function $F$, and hence density $f$.
$\blacksquare$

**Remarks.** An equivalent, purely geometric argument reaches the same
conclusion without any integration. The pair $(Y, U\!\cdot\! Mg(Y))$ is
uniformly distributed over the two-dimensional region lying under the
curve $Mg$, because $Y$ has marginal density $g$ and, given $Y=y$, the
second coordinate is uniform on $(0, Mg(y))$ by construction; a
standard fact about the uniform distribution on a planar region under
a curve is that its horizontal marginal is proportional to the height
of the curve. Restricting attention to the sub-region lying under $f$
(the accepted points) therefore gives a distribution uniform under
$f$, whose horizontal marginal is $f$ itself, exactly reproducing the
figure in the Intuition section. Both arguments are standard; see
Devroye (1986, Ch. II) and Robert & Casella (2004, Ch. 2).

**Examples.** Full worked numerical examples appear later in this
unit for both a bounded target (Beta) and an unbounded target
(half-Normal).

**Applications.** Because the proof of (4.5) only ever uses $f$
through the integral $\int f$, the same argument goes through
unchanged if $f$ is replaced everywhere by an *unnormalized* density
$\tilde f = c f$ for an unknown constant $c > 0$, provided $M$ is
re-chosen to satisfy $\tilde f \le M g$. The accepted variate is still
exactly distributed as $f$, without $c$ ever being computed. This
property is the reason acceptance-rejection is so widely used in
Bayesian computation, where a posterior density is typically available
only up to its normalizing constant.

## Theorem 4.2: Number of Trials Until Acceptance

**Statement.** Under the conditions of Theorem 4.1, the number of
independent trials $N$ required until the first acceptance (inclusive
of the accepted trial) follows a Geometric distribution with success
probability $1/M$,

$$
P(N = k) = \Bigl(1-\frac1M\Bigr)^{k-1}\frac1M, \qquad k=1,2,\dots,
$$

and hence $E(N) = M$.

**Proof.** Successive trials are independent and identically
distributed, each with acceptance probability $1/M$ by (4.4). The
number of independent Bernoulli$(1/M)$ trials up to and including the
first success is, by definition, Geometric$(1/M)$, with mean $1/(1/M)
= M$. $\blacksquare$

**Remarks.** Theorem 4.2 is the formal justification for treating $M$
as the "cost" of the algorithm: on average, $M$ candidate draws from
$g$ are needed to produce a single accepted draw from $f$. This is
precisely why the Choice of $M$ (Section 8) is not a minor technical
detail but the central design decision of the method.

# Mathematical Derivations

## The Optimal (Minimal) Bounding Constant

Since $E(N) = M$ by Theorem 4.2, computational efficiency is directly
governed by how small $M$ can be made while still satisfying $f(x)
\le Mg(x)$ for every $x$. Rearranging (4.1),

$$
M \ge \sup_{x \,:\, g(x) > 0} \frac{f(x)}{g(x)},
$$

so the smallest valid choice is

$$
M^{*} = \sup_{x} \frac{f(x)}{g(x)},
\tag{4.6}
$$

provided this supremum is finite. Using $M = M^{*}$ maximizes the
acceptance probability $1/M$ for the given proposal $g$; a poorer
choice of $g$, even paired with its own optimal $M^{*}$, can still
give a much larger $M^{*}$ than a better-matched proposal would. Two
practical consequences follow immediately. First, $g$ should have the
same support as $f$ (or larger), since $f(x) \le Mg(x)$ is impossible
to satisfy wherever $g(x) = 0$ but $f(x) > 0$. Second, $g$ should have
tails at least as heavy as those of $f$; if $g$'s tail decays faster
than $f$'s tail, the ratio $f(x)/g(x)$ diverges and no finite $M$
exists at all, so the method cannot be used in that pairing — a
concrete instance is attempting to generate Cauchy variates using a
Normal proposal, which fails because the Cauchy density decays only
polynomially while the Normal density decays like $e^{-x^2/2}$.

## Exponential Tilting

For distributions whose interesting behaviour lies deep in a tail —
extreme value analysis, ruin probabilities in insurance mathematics,
rare-event simulation in reliability engineering — the natural
proposal $g$ equal to the parent distribution itself gives a
disastrously small acceptance probability, because almost all mass of
$g$ lies away from the tail of interest. **Exponential tilting**
(also called exponential twisting, or the Esscher transform in
actuarial science) constructs a family of proposal densities better
matched to a target region by re-weighting a base density with an
exponential factor.

**Definition.** Given a density $h$ with moment generating function
$M_h(\theta) = E_h[e^{\theta X}] < \infty$ on some interval of
$\theta$, the **exponentially tilted density** with tilting parameter
$\theta$ is

$$
h_\theta(x) = \frac{e^{\theta x}\, h(x)}{M_h(\theta)}.
\tag{4.7}
$$

Increasing $\theta$ shifts probability mass of $h_\theta$ toward
larger values of $x$, which makes $h_\theta$ an effective proposal for
sampling from the right tail of a related target density; the tilting
parameter $\theta$ is a free design choice, and its optimal value is
the one minimizing the resulting bounding constant $M^{*}$ in (4.6).

**Worked construction: sampling the right tail of the standard
Normal.** Let $Z$ be a target variable with the standard Normal
density $\varphi(z) = (2\pi)^{-1/2} e^{-z^2/2}$, and suppose the goal
is to simulate from $Z$ conditional on $Z > a$ for a large threshold
$a > 0$, i.e. from the truncated density

$$
f(x) = \frac{\varphi(x)}{Q(a)}, \qquad x > a,
\qquad Q(a) = P(Z>a) = 1-\Phi(a).
$$

Sampling $Z$ unconditionally and discarding all draws with $Z \le a$
works in principle but has acceptance probability $Q(a)$, which
becomes vanishingly small as $a$ grows — for $a=4$, roughly one draw
in $30{,}000$ is retained. Exponential tilting of the shifted
Exponential family provides a dramatically better proposal. Take as
the base density the Exponential$(1)$ density shifted to start at $a$,
and tilt with parameter $\theta = a$, giving the proposal

$$
g(x) = a\,e^{-a(x-a)}, \qquad x > a,
$$

i.e. $Y = a + E$ with $E \sim \text{Exponential}(\text{rate}=a)$.
Substituting into the ratio $f(x)/g(x)$ and completing the square,

$$
\begin{aligned}
\frac{f(x)}{g(x)}
&= \frac{1}{Q(a)\,a\sqrt{2\pi}}\, \exp\Bigl(-\tfrac{x^2}{2} + ax\Bigr) \\
&= \frac{1}{Q(a)\,a\sqrt{2\pi}}\, \exp\Bigl(-\tfrac{(x-a)^2}{2} - \tfrac{a^2}{2}\Bigr) ,
\end{aligned}
\tag{4.8}
$$

using $-\tfrac{x^2}{2}+ax = -\tfrac{(x-a)^2}{2} - \tfrac{a^2}{2}$,
which is verified by expanding $(x-a)^2 = x^2 - 2ax + a^2$. Since
$x>a$, the factor $\exp(-(x-a)^2/2)$ is at most $1$, attained only in
the limit $x \to a^{+}$, so the supremum of (4.8) is

$$
M^{*} = \frac{e^{-a^2/2}}{Q(a)\,a\sqrt{2\pi}} = \frac{\varphi(a)}{a\,Q(a)}.
\tag{4.9}
$$

Dividing (4.8) by (4.9) shows that the acceptance ratio has the
remarkably simple form

$$
\frac{f(x)}{M^{*}g(x)} = e^{-(x-a)^2/2},
\tag{4.10}
$$

which depends only on the *excess* $x-a$ of the candidate above the
threshold, and involves no evaluation of $\Phi$ or $Q$ at run time.
Because $-\log U \sim \text{Exponential}(1)$ whenever $U \sim
\text{Uniform}(0,1)$, the acceptance rule $U \le e^{-(x-a)^2/2}$ is
equivalent to $-\log U \ge (x-a)^2/2$, so drawing a second, independent
Exponential$(1)$ variate $E_2$ and accepting whenever $E_2 \ge
(Y-a)^2/2$ implements the algorithm with no uniform variate and no
transcendental function evaluation beyond squaring — the classical
"two exponentials" algorithm of Marsaglia (1964), reproduced as
Algorithm 4.3 below.

By the Mills ratio approximation $Q(a) \sim \varphi(a)/a$ as $a \to
\infty$, expression (4.9) shows $M^{*} \to 1$ as $a \to \infty$: the
larger the threshold, the *more* efficient the tilted algorithm
becomes, in sharp contrast to naive rejection from the untilted
Normal, whose acceptance probability $Q(a) \to 0$. This inversion of
the usual difficulty — the tail becomes easier, not harder, to sample
— is the defining benefit of exponential tilting for rare-event
simulation.

# Algorithms

## Algorithm 4.1: General Acceptance-Rejection Algorithm

**Purpose.** Generate a sample from a target density $f$ using only a
proposal density $g$ and a valid bounding constant $M$, without
requiring a closed-form quantile function for $f$.

**Input.** Target density (or unnormalized density) $f$; proposal
density $g$, from which variates can be simulated; bounding constant
$M$ satisfying $f(x) \le Mg(x)$ on the support; required sample size
$n$.

**Output.** A vector $x_1, \dots, x_n$ of simulated observations from
$f$.

**Step-by-step procedure.**

1. Set the accepted-count to $0$ and initialize an empty output
   vector.
2. While fewer than $n$ values have been accepted:
   (a) draw a candidate $Y \sim g$;
   (b) draw $U \sim \text{Uniform}(0,1)$, independent of $Y$;
   (c) if $U \le f(Y)/(Mg(Y))$, append $Y$ to the output and increment
       the accepted-count; otherwise discard $Y$ and $U$ and repeat.
3. Return the vector of $n$ accepted values.

**Pseudo-code.**

```
Algorithm: Acceptance-Rejection
Input:  f, g, M, n
Output: vector x of length n

1. accepted <- 0
2. x <- empty vector
3. While accepted < n:
       y <- draw from g
       u <- draw from Uniform(0,1)
       If u <= f(y) / (M * g(y)):
           append y to x
           accepted <- accepted + 1
4. Return x
```

**Flowchart description.** Start; draw a candidate from the proposal;
draw an independent uniform variate; compare the uniform variate to
the acceptance ratio; if it does not exceed the ratio, record the
candidate as an accepted observation, otherwise discard both draws;
repeat until the required number of observations has been collected;
end.

**Complexity.** Each iteration costs $O(1)$ (one draw from $g$, one
uniform draw, one ratio evaluation), and by Theorem 4.2 the expected
number of iterations per accepted value is $M$, so the expected total
cost to produce $n$ observations is $O(Mn)$.

**Practical considerations.** All density evaluations should be
performed on the log scale, comparing $\log U \le \log f(Y) - \log M -
\log g(Y)$, whenever $f$ or $g$ can take very small values, to avoid
floating-point underflow.

## Algorithm 4.2: Acceptance-Rejection with a Uniform Envelope (Bounded Target)

**Purpose.** Generate a sample from a bounded density on a bounded
interval using the simplest possible envelope: a Uniform distribution
on the same interval, scaled by the maximum height of the target.

**Input.** Target density $f$ on $[0,1]$ with known maximum value
$f_{\max}$; sample size $n$.

**Output.** A vector of $n$ simulated observations from $f$.

**Step-by-step procedure.**

1. Set $M = f_{\max}$ and $g(x) = 1$ on $[0,1]$.
2. Apply Algorithm 4.1 with this $g$ and $M$.

**Pseudo-code.**

```
Algorithm: Acceptance-Rejection, Uniform Envelope
Input:  f, f_max, n
Output: vector x of length n

1. M <- f_max
2. accepted <- 0
3. x <- empty vector
4. While accepted < n:
       y <- Uniform(0,1)
       u <- Uniform(0,1)
       If u <= f(y) / M:
           append y to x
           accepted <- accepted + 1
5. Return x
```

**Flowchart description.** Identical to Algorithm 4.1, specialized so
that generating the candidate $Y$ is itself a single uniform draw and
the envelope height is the constant $f_{\max}$.

**Complexity.** $O(M n)$ expected time, with $M = f_{\max}$ for the
Uniform$(0,1)$ envelope; a worked value of $M$ for a specific target
(the Beta$(2,2)$ density) is given in the Numerical Examples section.

**Practical considerations.** This envelope is only viable when the
target is bounded and has bounded support; it becomes inefficient when
$f$ is sharply peaked, since $f_{\max}$ (and hence $M$) grows large
relative to the typical height of $f$ away from its mode.

## Algorithm 4.3: Exponentially Tilted Envelope for Normal Tail Simulation

**Purpose.** Efficiently generate observations from the standard
Normal distribution conditioned on exceeding a threshold $a$, using
the tilted Exponential envelope derived in Section 7.2, via the
"two exponentials" implementation that avoids evaluating $\varphi$,
$\Phi$, or any exponential function beyond the two auxiliary draws.

**Input.** Threshold $a > 0$; sample size $n$.

**Output.** A vector of $n$ simulated observations from the standard
Normal distribution truncated to $(a, \infty)$.

**Step-by-step procedure.**

1. While fewer than $n$ values have been accepted:
   (a) draw $E_1 \sim \text{Exponential}(\text{rate}=a)$ and set
       candidate $Y = a + E_1$;
   (b) draw $E_2 \sim \text{Exponential}(\text{rate}=1)$,
       independently;
   (c) if $E_2 \ge E_1^2/2$, accept $Y$; otherwise discard both draws.
2. Return the vector of accepted values.

**Pseudo-code.**

```
Algorithm: Tilted-Exponential Normal Tail Sampler
Input:  a, n
Output: vector x of length n

1. accepted <- 0
2. x <- empty vector
3. While accepted < n:
       e1 <- Exponential(rate = a)
       e2 <- Exponential(rate = 1)
       If e2 >= e1^2 / 2:
           append (a + e1) to x
           accepted <- accepted + 1
4. Return x
```

**Flowchart description.** Start; draw two independent exponential
variates; compare the second against half the square of the first;
if the check passes, report the threshold plus the first exponential
as an accepted tail observation; otherwise discard both and repeat;
end.

**Complexity.** $O(M^{*} n)$ expected time, with $M^{*}$ given
explicitly by (4.9); crucially, $M^{*} \to 1$ as $a \to \infty$, so
unlike Algorithms 4.1-4.2, this method becomes *more* efficient as the
target region becomes more extreme.

**Practical considerations.** For $a \le 0$, the shifted-Exponential
envelope is not the natural choice (a direct Normal draw, discarding
values below the threshold, is simpler and already efficient when the
threshold is not deep in the tail); the tilted method is intended
specifically for moderate-to-large positive thresholds.

# R Programming

```{r ar-general-function}
set.seed(123)

#' Generate variates by the general acceptance-rejection algorithm
#'
#' Purpose: implements Algorithm 4.1, drawing candidates from a
#' user-supplied proposal generator and accepting them according to
#' the ratio of the target density to the scaled proposal density.
#'
#' Arguments:
#'   sample_size        number of accepted observations required
#'   target_density     function f(x) evaluating the target density
#'   proposal_density    function g(x) evaluating the proposal density
#'   proposal_sampler    function() returning one draw from g
#'   bounding_constant   the constant M with f(x) <= M * g(x)
#'
#' Returns:
#'   a list with the accepted sample (`sample`) and the empirical
#'   acceptance rate (`acceptance_rate`)
#'
#' Example:
#'   accept_reject(
#'     sample_size = 1000,
#'     target_density = function(x) 6 * x * (1 - x),
#'     proposal_density = function(x) rep(1, length(x)),
#'     proposal_sampler = function() runif(1),
#'     bounding_constant = 1.5
#'   )
accept_reject <- function(sample_size, target_density, proposal_density,
                           proposal_sampler, bounding_constant) {
  accepted_values <- numeric(sample_size)
  accepted_count <- 0
  total_trials <- 0

  while (accepted_count < sample_size) {
    candidate <- proposal_sampler()
    uniform_draw <- runif(1)
    total_trials <- total_trials + 1

    acceptance_ratio <- target_density(candidate) /
      (bounding_constant * proposal_density(candidate))

    if (uniform_draw <= acceptance_ratio) {
      accepted_count <- accepted_count + 1
      accepted_values[accepted_count] <- candidate
    }
  }

  list(
    sample = accepted_values,
    acceptance_rate = sample_size / total_trials
  )
}

#' Generate right-tail standard Normal variates via exponential tilting
#'
#' Purpose: implements Algorithm 4.3, the two-exponentials method for
#' simulating Z | Z > a efficiently for large thresholds a.
#'
#' Arguments:
#'   sample_size    number of accepted observations required
#'   threshold      the truncation threshold a (a > 0)
#'
#' Returns:
#'   a list with the accepted sample (`sample`) and the empirical
#'   acceptance rate (`acceptance_rate`)
#'
#' Example:
#'   tilted_normal_tail(sample_size = 1000, threshold = 2)
tilted_normal_tail <- function(sample_size, threshold) {
  accepted_values <- numeric(sample_size)
  accepted_count <- 0
  total_trials <- 0

  while (accepted_count < sample_size) {
    first_exponential <- rexp(1, rate = threshold)
    second_exponential <- rexp(1, rate = 1)
    total_trials <- total_trials + 1

    if (second_exponential >= first_exponential^2 / 2) {
      accepted_count <- accepted_count + 1
      accepted_values[accepted_count] <- threshold + first_exponential
    }
  }

  list(
    sample = accepted_values,
    acceptance_rate = sample_size / total_trials
  )
}
```

# Simulation Studies

## Half-Normal Density via an Exponential(1) Envelope

The half-Normal density $f(x) = \sqrt{2/\pi}\, e^{-x^2/2}$, $x>0$, is
generated using the Exponential$(1)$ proposal $g(x) = e^{-x}$, with
bounding constant $M^{*} = \sqrt{2e/\pi}$ derived in the Numerical
Examples section below.

```{r half-normal-simulation}
set.seed(123)

m_half_normal <- sqrt(2 * exp(1) / pi)

half_normal_result <- accept_reject(
  sample_size = 20000,
  target_density = function(x) sqrt(2 / pi) * exp(-x^2 / 2),
  proposal_density = function(x) exp(-x),
  proposal_sampler = function() rexp(1, rate = 1),
  bounding_constant = m_half_normal
)

tibble(
  Quantity = c("Theoretical M", "Theoretical acceptance rate (1/M)", "Empirical acceptance rate"),
  Value = c(m_half_normal, 1 / m_half_normal, half_normal_result$acceptance_rate)
) %>%
  kable(digits = 4, caption = "Half-Normal acceptance-rejection: theory vs simulation")
```

The empirical acceptance rate closely tracks the theoretical value
$1/M^{*} \approx 0.760$, confirming Theorem 4.2 directly.

## Beta(2,2) Density via a Uniform Envelope

```{r beta-simulation}
set.seed(123)

beta_result <- accept_reject(
  sample_size = 20000,
  target_density = function(x) 6 * x * (1 - x),
  proposal_density = function(x) rep(1, length(x)),
  proposal_sampler = function() runif(1),
  bounding_constant = 1.5
)

tibble(
  Quantity = c("Theoretical M", "Theoretical acceptance rate (1/M)", "Empirical acceptance rate"),
  Value = c(1.5, 1 / 1.5, beta_result$acceptance_rate)
) %>%
  kable(digits = 4, caption = "Beta(2,2) acceptance-rejection: theory vs simulation")
```

## Exponential Tilting vs. Naive Rejection for the Normal Tail

This simulation contrasts the tilted "two exponentials" algorithm
against the naive strategy of drawing standard Normal variates and
discarding those at or below the threshold, for a moderately deep
threshold $a=3$.

```{r tilting-comparison}
set.seed(123)

threshold_value <- 3
q_theoretical <- 1 - pnorm(threshold_value)

# Naive method: draw standard Normal variates, keep only those above a
naive_trials <- 200000
naive_draws <- rnorm(naive_trials)
naive_acceptance_rate <- mean(naive_draws > threshold_value)

# Tilted method
tilted_result <- tilted_normal_tail(sample_size = 5000, threshold = threshold_value)

m_tilted_theoretical <- dnorm(threshold_value) / (threshold_value * q_theoretical)

comparison_tbl <- tibble(
  Method = c("Naive (discard Z <= a)", "Exponentially tilted (two exponentials)"),
  Theoretical_Acceptance_Rate = c(q_theoretical, 1 / m_tilted_theoretical),
  Empirical_Acceptance_Rate = c(naive_acceptance_rate, tilted_result$acceptance_rate)
)

kable(comparison_tbl, digits = 5,
      caption = "Naive rejection vs. exponentially tilted rejection for Z | Z > 3")
```

The tilted method accepts the overwhelming majority of its candidates,
while the naive method accepts roughly one candidate in every
$1/Q(3) \approx 741$; the tilted algorithm therefore requires several
orders of magnitude fewer total random draws to produce the same
number of tail observations, exactly as predicted by the asymptotic
argument following (4.9).

# Graphs

```{r density-overlay-plots, fig.cap="Simulated samples (histograms) against theoretical target densities (curves) for the half-Normal (left) and Beta(2,2) (right) acceptance-rejection examples."}
set.seed(123)

p_half_normal <- ggplot(tibble(x = half_normal_result$sample), aes(x = x)) +
  geom_histogram(aes(y = after_stat(density)), bins = 40, fill = "steelblue", colour = "white") +
  stat_function(fun = function(x) sqrt(2 / pi) * exp(-x^2 / 2), colour = "darkred", linewidth = 1) +
  labs(title = "Half-Normal via Exponential(1) envelope", x = "x", y = "Density") +
  theme_minimal(base_size = 11)

p_beta <- ggplot(tibble(x = beta_result$sample), aes(x = x)) +
  geom_histogram(aes(y = after_stat(density)), bins = 40, fill = "darkgreen", colour = "white") +
  stat_function(fun = function(x) 6 * x * (1 - x), colour = "darkred", linewidth = 1) +
  labs(title = "Beta(2,2) via Uniform envelope", x = "x", y = "Density") +
  theme_minimal(base_size = 11)

p_half_normal + p_beta
```

```{r tail-overlay-plot, fig.cap="Simulated tilted-envelope sample against the exact truncated Normal density for Z | Z > 3."}
set.seed(123)

truncated_normal_density <- function(x, threshold) {
  dnorm(x) / (1 - pnorm(threshold))
}

ggplot(tibble(x = tilted_result$sample), aes(x = x)) +
  geom_histogram(aes(y = after_stat(density)), bins = 30, fill = "purple", colour = "white", alpha = 0.7) +
  stat_function(
    fun = truncated_normal_density, args = list(threshold = threshold_value),
    colour = "black", linewidth = 1
  ) +
  labs(
    title = "Exponentially tilted sample vs. exact truncated Normal density",
    x = "x", y = "Density"
  ) +
  theme_minimal(base_size = 11)
```

# Numerical Examples

**Example 1: Bounding constant and one iteration for the half-Normal
density.**

*Problem.* Using the target $f(x) = \sqrt{2/\pi}\,e^{-x^2/2}$, $x>0$,
and proposal $g(x)=e^{-x}$, derive $M^{*}$ and determine whether the
candidate $y=0.8$ with $u=0.30$ is accepted.

*Solution.* The ratio is
$f(x)/g(x) = \sqrt{2/\pi}\,\exp(-x^2/2+x)$. Writing the exponent as
$-\tfrac12(x-1)^2 + \tfrac12$ (completing the square as in (4.8)-(4.9)
with $a=1$), the maximum over $x>0$ occurs at $x=1$, giving

$$
M^{*} = \sqrt{2/\pi}\,e^{1/2} = \sqrt{2e/\pi} \approx 1.3155.
$$

For the candidate $y=0.8$: $f(0.8) = \sqrt{2/\pi}\,e^{-0.32} =
0.7979\times 0.7261 = 0.5794$, and $M^{*}g(0.8) = 1.3155\times
e^{-0.8} = 1.3155\times 0.4493 = 0.5911$. The acceptance ratio is
$0.5794/0.5911 = 0.9802$. Since $u=0.30 \le 0.9802$, the candidate is
**accepted**, and $x=0.8$ is reported.

*Interpretation.* Because $0.8$ is close to the mode $1$ of the ratio
$f/g$, the acceptance ratio is close to its maximum value of $1$; a
candidate far from this mode would have a much smaller acceptance
ratio and would be rejected more often, which is exactly the pattern
verified in bulk by the simulation study above.

**Example 2: Manual acceptance-rejection trial for Beta(2,2).**

*Problem.* Using $f(x) = 6x(1-x)$ on $(0,1)$ and the Uniform$(0,1)$
envelope with $M=1.5$, determine the outcome for the candidate
$y=0.9$ with $u=0.5$, and separately for $y=0.9$ with $u=0.2$.

*Solution.* $f(0.9) = 6(0.9)(0.1) = 0.54$, and $Mg(0.9) = 1.5(1) =
1.5$, so the acceptance ratio is $0.54/1.5 = 0.36$. With $u=0.5$,
since $0.5 > 0.36$, the candidate is **rejected**. With $u=0.2$,
since $0.2 \le 0.36$, the candidate is **accepted**, and $x=0.9$ is
reported.

*Interpretation.* The value $y=0.9$ lies far from the mode of the
Beta$(2,2)$ density (which is at $x=0.5$), so its acceptance ratio of
$0.36$ is well below $1$; roughly $64\%$ of candidates equal to $0.9$
would be rejected, consistent with the overall acceptance rate of
$1/1.5 \approx 0.667$ averaged across all candidate locations.

**Example 3: Exponential tilting for the Normal tail at $a=2$.**

*Problem.* Compute the optimal bounding constant $M^{*}$ and the
acceptance probability for simulating $Z \mid Z>2$ using the
exponentially tilted envelope with $\theta=a=2$.

*Solution.* From (4.9), $M^{*} = \varphi(2)/(2\,Q(2))$. Using
$\varphi(2) = (2\pi)^{-1/2}e^{-2} = 0.05399$ and
$Q(2) = 1-\Phi(2) = 0.02275$,

$$
M^{*} = \frac{0.05399}{2 \times 0.02275} = \frac{0.05399}{0.04550}
= 1.1866,
$$

so the acceptance probability is $1/M^{*} \approx 0.8427$. By
contrast, the naive method of drawing $Z\sim N(0,1)$ and discarding
$Z \le 2$ has acceptance probability $Q(2) = 0.02275$ — the tilted
method is roughly $37$ times more efficient at this threshold, and
this efficiency ratio grows without bound as the threshold increases
further.

*Interpretation.* The R simulation for threshold $a=3$ above shows the
same phenomenon even more starkly, since the naive acceptance rate
$Q(3) \approx 0.00135$ falls rapidly with $a$ while the tilted
acceptance rate stays comfortably above $0.9$.

# Applications

- **Statistics.** Exact simulation from non-standard likelihoods and
  from densities specified only up to a normalizing constant, a
  routine situation in maximum likelihood and Bayesian computation.
- **Machine Learning.** Sampling from complex latent-variable models
  and from priors or posteriors lacking a closed-form quantile
  function; acceptance-rejection also underlies certain
  reparameterization strategies for gradient-based training with
  non-standard noise distributions.
- **Data Science.** Generating synthetic data that matches an
  empirically estimated density for stress-testing pipelines, when
  the estimated density has no simple quantile function.
- **Finance.** Simulating extreme portfolio losses and tail risk
  measures (Value-at-Risk, Expected Shortfall) efficiently using
  exponential tilting rather than brute-force Monte Carlo.
- **Reliability Engineering.** Simulating rare early-failure or
  extreme-stress events for structural or component reliability
  analysis, where naive simulation would require an infeasible number
  of trials.
- **Monte Carlo Methods.** Acceptance-rejection is one of the two
  foundational exact-sampling techniques (alongside inverse
  transformation) taught in every Monte Carlo methods course, and
  importance sampling — a closely related variance-reduction
  technique — borrows the same tilting idea used in Section 7.2.
- **Bayesian Statistics.** Exact posterior sampling when the posterior
  kernel is bounded by a tractable envelope, and as the textbook
  benchmark against which the bias-variance trade-offs of Markov
  chain Monte Carlo approximations are explained.
- **Operations Research.** Simulating rare congestion or failure
  events in queueing and network reliability models.
- **Engineering.** Monte Carlo reliability analysis of systems
  subjected to random loads whose extreme-value behaviour, rather than
  typical behaviour, determines the design margin.

# Advantages

- The method is exact: for any valid $(g, M)$ pair, the accepted
  variate has precisely the target distribution, with no approximation
  error, unlike Markov chain Monte Carlo methods, which converge only
  asymptotically.
- No quantile function, and no closed-form cumulative distribution
  function, is ever required.
- The target density need only be known up to an unknown
  multiplicative constant (Theorem 4.1, Applications remark), which is
  essential for Bayesian posterior simulation.
- The method extends without modification to multivariate targets,
  provided a suitable multivariate envelope and bounding constant can
  be found.

# Limitations

- Efficiency is entirely governed by $M$; a poorly matched envelope
  can make the method arbitrarily slow, and in high dimensions, even a
  well-matched envelope typically forces $M$ to grow rapidly with
  dimension, a manifestation of the curse of dimensionality.
- A finite $M$ may not exist at all if the proposal's tail is lighter
  than the target's tail, as in the Cauchy-from-Normal example of
  Section 6.
- Finding the optimal $M^{*} = \sup_x f(x)/g(x)$ analytically requires
  calculus that may be intractable for complicated densities, forcing
  numerical optimization and the risk of an invalid (too small) $M$
  if the optimization is not exhaustive.
- The number of iterations required is random (Theorem 4.2), so,
  unlike inverse transformation, the running time of a single call to
  the algorithm is not fixed in advance.

# Practical Tips

- **Always verify the envelope condition over the entire support, not
  just near the mode.** A very common implementation bug is checking
  $f(x) \le Mg(x)$ only at a handful of sample points; if the
  inequality fails anywhere, the output is silently biased with no
  runtime error to signal the mistake.
- **Match tail behaviour first, then optimize $M$.** An envelope whose
  tail decays no faster than the target's tail is a prerequisite for
  a finite $M$ to exist at all; only after this is satisfied does
  minimizing $M^{*}$ become the relevant optimization.
- **Work on the log scale for extreme densities.** Comparing $\log U$
  against $\log f(Y) - \log M - \log g(Y)$ avoids underflow when $f$
  or $g$ take values far from $1$, which is common for high-dimensional
  or heavy-tailed densities.
- **Use exponential tilting whenever the target region is a distant
  tail.** As demonstrated in Section 10.3, naive rejection from the
  parent distribution becomes catastrophically inefficient in the
  tail, while a tilted envelope becomes more efficient the further out
  the tail region lies.
- **Monitor the empirical acceptance rate during development.** A
  large discrepancy between the theoretical $1/M$ and the observed
  acceptance rate usually indicates an error in the density functions
  or in the bounding constant, rather than ordinary sampling
  variability.

# Summary

- The acceptance-rejection method simulates from a target density $f$
  using only a proposal density $g$ and a bounding constant $M$ with
  $f \le Mg$, requiring neither a quantile function nor knowledge of
  the target's normalizing constant.
- A candidate $Y\sim g$ is accepted as an exact draw from $f$ whenever
  an independent Uniform$(0,1)$ draw does not exceed $f(Y)/(Mg(Y))$;
  Theorem 4.1 proves this gives exactly the target distribution.
- The acceptance probability of each trial is exactly $1/M$, and the
  number of trials until acceptance is Geometric$(1/M)$ with mean $M$
  (Theorem 4.2), which makes minimizing $M$ the central design goal.
- The optimal bounding constant is $M^{*} = \sup_x f(x)/g(x)$; a
  finite $M^{*}$ requires the proposal's tail to be at least as heavy
  as the target's.
- Exponential tilting re-weights a base density by $e^{\theta x}$ to
  construct proposals well matched to a target tail region; applied to
  the standard Normal tail, it yields the elegant "two exponentials"
  algorithm whose efficiency improves, rather than worsens, as the
  threshold moves further into the tail.

# Key Formulae

- Bounding condition: $f(x) \le Mg(x)$ for all $x$ in the support.
- Acceptance rule: accept $Y$ if $U \le f(Y)/(Mg(Y))$.
- Acceptance probability per trial: $1/M$.
- Optimal bounding constant: $M^{*} = \sup_x f(x)/g(x)$.
- Trials until acceptance: $N \sim \text{Geometric}(1/M)$,
  $E(N) = M$.
- Exponentially tilted density: $h_\theta(x) = e^{\theta x}h(x)/M_h(\theta)$.
- Tilted Normal-tail envelope ($\theta=a$): $g(x)=ae^{-a(x-a)}$,
  $x>a$; acceptance ratio $\exp(-(x-a)^2/2)$;
  $M^{*} = \varphi(a)/(aQ(a))$.

# Key Terms

- **Acceptance-rejection method.** A simulation technique that
  generates exact draws from a target density by accepting or
  rejecting candidates drawn from a proposal density, according to a
  ratio of densities.
- **Bounding constant ($M$).** The smallest constant such that the
  target density never exceeds $M$ times the proposal density.
- **Envelope (proposal, instrumental) density.** The density $g$
  actually used to generate candidates.
- **Exponential tilting.** Re-weighting a density by an exponential
  factor $e^{\theta x}$ to shift its mass toward a region of interest,
  most often a distribution's tail.
- **Mills ratio.** The ratio $Q(a)/\varphi(a)$ (or its reciprocal),
  describing the asymptotic decay of the Normal tail probability.
- **Rejection sampling.** A synonym for the acceptance-rejection
  method.
- **Target density.** The density $f$ from which a sample is
  ultimately desired.
- **Truncated distribution.** A distribution restricted to a subset of
  its original support and renormalized to integrate to one.

# Exercises

## Multiple Choice Questions

1. The acceptance-rejection method requires
   (a) a closed-form quantile function for the target density
   (b) a proposal density and a valid bounding constant $M$
   (c) the target density to be Normal
   (d) the moment generating function of the target

2. The probability that a single trial is accepted equals
   (a) $M$
   (b) $1/M$
   (c) $f(Y)$
   (d) $g(Y)$

3. The number of trials until the first acceptance follows a
   (a) Binomial distribution
   (b) Poisson distribution
   (c) Geometric distribution
   (d) Normal distribution

4. The optimal bounding constant $M^{*}$ is defined as
   (a) $\inf_x f(x)/g(x)$
   (b) $\sup_x f(x)/g(x)$
   (c) $\int f(x)/g(x)\,dx$
   (d) $E[f(X)/g(X)]$

5. A finite bounding constant fails to exist when
   (a) the proposal's tail is heavier than the target's tail
   (b) the proposal's tail is lighter than the target's tail
   (c) the target is bounded
   (d) the proposal is Uniform

6. Exponential tilting re-weights a base density by
   (a) $\log(x)$
   (b) $e^{\theta x}$
   (c) $x^\theta$
   (d) $1/x$

7. For the tilted Normal-tail envelope with $\theta=a$, the
   acceptance ratio simplifies to
   (a) $e^{-(x-a)^2/2}$
   (b) $e^{-x^2/2}$
   (c) $a e^{-ax}$
   (d) $\Phi(x)-\Phi(a)$

8. As the threshold $a$ increases, the tilted-envelope acceptance
   probability
   (a) decreases toward zero
   (b) stays exactly constant
   (c) increases toward one
   (d) becomes undefined

9. The acceptance-rejection method remains exact even when the target
   density is known only up to
   (a) a location shift
   (b) an unknown multiplicative constant
   (c) an unknown scale parameter only
   (d) it cannot be used in this situation

10. Working on the log scale during acceptance comparison primarily
    guards against
    (a) slow convergence of Markov chains
    (b) floating-point underflow
    (c) violation of the Hull-Dobell theorem
    (d) loss of the Markov property

**Answers.** 1(b), 2(b), 3(c), 4(b), 5(b), 6(b), 7(a), 8(c), 9(b),
10(b).

## Short Answer Questions

1. Explain why the inverse transformation method is sometimes
   unusable, motivating the need for acceptance-rejection.
2. State the bounding condition that $M$, $f$, and $g$ must satisfy,
   and explain in words what happens if this condition is violated
   somewhere in the support.
3. Derive the unconditional acceptance probability of a single trial
   from first principles.
4. Explain, without invoking the formal proof, the geometric picture
   underlying the correctness of the method.
5. Why does the choice $M = M^{*} = \sup_x f(x)/g(x)$ minimize the
   expected number of trials per accepted observation?
6. Give an example of a target-proposal pairing for which no finite
   $M$ exists, and explain why.
7. Define exponential tilting and explain, intuitively, why increasing
   $\theta$ shifts mass toward larger values.
8. Why does the acceptance-rejection method remain exact when the
   target density is known only up to a multiplicative constant?
9. Explain why the tilted "two exponentials" algorithm requires no
   evaluation of $\Phi$ or $\varphi$ at run time.
10. Why is working on the log scale recommended for high-dimensional
    or heavy-tailed acceptance-rejection implementations?

## Long Answer Questions

1. State and prove Theorem 4.1, explaining each step of both the
   algebraic derivation and, separately, the geometric argument given
   in the Remarks.
2. State and prove Theorem 4.2, and explain its practical implications
   for the choice of $M$.
3. Derive the optimal bounding constant and acceptance ratio for the
   half-Normal density generated from an Exponential$(1)$ proposal,
   showing every algebraic step.
4. Derive the tilted envelope, the optimal-form bounding constant
   $M^{*}$, and the simplified acceptance ratio for simulating the
   standard Normal tail beyond a threshold $a$, using tilting
   parameter $\theta=a$.
5. Explain, with reference to the Mills ratio, why the tilted-envelope
   method becomes more efficient as the threshold $a$ increases, while
   the naive method becomes less efficient.
6. Compare the acceptance-rejection method with the inverse
   transformation method in terms of generality, computational cost,
   and the information required about the target distribution.
7. Discuss the curse of dimensionality as it affects
   acceptance-rejection sampling in high-dimensional settings, and
   explain why Markov chain Monte Carlo methods are often preferred in
   that regime despite being only asymptotically exact.
8. Explain how the acceptance-rejection method is used for Bayesian
   posterior simulation when the normalizing constant of the posterior
   is unknown, and why this is not a drawback under Theorem 4.1.
9. Discuss common implementation errors in acceptance-rejection
   sampling and the diagnostic role of comparing empirical and
   theoretical acceptance rates.
10. Trace the historical development of the acceptance-rejection
    method from von Neumann's original proposal to its modern role as
    a benchmark for exact simulation, including the role of
    exponential tilting in extending its reach to rare-event
    simulation.

## Programming Exercises

1. Implement `accept_reject()` from first principles (already provided
   in this unit) and use it to reproduce the manual calculation of
   Example 2 by fixing the random seed appropriately.
2. Write an R function that numerically estimates $M^{*} =
   \sup_x f(x)/g(x)$ for arbitrary $f$ and $g$ using `optimize()`, and
   verify it against the analytic value $\sqrt{2e/\pi}$ for the
   half-Normal example.
3. Implement the tilted "two exponentials" algorithm from first
   principles (already provided) and use it to estimate $P(Z>4 \mid
   Z>4)$-style tail quantiles (e.g. the 99th percentile of the
   truncated distribution) from a large tilted sample.
4. Using `system.time()`, compare the total number of random draws
   consumed by the naive and tilted methods to produce $1000$
   observations of $Z \mid Z>3$, and report the ratio.
5. Write an R function implementing acceptance-rejection for the
   Beta$(\alpha,\beta)$ density with a Uniform envelope for general
   $\alpha,\beta \ge 1$, computing the required $M$ analytically from
   the mode of the Beta density.
6. Simulate $n=5000$ observations from the target
   $f(x) \propto e^{-x}\bigl(1+\sin^2(3x)\bigr)$, $x>0$ (an unnormalized
   density with no closed form for its normalizing constant), using an
   Exponential$(1)$ proposal, and verify graphically that the
   histogram matches $f$ up to a constant.
7. Write a function that logs, for each rejected candidate, its value
   and its acceptance ratio, and produce a plot showing the
   distribution of acceptance ratios across many rejected candidates
   for the Beta$(2,2)$ example.
8. Implement a general-purpose acceptance-rejection sampler that works
   entirely on the log scale (log-target, log-proposal, log-$M$), and
   verify it reproduces the same output distribution as the ordinary-
   scale implementation for the half-Normal example.
9. Investigate empirically what happens when an invalid (too small)
   $M$ is used for the Beta$(2,2)$ example — implement the sampler
   with $M=1$ instead of $M=1.5$, generate a large sample, and use a
   Kolmogorov-Smirnov test (`ks.test()`) to demonstrate the resulting
   bias.
10. Write an R function comparing, across a grid of thresholds
    $a \in \{1,2,3,4,5\}$, the theoretical acceptance probabilities of
    the naive and tilted methods, and produce a `ggplot2` figure on a
    logarithmic y-axis showing how the gap widens with $a$.

## Mini Projects

1. **A general-purpose tilted-envelope tail sampler.** Extend the
   Normal-tail tilting construction of this unit to the Exponential
   and Gamma distributions: for each, derive (or look up and verify
   numerically) a tilted proposal suited to simulating the extreme
   right tail, implement the corresponding acceptance-rejection
   sampler in R, and benchmark its acceptance rate against naive
   rejection across a range of thresholds. Summarize your findings,
   including the algebraic derivation for at least one distribution
   beyond the Normal, in a short report.

2. **Diagnosing and repairing a biased sampler.** Take the Beta$(2,2)$
   acceptance-rejection example and deliberately introduce three
   different bugs, one at a time: (i) an invalid (too small) $M$; (ii)
   a proposal density with lighter support than the target (e.g. a
   Uniform$(0.1, 0.9)$ proposal); (iii) reusing the same uniform draw
   $U$ both to generate the candidate and to decide acceptance instead
   of drawing two independent uniforms. For each bug, generate a large
   sample, compare it to the true Beta$(2,2)$ density both graphically
   and using a formal goodness-of-fit test, and explain, from first
   principles, why each bug produces the biased behaviour observed.

**Difficulty labels.** Multiple choice, short answer: **Easy**. Long
answer questions 3, 6, 7, 9, programming exercises 1-6: **Medium**.
Long answer questions 1, 2, 4, 5, 8, 10, programming exercises 7-10:
**Medium to Advanced**. Mini projects: **Advanced**.

# References

Devroye, L. (1986). *Non-uniform random variate generation*.
Springer-Verlag.

Marsaglia, G. (1964). Generating a variable from the tail of the
normal distribution. *Technometrics*, 6(1), 101-102.

Robert, C. P., & Casella, G. (2004). *Monte Carlo statistical
methods* (2nd ed.). Springer.

Ripley, B. D. (1987). *Stochastic simulation*. Wiley.

Rizzo, M. L. (2019). *Statistical computing with R* (2nd ed.). CRC
Press.

Ross, S. M. (2020). *Simulation* (6th ed.). Academic Press.

von Neumann, J. (1951). Various techniques used in connection with
random digits. In *Monte Carlo Method* (National Bureau of Standards
Applied Mathematics Series 12, pp. 36-38). U.S. Government Printing
Office.
