---
title: "Unit 3.3: The Inverse Transformation Method"
author: "Dr. Manoj C. Patil"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: false
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
    code_folding: hide
    theme: cosmo
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    echo = TRUE,
    warning = FALSE,
    message = FALSE,
    fig.align = "center",
    fig.width = 7,
    fig.height = 4.5
)

set.seed(123)
library(tidyverse)
library(ggplot2)
library(dplyr)
library(knitr)
library(kableExtra)
library(gridExtra)
library(patchwork)
```

# Learning Objectives

By the end of this section, a student should be able to

- explain why simulation of an arbitrary random variable can be reduced
  to simulation of a single Uniform$(0,1)$ random variable,
- state and prove the probability integral transformation theorem,
- define the generalized inverse (quantile function) of a cumulative
  distribution function and describe its essential properties,
- derive the distribution of the quantile function evaluated at a
  Uniform$(0,1)$ random variable,
- construct the inverse transformation algorithm for continuous random
  variables and implement it in R,
- construct the inverse transformation algorithm for discrete random
  variables and implement it in R,
- assess the computational advantages and limitations of the method
  relative to alternative generation techniques.



# Prerequisites

A reader should already be comfortable with

- the definition and basic properties of a cumulative distribution
  function (cdf), including right-continuity and monotonicity,
- the notions of a continuous and a discrete random variable,
- elementary calculus, including differentiation and integration of
  standard functions,
- the `runif()` function in R and vectorized R operations.



# Introduction

Every simulation of a stochastic system eventually reduces to a very
small number of primitive operations, the most fundamental of which is
the generation of a variate that behaves, for practical purposes, like
a draw from the Uniform$(0,1)$ distribution. Unit 3.1 discussed how such
uniform variates are produced algorithmically using congruential
generators. The question addressed in this section is the converse
one: given a stream of Uniform$(0,1)$ numbers, how does one convert it
into a stream of numbers that follow a prescribed target distribution,
such as an Exponential, a Gamma, or a Binomial distribution?

The inverse transformation method answers this question in essentially
the most direct way possible. It exploits the fact that every random
variable can be described completely by its cumulative distribution
function, and that this function itself carries enough information to
be inverted and used as a generating mechanism. The idea has a long
history in statistics: it underlies the classical technique of
generating an observation by first drawing a random probability and
then reading off the corresponding quantile, a procedure that predates
electronic computing and appears in early Monte Carlo work carried out
by physicists such as Ulam and von Neumann in the 1940s in connection
with the Manhattan Project. Later authors, notably Fishman (1996) and
Devroye (1986), placed the method on a rigorous footing and catalogued
its use across a wide range of standard distributions.

The practical importance of the method lies in its generality and its
conceptual simplicity. Whenever the inverse cdf can be written down in
closed form, or computed cheaply by numerical inversion, the method
provides an exact, unbiased, and easily coded generator. It is for this
reason the default strategy taught first in any course on stochastic
simulation, and it forms the conceptual foundation against which more
elaborate techniques, such as acceptance-rejection and composition
methods, are later compared.



# Intuition

Consider a continuous random variable $X$ with a strictly increasing
cumulative distribution function $F_X$. Because $F_X$ is strictly
increasing, it maps the real line onto the interval $(0,1)$ in a
one-to-one fashion, and the value $F_X(x)$ can be read as "the
proportion of the distribution lying at or below $x$." This proportion
is itself a number between $0$ and $1$, and a remarkable fact is that
if $X$ is random with distribution $F_X$, then the transformed
quantity $F_X(X)$ is random with the Uniform$(0,1)$ distribution,
regardless of what $F_X$ actually is. Intuitively, $F_X$ stretches and
compresses different regions of the real line so that, after the
transformation, every sub-interval of $(0,1)$ of a given length
receives exactly the same amount of probability mass.

This observation can be turned around. If stretching $X$ through
$F_X$ produces a uniform variate, then the reverse operation —
starting from a uniform variate $U$ and applying the inverse function
$F_X^{-1}$ — must produce a variate distributed as $X$. Generating $X$
therefore reduces to two steps: draw $U$ uniformly on $(0,1)$, and
report $x = F_X^{-1}(u)$ as the simulated value. A useful mental image
is that of reading a vertical line at height $u$ across the graph of
$F_X$ until it meets the curve, and then dropping down to the
corresponding value on the horizontal axis; that horizontal value is
the simulated observation.

```{r intuition-figure, echo=FALSE, fig.cap="Geometric idea of the inverse transformation: a draw u on the vertical axis is mapped, through the cdf, to the simulated value x on the horizontal axis."}
x_grid <- seq(-4, 4, length.out = 400)
cdf_values <- pnorm(x_grid)

u_example <- 0.75
x_example <- qnorm(u_example)

ggplot(data.frame(x = x_grid, F = cdf_values), aes(x = x, y = F)) +
    geom_line(colour = "steelblue", linewidth = 1) +
    geom_segment(
        aes(x = -4, xend = x_example, y = u_example, yend = u_example),
        linetype = "dashed", colour = "darkred"
    ) +
    geom_segment(
        aes(x = x_example, xend = x_example, y = 0, yend = u_example),
        linetype = "dashed", colour = "darkred"
    ) +
    labs(
        title = expression(paste("Reading the inverse cdf at ", u == 0.75)),
        x = "x", y = expression(F[X](x))
    ) +
    theme_minimal(base_size = 12)
```

The situation for a discrete random variable is only slightly more
delicate, because $F_X$ is then a step function rather than a smooth
curve, so its inverse cannot be defined by simple algebraic inversion.
The remedy is to define the inverse in a slightly generalized sense,
described formally below, which continues to make sense whether $F_X$
is continuous, discrete, or a mixture of the two.



# Formal Definitions

## The Quantile Function

**Definition.** For a cumulative distribution function $F_X$, the
**quantile function**, or **generalized inverse**, is defined for
$0 < u < 1$ by

\[
F_X^{-1}(u) \;=\; \inf\{\, x \in \mathbb{R} : F_X(x) \ge u \,\}.
\]

where

- $u$ denotes a probability level in $(0,1)$,
- $F_X(x)$ denotes the cumulative distribution function of $X$
  evaluated at $x$,
- $\inf$ denotes the infimum, that is, the greatest lower bound of the
  indicated set.

When $F_X$ is continuous and strictly increasing, $F_X^{-1}$ coincides
with the ordinary functional inverse of $F_X$, so that
$F_X\bigl(F_X^{-1}(u)\bigr) = u$ for every $u \in (0,1)$. When $F_X$
has jumps, as it does for a discrete random variable, or flat
segments, the generalized definition still produces a single
well-defined value $F_X^{-1}(u)$ for every $u$, namely the smallest
$x$ at which the cdf reaches or exceeds the level $u$.

## Properties of the Quantile Function

The quantile function possesses three properties that make it suitable
as a generating device.

- **Property 1 (Non-decreasing).** $F_X^{-1}$ is a non-decreasing
  function of $u$ on $(0,1)$, because $F_X$ itself is non-decreasing.
- **Property 2 (Left-continuity).** $F_X^{-1}$ is left-continuous, a
  mirror image of the right-continuity of $F_X$.
- **Property 3 (Inversion identity).** For every $x \in \mathbb{R}$
  and every $u \in (0,1)$,
  \[
  F_X^{-1}(u) \le x \quad \Longleftrightarrow \quad u \le F_X(x).
  \]
  This equivalence, sometimes called the **compatibility relation**,
  is the identity on which the probability integral transformation
  theorem is built, and it holds regardless of whether $F_X$ is
  continuous or discrete.



# Theorems

## Theorem 3.1: Probability Integral Transformation

**Statement.** Let $X$ be a continuous random variable with cdf
$F_X$. Then the random variable

\[
U = F_X(X)
\]

follows the Uniform$(0,1)$ distribution.

**Assumptions.**

- $F_X$ is continuous on $\mathbb{R}$.
- $F_X$ takes values only in $(0,1)$ over the support of $X$ (the
  boundary values $0$ and $1$ occur with probability zero for a
  continuous distribution).

**Proof.** Let $u \in (0,1)$. Using the compatibility relation from
Property 3,

\[
\begin{aligned}
P(U \le u) &= P\bigl(F_X(X) \le u\bigr) \\
&= P\bigl(X \le F_X^{-1}(u)\bigr) \qquad \text{(by continuity of } F_X \text{, equality holds a.s.)}\\
&= F_X\bigl(F_X^{-1}(u)\bigr) \\
&= u.
\end{aligned}
\]

The final equality uses the fact that, for a continuous cdf, applying
$F_X$ to its own generalized inverse returns the original argument.
Since $P(U \le u) = u$ for every $u \in (0,1)$, $U$ has the
Uniform$(0,1)$ cdf, and the theorem follows. $\blacksquare$

**Remarks.** The theorem is often summarized by saying that "the cdf
of a continuous random variable, evaluated at that random variable,
is uniform." It is the theoretical engine behind numerous statistical
procedures beyond simulation, including goodness-of-fit testing (the
Kolmogorov–Smirnov test of Section 3.2 relies on exactly this fact)
and copula theory.

**Examples.** If $X \sim \text{Exponential}(\lambda)$ with
$F_X(x) = 1 - e^{-\lambda x}$ for $x>0$, then
$U = 1 - e^{-\lambda X}$ is Uniform$(0,1)$; this can be verified
numerically by simulating a large exponential sample and applying the
transformation.

**Applications.** Probability integral transformation is the basis of
copula-based dependence modelling, quantile-quantile plotting for
distributional diagnostics, and simulation-based p-value calibration.

## Theorem 3.2: The Inverse Transformation Theorem

**Statement.** Let $U \sim \text{Uniform}(0,1)$, and let $F_X$ be any
cumulative distribution function with generalized inverse
$F_X^{-1}$. Then the random variable

\[
X = F_X^{-1}(U)
\]

has cumulative distribution function $F_X$.

**Assumptions.**

- $F_X$ is a valid cdf (non-decreasing, right-continuous, with limits
  $0$ and $1$ at $-\infty$ and $+\infty$ respectively).
- No continuity or strict monotonicity of $F_X$ is required; the
  theorem holds for continuous, discrete, and mixed distributions
  alike.

**Proof.** Fix $x \in \mathbb{R}$. Using the compatibility relation of
Property 3,

\[
\begin{aligned}
P(X \le x) &= P\bigl(F_X^{-1}(U) \le x\bigr) \\
&= P\bigl(U \le F_X(x)\bigr) \\
&= F_U\bigl(F_X(x)\bigr) \\
&= F_X(x),
\end{aligned}
\]

where the third equality uses the Uniform$(0,1)$ cdf
$F_U(t) = t$ for $t \in (0,1)$, and $F_X(x) \in [0,1]$ automatically.
Since $P(X \le x) = F_X(x)$ for every $x$, the random variable
$X = F_X^{-1}(U)$ has the required distribution. $\blacksquare$

**Remarks.** This theorem, rather than Theorem 3.1, is the one
actually used for generation: it converts a single recipe — draw
$U$, evaluate $F_X^{-1}$ — into a variate from any target
distribution, continuous or discrete, as long as $F_X^{-1}$ can be
evaluated. The two theorems are converses of one another, and together
they establish a full equivalence between "being distributed as $X$"
and "being obtainable from a uniform variate via $F_X^{-1}$."

**Examples.** These are developed in full in the worked examples
below for both the continuous and the discrete case.

**Applications.** The inverse transformation theorem underlies random
variate generation in nearly every general-purpose statistical
package, is used to couple correlated simulations sharing common
random numbers (a variance-reduction technique), and provides the
generating mechanism for parametric bootstrap resampling from a fitted
distribution.



# Mathematical Derivations

## Distribution of the Quantile Function at a Uniform Argument

It is instructive to see explicitly why the derivation above does not
depend on continuity. Write $G(u) = F_X^{-1}(u)$ for brevity, and
recall the compatibility relation

\[
G(u) \le x \iff u \le F_X(x), \qquad x \in \mathbb{R},\; u \in (0,1).
\]

For a fixed $x$, define $u_0 = F_X(x)$. The relation states that the
event $\{G(U) \le x\}$ and the event $\{U \le u_0\}$ are the same
event, as sets of outcomes, whatever the shape of $F_X$. Since $U$ is
Uniform$(0,1)$,

\[
P(U \le u_0) = u_0 = F_X(x),
\]

and therefore $P(G(U) \le x) = F_X(x)$ for every real $x$, which is
precisely the statement that $G(U) = F_X^{-1}(U)$ has cdf $F_X$. The
argument used only the compatibility relation, so it is equally valid
when $F_X$ has jumps (the discrete case) or flat stretches (a
distribution with an atom followed by a continuous part). This is the
reason the inverse transformation method extends, without
modification in principle, from continuous to discrete random
variables; only the practical evaluation of $F_X^{-1}$ changes.

## Continuous Case: Explicit Inversion

When $F_X$ is continuous and strictly increasing on the support of
$X$, $F_X^{-1}$ can, in many standard families, be written down by
solving $F_X(x) = u$ algebraically for $x$.

**Derivation for the Exponential distribution.** Let
$X \sim \text{Exponential}(\lambda)$, so that
$F_X(x) = 1 - e^{-\lambda x}$ for $x > 0$. Setting $F_X(x) = u$,

\[
\begin{aligned}
1 - e^{-\lambda x} &= u \\
e^{-\lambda x} &= 1 - u \\
-\lambda x &= \log(1-u) \\
x &= -\frac{1}{\lambda}\log(1-u).
\end{aligned}
\]

Hence $F_X^{-1}(u) = -\lambda^{-1}\log(1-u)$. Because $U$ and $1-U$
have the same Uniform$(0,1)$ distribution, it is equally valid, and
slightly more economical in code, to generate

\[
X = -\frac{1}{\lambda}\log(U).
\]

Derivation for the density $f(x) = 3x^2$, $0<x<1$. Integrating,
$F_X(x) = x^3$ for $0<x<1$. Setting $x^3 = u$ gives

\[
F_X^{-1}(u) = u^{1/3}.
\]

## Discrete Case: Inversion by Search

When $X$ is discrete with possible values $x_1 < x_2 < x_3 < \cdots$
at which $F_X$ jumps, the compatibility relation specializes to

\[
F_X^{-1}(u) = x_i \quad \text{where} \quad F_X(x_{i-1}) < u \le F_X(x_i).
\]

Generating $X$ therefore requires locating the first value $x_i$ at
which the cumulative sum of probabilities reaches or exceeds $u$.
Unlike the continuous case, there is generally no closed-form
algebraic shortcut, and the inversion is instead carried out by
searching a table of cumulative probabilities, either sequentially or,
for large tables, using a fast lookup such as a binary search.

**Derivation for a two-point (Bernoulli) distribution.** Let
$X \sim \text{Bernoulli}(p)$, so $F_X(0) = 1-p$ and $F_X(1) = 1$.
Then

\[
F_X^{-1}(u) =
\begin{cases}
0, & 0 < u \le 1-p, \\
1, & 1-p < u < 1.
\end{cases}
\]

**Derivation for a geometric distribution.** Let
$X \sim \text{Geometric}(p)$ on $\{0,1,2,\dots\}$ with
$q = 1-p$ and cdf $F_X(x) = 1 - q^{x+1}$. Solving
$F_X(x-1) < u \le F_X(x)$,

\[
1 - q^{x} < u \le 1 - q^{x+1}
\quad \Longleftrightarrow \quad
x < \frac{\log(1-u)}{\log q} \le x+1,
\]

so that $x + 1 = \big\lceil \log(1-u)/\log q \big\rceil$, where
$\lceil t \rceil$ denotes the ceiling function (the smallest integer
not less than $t$). Because $U$ and $1-U$ are identically distributed,
this simplifies in implementation to

\[
X = \left\lfloor \frac{\log U}{\log q} \right\rfloor.
\]



# Algorithms

## Algorithm 3.1: Inverse Transformation Method, Continuous Case

**Purpose.** Generate a random sample from a continuous distribution
whose quantile function $F_X^{-1}$ is available in closed form or can
be evaluated numerically.

**Input.** Sample size $n$; the quantile function $F_X^{-1}(\cdot)$
(or the parameters needed to evaluate it).

**Output.** A vector $x_1,\dots,x_n$ of simulated observations from
the target distribution.

**Step-by-step procedure.**

1. Derive the quantile function $F_X^{-1}(u)$ analytically by solving
   $F_X(x) = u$ for $x$.
2. Implement $F_X^{-1}(u)$ as an R expression or function.
3. For $i = 1,\dots,n$:
   (a) draw $u_i \sim \text{Uniform}(0,1)$;
   (b) compute $x_i = F_X^{-1}(u_i)$.
4. Return the vector $(x_1,\dots,x_n)$.

**Pseudo-code.**

```
Algorithm: Inverse Transform (Continuous)
Input:  n, quantile function Finv
Output: vector x of length n

1. Generate u[1], ..., u[n] each Uniform(0,1)
2. For i = 1 to n:
       x[i] <- Finv(u[i])
3. Return x
```

**Flowchart description.** Start, draw a uniform vector of length
$n$, apply the (vectorized) quantile function element-wise, obtain the
simulated sample, end.

**Complexity.** Generating $n$ uniform variates costs $O(n)$, and
evaluating a closed-form quantile function at $n$ points also costs
$O(n)$ under vectorization, so the overall time complexity is
$O(n)$. Memory usage is $O(n)$ for storage of the sample.

**Advantages.** Extremely fast when a closed form is available;
requires only one uniform draw per observation; naturally vectorized
in R.

**Limitations.** Requires an explicit or numerically invertible
quantile function; some distributions (e.g. the standard Normal) have
no elementary closed form, and rely on numerical approximations
implemented in built-in functions such as `qnorm()`.

**Applications.** Simulating waiting times (Exponential), simulating
order statistics via the Beta quantile function, generating covariates
in Monte Carlo studies, and parametric bootstrap resampling.

## Algorithm 3.2: Inverse Transformation Method, Discrete Case

**Purpose.** Generate a random sample from a discrete distribution
supported on a countable set of values, by searching a table of
cumulative probabilities.

**Input.** Sample size $n$; the support points
$x_1 < x_2 < \cdots < x_k$; the corresponding cumulative
probabilities $F_X(x_1),\dots,F_X(x_k)$.

**Output.** A vector $x_1^{*},\dots,x_n^{*}$ of simulated observations.

**Step-by-step procedure.**

1. Compute (or look up) the cumulative probability vector
   $F = \bigl(F_X(x_1),\dots,F_X(x_k)\bigr)$.
2. For $i = 1,\dots,n$:
   (a) draw $u_i \sim \text{Uniform}(0,1)$;
   (b) find the smallest index $j$ such that $F_X(x_j) \ge u_i$;
   (c) set $x_i^{*} = x_j$.
3. Return the vector $(x_1^{*},\dots,x_n^{*})$.

**Pseudo-code.**

```
Algorithm: Inverse Transform (Discrete)
Input:  n, support (x_1,...,x_k), cdf values (F_1,...,F_k)
Output: vector x of length n

1. Generate u[1], ..., u[n] each Uniform(0,1)
2. For i = 1 to n:
       j <- smallest index such that F[j] >= u[i]
       x[i] <- support[j]
3. Return x
```

**Flowchart description.** Start, build or retrieve the cumulative
probability table, draw a uniform vector, for each uniform value scan
(or binary-search) the table for the first cumulative probability at
or above it, record the corresponding support point, end.

**Complexity.** With linear search of a table of size $k$, generating
$n$ observations costs $O(nk)$ in the worst case. Using a
pre-sorted table with binary search reduces this to $O(n \log k)$.
Vectorized R implementations exploiting `findInterval()` achieve the
$O(n \log k)$ bound efficiently in practice.

**Advantages.** Works for any discrete distribution with a computable
cdf, including distributions without closed-form quantile expressions;
straightforward to implement using base R search utilities.

**Limitations.** Slower than direct continuous inversion because of
the search step; for distributions with unbounded support (e.g. the
Poisson or the logarithmic distribution used later in this unit) the
cumulative table must be truncated or extended adaptively.

**Applications.** Simulating categorical survey responses, generating
Bernoulli and Binomial trials, bootstrap resampling from an empirical
distribution, simulating claim counts in actuarial risk models.



# R Programming

The following code implements both algorithms. Each example follows
the pattern: state the target distribution, derive or specify the
quantile function, generate the sample, and compare it graphically and
numerically against the theoretical distribution.

## Continuous Case: Density $f(x) = 3x^2$, $0 < x < 1$

```{r continuous-example-1}
set.seed(123)

## Target density f(x) = 3x^2 on (0,1) has cdf F(x) = x^3
## so the quantile function is F^{-1}(u) = u^{1/3}

sample_size <- 1000
uniform_draws <- runif(n = sample_size)
simulated_x <- uniform_draws^(1 / 3)

summary(simulated_x)
```

```{r continuous-example-1-plot, fig.cap="Simulated density (histogram) versus the theoretical density f(x) = $3x^2$ superimposed as a curve."}
theoretical_grid <- seq(0, 1, by = 0.01)
theoretical_density <- 3 * theoretical_grid^2

ggplot(data.frame(x = simulated_x), aes(x = x)) +
    geom_histogram(
        aes(y = after_stat(density)),
        bins = 30, fill = "steelblue", colour = "white", alpha = 0.8
    ) +
    geom_line(
        data = data.frame(x = theoretical_grid, y = theoretical_density),
        aes(x = x, y = y),
        colour = "darkred", linewidth = 1
    ) +
    labs(
        title = expression(paste("Simulated vs. theoretical density, ", f(x) == 3 * x^2)),
        x = "x", y = "Density"
    ) +
    theme_minimal(base_size = 12)
```

## Continuous Case: Exponential Distribution

```{r continuous-example-2}
set.seed(123)

## X ~ Exponential(rate = lambda); F^{-1}(u) = -log(u) / lambda
sample_size <- 1000
lambda <- 2

uniform_draws <- runif(n = sample_size)
simulated_exp <- -log(uniform_draws) / lambda

sample_mean <- mean(simulated_exp)
theoretical_mean <- 1 / lambda

kable(
    data.frame(
        Quantity = c("Sample mean", "Theoretical mean (1/lambda)"),
        Value = round(c(sample_mean, theoretical_mean), 4)
    ),
    caption = "Comparison of sample and theoretical mean, Exponential(2)"
) %>%
    kable_styling(full_width = FALSE)
```

## Discrete Case: Bernoulli Distribution

```{r discrete-example-1}
set.seed(123)

## X ~ Bernoulli(p); F(0) = 1-p, F(1) = 1
sample_size <- 1000
success_prob <- 0.4

uniform_draws <- runif(n = sample_size)
simulated_bernoulli <- as.integer(uniform_draws > (1 - success_prob))

kable(
    data.frame(
        Quantity = c("Sample mean", "Theoretical mean (p)",
                      "Sample variance", "Theoretical variance p(1-p)"),
        Value = round(
            c(mean(simulated_bernoulli), success_prob,
              var(simulated_bernoulli), success_prob * (1 - success_prob)),
            4
        )
    ),
    caption = "Comparison of sample and theoretical moments, Bernoulli(0.4)"
) %>%
    kable_styling(full_width = FALSE)
```

## Discrete Case: Geometric Distribution

```{r discrete-example-2}
set.seed(123)

## X ~ Geometric(p) on {0,1,2,...}; using X = floor(log(u) / log(1-p))
sample_size <- 1000
success_prob <- 0.25

uniform_draws <- runif(n = sample_size)
simulated_geometric <- floor(log(uniform_draws) / log(1 - success_prob))

sample_mean <- mean(simulated_geometric)
theoretical_mean <- (1 - success_prob) / success_prob

kable(
    data.frame(
        Quantity = c("Sample mean", "Theoretical mean ((1-p)/p)"),
        Value = round(c(sample_mean, theoretical_mean), 4)
    ),
    caption = "Comparison of sample and theoretical mean, Geometric(0.25)"
) %>%
    kable_styling(full_width = FALSE)
```

## Discrete Case: General Table Lookup with `findInterval()`

```{r discrete-example-3}
set.seed(123)

## General discrete distribution on support {1,2,3,4} with given pmf
support_points <- c(1, 2, 3, 4)
pmf_values <- c(0.1, 0.2, 0.3, 0.4)
cdf_values <- cumsum(pmf_values)

sample_size <- 1000
uniform_draws <- runif(n = sample_size)

## findInterval locates, for each u, the index of the first cdf value >= u
indices <- findInterval(
    x = uniform_draws,
    vec = cdf_values,
    left.open = TRUE
) + 1
simulated_general <- support_points[indices]

table(simulated_general) / sample_size
```



# Simulation Studies

## Purpose

The purpose of this simulation study is to verify, for both a
continuous and a discrete example, that the empirical distribution
produced by the inverse transformation algorithm converges to the
theoretical target distribution as the sample size increases, and to
illustrate the rate of that convergence.

## Code

```{r simulation-study}
set.seed(123)

sample_sizes <- c(50, 200, 1000, 5000)
lambda <- 1

convergence_results <- map_dfr(sample_sizes, function(n) {
    u <- runif(n)
    x <- -log(u) / lambda
    tibble(
        n = n,
        sample_mean = mean(x),
        sample_sd = sd(x)
    )
})

kable(
    convergence_results,
    col.names = c("Sample size (n)", "Sample mean", "Sample SD"),
    caption = "Convergence of sample moments to theoretical values (mean = 1, sd = 1) as n increases"
) %>%
    kable_styling(full_width = FALSE)
```

## Output

```{r simulation-study-plot, fig.cap="Kernel density estimates of simulated Exponential(1) samples at increasing sample sizes, compared with the theoretical density."}
set.seed(123)

simulated_data <- map_dfr(sample_sizes, function(n) {
    u <- runif(n)
    x <- -log(u) / lambda
    tibble(n = factor(n), x = x)
})

ggplot(simulated_data, aes(x = x, colour = n)) +
    geom_density(linewidth = 0.9) +
    stat_function(
        fun = dexp, args = list(rate = lambda),
        colour = "black", linetype = "dashed", linewidth = 1
    ) +
    labs(
        title = "Convergence of simulated density to the Exponential(1) target",
        x = "x", y = "Density", colour = "Sample size"
    ) +
    theme_minimal(base_size = 12)
```

## Interpretation and Discussion

The table shows that the sample mean and sample standard deviation
approach the theoretical values of $1$ and $1$ (for a Rate$=1$
Exponential distribution) as $n$ grows, with the fluctuation around
these targets shrinking approximately in proportion to
$1/\sqrt{n}$, consistent with the law of large numbers and the central
limit theorem applied to the sample mean. The density plot exhibits
the same pattern visually: the estimated curve for $n=50$ is jagged
and departs noticeably from the theoretical dashed curve, whereas the
curve for $n=5000$ is smooth and tracks the theoretical density
closely. This confirms that the inverse transformation algorithm is
not merely producing values with the correct support, but is
reproducing the full shape of the target distribution, as guaranteed
by Theorem 3.2. The exercise also illustrates a general principle in
simulation practice: convergence diagnostics of this kind should
always accompany a newly implemented generator before it is trusted in
downstream analyses.



# Graphs

The figures embedded in the R Programming and Simulation Studies
sections above were produced entirely in R using `ggplot2`, following
the house style of this course. Each figure carries a title, axis
labels, and, where more than one series is displayed, a legend
distinguishing the series (see the convergence plot, where colour
encodes sample size).



# Numerical Examples

## Example 1: Density $f(x) = 3x^2$

**Problem.** Using the inverse transformation method, describe how a
single observation is generated from the density $f(x) = 3x^2$,
$0<x<1$, given the uniform draw $u = 0.729$.

**Given.** $F_X(x) = x^3$, hence $F_X^{-1}(u) = u^{1/3}$; $u = 0.729$.

**Solution.**

\[
x = u^{1/3} = 0.729^{1/3} = 0.9.
\]

**Interpretation.** The value $u=0.729$, interpreted as a cumulative
probability, corresponds to the point $x=0.9$ under this distribution,
meaning that $72.9\%$ of the probability mass of $X$ lies at or below
$0.9$, consistent with $F_X(0.9) = 0.9^3 = 0.729$.

## Example 2: Exponential Distribution

**Problem.** Using the inverse transformation method with rate
$\lambda = 0.5$, find the simulated observation corresponding to the
uniform draw $u = 0.2$.

**Given.** $F_X^{-1}(u) = -\log(u)/\lambda$; $\lambda = 0.5$;
$u = 0.2$.

**Solution.**

\[
x = -\frac{\log(0.2)}{0.5} = -\frac{-1.6094}{0.5} = 3.2189.
\]

**Interpretation.** A relatively small uniform draw of $0.2$ produces
a moderately large exponential observation, illustrating that small
values of $u$ near zero are mapped to the lower tail of $F_X$, which
for the exponential distribution corresponds to values further into
the right tail because $F_X$ rises slowly for small $x$ relative to
$\lambda = 0.5$.

## Example 3: Bernoulli Distribution

**Problem.** With success probability $p = 0.4$, determine the
simulated value of $X \sim \text{Bernoulli}(0.4)$ corresponding to
uniform draws $u = 0.55$ and $u = 0.70$.

**Given.** $F_X(0) = 0.6$, $F_X(1) = 1$.

**Solution.** For $u = 0.55$: since $0.55 \le 0.6 = F_X(0)$, the
generated value is $x = 0$. For $u = 0.70$: since $0.70 > 0.6$, the
generated value is $x = 1$.

**Interpretation.** Uniform draws at or below $F_X(0) = 1-p = 0.6$
are mapped to the outcome $0$, and draws above that threshold are
mapped to $1$; the length of each sub-interval of $(0,1)$ mapped to a
given outcome equals the probability of that outcome, here $0.6$ for
outcome $0$ and $0.4$ for outcome $1$.

## Example 4: Geometric Distribution

**Problem.** With $p = 0.25$, determine the simulated value of
$X \sim \text{Geometric}(0.25)$ on $\{0,1,2,\dots\}$ corresponding to
the uniform draw $u = 0.6$.

**Given.** $q = 1-p = 0.75$; $X = \lfloor \log(u)/\log(q) \rfloor$.

**Solution.**

\[
x = \left\lfloor \frac{\log(0.6)}{\log(0.75)} \right\rfloor
= \left\lfloor \frac{-0.5108}{-0.2877} \right\rfloor
= \lfloor 1.7756 \rfloor = 1.
\]

**Interpretation.** The generated value $x=1$ means that, under this
draw, the process required one "failure" before the first "success" in
the underlying Bernoulli trial sequence interpretation of the
geometric distribution.



# Applications

**Statistics.** The inverse transformation method provides the
standard mechanism for parametric bootstrap resampling, in which
observations are drawn from a fitted parametric cdf rather than from
the empirical distribution.

**Machine Learning.** Data augmentation pipelines that need to inject
synthetic covariates from a specified marginal distribution frequently
rely on inverse transformation, particularly within copula-based
generative models.

**Data Science.** Exploratory simulation of "what-if" scenarios, such
as generating synthetic customer arrival times from a fitted
distribution, uses the method directly.

**Finance.** Simulation of asset return distributions, value-at-risk
estimation via Monte Carlo, and simulation of interarrival times of
trades in market microstructure models all employ inverse
transformation for the marginal distributions involved.

**Reliability.** Failure-time simulation from Exponential, Weibull, or
Gamma lifetime models is a direct application, since these
distributions admit closed-form or easily computed quantile functions.

**Monte Carlo.** The method is the workhorse for generating the
primitive random inputs of a Monte Carlo integration or estimation
scheme whenever the required marginal distribution is available in
closed form.

**Bayesian Statistics.** Inverse transformation is used to sample from
full conditional distributions in Gibbs sampling steps whenever those
conditionals belong to a standard family with a tractable quantile
function.

**Operations Research.** Discrete-event simulation of queueing
systems relies on the discrete and continuous inverse transformation
algorithms to generate service times, interarrival times, and
categorical routing decisions.

**Engineering.** Simulation of component lifetimes and stress-test
scenarios in reliability engineering commonly draws on the continuous
inverse transformation algorithm applied to lifetime distributions.



# Advantages

- Conceptually simple and directly justified by Theorem 3.2.
- Requires exactly one uniform draw per simulated observation, which
  is efficient when uniform generation is the bottleneck.
- Exact: the simulated variate has precisely the target distribution,
  with no approximation error beyond floating-point rounding.
- Naturally vectorizable in R, giving fast implementations for large
  sample sizes.
- Extends without conceptual change from continuous to discrete
  distributions via the generalized inverse.



# Limitations

- Requires an explicit or efficiently computable quantile function;
  some important distributions, such as the standard Normal, have no
  elementary closed-form inverse and must rely on numerical
  approximation algorithms internally (as implemented in `qnorm()`).
- For discrete distributions with large or unbounded support, table
  search adds computational overhead relative to specialized
  generators.
- Distributions defined only implicitly, for example through a
  density known up to a normalizing constant, are typically not
  amenable to this method and require alternatives such as
  acceptance-rejection (Unit 3.4).



# Practical Tips

- Always verify that the derived quantile function is applied to the
  correct branch of the cdf; sign errors in solving $F_X(x)=u$ are a
  common source of bugs.
- Exploit the identity that $U$ and $1-U$ share the same
  Uniform$(0,1)$ distribution to simplify formulas, as done for the
  Exponential and Geometric cases above.
- When implementing the discrete case, prefer `findInterval()` over an
  explicit loop for speed; a naive `for` loop searching a cumulative
  table becomes slow for large sample sizes.
- Always set a seed with `set.seed()` before a simulation intended to
  be reproducible, and document the seed value used.
- Check the boundary behaviour of a custom quantile function at
  $u \to 0^{+}$ and $u \to 1^{-}$ to ensure it does not produce
  `NaN` or `Inf` values inadvertently, particularly for distributions
  with unbounded support.



# Common Mistakes

Students often confuse the **cdf** with the **quantile function**,
attempting to plug a value of $x$ into $F_X^{-1}$ instead of a
probability $u$. It is important to remember that $F_X$ maps values of
the random variable to probabilities, while $F_X^{-1}$ maps
probabilities back to values of the random variable; the two functions
have different domains and ranges, and reversing them by mistake
produces a nonsensical simulated sample. Another frequent error is
neglecting the discrete case's need for a search step, and instead
attempting to apply an algebraic inverse formula that does not exist
because $F_X$ is not continuous.



# Summary

- Every continuous random variable $X$ satisfies $F_X(X) \sim
  \text{Uniform}(0,1)$ (Theorem 3.1), and conversely,
  $F_X^{-1}(U) \sim X$ for $U \sim \text{Uniform}(0,1)$ (Theorem
  3.2), where $F_X^{-1}$ is the generalized inverse (quantile
  function) of $F_X$.
- The inverse transformation algorithm reduces simulation of any
  distribution with a computable quantile function to a single
  uniform draw followed by evaluation of $F_X^{-1}$.
- In the continuous case, $F_X^{-1}$ is often available in closed
  form by algebraically solving $F_X(x) = u$.
- In the discrete case, $F_X^{-1}(u)$ is obtained by locating the
  smallest support point $x_i$ for which $F_X(x_i) \ge u$, typically
  via table search.
- The method is exact and simple, but its efficiency and applicability
  depend on how cheaply $F_X^{-1}$ can be evaluated.



# Key Formulae

| Concept | Formula |
|||
| Quantile function | $F_X^{-1}(u) = \inf\{x : F_X(x) \ge u\}$ |
| Compatibility relation | $F_X^{-1}(u) \le x \iff u \le F_X(x)$ |
| Probability integral transform | $F_X(X) \sim \text{Uniform}(0,1)$ |
| Inverse transform generator | $X = F_X^{-1}(U),\; U \sim \text{Uniform}(0,1)$ |
| Exponential quantile function | $F_X^{-1}(u) = -\lambda^{-1}\log(1-u)$ |
| Discrete inversion rule | $F_X^{-1}(u) = x_i$ where $F_X(x_{i-1}) < u \le F_X(x_i)$ |
| Geometric quantile function | $F_X^{-1}(u) = \big\lceil \log(1-u)/\log(1-p) \big\rceil - 1$ |



# Key Terms

- **Cumulative distribution function (cdf):** the function
  $F_X(x) = P(X \le x)$.
- **Quantile function (generalized inverse):** the function
  $F_X^{-1}(u) = \inf\{x : F_X(x) \ge u\}$.
- **Probability integral transformation:** the result that
  $F_X(X) \sim \text{Uniform}(0,1)$ for continuous $X$.
- **Inverse transformation method:** a simulation technique that
  generates $X = F_X^{-1}(U)$ from a uniform variate $U$.
- **Support:** the set of values that a random variable can take with
  positive probability (discrete case) or positive density
  (continuous case).
- **Compatibility relation:** the equivalence
  $F_X^{-1}(u) \le x \iff u \le F_X(x)$.



# Exercises

## Multiple Choice Questions

1. The inverse transformation method requires which of the following
   as its only source of randomness?
   (a) A Normal(0,1) variate
   (b) A Uniform(0,1) variate
   (c) A Poisson variate
   (d) A Bernoulli variate

2. If $X$ is continuous with cdf $F_X$, then $F_X(X)$ follows
   (a) the distribution of $X$
   (b) the Uniform(0,1) distribution
   (c) the standard Normal distribution
   (d) the Exponential(1) distribution

3. The quantile function $F_X^{-1}(u)$ is defined as
   (a) $\sup\{x : F_X(x) \le u\}$
   (b) $\inf\{x : F_X(x) \ge u\}$
   (c) the derivative of $F_X$
   (d) the reciprocal of $F_X$

4. For $X \sim \text{Exponential}(\lambda)$, the quantile function is
   (a) $-\lambda \log(1-u)$
   (b) $-\lambda^{-1}\log(1-u)$
   (c) $\lambda^{-1} e^{-u}$
   (d) $\lambda u$

5. In the discrete case, $F_X^{-1}(u) = x_i$ whenever
   (a) $u = F_X(x_i)$ exactly
   (b) $F_X(x_{i-1}) < u \le F_X(x_i)$
   (c) $F_X(x_i) < u \le F_X(x_{i-1})$
   (d) $u < F_X(x_{i-1})$

6. Which R function is most efficient for locating the correct
   discrete category during vectorized inverse transformation?
   (a) `sapply()` with a manual loop
   (b) `findInterval()`
   (c) `paste()`
   (d) `Recall()`

7. Which identity is commonly exploited to simplify quantile function
   formulas?
   (a) $U$ and $U^2$ are identically distributed
   (b) $U$ and $1-U$ are identically distributed
   (c) $U$ and $\log U$ are identically distributed
   (d) $U$ is always symmetric about $1$

8. The time complexity of continuous inverse transformation for $n$
   observations, assuming a closed-form quantile function, is
   (a) $O(n^2)$
   (b) $O(n \log n)$
   (c) $O(n)$
   (d) $O(\log n)$

9. Which distribution does NOT have a simple closed-form quantile
   function?
   (a) Exponential
   (b) Uniform
   (c) standard Normal
   (d) two-point (Bernoulli)

10. The compatibility relation states that
    (a) $F_X^{-1}(u) \le x \iff u \le F_X(x)$
    (b) $F_X^{-1}(u) \ge x \iff u \ge F_X(x)$
    (c) $F_X(u) \le x \iff F_X^{-1}(x) \le u$
    (d) $F_X$ and $F_X^{-1}$ are always equal

**Answers.** 1(b), 2(b), 3(b), 4(b), 5(b), 6(b), 7(b), 8(c), 9(c), 10(a).

## Fill in the Blanks

1. The inverse transformation method converts a __________ variate
   into a variate from any target distribution.
2. The function $F_X^{-1}(u) = \inf\{x : F_X(x) \ge u\}$ is called the
   __________ function.
3. The probability integral transformation states that
   $F_X(X)$ follows the __________ distribution.
4. For $X \sim \text{Exponential}(\lambda)$, the quantile function is
   $F_X^{-1}(u) = $ __________.
5. In the discrete case, $F_X^{-1}(u)$ is found by __________ a
   cumulative probability table.
6. The R function __________ efficiently performs table lookup for
   vectorized discrete inversion.
7. $U$ and __________ share the same Uniform(0,1) distribution.
8. The quantile function is always __________ (increasing/decreasing)
   in $u$.
9. Theorem 3.2 shows that $F_X^{-1}(U)$ has cdf equal to __________.
10. The standard Normal distribution requires __________ inversion
    because no elementary closed form exists.

**Answers.** 1. uniform; 2. quantile (generalized inverse); 3.
Uniform(0,1); 4. $-\lambda^{-1}\log(1-u)$; 5. searching; 6.
`findInterval()`; 7. $1-U$; 8. non-decreasing; 9. $F_X$; 10. numerical.

## True or False

1. The inverse transformation method can be applied to discrete
   distributions. (True/False)
2. $F_X(X)$ is Uniform(0,1) only when $X$ is discrete. (True/False)
3. The quantile function is always continuous. (True/False)
4. $U$ and $1-U$ have the same distribution when $U \sim
   \text{Uniform}(0,1)$. (True/False)
5. The exponential quantile function requires numerical root-finding.
   (True/False)
6. The discrete inverse transformation algorithm never requires a
   search step. (True/False)
7. Theorem 3.2 holds regardless of whether $F_X$ is continuous.
   (True/False)
8. `findInterval()` can be used to vectorize discrete inverse
   transformation. (True/False)
9. The inverse transformation method always requires more than one
   uniform draw per observation. (True/False)
10. The quantile function is non-decreasing in $u$. (True/False)

**Answers.** 1. True; 2. False; 3. False; 4. True; 5. False; 6.
False; 7. True; 8. True; 9. False; 10. True.

## Short Answer Questions

1. State the probability integral transformation theorem.
2. Define the quantile function using the infimum notation.
3. Explain why $U$ and $1-U$ can be used interchangeably in
   simplifying quantile functions.
4. Why does the inverse transformation method require a search
   procedure in the discrete case but not in the continuous case?
5. What is the compatibility relation, and why is it central to the
   proof of Theorem 3.2?
6. Give one example of a distribution for which the quantile function
   has no elementary closed form.
7. Explain the geometric interpretation of reading a horizontal line
   at height $u$ across the graph of $F_X$.
8. What is the time complexity of the discrete algorithm using a
   linear search, and how is it improved?
9. Why is the inverse transformation method described as "exact"?
10. Describe one application of the inverse transformation method in
    reliability engineering.

## Long Answer Questions

1. Derive the probability integral transformation theorem, and
   explain its role in goodness-of-fit testing.
2. State and prove the inverse transformation theorem for a general
   cdf, without assuming continuity.
3. Derive the quantile function for the Exponential distribution and
   explain the simplification obtained using the identity between
   $U$ and $1-U$.
4. Derive the discrete inverse transformation formula for the
   Geometric distribution, showing every algebraic step.
5. Compare the computational complexity of the continuous and
   discrete inverse transformation algorithms, and explain the source
   of the difference.
6. Explain, with a worked numerical example, how the inverse
   transformation method is used for parametric bootstrap resampling.
7. Discuss the limitations of the inverse transformation method and
   describe at least one alternative technique appropriate when these
   limitations apply.
8. Explain how `findInterval()` implements the discrete inverse
   transformation algorithm, and compare it with a manual `for` loop
   implementation in terms of both correctness and speed.
9. Discuss the historical development of the inverse transformation
   method and its relationship to early Monte Carlo simulation.
10. Using the compatibility relation, prove that the quantile function
    is non-decreasing and left-continuous.

## Numerical Problems

1. Given $f(x) = 2x$, $0<x<1$, derive $F_X^{-1}(u)$ and compute the
   simulated value for $u = 0.36$.
2. For $X \sim \text{Exponential}(\lambda = 3)$, compute the simulated
   value corresponding to $u = 0.1$.
3. For $X \sim \text{Bernoulli}(p=0.7)$, determine the simulated
   values corresponding to $u = 0.2, 0.5, 0.9$.
4. For $X \sim \text{Geometric}(p=0.3)$, compute the simulated value
   corresponding to $u = 0.4$.
5. Given the discrete support $\{1,2,3\}$ with probabilities
   $0.2, 0.5, 0.3$, compute the simulated value for $u = 0.65$.
6. For the density $f(x) = 4x^3$, $0<x<1$, derive $F_X^{-1}(u)$ and
   compute the value for $u=0.81$.
7. For $X \sim \text{Uniform}(a,b)$, derive the general quantile
   function and simulate a value for $a=2, b=10, u=0.25$.
8. For a two-point distribution with $P(X=0)=0.55$, $P(X=1)=0.45$,
   compute the generated value for $u=0.5$ and $u=0.6$.
9. For $X$ with cdf $F_X(x) = 1 - e^{-x^2}$, $x>0$ (Rayleigh-type),
   derive $F_X^{-1}(u)$ and compute the value for $u=0.5$.
10. For $X \sim \text{Geometric}(p=0.5)$, compute the simulated values
    corresponding to $u = 0.1, 0.3, 0.5, 0.7, 0.9$ and tabulate them.

## Programming Exercises

1. Write an R function `qcubic_inverse(u)` that returns
   $F_X^{-1}(u) = u^{1/3}$ and use it to simulate $n=1000$
   observations from $f(x)=3x^2$.
2. Write an R function to simulate $n$ observations from
   $\text{Exponential}(\lambda)$ using only `runif()` and elementary
   arithmetic, and compare the output to `rexp()` using a Q-Q plot.
3. Implement the discrete inverse transformation algorithm from first
   principles (without `findInterval()`) using a `for` loop, and
   benchmark it against a vectorized `findInterval()` implementation
   using `system.time()`.
4. Write an R function `rgeom_custom(n, p)` implementing the inverse
   transformation algorithm for the Geometric distribution, and verify
   its output against `rgeom(n, p)` using summary statistics.
5. Write an R function that simulates from an arbitrary discrete
   distribution given a vector of support points and a vector of
   probabilities, validating that the probabilities sum to one.
6. Simulate $n=1000$ Bernoulli($p=0.4$) variates using inverse
   transformation and verify the sample proportion using a one-sample
   proportion test (`prop.test()`).
7. Write an R function implementing the quantile function for the
   Uniform$(a,b)$ distribution and use it to simulate a sample,
   confirming the sample range lies within $(a,b)$.
8. Using `ggplot2`, produce a histogram overlaid with the theoretical
   density for a distribution of your choice, generated via inverse
   transformation.
9. Write an R script that simulates the Logarithmic$(\theta)$
   distribution using inverse transformation by constructing and
   searching a cumulative probability vector, extending the vector
   adaptively if the initial length is insufficient.
10. Write an R function that performs a Kolmogorov-Smirnov test
    (`ks.test()`) comparing a simulated sample obtained via inverse
    transformation to the theoretical target distribution, and report
    the p-value.

## Mini Projects

1. **Simulation-based validation of custom generators.** Implement
   inverse transformation generators for three distributions not
   directly available in base R (for example, the Logarithmic
   distribution, a discrete triangular distribution, and a truncated
   Exponential distribution). For each, validate the generator using
   graphical comparison (density or bar plots) and a formal
   goodness-of-fit test, and summarize your findings in a short
   report.

2. **Parametric bootstrap via inverse transformation.** Fit a
   parametric distribution (e.g. Gamma or Weibull) to a real or
   simulated dataset, and implement a parametric bootstrap procedure
   that resamples from the fitted distribution using the inverse
   transformation method rather than `rgamma()` or `rweibull()`
   directly. Compare bootstrap confidence intervals for the mean
   obtained from your custom generator against those obtained from
   the built-in generator, and comment on any discrepancies.

**Difficulty labels.** Multiple choice, fill in the blanks, true or
false: **Easy**. Short answer, numerical problems 1–6: **Medium**.
Long answer, numerical problems 7–10, programming exercises: **Medium
to Advanced**. Mini projects: **Advanced**.



# References

Devroye, L. (1986). *Non-uniform random variate generation*.
Springer-Verlag.

Fishman, G. S. (1996). *Monte Carlo: Concepts, algorithms, and
applications*. Springer-Verlag.

Rizzo, M. L. (2019). *Statistical computing with R* (2nd ed.). CRC
Press.

Ross, S. M. (2020). *Simulation* (6th ed.). Academic Press.

