---
title: "Unit 3.6: Multivariate Random Variable Generation"
author: "Dr. Manoj C. Patil"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: false
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
    code_folding: hide
    theme: cosmo
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    echo = TRUE,
    warning = FALSE,
    message = FALSE,
    fig.align = "center",
    fig.width = 7,
    fig.height = 4.5
)

set.seed(123)
library(tidyverse)
library(ggplot2)
library(dplyr)
library(knitr)
library(kableExtra)
library(gridExtra)
library(patchwork)
```

# Learning Objectives

By the end of this unit, a student should be able to

- explain why generating a random vector generally requires more than
  generating each coordinate separately, and how dependence between
  coordinates must be built into the generating mechanism,
- state and prove the correctness of the sequential conditional
  distribution method for simulating a bivariate or multivariate
  random vector,
- derive the linear (Cholesky-based) representation of a Normal random
  vector with a prescribed covariance matrix, and prove its
  correctness,
- implement both the sequential conditional method and the Cholesky
  method in R, for bivariate and higher-dimensional targets,
- relate the sequential conditional method to Gibbs sampling, and
  understand the distinction between the two,
- evaluate the strengths and limitations of each multivariate
  generation strategy in applied settings.

# Prerequisites

A reader should already be comfortable with

- the inverse transformation, acceptance-rejection, and
  distributional-relationship methods for univariate generation
  (Units 3.3-3.5),
- the definition of a joint density and of a conditional density, and
  the multiplication rule $f(x,y) = f_X(x)\,f_{Y\mid X}(y\mid x)$,
- basic matrix algebra, including symmetric positive definite matrices
  and the notion of a triangular matrix,
- the multivariate Normal distribution and its moment generating
  function, at least at the level of a first course in mathematical
  statistics.

# Introduction

Every method developed so far in this course generates a single
number at a time. Many applications, however, require a *vector* of
numbers that are not independent of one another: the systolic and
diastolic blood pressure of a patient, the returns of several assets
in a portfolio on a given day, or the coordinates of a point at which
several structural loads act simultaneously. Generating each
coordinate on its own, using the univariate methods of Units 3.3-3.5,
would produce a vector with the correct marginal behaviour in each
coordinate but with no dependence between coordinates whatsoever — a
serious flaw whenever the dependence itself is the feature of
scientific interest, as it usually is.

The mathematical resolution is old and elegant. Any joint density can
be factored, by repeated application of the definition of conditional
probability, into a product of a marginal density and a chain of
conditional densities — a fact usually first encountered as the
"multiplication rule" or "chain rule" of probability. This
factorization turns the problem of generating a dependent vector into
a sequence of ordinary univariate generation problems: simulate the
first coordinate from its marginal distribution, then simulate the
second coordinate from its distribution *conditional on* the value
just simulated for the first, and so on. When the target vector is
Gaussian, an even more economical route is available: any Normal
vector with a prescribed covariance matrix can be built directly by
a single linear transformation of independent standard Normal
variables, using the Cholesky decomposition of the covariance matrix
— a technique introduced by André-Louis Cholesky in the early
twentieth century for solving systems of normal equations in geodesy,
and now a standard tool throughout computational statistics.

The practical importance of multivariate generation is difficult to
overstate. Simulating correlated asset returns is the foundation of
portfolio risk analysis and derivative pricing in finance; simulating
correlated component lifetimes is essential in reliability engineering
whenever components share a common operating environment; and
multivariate posterior simulation, whether by direct construction or
by the iterative Gibbs sampling algorithm introduced at the end of
this unit, is a central tool of modern Bayesian statistics. This unit
develops both the general sequential conditional method and its
specialization to the Normal distribution via the Cholesky
decomposition.

# Intuition

Suppose the goal is to simulate a pair of correlated measurements,
such as a student's midterm score and final exam score, where a
higher midterm score tends to be associated with a higher final score.
Simulating the two scores completely independently would fail to
capture this association at all. A natural fix suggests itself:
first simulate the midterm score from its own (marginal) distribution,
observe the value obtained, and then simulate the final score from a
distribution that has been *shifted* to reflect the midterm value just
drawn — a higher midterm value should shift the distribution of the
final score upward. This two-step recipe — draw the first coordinate,
then draw the second coordinate from a distribution that depends on
the first — is exactly the sequential conditional method, and it
generalizes immediately to three, four, or any number of coordinates,
each drawn conditional on all those drawn before it.

For the particular case of Normally distributed measurements, there is
a second, purely linear-algebraic way of seeing the same idea. Two
independent standard Normal quantities, call them $W_1$ and $W_2$,
carry no association with each other. Forming a new pair
$X_1 = W_1$ and $X_2 = \rho\, W_1 + \sqrt{1-\rho^2}\, W_2$ blends a
fraction $\rho$ of the first quantity into the second; the larger
$\rho$ is, the more the two resulting quantities move together. This
blending construction is precisely what the Cholesky decomposition
automates in arbitrary dimension: it finds the "recipe" of linear
blends, one per coordinate, that turns a batch of independent standard
Normal quantities into a Normal vector with any desired covariance
structure.

```{r intuition-figure, echo=FALSE, fig.cap="Independent standard Normal pairs (left) versus a linear blend with rho = 0.6 (right); the blend introduces visible positive association without changing either marginal distribution."}
set.seed(123)
n_demo <- 600
w1 <- rnorm(n_demo)
w2 <- rnorm(n_demo)
rho_demo <- 0.6
x2_blend <- rho_demo * w1 + sqrt(1 - rho_demo^2) * w2

p_independent <- ggplot(tibble(w1 = w1, w2 = w2), aes(x = w1, y = w2)) +
  geom_point(alpha = 0.4, size = 0.8, colour = "steelblue") +
  labs(title = "Independent pair (rho = 0)", x = expression(W[1]), y = expression(W[2])) +
  theme_minimal(base_size = 11)

p_blended <- ggplot(tibble(w1 = w1, x2 = x2_blend), aes(x = w1, y = x2)) +
  geom_point(alpha = 0.4, size = 0.8, colour = "firebrick") +
  labs(title = "Linear blend (rho = 0.6)", x = expression(X[1]), y = expression(X[2])) +
  theme_minimal(base_size = 11)

p_independent + p_blended
```

# Formal Definitions

## Random Vectors and Joint Densities

**Definition.** A **random vector** $\mathbf{X} = (X_1,\dots,X_p)'$ is
a collection of $p$ random variables considered jointly, with joint
density (or joint probability mass function, in the discrete case)
$f(x_1,\dots,x_p)$. A **bivariate distribution** is the special case
$p=2$.

## Conditional Densities and the Chain Rule

**Definition.** For a joint density $f(x_1,\dots,x_p)$, the
**conditional density** of $X_i$ given $X_1=x_1,\dots,X_{i-1}=x_{i-1}$
is

$$
f_i(x_i \mid x_1,\dots,x_{i-1})
= \frac{f(x_1,\dots,x_i)}{f(x_1,\dots,x_{i-1})},
$$

wherever the denominator is positive, with $f(x_1,\dots,x_i)$ denoting
the joint density of the first $i$ coordinates. Repeated application
of this definition gives the **chain rule factorization**

$$
f(x_1,\dots,x_p) = f_1(x_1)\, f_2(x_2\mid x_1)\, f_3(x_3\mid x_1,x_2)
\cdots f_p(x_p \mid x_1,\dots,x_{p-1}),
\tag{6.1}
$$

where $f_1$ is the marginal density of $X_1$ alone.

where

- $f_1(x_1)$ denotes the marginal density of the first coordinate,
- $f_i(x_i \mid x_1,\dots,x_{i-1})$ denotes the density of the $i$-th
  coordinate conditional on the values of all preceding coordinates,
- the ordering of coordinates $1,\dots,p$ is arbitrary; any ordering
  yields a valid factorization of the same joint density, though the
  individual conditional densities differ across orderings.

## The Multivariate Normal Distribution

**Definition.** A random vector $\mathbf{X}=(X_1,\dots,X_p)'$ has the
**multivariate Normal distribution** with mean vector $\boldsymbol\mu$
and covariance matrix $\boldsymbol\Sigma$ (symmetric and positive
definite), written $\mathbf{X}\sim N_p(\boldsymbol\mu,\boldsymbol\Sigma)$,
if its moment generating function is

$$
M_{\mathbf{X}}(\mathbf{t}) = \exp\Bigl(\mathbf{t}'\boldsymbol\mu
+ \tfrac12 \mathbf{t}'\boldsymbol\Sigma\mathbf{t}\Bigr), \qquad
\mathbf{t}\in\mathbb{R}^p.
\tag{6.2}
$$

**Definition (Cholesky decomposition).** For a symmetric positive
definite matrix $\boldsymbol\Sigma$, the **Cholesky decomposition** is
the unique factorization $\boldsymbol\Sigma = \mathbf{L}\mathbf{L}'$,
where $\mathbf{L}$ is lower triangular with strictly positive diagonal
entries.

# Theorems

## Theorem 6.1: Correctness of the Sequential Conditional Method

**Statement.** Let $f(x_1,\dots,x_p)$ admit the chain rule
factorization (6.1). If $X_1$ is drawn from $f_1$, and for
$i=2,\dots,p$, $X_i$ is drawn from $f_i(\cdot \mid X_1,\dots,X_{i-1})$
using the values already drawn for $X_1,\dots,X_{i-1}$, then the
resulting vector $(X_1,\dots,X_p)$ has joint density
$f(x_1,\dots,x_p)$.

**Assumptions.** Each conditional density $f_i(\cdot \mid
x_1,\dots,x_{i-1})$ is a valid density (or pmf) for every value of the
conditioning arguments actually reachable by the construction.

**Proof.** By induction on $p$. For $p=1$, the claim is immediate:
$X_1 \sim f_1$ by construction. Assume the claim holds for $p-1$
coordinates, so that $(X_1,\dots,X_{p-1})$, generated sequentially, has
joint density $f(x_1,\dots,x_{p-1})$. By construction, $X_p$ is drawn,
conditional on $(X_1,\dots,X_{p-1})=(x_1,\dots,x_{p-1})$, from
$f_p(\cdot\mid x_1,\dots,x_{p-1})$; by the definition of a conditional
density, this means the joint density of $(X_1,\dots,X_p)$ satisfies

$$
f(x_1,\dots,x_p) = f(x_1,\dots,x_{p-1})\, f_p(x_p\mid x_1,\dots,x_{p-1}),
$$

which is exactly the chain rule factorization (6.1) for $p$
coordinates. By the induction hypothesis, $f(x_1,\dots,x_{p-1})$ is
correct, so $(X_1,\dots,X_p)$ has joint density $f(x_1,\dots,x_p)$,
completing the induction. $\blacksquare$

**Remarks.** The theorem places no restriction on the ordering of
coordinates: any ordering of $X_1,\dots,X_p$ yields a valid
factorization and a correct algorithm, although the individual
conditional densities required will generally differ across orderings,
and some orderings may make those conditionals far easier (or far
harder) to derive and simulate than others. This method should not be
confused with **Gibbs sampling**, an iterative Markov chain Monte
Carlo technique that repeatedly cycles through the *full* conditional
distributions $f(x_i \mid x_1,\dots,x_{i-1},x_{i+1},\dots,x_p)$
(conditioning on *all other* coordinates, not only the preceding
ones) and converges only asymptotically to the joint distribution.
Theorem 6.1 describes a single-pass, exact method that requires the
successive conditionals of the chain rule; Gibbs sampling is used
precisely in settings where those successive conditionals are
unavailable in closed form but the full conditionals are.

**Examples.** Full worked examples are given in the Numerical Examples
section below, for both a continuous and a discrete bivariate target.

**Applications.** The sequential conditional method underlies
simulation from graphical models and Bayesian networks, in which the
joint distribution is specified natively as a product of conditional
distributions along a directed graph, and is the exact-sampling
ancestor of the approximate Gibbs sampling algorithm used throughout
Bayesian computation.

## Theorem 6.2: The Cholesky Construction of a Multivariate Normal Vector

**Statement.** Let $\mathbf{Z}=(Z_1,\dots,Z_p)'$ have independent
standard Normal components, so $\mathbf{Z}\sim N_p(\mathbf{0},
\mathbf{I}_p)$. Let $\boldsymbol\Sigma$ be symmetric positive
definite with Cholesky decomposition $\boldsymbol\Sigma =
\mathbf{L}\mathbf{L}'$. Then

$$
\mathbf{X} = \boldsymbol\mu + \mathbf{L}\mathbf{Z} \;\sim\;
N_p(\boldsymbol\mu, \boldsymbol\Sigma).
\tag{6.3}
$$

**Assumptions.** $\boldsymbol\Sigma$ is symmetric positive definite,
guaranteeing that the Cholesky decomposition exists and is unique with
a strictly positive diagonal.

**Proof.** Since $\mathbf{Z}\sim N_p(\mathbf{0},\mathbf{I}_p)$, its
moment generating function is
$M_{\mathbf{Z}}(\mathbf{t}) = \exp(\mathbf{t}'\mathbf{t}/2)$ (the
product of $p$ independent standard Normal moment generating
functions). For $\mathbf{X}=\boldsymbol\mu+\mathbf{L}\mathbf{Z}$,

$$
\begin{aligned}
M_{\mathbf{X}}(\mathbf{t}) &= E\bigl[\exp(\mathbf{t}'\mathbf{X})\bigr]
= E\bigl[\exp\bigl(\mathbf{t}'\boldsymbol\mu + \mathbf{t}'\mathbf{L}\mathbf{Z}\bigr)\bigr] \\
&= \exp(\mathbf{t}'\boldsymbol\mu)\, E\bigl[\exp\bigl((\mathbf{L}'\mathbf{t})'\mathbf{Z}\bigr)\bigr]
= \exp(\mathbf{t}'\boldsymbol\mu)\, M_{\mathbf{Z}}(\mathbf{L}'\mathbf{t}) \\
&= \exp(\mathbf{t}'\boldsymbol\mu)\, \exp\Bigl(\tfrac12 (\mathbf{L}'\mathbf{t})'(\mathbf{L}'\mathbf{t})\Bigr)
= \exp\Bigl(\mathbf{t}'\boldsymbol\mu + \tfrac12 \mathbf{t}'\mathbf{L}\mathbf{L}'\mathbf{t}\Bigr) \\
&= \exp\Bigl(\mathbf{t}'\boldsymbol\mu + \tfrac12 \mathbf{t}'\boldsymbol\Sigma\mathbf{t}\Bigr),
\end{aligned}
$$

using $\mathbf{L}\mathbf{L}' = \boldsymbol\Sigma$ in the last step.
This is exactly the moment generating function (6.2) of
$N_p(\boldsymbol\mu,\boldsymbol\Sigma)$, and since the moment
generating function determines the distribution uniquely,
$\mathbf{X}\sim N_p(\boldsymbol\mu,\boldsymbol\Sigma)$. $\blacksquare$

**Remarks.** The bivariate blending construction of the Intuition
section, $X_1=W_1$, $X_2=\rho W_1+\sqrt{1-\rho^2}W_2$, is precisely
the $p=2$ special case of (6.3) with correlation matrix

$$
\boldsymbol\Sigma = \begin{pmatrix}1&\rho\\\rho&1\end{pmatrix}
\qquad\text{and Cholesky factor}\qquad
\mathbf{L}=\begin{pmatrix}1&0\\\rho&\sqrt{1-\rho^2}\end{pmatrix},
$$

as verified directly by matrix multiplication:

$$
\mathbf{L}\mathbf{L}' = \begin{pmatrix}1&\rho\\\rho&\rho^2+1-\rho^2\end{pmatrix}
= \begin{pmatrix}1&\rho\\\rho&1\end{pmatrix}.
$$

**Examples.** Worked numerically, for both $p=2$ and $p=3$, in the
Numerical Examples section.

**Applications.** The Cholesky construction is the standard technique
for simulating correlated asset returns in quantitative finance,
correlated random effects in multilevel models, and correlated
structural loads in engineering reliability analysis.

# Mathematical Derivations

## Equivalence of the Conditional Method and the Cholesky Method for the Normal Distribution

For a Normal vector, the general chain-rule factorization of Theorem
6.1 and the Cholesky construction of Theorem 6.2 are not merely two
unrelated techniques; they compute exactly the same generating recipe.
The rows of the Cholesky factor $\mathbf{L}$ are precisely the
coefficients of the successive conditional-mean regressions implied
by (6.1): writing $\mathbf{X}=\boldsymbol\mu+\mathbf{L}\mathbf{Z}$
coordinate by coordinate,

$$
X_i = \mu_i + \sum_{j=1}^{i} L_{ij} Z_j
= \Bigl(\mu_i + \sum_{j=1}^{i-1} L_{ij}Z_j\Bigr) + L_{ii}Z_i,
\qquad i=1,\dots,p,
$$

shows that, conditional on $X_1,\dots,X_{i-1}$ (equivalently, on
$Z_1,\dots,Z_{i-1}$, since the map between them is triangular and
invertible), $X_i$ is Normal with mean equal to the linear regression
of $X_i$ on $X_1,\dots,X_{i-1}$ and variance $L_{ii}^2$. This is
exactly the conditional density $f_i(x_i\mid x_1,\dots,x_{i-1})$
required by (6.1), specialized to the Gaussian case. In other words,
for the Normal distribution, sequentially applying Theorem 6.1 with
Normal conditional distributions and applying Theorem 6.2 via the
Cholesky decomposition are two descriptions of the identical
algorithm; the Cholesky decomposition is simply the closed-form,
matrix-algebraic solution of the successive linear regressions that
the conditional method would otherwise require deriving by hand, one
coordinate at a time (Pourahmadi, 1999).

# Algorithms

## Algorithm 6.1: General Sequential Conditional Distribution Method

**Purpose.** Generate a random vector $(X_1,\dots,X_p)$ with a
specified joint distribution, given the marginal density of $X_1$ and
the successive conditional densities implied by the chain rule (6.1).

**Input.** Marginal sampler for $X_1$; conditional samplers for
$X_i \mid X_1,\dots,X_{i-1}$, $i=2,\dots,p$; required sample size $n$.

**Output.** A matrix of $n$ simulated vectors, each of dimension $p$.

**Step-by-step procedure.**

1. For each of $n$ replications:
   (a) draw $x_1$ from the marginal density of $X_1$;
   (b) for $i=2,\dots,p$, draw $x_i$ from the conditional density of
       $X_i$ given the values $x_1,\dots,x_{i-1}$ already drawn;
2. Return the collection of $n$ vectors $(x_1,\dots,x_p)$.

**Pseudo-code.**

```
Algorithm: Sequential Conditional Method
Input:  marginal sampler for X1, conditional samplers for X2..Xp, n
Output: n x p matrix of simulated vectors

1. For k = 1 to n:
       x[1] <- draw from marginal of X1
       For i = 2 to p:
           x[i] <- draw from conditional of Xi given x[1..i-1]
       Store x as row k of output
2. Return output
```

**Flowchart description.** Start; draw the first coordinate from its
marginal distribution; for each remaining coordinate, in order, draw
from its distribution conditional on all coordinates already drawn;
after all $p$ coordinates are drawn, record the completed vector;
repeat for $n$ replications; end.

**Complexity.** $O(np)$ time overall, provided each individual
conditional draw costs $O(1)$; the memory required to store the output
is $O(np)$.

**Practical considerations.** The choice of ordering among the $p$
coordinates does not affect correctness (Theorem 6.1) but can greatly
affect how tractable the required conditional densities are; an
ordering that produces simple, standard conditional distributions
(such as the Normal-Normal structure of Section 7) should always be
preferred when one is available.

## Algorithm 6.2: Bivariate Normal Generation via Conditional Distributions

**Purpose.** Generate pairs $(X_1,X_2)$ from the standard bivariate
Normal distribution with correlation $\rho$, using the sequential
conditional method specialized to the Normal-Normal case.

**Input.** Correlation $\rho \in (-1,1)$; sample size $n$.

**Output.** An $n\times 2$ matrix of simulated pairs.

**Step-by-step procedure.**

1. For $k=1,\dots,n$: draw $w_{1} \sim N(0,1)$ and $w_2 \sim N(0,1)$
   independently; set $x_1 = w_1$ and
   $x_2 = \rho\, w_1 + \sqrt{1-\rho^2}\, w_2$.
2. Return the $n$ pairs $(x_1,x_2)$.

**Pseudo-code.**

```
Algorithm: Bivariate Normal, Conditional Method
Input:  rho, n
Output: n x 2 matrix

1. For k = 1 to n:
       w1 <- draw from N(0,1)
       w2 <- draw from N(0,1)
       x1 <- w1
       x2 <- rho * w1 + sqrt(1 - rho^2) * w2
2. Return matrix of (x1, x2) pairs
```

**Flowchart description.** Start; draw two independent standard Normal
variates; report the first unchanged as $X_1$; blend the two variates
using the correlation coefficient to obtain $X_2$; repeat for $n$
replications; end.

**Complexity.** $O(n)$, requiring two Normal draws per replication.

**Practical considerations.** The formula requires $\rho\in(-1,1)$
strictly; at $\rho=\pm1$ the conditional variance $1-\rho^2$ vanishes
and $X_2$ becomes a deterministic (degenerate) function of $X_1$,
which is mathematically valid but numerically edge-case behaviour
that should be handled explicitly if $|\rho|=1$ is a possibility in
an application.

## Algorithm 6.3: Multivariate Normal Generation via Cholesky Decomposition

**Purpose.** Generate vectors from $N_p(\boldsymbol\mu,
\boldsymbol\Sigma)$ for general dimension $p$, using Theorem 6.2.

**Input.** Mean vector $\boldsymbol\mu$; symmetric positive definite
covariance matrix $\boldsymbol\Sigma$; sample size $n$.

**Output.** An $n \times p$ matrix of simulated vectors.

**Step-by-step procedure.**

1. Compute the Cholesky factor $\mathbf{L}$ satisfying
   $\boldsymbol\Sigma = \mathbf{L}\mathbf{L}'$.
2. For $k=1,\dots,n$: draw $\mathbf{z} \sim N_p(\mathbf{0},
   \mathbf{I}_p)$ (i.e. $p$ independent standard Normal draws); set
   $\mathbf{x} = \boldsymbol\mu + \mathbf{L}\mathbf{z}$.
3. Return the $n$ vectors $\mathbf{x}$.

**Pseudo-code.**

```
Algorithm: Multivariate Normal, Cholesky Method
Input:  mu (length p), Sigma (p x p), n
Output: n x p matrix

1. L <- Cholesky factor of Sigma, with Sigma = L L'
2. For k = 1 to n:
       z <- p independent draws from N(0,1)
       x <- mu + L %*% z
3. Return matrix of x vectors
```

**Flowchart description.** Start; compute the Cholesky factor of the
target covariance matrix once; for each replication, draw a vector of
independent standard Normal variates and apply the fixed linear
transformation $\boldsymbol\mu+\mathbf{L}\mathbf{z}$; repeat for $n$
replications; end.

**Complexity.** The one-time Cholesky decomposition costs
$O(p^3)$; each subsequent replication costs $O(p^2)$ for the
matrix-vector multiplication, so generating $n$ vectors costs
$O(p^3 + np^2)$ overall.

**Practical considerations.** $\boldsymbol\Sigma$ must be symmetric
positive definite for the Cholesky decomposition to exist; in R, the
built-in `chol()` function raises an informative error if the supplied
matrix is not positive definite, which is a useful validation step
before relying on a covariance matrix assembled by hand from
individually plausible pairwise correlations that may not be jointly
consistent.

# R Programming

```{r generator-functions}
set.seed(123)

#' Generate bivariate Normal pairs via the conditional distribution method
#'
#' Purpose: implements Algorithm 6.2, blending two independent standard
#' Normal draws to obtain a correlated pair.
#'
#' Arguments:
#'   sample_size   number of pairs to generate
#'   correlation   the target correlation rho, in (-1, 1)
#'
#' Returns:
#'   a matrix with sample_size rows and 2 columns
#'
#' Example:
#'   generate_bivariate_normal(sample_size = 5, correlation = 0.6)
generate_bivariate_normal <- function(sample_size, correlation) {
  first_component <- rnorm(sample_size)
  independent_component <- rnorm(sample_size)

  second_component <- correlation * first_component +
    sqrt(1 - correlation^2) * independent_component

  cbind(x1 = first_component, x2 = second_component)
}

#' Generate multivariate Normal vectors via the Cholesky decomposition
#'
#' Purpose: implements Algorithm 6.3 for a general covariance matrix.
#'
#' Arguments:
#'   sample_size    number of vectors to generate
#'   mean_vector    numeric vector of length p
#'   covariance_matrix  p x p symmetric positive definite matrix
#'
#' Returns:
#'   a matrix with sample_size rows and length(mean_vector) columns
#'
#' Example:
#'   generate_multivariate_normal(
#'     sample_size = 5,
#'     mean_vector = c(0, 0, 0),
#'     covariance_matrix = diag(3)
#'   )
generate_multivariate_normal <- function(sample_size, mean_vector, covariance_matrix) {
  dimension <- length(mean_vector)

  # R's chol() returns the upper-triangular factor U with t(U) %*% U = Sigma;
  # the lower-triangular L used in Theorem 6.2 is L = t(U)
  cholesky_lower <- t(chol(covariance_matrix))

  standard_normal_matrix <- matrix(
    rnorm(sample_size * dimension), nrow = dimension, ncol = sample_size
  )

  transformed <- mean_vector + cholesky_lower %*% standard_normal_matrix

  t(transformed)
}

#' Generate a bivariate (Exponential, conditional Uniform) pair
#'
#' Purpose: demonstrates Algorithm 6.1 for a non-Gaussian bivariate
#' target, X ~ Exponential(rate), Y | X = x ~ Uniform(0, x).
#'
#' Arguments:
#'   sample_size   number of pairs to generate
#'   rate          rate parameter of the Exponential marginal
#'
#' Returns:
#'   a matrix with sample_size rows and 2 columns
#'
#' Example:
#'   generate_exponential_uniform_pair(sample_size = 5, rate = 1)
generate_exponential_uniform_pair <- function(sample_size, rate) {
  x_values <- rexp(sample_size, rate = rate)
  y_values <- runif(sample_size, min = 0, max = x_values)

  cbind(x = x_values, y = y_values)
}
```

# Simulation Studies

## Bivariate Normal via the Conditional Method

```{r bivariate-normal-simulation}
set.seed(123)

bivariate_sample <- generate_bivariate_normal(sample_size = 20000, correlation = 0.6)

tibble(
  Quantity = c("Target correlation", "Empirical correlation",
               "Empirical mean X1", "Empirical mean X2",
               "Empirical variance X1", "Empirical variance X2"),
  Value = c(
    0.6, cor(bivariate_sample[, "x1"], bivariate_sample[, "x2"]),
    mean(bivariate_sample[, "x1"]), mean(bivariate_sample[, "x2"]),
    var(bivariate_sample[, "x1"]), var(bivariate_sample[, "x2"])
  )
) %>%
  kable(digits = 4, caption = "Bivariate Normal via conditional method: theory vs simulation")
```

## Trivariate Normal via the Cholesky Method

```{r trivariate-normal-simulation}
set.seed(123)

target_mean <- c(0, 0, 0)
target_covariance <- matrix(
  c(1.0, 0.5, 0.3,
    0.5, 1.0, 0.4,
    0.3, 0.4, 1.0),
  nrow = 3, byrow = TRUE
)

trivariate_sample <- generate_multivariate_normal(
  sample_size = 20000, mean_vector = target_mean, covariance_matrix = target_covariance
)

empirical_covariance <- cov(trivariate_sample)

kable(target_covariance, digits = 3, caption = "Target covariance matrix")
kable(empirical_covariance, digits = 3, caption = "Empirical covariance matrix from 20000 simulated vectors")
```

## Non-Gaussian Bivariate Target: Exponential-Conditional-Uniform

```{r exp-uniform-simulation}
set.seed(123)

exp_uniform_sample <- generate_exponential_uniform_pair(sample_size = 20000, rate = 1)

ks_marginal_x <- ks.test(exp_uniform_sample[, "x"], "pexp", rate = 1)

tibble(
  Quantity = c("KS test p-value, X marginal vs Exponential(1)",
               "Proportion of pairs with Y <= X (should be 1)"),
  Value = c(ks_marginal_x$p.value, mean(exp_uniform_sample[, "y"] <= exp_uniform_sample[, "x"]))
) %>%
  kable(digits = 4, caption = "Exponential-Uniform bivariate construction: diagnostics")
```

# Graphs

```{r bivariate-normal-plot, fig.cap="Simulated bivariate Normal pairs with rho = 0.6, overlaid with theoretical 50%, 90%, and 99% density contours."}
set.seed(123)

grid_vals <- seq(-4, 4, length.out = 120)
grid_df <- expand.grid(x1 = grid_vals, x2 = grid_vals)

bivariate_normal_density <- function(x1, x2, rho) {
  z <- (x1^2 - 2 * rho * x1 * x2 + x2^2) / (1 - rho^2)
  (1 / (2 * pi * sqrt(1 - rho^2))) * exp(-z / 2)
}

grid_df$density <- bivariate_normal_density(grid_df$x1, grid_df$x2, rho = 0.6)

ggplot() +
  geom_point(
    data = as_tibble(bivariate_sample), aes(x = x1, y = x2),
    alpha = 0.15, size = 0.6, colour = "steelblue"
  ) +
  geom_contour(
    data = grid_df, aes(x = x1, y = x2, z = density),
    colour = "darkred", breaks = c(0.01, 0.05, 0.1)
  ) +
  labs(
    title = "Simulated bivariate Normal (rho = 0.6) with theoretical density contours",
    x = expression(X[1]), y = expression(X[2])
  ) +
  theme_minimal(base_size = 11)
```

```{r trivariate-pairwise-plots, fig.cap="Pairwise scatterplots of the simulated trivariate Normal sample, illustrating the three pairwise correlations of the target covariance matrix (0.5, 0.3, 0.4)."}
trivariate_df <- as_tibble(trivariate_sample, .name_repair = "unique") %>%
  rename(x1 = 1, x2 = 2, x3 = 3)

p_12 <- ggplot(trivariate_df, aes(x = x1, y = x2)) +
  geom_point(alpha = 0.15, size = 0.5, colour = "steelblue") +
  labs(title = "X1 vs X2 (rho = 0.5)") +
  theme_minimal(base_size = 10)

p_13 <- ggplot(trivariate_df, aes(x = x1, y = x3)) +
  geom_point(alpha = 0.15, size = 0.5, colour = "darkgreen") +
  labs(title = "X1 vs X3 (rho = 0.3)") +
  theme_minimal(base_size = 10)

p_23 <- ggplot(trivariate_df, aes(x = x2, y = x3)) +
  geom_point(alpha = 0.15, size = 0.5, colour = "purple") +
  labs(title = "X2 vs X3 (rho = 0.4)") +
  theme_minimal(base_size = 10)

p_12 + p_13 + p_23
```

```{r exp-uniform-plot, fig.cap="Simulated (X, Y) pairs from X ~ Exponential(1), Y | X ~ Uniform(0, X); every point lies below the line y = x by construction."}
ggplot(as_tibble(exp_uniform_sample), aes(x = x, y = y)) +
  geom_point(alpha = 0.15, size = 0.5, colour = "darkorange") +
  geom_abline(slope = 1, intercept = 0, colour = "black", linetype = "dashed") +
  coord_cartesian(xlim = c(0, 5), ylim = c(0, 5)) +
  labs(title = "Exponential(1) with conditional Uniform(0, X)", x = "X", y = "Y") +
  theme_minimal(base_size = 11)
```

# Numerical Examples

**Example 1: Bivariate Normal via the conditional method.**

*Problem.* With $\rho=0.6$, and draws $w_1=0.5$, $w_2=0.5$, compute
the resulting pair $(x_1,x_2)$.

*Solution.* $x_1 = w_1 = 0.5$, and

$$
x_2 = \rho w_1 + \sqrt{1-\rho^2}\, w_2
= 0.6(0.5) + \sqrt{1-0.36}(0.5)
= 0.3 + (0.8)(0.5) = 0.3+0.4 = 0.7.
$$

*Interpretation.* The pair $(0.5, 0.7)$ shows the expected positive
association: because $\rho=0.6>0$, a positive value of $x_1$ pulls the
generated $x_2$ toward positive values as well, though the independent
component $w_2$ still contributes its own variability.

**Example 2: Trivariate Normal via the Cholesky method.**

*Problem.* Using the target covariance matrix of Section 10.2, with
Cholesky factor $\mathbf{L}$ (computed in R above) and standard Normal
draws $\mathbf{z}=(0.2,-0.1,0.5)'$, compute $\mathbf{x}=\mathbf{L}\mathbf{z}$
(the mean vector is $\mathbf{0}$).

*Solution.* Row by row,

$$
\begin{aligned}
x_1 &= L_{11} z_1 = 1.000(0.2) = 0.2000, \\
x_2 &= L_{21}z_1 + L_{22}z_2 = 0.500(0.2) + 0.866(-0.1) = 0.100 - 0.0866 = 0.0134, \\
x_3 &= L_{31}z_1 + L_{32}z_2 + L_{33}z_3
= 0.300(0.2) + 0.289(-0.1) + 0.909(0.5) \\
&= 0.060 - 0.0289 + 0.4546 = 0.4857.
\end{aligned}
$$

*Interpretation.* The lower-triangular structure of $\mathbf{L}$ means
each coordinate is built from the standard Normal draws up to and
including its own index, exactly mirroring the sequential conditional
construction of Section 7: $X_1$ depends only on $Z_1$, $X_2$ depends
on $Z_1$ and $Z_2$, and $X_3$ depends on all three.

**Example 3: The Exponential-conditional-Uniform bivariate
construction.**

*Problem.* With rate $1$ for the Exponential marginal, and draws
$u_1=0.3935$ (for $X$) and $u_2=0.4$ (for $Y\mid X$), compute the
resulting pair.

*Solution.* By inverse transformation,
$x = -\ln(1-u_1) = -\ln(0.6065) = 0.5000$. Given $x=0.5$, $Y\sim
\text{Uniform}(0, 0.5)$, so $y = u_2 \cdot x = 0.4 \times 0.5 = 0.2000$.

*Interpretation.* By construction, $Y$ can never exceed $X$, which is
exactly the triangular region visible in the scatterplot above; this
bivariate structure arises naturally, for instance, as the joint
distribution of a component's total lifetime ($X$) and the time of a
uniformly-timed random inspection that occurs sometime during that
lifetime ($Y$).

**Example 4: A discrete conditional construction (Poisson thinning).**

*Problem.* Given a realized count $X=4$, and independent Uniform draws
$u=(0.3, 0.7, 0.2, 0.9)$ used to construct
$Y \mid X=4 \sim \text{Binomial}(4, 0.5)$ as a sum of four
Bernoulli$(0.5)$ trials (success if the corresponding $u_i \le 0.5$),
determine $Y$.

*Solution.* Successes occur for $u_1=0.3$ and $u_3=0.2$ (both
$\le 0.5$), and failures for $u_2=0.7$ and $u_4=0.9$. Hence
$Y = 1+0+1+0 = 2$.

*Interpretation.* This construction, in which each of $X$ "events" is
independently retained with probability $p$, is known as **Poisson
thinning** when $X$ itself is Poisson; it is used to model, for
example, the number of defective items among a random number of items
produced, and is a discrete illustration of exactly the same
sequential conditional principle used for the continuous examples
above.

# Applications

- **Statistics.** Multivariate analysis of variance, multivariate
  regression diagnostics, and copula-based dependence modelling all
  require simulating correlated random vectors for validation and
  power studies.
- **Machine Learning.** Generating correlated synthetic features for
  benchmarking algorithms, and simulating from Gaussian graphical
  models, both rely directly on the methods of this unit.
- **Data Science.** Producing realistic synthetic multivariate
  datasets that preserve a specified correlation structure, for
  privacy-preserving data sharing or for stress-testing analytics
  pipelines.
- **Finance.** Simulating correlated asset returns via the Cholesky
  method is the standard first step in portfolio value-at-risk
  estimation and in Monte Carlo pricing of multi-asset derivatives.
- **Reliability Engineering.** Simulating dependent component
  lifetimes, when components share a common operating environment or
  a common manufacturing batch, requires exactly the multivariate
  techniques of this unit rather than independent univariate
  generation.
- **Monte Carlo Methods.** Variance reduction techniques such as
  antithetic and common random number methods often require
  deliberately correlated, rather than independent, simulated vectors.
- **Bayesian Statistics.** Direct simulation from a multivariate
  posterior via the sequential conditional method, when the necessary
  conditionals are tractable, and Gibbs sampling (Remarks to Theorem
  6.1), when only full conditionals are available, are both
  applications of the chain-rule idea developed in this unit.
- **Operations Research.** Simulating correlated demand across
  multiple products or locations in inventory and supply chain
  models.
- **Engineering.** Simulating correlated structural loads (e.g. wind
  and seismic loads acting jointly on a structure) for reliability-
  based design.

# Advantages

- The sequential conditional method is completely general: it applies
  to any joint distribution for which the required conditional
  densities can be derived and simulated, continuous, discrete, or
  mixed.
- The Cholesky method is exact, numerically stable, and computationally
  cheap relative to the dimension of the problem, making it the
  standard tool for Gaussian (and Gaussian-copula) multivariate
  simulation.
- Both methods are exact, in the same sense as the univariate methods
  of earlier units: no asymptotic or iterative approximation is
  involved.

# Limitations

- The sequential conditional method requires closed-form (or at least
  efficiently simulable) conditional distributions at every stage;
  for many multivariate targets, deriving these conditionals is itself
  a difficult problem, which is precisely the situation that motivates
  Gibbs sampling and other Markov chain Monte Carlo methods.
- The Cholesky method, as presented, is specific to the (multivariate)
  Normal distribution; constructing a non-Normal random vector with
  prescribed marginals and a prescribed correlation structure requires
  additional machinery, such as copula methods, that lie beyond the
  scope of this unit.
- The one-time cost of the Cholesky decomposition itself is $O(p^3)$,
  which can become a computational bottleneck for very high-dimensional
  covariance matrices, such as those arising in spatial statistics
  with thousands of locations.

# Practical Tips

- **Always validate that a covariance matrix is symmetric positive
  definite before decomposing it.** A matrix built by hand from
  individually plausible pairwise correlations can fail to be jointly
  consistent (positive definite); R's `chol()` raises an error in this
  case, which should be treated as a genuine modelling warning rather
  than worked around.
- **Remember that ordering does not affect correctness, but does
  affect tractability, in the sequential conditional method.** If the
  natural ordering of coordinates leads to an intractable conditional
  distribution, consider whether a different ordering yields simpler
  conditionals for the same joint target.
- **Use `chol()`'s return convention correctly.** R's `chol()` returns
  an *upper* triangular matrix $\mathbf{U}$ with
  $\mathbf{U}'\mathbf{U}=\boldsymbol\Sigma$; the lower-triangular
  $\mathbf{L}$ of Theorem 6.2 is obtained as `t(chol(Sigma))`, and
  confusing the two conventions is a common source of silent
  transposition errors.
- **Do not confuse the sequential conditional method with Gibbs
  sampling.** The former is a single-pass exact algorithm; the latter
  is an iterative, only asymptotically exact Markov chain algorithm,
  used precisely when the former's requirements cannot be met.
- **Vectorize the Cholesky-based generator.** As implemented in
  `generate_multivariate_normal()`, generating the entire matrix of
  standard Normal draws at once and applying a single matrix
  multiplication is far faster in R than looping over replications.

# Summary

- Generating a random vector generally requires an explicit mechanism
  for dependence between coordinates; generating each coordinate
  independently destroys any dependence structure entirely.
- The chain rule factors any joint density into a marginal density and
  a sequence of conditional densities; Theorem 6.1 proves that drawing
  coordinates sequentially from this factorization reproduces the
  exact joint distribution, for any valid ordering of coordinates.
- For the multivariate Normal distribution, the Cholesky decomposition
  of the covariance matrix gives a direct, single-step linear
  transformation of independent standard Normal variables that
  reproduces the exact target distribution (Theorem 6.2), and this
  construction is shown to coincide exactly with the sequential
  conditional method specialized to Normal conditionals.
- The sequential conditional method should be distinguished carefully
  from Gibbs sampling, an iterative approximate technique used when
  the successive conditionals required by the chain rule are not
  available in closed form.

# Key Formulae

- Chain rule factorization:
  $f(x_1,\dots,x_p) = f_1(x_1)\prod_{i=2}^p f_i(x_i\mid x_1,\dots,x_{i-1})$.
- Multivariate Normal moment generating function:
  $M_{\mathbf{X}}(\mathbf{t}) = \exp(\mathbf{t}'\boldsymbol\mu +
  \tfrac12\mathbf{t}'\boldsymbol\Sigma\mathbf{t})$.
- Cholesky construction: $\boldsymbol\Sigma=\mathbf{L}\mathbf{L}'$,
  $\mathbf{X}=\boldsymbol\mu+\mathbf{L}\mathbf{Z}$,
  $\mathbf{Z}\sim N_p(\mathbf{0},\mathbf{I}_p)$.
- Bivariate Normal blend ($p=2$):
  $X_1=W_1$, $X_2=\rho W_1+\sqrt{1-\rho^2}W_2$.

# Key Terms

- **Bivariate distribution.** The joint distribution of two random
  variables considered together.
- **Chain rule (multiplication rule).** The factorization of a joint
  density into a marginal density and a sequence of conditional
  densities.
- **Cholesky decomposition.** The factorization
  $\boldsymbol\Sigma=\mathbf{L}\mathbf{L}'$ of a symmetric positive
  definite matrix into a lower triangular factor and its transpose.
- **Conditional density.** The density of one random variable given
  the observed values of others.
- **Gibbs sampling.** An iterative Markov chain Monte Carlo method that
  cycles through full conditional distributions; distinguished from
  the exact, single-pass sequential conditional method of this unit.
- **Multivariate Normal distribution.** The joint distribution of a
  random vector whose moment generating function has the exponential
  quadratic form (6.2).
- **Random vector.** A collection of random variables considered
  jointly, with a specified joint distribution.
- **Sequential conditional method.** Generating a random vector one
  coordinate at a time, each conditional on the coordinates already
  generated, following the chain rule factorization.

# Exercises

## Multiple Choice Questions

1. The chain rule factorizes a joint density into
   (a) a product of independent marginal densities
   (b) a marginal density and a sequence of conditional densities
   (c) a sum of conditional densities
   (d) a ratio of two marginal densities

2. According to Theorem 6.1, the ordering of coordinates in the
   sequential conditional method
   (a) must always be the natural numerical order
   (b) does not affect correctness, though it may affect tractability
   (c) must be chosen to minimize variance
   (d) is irrelevant to both correctness and tractability

3. The Cholesky decomposition factors a covariance matrix as
   (a) $\boldsymbol\Sigma = \mathbf{L} + \mathbf{L}'$
   (b) $\boldsymbol\Sigma = \mathbf{L}\mathbf{L}'$
   (c) $\boldsymbol\Sigma = \mathbf{L}^{-1}$
   (d) $\boldsymbol\Sigma = \mathbf{L}\mathbf{L}$

4. In the bivariate Normal blend $X_2 = \rho W_1 + \sqrt{1-\rho^2}W_2$,
   the role of $\sqrt{1-\rho^2}$ is to
   (a) shift the mean of $X_2$
   (b) ensure $X_2$ has variance $1$
   (c) ensure $X_1$ and $X_2$ are independent
   (d) ensure $\rho$ lies in $(-1,1)$

5. Gibbs sampling differs from the sequential conditional method of
   this unit primarily because Gibbs sampling
   (a) is exact after a single pass
   (b) uses full conditionals and iterates, converging only
       asymptotically
   (c) does not require any conditional distributions
   (d) only applies to discrete distributions

6. The one-time cost of the Cholesky decomposition of a $p\times p$
   covariance matrix is
   (a) $O(p)$
   (b) $O(p^2)$
   (c) $O(p^3)$
   (d) $O(\log p)$

7. R's `chol()` function returns
   (a) the lower triangular factor $\mathbf{L}$ directly
   (b) the upper triangular factor $\mathbf{U}$ with $\mathbf{U}'\mathbf{U}=\boldsymbol\Sigma$
   (c) the inverse of the covariance matrix
   (d) the eigenvalues of the covariance matrix

8. In the Exponential-conditional-Uniform construction of this unit,
   the generated pair $(X,Y)$ always satisfies
   (a) $Y \ge X$
   (b) $Y \le X$
   (c) $Y = X$
   (d) no fixed relationship

9. A necessary condition for the Cholesky decomposition of
   $\boldsymbol\Sigma$ to exist is that $\boldsymbol\Sigma$ be
   (a) diagonal
   (b) symmetric positive definite
   (c) upper triangular
   (d) a correlation matrix specifically

10. Poisson thinning, as illustrated in this unit, is an example of
    (a) the Cholesky method
    (b) a discrete sequential conditional construction
    (c) the inverse transformation method applied to a continuous
        distribution
    (d) acceptance-rejection

**Answers.** 1(b), 2(b), 3(b), 4(b), 5(b), 6(c), 7(b), 8(b), 9(b),
10(b).

## Short Answer Questions

1. State the chain rule factorization of a joint density for $p=3$
   coordinates.
2. Explain why generating each coordinate of a random vector
   independently generally fails to reproduce a specified joint
   distribution.
3. Explain, in your own words, why the ordering of coordinates does
   not affect the correctness of the sequential conditional method.
4. State the bivariate Normal blending formula and identify the roles
   of $\rho$ and $\sqrt{1-\rho^2}$.
5. Explain what it means for a matrix to be symmetric positive
   definite, and why this property is required for the Cholesky
   decomposition to exist.
6. Distinguish the sequential conditional method from Gibbs sampling
   in one or two sentences.
7. Why must R's `chol()` output be transposed before it can be used
   directly as the $\mathbf{L}$ of Theorem 6.2?
8. Explain, using Example 3, why $Y \le X$ always holds in the
   Exponential-conditional-Uniform construction.
9. Give one application area, distinct from those emphasized in this
   unit, where correlated random vector simulation is essential.
10. Explain why the computational cost of the Cholesky method has two
    distinct components with different growth rates in $p$ and in $n$.

## Long Answer Questions

1. State and prove Theorem 6.1 (correctness of the sequential
   conditional method) by induction, and explain why the base case and
   inductive step together establish the result for general $p$.
2. State and prove Theorem 6.2 (the Cholesky construction of the
   multivariate Normal), using the moment generating function
   argument in full.
3. Derive the equivalence between the sequential conditional method
   and the Cholesky method for the Normal distribution, as presented
   in the Mathematical Derivations section, and explain the role of
   the triangular structure of $\mathbf{L}$.
4. Derive the bivariate Normal blending construction
   $X_2=\rho W_1+\sqrt{1-\rho^2}W_2$ from first principles, verifying
   both the variance and the correlation of the resulting pair.
5. Compare the sequential conditional method and the Cholesky method
   in terms of generality, computational cost, and the class of
   distributions each can generate.
6. Discuss the distinction between the sequential conditional method
   and Gibbs sampling in detail, including the type of conditional
   distribution each requires and the exactness (or lack thereof) of
   each method.
7. Explain, with reference to a concrete example, why a covariance
   matrix assembled from individually valid pairwise correlations can
   fail to be positive definite, and describe how this failure would
   be detected in practice.
8. Discuss at least two applications of multivariate random variable
   generation in finance or reliability engineering, explaining
   specifically how correlation between coordinates affects the
   quantity of ultimate interest in each application.
9. Explain how the Exponential-conditional-Uniform construction of
   this unit could be adapted to model the joint distribution of a
   component's lifetime and the time of a randomly timed inspection,
   and discuss what modification would be needed if the inspection
   time were not uniformly distributed over the lifetime.
10. Discuss the computational and numerical considerations that arise
    when extending the Cholesky method to high-dimensional covariance
    matrices, and describe at least one practical mitigation.

## Programming Exercises

1. Implement `generate_bivariate_normal()` from first principles
   (already provided) and reproduce the manual calculation of Example
   1 by fixing the random seed appropriately.
2. Implement `generate_multivariate_normal()` from first principles
   (already provided) and reproduce the manual calculation of Example
   2 by fixing the random seed and manually extracting the standard
   Normal draws used.
3. Write an R function that checks whether a supplied matrix is
   symmetric positive definite (using `chol()` inside a
   `tryCatch()`), and test it on both a valid and an invalid
   covariance matrix.
4. Using `MASS::mvrnorm()` or `mvtnorm::rmvnorm()` as a benchmark,
   verify that `generate_multivariate_normal()` produces samples with
   statistically indistinguishable covariance structure for a
   4-dimensional target of your choosing.
5. Implement `generate_exponential_uniform_pair()` from first
   principles (already provided) and empirically verify that the
   conditional distribution of $Y$ given $X$ in a narrow bin around
   $x=1$ is approximately Uniform$(0,1)$.
6. Write an R function implementing the sequential conditional method
   for a three-stage discrete Markov-chain-like construction:
   $X_1 \sim \text{Bernoulli}(0.5)$,
   $X_2 \mid X_1 \sim \text{Bernoulli}(0.7 \text{ if } X_1=1
   \text{ else } 0.2)$,
   $X_3 \mid X_2 \sim \text{Bernoulli}(0.7 \text{ if } X_2=1 \text{
   else } 0.2)$, and estimate $P(X_3=1)$ from a large simulated
   sample.
7. Reproduce the trivariate pairwise scatterplot figure of this unit
   for a covariance matrix of your own choosing, and verify the
   empirical correlations against the target values.
8. Write a function that generates from a bivariate distribution with
   Exponential marginals and a specified dependence via the
   composition idea of Unit 3.5 combined with the conditional method
   of this unit (for example, a shared-shock construction), and
   examine the resulting correlation empirically.
9. Benchmark, using `system.time()`, the Cholesky method against a
   naive coordinate-by-coordinate conditional-regression
   implementation for simulating a 10-dimensional Normal vector, and
   discuss the results.
10. Implement Poisson thinning in R: simulate $X \sim
    \text{Poisson}(\lambda)$, then $Y \mid X=x \sim
    \text{Binomial}(x,p)$, and verify empirically that the marginal
    distribution of $Y$ is $\text{Poisson}(\lambda p)$, a classical
    result in point process theory.

## Mini Projects

1. **Simulating correlated asset returns.** Using the Cholesky method,
   simulate ten years of daily returns for four correlated assets with
   a covariance matrix estimated from (or plausibly specified for)
   real or illustrative data. Compute the simulated portfolio's
   value-at-risk at the $95\%$ and $99\%$ confidence levels for an
   equally weighted portfolio, and compare these figures to those
   obtained if the assets were (incorrectly) simulated as independent.
   Discuss, quantitatively, how ignoring correlation distorts the risk
   assessment.

2. **From sequential conditionals to Gibbs sampling.** Implement the
   exact sequential conditional method for a bivariate Normal target
   with a known correlation $\rho$ (as in Algorithm 6.2), and
   separately implement a Gibbs sampler for the same target using its
   full conditional distributions
   $X_1 \mid X_2=x_2 \sim N(\rho x_2, 1-\rho^2)$ and
   $X_2 \mid X_1=x_1 \sim N(\rho x_1, 1-\rho^2)$, iterated for a large
   number of steps. Compare the two resulting samples graphically and
   using summary statistics, discuss the burn-in behaviour of the
   Gibbs sampler, and explain why the exact method is preferable
   whenever it is available.

**Difficulty labels.** Multiple choice, short answer: **Easy**. Long
answer questions 3, 4, 7, 8, 9, programming exercises 1-6: **Medium**.
Long answer questions 1, 2, 5, 6, 10, programming exercises 7-10:
**Medium to Advanced**. Mini projects: **Advanced**.

# References

Casella, G., & Berger, R. L. (2002). *Statistical inference* (2nd
ed.). Duxbury.

Golub, G. H., & Van Loan, C. F. (2013). *Matrix computations* (4th
ed.). Johns Hopkins University Press.

Pourahmadi, M. (1999). Joint mean-covariance models with applications
to longitudinal data: Unconstrained parameterisation. *Biometrika*,
86(3), 677-690.

Ripley, B. D. (1987). *Stochastic simulation*. Wiley.

Rizzo, M. L. (2019). *Statistical computing with R* (2nd ed.). CRC
Press.

Ross, S. M. (2020). *Simulation* (6th ed.). Academic Press.
