---
title: "Unit 3.5: Generation Using Distributional Relationships"
author: "Dr. Manoj C. Patil"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: false
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
    code_folding: hide
    theme: cosmo
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    echo = TRUE,
    warning = FALSE,
    message = FALSE,
    fig.align = "center",
    fig.width = 7,
    fig.height = 4.5
)

set.seed(123)
library(tidyverse)
library(ggplot2)
library(dplyr)
library(knitr)
library(kableExtra)
library(gridExtra)
library(patchwork)
```

# Learning Objectives

By the end of this unit, a student should be able to

- explain what is meant by a distributional relationship, and why
  exploiting such relationships can be more efficient than inverse
  transformation or acceptance-rejection,
- state and apply the composition method for simulating from finite
  and continuous mixture distributions,
- state and apply the convolution method for simulating from
  distributions expressible as sums of independent, more elementary
  random variables,
- derive the Chi-square distribution as a sum of squared standard
  Normal variables, and generate it accordingly,
- derive the density of Student's $t$ distribution from its defining
  ratio representation and implement the corresponding generator,
- derive the density of the $F$ distribution from its defining ratio
  representation and implement the corresponding generator,
- assess when a distributional-relationship generator is preferable to
  a general-purpose method, and recognize its computational
  limitations.

# Prerequisites

A reader should already be comfortable with

- the inverse transformation and acceptance-rejection methods (Units
  3.3 and 3.4),
- moment generating functions and their uniqueness property for
  determining a distribution,
- the change-of-variables technique for deriving the density of a
  transformed random variable, including the bivariate Jacobian
  transformation,
- the definitions of the Gamma and Beta functions, and of the Normal,
  Exponential, and Gamma distributions.

# Introduction

The methods developed in Units 3.3 and 3.4 are general-purpose: given
essentially any target density, an inverse transformation or an
acceptance-rejection scheme can, in principle, be constructed. General
purpose, however, does not mean most efficient. A very large number of
distributions that arise constantly in statistical practice —
Chi-square, Student's $t$, the $F$ distribution, Binomial, Negative
Binomial, Erlang, and finite or continuous mixture models — are not
arbitrary densities plucked at random; they are *defined*, or are
provably equivalent to, explicit algebraic combinations of simpler,
already-familiar random variables. Whenever such a relationship is
known, it typically gives a generator that is simpler to code, faster
to run, and easier to verify than a general-purpose alternative.

This observation has a long history in the theory of distributions.
The Chi-square distribution was introduced by Karl Pearson in 1900
precisely as the distribution of a sum of squared standardized
Normal deviates, arising from his goodness-of-fit statistic; Student's
$t$ distribution was derived by William Sealy Gosset in 1908 (writing
under the pseudonym "Student") as the distribution of a standardized
sample mean when the population variance must itself be estimated,
and its ratio representation was placed on firm footing shortly
afterward by Ronald Fisher; the $F$ distribution, named in Fisher's
honour by George Snedecor, arises as the ratio of two independent
variance estimates and is the backbone of the analysis of variance.
Each of these three distributions is generated most naturally not by
inverting its own (often intractable) cumulative distribution
function, but by simulating the simpler quantities from which it is
*defined* and combining them exactly as the definition prescribes.

The practical importance of this unit is therefore twofold. First, it
supplies fast, exact generators for three distributions — Chi-square,
$t$, and $F$ — that are used in essentially every classical hypothesis
test taught in a statistics curriculum, and that any simulation study
of test size or power must be able to generate rapidly and correctly.
Second, and more broadly, it introduces two general strategies —
composition and convolution — that extend far beyond these three
named distributions, underlying the simulation of finite mixture
models in cluster analysis, hierarchical Bayesian models, and
compound distributions in actuarial and reliability applications.

# Intuition

Consider trying to describe how tall a randomly chosen adult is. A
single Normal distribution is a reasonable first approximation, but a
population is really a blend of sub-populations — for instance, a
blend of the height distributions of men and of women, each
Normal but with different means, combined according to the
proportion of each group in the population. Simulating a height
observation the "hard way" would mean writing down the density of this
blended distribution and inverting its (rather complicated) cumulative
distribution function. The easy way exploits the structure directly:
first decide, at random and with the correct probability, which
sub-population the observation belongs to, and only then simulate a
height from that sub-population's own — much simpler — Normal
distribution. This two-stage recipe, decide-the-component and then
simulate-within-the-component, is the entire idea behind the
**composition method**, and it works because the resulting mixture
automatically has exactly the blended distribution being sought.

A second, equally natural idea arises when a quantity is itself built
up as an accumulation of independent smaller effects. The total number
of typographical errors in a book, if each of many independent pages
contributes its own small and independent error count, is naturally
generated by simulating each page's count separately and adding them
up, rather than by directly inverting the cumulative distribution
function of the (potentially complicated) total. This is the
**convolution method**: when a target variable is *known* to equal a
sum of independent, simpler variables, simulate each summand and add.
The Chi-square distribution with $\nu$ degrees of freedom, defined as
the sum of $\nu$ independent squared standard Normal variables, is the
cleanest possible illustration of this idea, and it is also the
building block, via simple ratios, for both the $t$ and $F$
distributions developed later in this unit.

# Formal Definitions

## Distributional Relationships

**Definition.** A **distributional relationship** is a known
probabilistic identity linking a target random variable $X$ to one or
more other random variables $Y_1, \dots, Y_k$ whose distributions are
already easy to simulate (for example, by inverse transformation or by
a built-in generator), expressed either as a deterministic function

$$
X = \phi(Y_1, \dots, Y_k)
$$

of independent (or otherwise jointly specified) building blocks, or as
a probabilistic mixture over a random index or parameter. Generation
by distributional relationship means simulating the $Y_i$ (or the
mixing index/parameter) according to their own distributions and then
applying the known relationship $\phi$ (or the corresponding
conditional generation step) to obtain a draw of $X$.

## The Composition Method

**Definition (discrete mixture).** Suppose the target density can be
written as a finite mixture

$$
f(x) = \sum_{j=1}^{k} p_j\, f_j(x), \qquad p_j \ge 0,\ \sum_{j=1}^k p_j = 1,
\tag{5.1}
$$

where each component density $f_j$ is easy to simulate. The
**composition method** generates $X$ by first drawing a component
index $J \in \{1,\dots,k\}$ with $P(J=j) = p_j$, and then, conditional
on $J=j$, drawing $X \sim f_j$.

**Definition (continuous mixture).** More generally, if

$$
f(x) = \int f(x\mid\theta)\, w(\theta)\, d\theta,
\tag{5.2}
$$

for a mixing density $w$ and a family of conditional densities
$f(\cdot\mid\theta)$, the composition method draws $\Theta \sim w$ and
then, conditional on $\Theta=\theta$, draws $X \sim f(\cdot\mid\theta)$.

where

- $p_j$ (or $w(\theta)$) denotes the probability (or density) of
  selecting component $j$ (or parameter value $\theta$),
- $f_j$ (or $f(\cdot\mid\theta)$) denotes the density from which $X$
  is drawn once the component (or parameter) has been selected.

## The Convolution Method

**Definition.** Suppose the target random variable $X$ is known to
satisfy

$$
X = Y_1 + Y_2 + \cdots + Y_n,
\tag{5.3}
$$

where $Y_1,\dots,Y_n$ are independent random variables with known,
easily simulated distributions, and the distribution of $X$ is
precisely the $n$-fold convolution of the distributions of the $Y_i$.
The **convolution method** generates $X$ by drawing each $Y_i$
independently and returning their sum.

## Mixture Distributions

**Definition.** A **mixture distribution** is precisely a distribution
of the form (5.1) or (5.2): the marginal distribution of a variable
whose generating mechanism first selects, at random, a sub-population
(or parameter value), and only then generates the observation from
within that sub-population. The composition method (Section 5.2) is
the direct simulation counterpart of this definition; simulating a
mixture is, definitionally, an application of the composition method.

## The Chi-square Distribution

**Definition.** If $Z_1,\dots,Z_\nu$ are independent standard Normal
random variables, the **Chi-square distribution with $\nu$ degrees of
freedom** is the distribution of

$$
X = \sum_{i=1}^{\nu} Z_i^2,
\tag{5.4}
$$

written $X \sim \chi^2_\nu$. This is, by construction, a convolution
representation: $\chi^2_\nu$ is the $\nu$-fold convolution of the
$\chi^2_1$ distribution (the distribution of a single $Z_i^2$) with
itself.

## Student's $t$ Distribution

**Definition.** If $Z \sim N(0,1)$ and $V \sim \chi^2_\nu$ are
independent, **Student's $t$ distribution with $\nu$ degrees of
freedom** is the distribution of

$$
T = \frac{Z}{\sqrt{V/\nu}},
\tag{5.5}
$$

written $T \sim t_\nu$.

## The $F$ Distribution

**Definition.** If $U \sim \chi^2_{d_1}$ and $V \sim \chi^2_{d_2}$ are
independent, the **$F$ distribution with $(d_1,d_2)$ degrees of
freedom** is the distribution of

$$
F = \frac{U/d_1}{V/d_2},
\tag{5.6}
$$

written $F \sim F_{d_1,d_2}$.

# Theorems

## Theorem 5.1: Correctness of the Composition Method

**Statement.** If $X$ is generated by the composition method from
(5.1) — draw $J$ with $P(J=j)=p_j$, then draw $X\sim f_J$ — the
marginal density of $X$ is exactly $f(x) = \sum_{j=1}^k p_j f_j(x)$.
The analogous statement holds for the continuous-mixture form (5.2).

**Assumptions.** The component densities $f_j$ (or $f(\cdot\mid\theta)$)
and the mixing probabilities $p_j$ (or density $w$) are all valid
probability functions, and $J$ (or $\Theta$) is generated independently
of the subsequent within-component draw, conditional on its own value.

**Proof.** By the law of total probability, for any measurable set
$A$,

$$
P(X \in A) = \sum_{j=1}^{k} P(X \in A \mid J=j)\, P(J=j)
= \sum_{j=1}^{k} p_j \int_A f_j(x)\, dx
= \int_A \Bigl(\sum_{j=1}^{k} p_j f_j(x)\Bigr) dx,
$$

so $X$ has density $f(x) = \sum_j p_j f_j(x)$, as required. The
continuous case follows identically with the sum replaced by an
integral over $\theta$ and $P(J=j)$ replaced by $w(\theta)\,d\theta$.
$\blacksquare$

**Remarks.** The proof uses nothing beyond the law of total
probability; the entire content of the method is the *algorithmic*
observation that this decomposition can be turned around and used as
a two-stage generation recipe, rather than the mathematical content of
the theorem itself, which is elementary.

**Examples.** A full worked example for a two-component Exponential
mixture is given in the Numerical Examples section.

**Applications.** The composition method is the standard technique for
simulating from Gaussian mixture models in cluster analysis, from
compound distributions in actuarial risk modelling (e.g. a claim-size
distribution that differs by policy type), and from hierarchical
Bayesian models, in which a parameter is itself drawn from a prior
before data are simulated conditional on it.

## Theorem 5.2: Additivity of Independent Gamma Variables

**Statement.** If $Y_1,\dots,Y_n$ are independent with
$Y_i \sim \text{Gamma}(\alpha_i, \beta)$ (shape $\alpha_i$, common
scale $\beta$), then

$$
X = \sum_{i=1}^{n} Y_i \;\sim\; \text{Gamma}\Bigl(\sum_{i=1}^n \alpha_i,\ \beta\Bigr).
$$

**Assumptions.** The $Y_i$ share a common scale parameter $\beta$;
additivity does not hold, in general, across Gamma variables with
different scales.

**Proof.** The moment generating function of $Y_i \sim
\text{Gamma}(\alpha_i,\beta)$ is $M_{Y_i}(t) = (1-\beta t)^{-\alpha_i}$
for $t < 1/\beta$. By independence, the moment generating function of
the sum is the product,

$$
M_X(t) = \prod_{i=1}^n M_{Y_i}(t) = \prod_{i=1}^n (1-\beta t)^{-\alpha_i}
= (1-\beta t)^{-\sum_i \alpha_i},
$$

which is exactly the moment generating function of a
$\text{Gamma}(\sum_i \alpha_i, \beta)$ random variable. Since the
moment generating function determines the distribution uniquely (when
it exists in a neighbourhood of $0$), $X \sim
\text{Gamma}(\sum_i\alpha_i,\beta)$. $\blacksquare$

**Remarks.** Since $Z^2 \sim \chi^2_1 = \text{Gamma}(1/2, 2)$ for
$Z \sim N(0,1)$ (derived in the Mathematical Derivations section), and
all $\chi^2_1$ variables share the common scale $\beta=2$, Theorem 5.2
immediately shows that a sum of $\nu$ independent $\chi^2_1$ variables
is $\text{Gamma}(\nu/2, 2) = \chi^2_\nu$, giving a rigorous
justification for the convolution-based Chi-square generator of
Section 8.

**Examples.** Verified numerically in the Simulation Studies section.

**Applications.** The same additivity underlies simulating an Erlang
(integer-shape Gamma) waiting time as a sum of independent Exponential
interarrival times in queueing and reliability models, and simulating
a Negative Binomial count as a sum of independent Geometric trial
counts.

# Mathematical Derivations

## The Density of $Z^2$ is $\chi^2_1$

Let $Z \sim N(0,1)$ and $X = Z^2$. For $x > 0$,

$$
\begin{aligned}
F_X(x) = P(Z^2 \le x) &= P(-\sqrt{x} \le Z \le \sqrt{x}) \\
&= \Phi(\sqrt{x}) - \Phi(-\sqrt{x}) = 2\Phi(\sqrt{x}) - 1.
\end{aligned}
$$

Differentiating with respect to $x$, using the chain rule and
$\Phi'(z) = \varphi(z) = (2\pi)^{-1/2}e^{-z^2/2}$,

$$
f_X(x) = 2\varphi(\sqrt{x})\cdot \frac{1}{2\sqrt{x}}
= \frac{1}{\sqrt{2\pi x}}\, e^{-x/2}, \qquad x>0.
\tag{5.7}
$$

This is exactly the Gamma$(1/2, 2)$ density
$f(x) = x^{-1/2}e^{-x/2}/(2^{1/2}\Gamma(1/2))$, using
$\Gamma(1/2)=\sqrt{\pi}$, confirming $Z^2 \sim \chi^2_1$. Applying
Theorem 5.2 to a sum of $\nu$ independent copies of $Z^2$ then gives
$\chi^2_\nu = \text{Gamma}(\nu/2,2)$, with density

$$
f_{\chi^2_\nu}(x) = \frac{1}{2^{\nu/2}\Gamma(\nu/2)}\, x^{\nu/2-1} e^{-x/2},
\qquad x>0.
\tag{5.8}
$$

## Derivation of the Student's $t$ Density from the Ratio Representation

Let $Z\sim N(0,1)$ and $V\sim\chi^2_\nu$ be independent, and let
$T = Z/\sqrt{V/\nu}$. Conditional on $V=v$, $T$ is a constant multiple
of $Z$, so

$$
T \mid V=v \;\sim\; N\Bigl(0, \frac{\nu}{v}\Bigr),
\qquad
f_{T\mid V}(t\mid v) = \sqrt{\frac{v}{2\pi\nu}}\, \exp\Bigl(-\frac{t^2 v}{2\nu}\Bigr).
$$

Multiplying by the $\chi^2_\nu$ density (5.8) and integrating out $v$,

$$
f_T(t) = \int_0^\infty \sqrt{\frac{v}{2\pi\nu}}\, e^{-t^2v/(2\nu)}
\cdot \frac{v^{\nu/2-1}e^{-v/2}}{2^{\nu/2}\Gamma(\nu/2)}\, dv
= \frac{1}{\sqrt{2\pi\nu}\,2^{\nu/2}\Gamma(\nu/2)}
\int_0^\infty v^{(\nu+1)/2 - 1}\, e^{-\frac{v}{2}\left(1+\frac{t^2}{\nu}\right)}\, dv.
$$

The remaining integral is a Gamma-function integral,
$\int_0^\infty v^{a-1}e^{-bv}\,dv = \Gamma(a)/b^{a}$, with
$a = (\nu+1)/2$ and $b = \tfrac12\bigl(1+t^2/\nu\bigr)$, giving

$$
\int_0^\infty v^{(\nu+1)/2-1} e^{-bv}\, dv
= \Gamma\Bigl(\frac{\nu+1}{2}\Bigr) \cdot \frac{2^{(\nu+1)/2}}
{\bigl(1+t^2/\nu\bigr)^{(\nu+1)/2}}.
$$

Substituting back and combining the powers of $2$
($2^{(\nu+1)/2}/2^{\nu/2} = \sqrt2$),

$$
f_T(t) = \frac{\Gamma\left(\frac{\nu+1}{2}\right)}
{\sqrt{\pi\nu}\; \Gamma\left(\frac{\nu}{2}\right)}
\Bigl(1+\frac{t^2}{\nu}\Bigr)^{-(\nu+1)/2}, \qquad t \in \mathbb{R},
\tag{5.9}
$$

which is exactly the standard Student's $t_\nu$ density.

## Derivation of the $F$ Density from the Ratio Representation

Let $U\sim\chi^2_{d_1}$ and $V\sim\chi^2_{d_2}$ be independent, and let
$F = (U/d_1)/(V/d_2) = (d_2 U)/(d_1 V)$. Transform $(U,V) \to (F,V)$
via $U = (d_1/d_2)\,F V$, holding $V$ fixed, so the Jacobian of the
transformation is $\partial u/\partial f = (d_1/d_2)\,v$. Substituting
into the product $f_U(u)f_V(v)$ (independence) and multiplying by the
Jacobian,

$$
f_{F,V}(f,v) = f_U\Bigl(\frac{d_1}{d_2}fv\Bigr)\, f_V(v)\, \frac{d_1}{d_2}v.
$$

Using the Chi-square density (5.8) for both factors and collecting
powers of $v$ (the same integral technique used for the $t$
derivation above),

$$
f_{F,V}(f,v) = C\, f^{d_1/2-1}\, v^{(d_1+d_2)/2-1}
\exp\Bigl(-\frac{v}{2}\cdot\frac{d_2+d_1 f}{d_2}\Bigr),
$$

for a constant $C$ collecting all factors not involving $v$. Applying
$\int_0^\infty v^{a-1}e^{-bv}\,dv=\Gamma(a)/b^a$ with $a=(d_1+d_2)/2$
and $b = (d_2+d_1f)/(2d_2)$, and simplifying the resulting powers of
$d_1$, $d_2$, and $2$ exactly as in the $t$-density derivation, yields

$$
f_F(f) = \frac{\Gamma\left(\frac{d_1+d_2}{2}\right)}
{\Gamma\left(\frac{d_1}{2}\right)\Gamma\left(\frac{d_2}{2}\right)}
\left(\frac{d_1}{d_2}\right)^{d_1/2} f^{\,d_1/2-1}
\Bigl(1+\frac{d_1}{d_2}f\Bigr)^{-(d_1+d_2)/2}, \qquad f>0,
\tag{5.10}
$$

which is exactly the standard $F_{d_1,d_2}$ density.

# Algorithms

## Algorithm 5.1: Composition Method

**Purpose.** Generate a sample from a mixture density
$f(x)=\sum_j p_j f_j(x)$ using component samplers that are already
available.

**Input.** Component probabilities $p_1,\dots,p_k$; component samplers
$f_1,\dots,f_k$; required sample size $n$.

**Output.** A vector $x_1,\dots,x_n$ of simulated observations from
$f$.

**Step-by-step procedure.**

1. Compute the cumulative probabilities
   $c_j = \sum_{i=1}^j p_i$, $j=1,\dots,k$.
2. For $i=1,\dots,n$: draw $U_i \sim \text{Uniform}(0,1)$; find the
   smallest $j$ with $U_i \le c_j$; draw $x_i$ from component $f_j$.
3. Return $(x_1,\dots,x_n)$.

**Pseudo-code.**

```
Algorithm: Composition Method
Input:  probabilities p[1..k], component samplers sample_j(), n
Output: vector x of length n

1. c[1..k] <- cumulative sum of p
2. For i = 1 to n:
       u <- Uniform(0,1)
       j <- smallest index with c[j] >= u
       x[i] <- sample_j()
3. Return x
```

**Flowchart description.** Start; draw a uniform variate to select a
component index via the cumulative probability table (an application
of the discrete inverse transformation algorithm of Unit 3.3); draw an
observation from the selected component's own generator; repeat until
$n$ observations are collected; end.

**Complexity.** $O(n)$ overall, provided component selection is
$O(\log k)$ (via table lookup) or $O(k)$ (via linear scan) and each
within-component draw is $O(1)$.

**Practical considerations.** For a large number of repeated draws
with the same fixed component probabilities, R's `sample()` function
performs component selection efficiently and should be preferred over
a hand-written cumulative-sum search.

## Algorithm 5.2: Convolution Method for the Chi-square Distribution

**Purpose.** Generate a Chi-square random variable with integer
degrees of freedom $\nu$ directly from its defining sum of squared
Normal variables.

**Input.** Degrees of freedom $\nu$ (positive integer); required
sample size $n$.

**Output.** A vector $x_1,\dots,x_n$ of simulated $\chi^2_\nu$
observations.

**Step-by-step procedure.**

1. For $i=1,\dots,n$: draw $Z_{i1},\dots,Z_{i\nu}$ independently from
   $N(0,1)$; set $x_i = \sum_{k=1}^\nu Z_{ik}^2$.
2. Return $(x_1,\dots,x_n)$.

**Pseudo-code.**

```
Algorithm: Convolution Method, Chi-square
Input:  nu, n
Output: vector x of length n

1. For i = 1 to n:
       z <- nu independent draws from N(0,1)
       x[i] <- sum(z^2)
2. Return x
```

**Flowchart description.** Start; for each of $n$ output values, draw
$\nu$ independent standard Normal variates, square and sum them;
collect the $n$ sums; end.

**Complexity.** $O(n\nu)$, since each output value requires $\nu$
Normal draws and $\nu$ squarings and additions; this is markedly more
expensive than a direct Gamma-based generator (Section 12) when $\nu$
is large.

**Practical considerations.** This method is pedagogically valuable
because it is a direct implementation of the *definition* of the
Chi-square distribution, but it should not be used for large $\nu$ in
production code, where a direct Gamma$(\nu/2,2)$ generator (which
extends naturally to non-integer $\nu$ as well) is both faster and
more general.

## Algorithm 5.3: Ratio Method for Student's $t$ Distribution

**Purpose.** Generate a $t_\nu$ random variable directly from its
defining ratio of an independent Normal and Chi-square variable.

**Input.** Degrees of freedom $\nu$; required sample size $n$.

**Output.** A vector $t_1,\dots,t_n$ of simulated $t_\nu$ observations.

**Step-by-step procedure.**

1. For $i=1,\dots,n$: draw $Z_i \sim N(0,1)$ and $V_i \sim \chi^2_\nu$
   independently; set $t_i = Z_i / \sqrt{V_i/\nu}$.
2. Return $(t_1,\dots,t_n)$.

**Pseudo-code.**

```
Algorithm: Ratio Method, Student's t
Input:  nu, n
Output: vector t of length n

1. For i = 1 to n:
       z <- draw from N(0,1)
       v <- draw from ChiSquare(nu)
       t[i] <- z / sqrt(v / nu)
2. Return t
```

**Flowchart description.** Start; for each output value, draw one
standard Normal variate and one independent Chi-square variate;
divide the Normal draw by the square root of the scaled Chi-square
draw; collect the results; end.

**Complexity.** $O(n)$, given access to $O(1)$-amortized Normal and
Chi-square generators.

**Practical considerations.** The Chi-square draw $V_i$ may itself use
Algorithm 5.2 (for small integer $\nu$) or a general Gamma-based
generator; either choice preserves exactness of the resulting $t_\nu$
distribution.

## Algorithm 5.4: Ratio Method for the $F$ Distribution

**Purpose.** Generate an $F_{d_1,d_2}$ random variable directly from
its defining ratio of two independent scaled Chi-square variables.

**Input.** Degrees of freedom $d_1$, $d_2$; required sample size $n$.

**Output.** A vector $f_1,\dots,f_n$ of simulated $F_{d_1,d_2}$
observations.

**Step-by-step procedure.**

1. For $i=1,\dots,n$: draw $U_i \sim \chi^2_{d_1}$ and
   $V_i \sim \chi^2_{d_2}$ independently; set
   $f_i = (U_i/d_1)/(V_i/d_2)$.
2. Return $(f_1,\dots,f_n)$.

**Pseudo-code.**

```
Algorithm: Ratio Method, F distribution
Input:  d1, d2, n
Output: vector f of length n

1. For i = 1 to n:
       u <- draw from ChiSquare(d1)
       v <- draw from ChiSquare(d2)
       f[i] <- (u / d1) / (v / d2)
2. Return f
```

**Flowchart description.** Start; for each output value, draw two
independent Chi-square variates with the respective degrees of
freedom; scale each by its own degrees of freedom and take the ratio;
collect the results; end.

**Complexity.** $O(n)$, given $O(1)$-amortized Chi-square generation.

**Practical considerations.** Because an $F_{d_1,d_2}$ variate is a
ratio of two positive quantities, numerical care is needed only in the
unlikely event that $V_i$ is generated as exactly zero, which has
probability zero analytically but should still be guarded against in
finite-precision arithmetic when $d_2$ is very small.

# R Programming

```{r generator-functions}
set.seed(123)

#' Generate a mixture sample using the composition method
#'
#' Purpose: implements Algorithm 5.1 for a finite mixture with
#' arbitrary component samplers.
#'
#' Arguments:
#'   sample_size            number of observations to generate
#'   component_probabilities  numeric vector of mixture weights, summing to 1
#'   component_samplers      a list of zero-argument functions, one per
#'                            component, each returning a single draw
#'
#' Returns:
#'   a numeric vector of length sample_size
#'
#' Example:
#'   generate_mixture(
#'     sample_size = 5,
#'     component_probabilities = c(0.3, 0.7),
#'     component_samplers = list(
#'       function() rexp(1, rate = 1),
#'       function() rexp(1, rate = 5)
#'     )
#'   )
generate_mixture <- function(sample_size, component_probabilities, component_samplers) {
  component_index <- sample(
    x = seq_along(component_probabilities),
    size = sample_size,
    replace = TRUE,
    prob = component_probabilities
  )

  vapply(
    component_index,
    function(index) component_samplers[[index]](),
    numeric(1)
  )
}

#' Generate Chi-square variates by the convolution method
#'
#' Purpose: implements Algorithm 5.2, summing squared standard Normal
#' variates directly from the definition of the Chi-square distribution.
#'
#' Arguments:
#'   sample_size       number of observations to generate
#'   degrees_of_freedom  positive integer degrees of freedom
#'
#' Returns:
#'   a numeric vector of length sample_size
#'
#' Example:
#'   generate_chisq_convolution(sample_size = 5, degrees_of_freedom = 3)
generate_chisq_convolution <- function(sample_size, degrees_of_freedom) {
  standard_normal_matrix <- matrix(
    rnorm(sample_size * degrees_of_freedom),
    nrow = sample_size, ncol = degrees_of_freedom
  )

  rowSums(standard_normal_matrix^2)
}

#' Generate Student's t variates by the ratio method
#'
#' Purpose: implements Algorithm 5.3, dividing an independent standard
#' Normal draw by the square root of a scaled Chi-square draw.
#'
#' Arguments:
#'   sample_size       number of observations to generate
#'   degrees_of_freedom  degrees of freedom of the target t distribution
#'
#' Returns:
#'   a numeric vector of length sample_size
#'
#' Example:
#'   generate_t_ratio(sample_size = 5, degrees_of_freedom = 5)
generate_t_ratio <- function(sample_size, degrees_of_freedom) {
  normal_draws <- rnorm(sample_size)
  chisq_draws <- generate_chisq_convolution(sample_size, degrees_of_freedom)

  normal_draws / sqrt(chisq_draws / degrees_of_freedom)
}

#' Generate F variates by the ratio method
#'
#' Purpose: implements Algorithm 5.4, forming the ratio of two
#' independent, appropriately scaled Chi-square draws.
#'
#' Arguments:
#'   sample_size    number of observations to generate
#'   df1            numerator degrees of freedom
#'   df2            denominator degrees of freedom
#'
#' Returns:
#'   a numeric vector of length sample_size
#'
#' Example:
#'   generate_f_ratio(sample_size = 5, df1 = 3, df2 = 5)
generate_f_ratio <- function(sample_size, df1, df2) {
  numerator_chisq <- generate_chisq_convolution(sample_size, df1)
  denominator_chisq <- generate_chisq_convolution(sample_size, df2)

  (numerator_chisq / df1) / (denominator_chisq / df2)
}
```

# Simulation Studies

## Composition Method: A Two-Component Exponential Mixture

```{r mixture-simulation}
set.seed(123)

mixture_sample <- generate_mixture(
  sample_size = 20000,
  component_probabilities = c(0.3, 0.7),
  component_samplers = list(
    function() rexp(1, rate = 1),
    function() rexp(1, rate = 5)
  )
)

mixture_density_theoretical <- function(x) {
  0.3 * dexp(x, rate = 1) + 0.7 * dexp(x, rate = 5)
}

tibble(
  Quantity = c("Theoretical mean (0.3/1 + 0.7/5)", "Empirical mean"),
  Value = c(0.3 / 1 + 0.7 / 5, mean(mixture_sample))
) %>%
  kable(digits = 4, caption = "Composition-method mixture: theoretical vs empirical mean")
```

## Convolution Method: Chi-square Generated from Squared Normals

```{r chisq-simulation}
set.seed(123)

chisq_sample <- generate_chisq_convolution(sample_size = 20000, degrees_of_freedom = 4)

ks_result_chisq <- ks.test(chisq_sample, "pchisq", df = 4)

tibble(
  Quantity = c("Theoretical mean (df)", "Empirical mean",
               "Theoretical variance (2 df)", "Empirical variance",
               "KS test p-value against chisq(df=4)"),
  Value = c(4, mean(chisq_sample), 8, var(chisq_sample), ks_result_chisq$p.value)
) %>%
  kable(digits = 4, caption = "Convolution-method Chi-square(4): theory vs simulation")
```

## Ratio Method: Student's $t$ and the $F$ Distribution

```{r t-f-simulation}
set.seed(123)

t_sample <- generate_t_ratio(sample_size = 20000, degrees_of_freedom = 8)
f_sample <- generate_f_ratio(sample_size = 20000, df1 = 4, df2 = 10)

ks_result_t <- ks.test(t_sample, "pt", df = 8)
ks_result_f <- ks.test(f_sample, "pf", df1 = 4, df2 = 10)

tibble(
  Distribution = c("t(df=8)", "F(4,10)"),
  Theoretical_Mean = c(0, 10 / (10 - 2)),
  Empirical_Mean = c(mean(t_sample), mean(f_sample)),
  KS_p_value = c(ks_result_t$p.value, ks_result_f$p.value)
) %>%
  kable(digits = 4, caption = "Ratio-method t and F distributions: theory vs simulation")
```

In every case the Kolmogorov-Smirnov test comparing the simulated
sample to the target theoretical distribution yields a large
$p$-value, showing no evidence against the samples having exactly the
intended distribution, as the theory of Sections 6-7 guarantees.

# Graphs

```{r mixture-histogram, fig.cap="Composition-method mixture sample against its theoretical density."}
ggplot(tibble(x = mixture_sample), aes(x = x)) +
  geom_histogram(aes(y = after_stat(density)), bins = 50, fill = "steelblue", colour = "white") +
  stat_function(fun = mixture_density_theoretical, colour = "darkred", linewidth = 1) +
  coord_cartesian(xlim = c(0, 4)) +
  labs(title = "Two-component Exponential mixture (composition method)", x = "x", y = "Density") +
  theme_minimal(base_size = 11)
```

```{r chisq-histogram, fig.cap="Convolution-method Chi-square(4) sample against the theoretical density."}
ggplot(tibble(x = chisq_sample), aes(x = x)) +
  geom_histogram(aes(y = after_stat(density)), bins = 50, fill = "darkgreen", colour = "white") +
  stat_function(fun = function(x) dchisq(x, df = 4), colour = "darkred", linewidth = 1) +
  labs(title = "Chi-square(4) via convolution of squared Normals", x = "x", y = "Density") +
  theme_minimal(base_size = 11)
```

```{r t-f-histograms, fig.cap="Ratio-method Student's t(8) (left) and F(4,10) (right) samples against their theoretical densities."}
p_t <- ggplot(tibble(x = t_sample), aes(x = x)) +
  geom_histogram(aes(y = after_stat(density)), bins = 50, fill = "purple", colour = "white", alpha = 0.7) +
  stat_function(fun = function(x) dt(x, df = 8), colour = "black", linewidth = 1) +
  coord_cartesian(xlim = c(-6, 6)) +
  labs(title = "Student's t(df=8)", x = "x", y = "Density") +
  theme_minimal(base_size = 11)

p_f <- ggplot(tibble(x = f_sample), aes(x = x)) +
  geom_histogram(aes(y = after_stat(density)), bins = 50, fill = "orange", colour = "white", alpha = 0.7) +
  stat_function(fun = function(x) df(x, df1 = 4, df2 = 10), colour = "black", linewidth = 1) +
  coord_cartesian(xlim = c(0, 6)) +
  labs(title = "F(4, 10)", x = "x", y = "Density") +
  theme_minimal(base_size = 11)

p_t + p_f
```

```{r t-tails-comparison, fig.cap="Student's t density for several degrees of freedom compared to the standard Normal, illustrating convergence as degrees of freedom increase."}
t_curve_df <- tibble(
  x = seq(-5, 5, length.out = 400)
) %>%
  mutate(
    df_1 = dt(x, df = 1),
    df_5 = dt(x, df = 5),
    df_30 = dt(x, df = 30),
    Normal = dnorm(x)
  ) %>%
  pivot_longer(cols = -x, names_to = "Distribution", values_to = "Density")

ggplot(t_curve_df, aes(x = x, y = Density, colour = Distribution, linetype = Distribution)) +
  geom_line(linewidth = 0.9) +
  labs(title = "Heavier t tails converge to the Normal as degrees of freedom grow") +
  theme_minimal(base_size = 11)
```

# Numerical Examples

**Example 1: Composition method for a two-component mixture.**

*Problem.* With component probabilities $p_1=0.3$ (Exponential rate
$1$) and $p_2=0.7$ (Exponential rate $5$), and selector draw
$u_1=0.20$ together with within-component draw $u_2=0.6321$, apply
the composition method.

*Solution.* Since $u_1 = 0.20 \le p_1 = 0.30$, component $1$ is
selected. Within component $1$, inverse transformation for
Exponential$(1)$ gives $x = -\ln(1-u_2) = -\ln(0.3679) = 1.000$.

*Interpretation.* Repeating with $u_1 = 0.85$ ($>0.30$, so component
$2$ is selected) and the same $u_2=0.6321$ gives
$x = -\ln(1-u_2)/5 = 1.000/5 = 0.200$: because component $2$ has a
much higher rate, its typical draws are correspondingly smaller,
exactly the qualitative behaviour visible in the histogram above.

**Example 2: Convolution method for $\chi^2_3$.**

*Problem.* Given three standard Normal draws $z_1=1.2$, $z_2=-0.5$,
$z_3=0.8$, compute the resulting $\chi^2_3$ observation.

*Solution.*
$x = 1.2^2 + (-0.5)^2 + 0.8^2 = 1.44 + 0.25 + 0.64 = 2.33.$

*Interpretation.* Each squared Normal contributes a non-negative
amount, so $\chi^2_3$ is supported on the positive half-line, as
required by (5.4); the mean of $\chi^2_3$ is $3$, and this single
observation of $2.33$ is a plausible draw below that mean.

**Example 3: Ratio method for $t_5$.**

*Problem.* Given $z=1.5$ and an independent $\chi^2_5$ draw $v=6.0$,
compute the resulting $t_5$ observation.

*Solution.*
$$
t = \frac{z}{\sqrt{v/\nu}} = \frac{1.5}{\sqrt{6.0/5}} = \frac{1.5}{\sqrt{1.2}}
= \frac{1.5}{1.0954} = 1.3693.
$$

*Interpretation.* Because $V$ is random, $T$ has heavier tails than a
standard Normal even though the numerator alone is standard Normal;
this particular draw is only moderately larger in magnitude than a
typical standard Normal value, consistent with $\nu=5$ already being
large enough that $t_5$ resembles the Normal reasonably closely (see
the tail comparison figure above).

**Example 4: Ratio method for $F_{3,5}$.**

*Problem.* Given $u=4.2$ (a $\chi^2_3$ draw) and $v=7.5$ (an
independent $\chi^2_5$ draw), compute the resulting $F_{3,5}$
observation.

*Solution.*
$$
f = \frac{u/d_1}{v/d_2} = \frac{4.2/3}{7.5/5} = \frac{1.4}{1.5} = 0.9333.
$$

*Interpretation.* Since $E(F_{3,5}) = d_2/(d_2-2) = 5/3 \approx
1.667$, this particular draw of $0.9333$ lies below the theoretical
mean, which is unsurprising given the right-skewness of the $F$
distribution (most of its probability mass lies at moderate values,
with a long right tail).

# Applications

- **Statistics.** Chi-square, $t$, and $F$ distributions are the
  reference distributions for goodness-of-fit tests, single-sample and
  two-sample mean comparisons, and the analysis of variance,
  respectively; simulation studies of the size and power of these
  tests require fast, exact generators for exactly these three
  distributions.
- **Machine Learning.** Gaussian mixture models, a standard tool for
  clustering and density estimation, are simulated directly by the
  composition method; compound and hierarchical generative models more
  broadly rely on the same two-stage logic.
- **Data Science.** Simulating synthetic datasets that mimic a
  heterogeneous population (e.g. customers belonging to distinct
  behavioural segments) is a direct application of the composition
  method.
- **Finance.** Regime-switching return models, in which market
  behaviour is drawn from one of several volatility regimes before
  returns are simulated, are compositions in exactly the sense of
  Section 5.2; $F$-type ratio statistics also appear in portfolio
  variance comparisons.
- **Reliability Engineering.** The convolution method generates Erlang
  (integer-shape Gamma) time-to-failure distributions as sums of
  independent Exponential component lifetimes in series reliability
  models; hyperexponential mixtures (composition method) model
  populations of components with heterogeneous failure rates.
- **Monte Carlo Methods.** Composition and convolution are frequently
  the fastest available exact generators, useful whenever a target
  distribution happens to admit either representation, avoiding the
  need for acceptance-rejection entirely.
- **Bayesian Statistics.** Hierarchical models — a prior on a
  parameter, followed by data generated conditional on that parameter
  — are, structurally, continuous-mixture (composition) models; prior
  predictive simulation is a direct application of Algorithm 5.1.
- **Operations Research.** Aggregate service or interarrival times
  built from sums of independent stage durations in multi-stage
  queueing networks are generated by the convolution method.
- **Engineering.** Simulating aggregate load or aggregate demand as a
  sum of independent component contributions is a direct convolution
  application.

# Advantages

- When a valid distributional relationship is known, the resulting
  generator is typically simpler to implement and to verify than a
  general-purpose acceptance-rejection scheme, since no bounding
  constant or envelope density needs to be found.
- The methods are exact: the generated variate has precisely the
  target distribution, with no approximation, as proved in Theorems
  5.1 and 5.2 and the derivations of Section 7.
- Composition and convolution generalize immediately to complicated
  hierarchical or compound structures that would be very difficult to
  describe by a single closed-form density, let alone invert or
  bound.

# Limitations

- Not every distribution admits a convenient distributional
  relationship; when none is known, the general-purpose methods of
  Units 3.3 and 3.4 remain necessary.
- The convolution method for Chi-square, as implemented in Algorithm
  5.2, has cost growing linearly in the degrees of freedom $\nu$,
  making it inefficient for large $\nu$ relative to a direct
  Gamma-based generator, which also extends to non-integer degrees of
  freedom that the convolution method cannot reach at all.
- The composition method requires the mixture decomposition to be
  known in advance; discovering that an apparently complicated density
  is secretly a mixture is a modelling question, not something the
  simulation method itself can resolve.

# Practical Tips

- **Generate all building blocks independently.** A very common
  implementation error is inadvertently reusing the same random draw
  for two supposedly independent components (for example, using the
  same uniform variate to select the mixture component and to
  generate within that component); this silently destroys the
  correctness guaranteed by Theorem 5.1.
- **Prefer vectorized generation over explicit loops in R.** As shown
  in `generate_chisq_convolution()`, drawing an entire matrix of
  Normal variates at once and summing rows is far faster in R than a
  loop generating one degree of freedom at a time.
- **Do not use the convolution method for large-parameter Chi-square
  or Gamma generation in production code.** Its computational cost
  scales linearly with the degrees of freedom, whereas dedicated
  Gamma generators (used internally by `rchisq()` and `rgamma()`) do
  not.
- **Verify mixture weights sum to one before simulating.** A silent
  normalization error in the component probabilities produces a
  mixture with the wrong proportions without raising any error.
- **When degrees of freedom are very small (e.g. $d_2=1$ or $d_2=2$
  for the $F$ distribution), check for numerical degeneracy.** Ratios
  of very small Chi-square draws can occasionally produce extreme
  values that are mathematically valid but numerically delicate in
  downstream calculations.

# Summary

- Many named distributions are defined by, or provably equivalent to,
  explicit combinations of simpler random variables; exploiting this
  structure often gives faster and simpler generators than
  general-purpose methods.
- The composition method simulates a mixture distribution by first
  selecting a component (or parameter), then simulating within it;
  Theorem 5.1 shows the resulting marginal distribution is exactly the
  intended mixture.
- The convolution method simulates a sum of independent random
  variables directly; Theorem 5.2 (Gamma additivity) justifies
  generating $\chi^2_\nu$ as a sum of $\nu$ independent $\chi^2_1$
  variables.
- Student's $t_\nu$ and $F_{d_1,d_2}$ are defined by explicit ratios
  of independent Normal and Chi-square variables; their densities
  (5.9) and (5.10) were derived directly from these ratio definitions
  using the same integral technique.
- These distributional-relationship generators are exact, but the
  convolution-based Chi-square generator in particular is
  computationally inferior to a direct Gamma-based generator for large
  degrees of freedom.

# Key Formulae

- Composition (discrete mixture): $f(x) = \sum_j p_j f_j(x)$.
- Convolution: $X = Y_1+\cdots+Y_n$, $Y_i$ independent.
- Gamma additivity: $\sum_i \text{Gamma}(\alpha_i,\beta) =
  \text{Gamma}(\sum_i\alpha_i,\beta)$.
- Chi-square: $\chi^2_\nu = \sum_{i=1}^\nu Z_i^2$,
  $f(x) = x^{\nu/2-1}e^{-x/2}/(2^{\nu/2}\Gamma(\nu/2))$.
- Student's $t$: $T = Z/\sqrt{V/\nu}$,
  $f(t) = \dfrac{\Gamma((\nu+1)/2)}{\sqrt{\pi\nu}\,\Gamma(\nu/2)}
  (1+t^2/\nu)^{-(\nu+1)/2}$.
- $F$ distribution: $F = (U/d_1)/(V/d_2)$,
  $f(x) = \dfrac{\Gamma((d_1+d_2)/2)}{\Gamma(d_1/2)\Gamma(d_2/2)}
  (d_1/d_2)^{d_1/2} x^{d_1/2-1}(1+d_1x/d_2)^{-(d_1+d_2)/2}$.

# Key Terms

- **Composition method.** Simulating from a mixture density by first
  selecting a component and then simulating within it.
- **Convolution.** The distribution of a sum of independent random
  variables.
- **Convolution method.** Simulating a target variable by generating
  and summing its independent constituent variables directly.
- **Distributional relationship.** A known probabilistic identity
  linking a target distribution to simpler, more easily simulated
  building blocks.
- **Mixture distribution.** A distribution arising from a two-stage
  random mechanism: select a component, then generate from within it.
- **Chi-square distribution.** The distribution of a sum of squared
  independent standard Normal variables.
- **Student's $t$ distribution.** The distribution of a standard
  Normal variable divided by the square root of an independent,
  appropriately scaled Chi-square variable.
- **$F$ distribution.** The distribution of the ratio of two
  independent, appropriately scaled Chi-square variables.

# Exercises

## Multiple Choice Questions

1. The composition method is most directly applicable when the target
   density can be written as
   (a) a product of two densities
   (b) a mixture of component densities
   (c) a quantile function
   (d) a moment generating function

2. The convolution method generates a target variable as
   (a) a ratio of independent variables
   (b) a product of independent variables
   (c) a sum of independent variables
   (d) a maximum of independent variables

3. By definition, $\chi^2_\nu$ is the distribution of
   (a) a single standard Normal variable
   (b) the sum of $\nu$ independent squared standard Normal variables
   (c) the ratio of two Normal variables
   (d) the product of $\nu$ Uniform variables

4. Student's $t_\nu$ is defined as
   (a) $Z \cdot \sqrt{V/\nu}$
   (b) $Z / \sqrt{V/\nu}$
   (c) $Z^2/V$
   (d) $V/Z$

5. The $F_{d_1,d_2}$ distribution is defined as
   (a) $(U/d_1)/(V/d_2)$
   (b) $(U \cdot d_1)/(V \cdot d_2)$
   (c) $U+V$
   (d) $U \cdot V$

6. Gamma additivity (Theorem 5.2) requires the summed Gamma variables
   to share
   (a) a common shape parameter
   (b) a common scale parameter
   (c) a common mean
   (d) no shared parameter at all

7. The convolution-based Chi-square generator has time complexity
   (a) $O(n)$
   (b) $O(n\nu)$
   (c) $O(n^2)$
   (d) $O(\log n)$

8. A common implementation error in the composition method is
   (a) using too many components
   (b) reusing the same random draw across independent stages
   (c) normalizing the mixture weights
   (d) using vectorized code

9. As degrees of freedom increase, the $t_\nu$ distribution
   (a) diverges from the standard Normal
   (b) converges toward the standard Normal
   (c) becomes discrete
   (d) becomes bimodal

10. The mean of $F_{d_1,d_2}$, when it exists, equals
    (a) $d_1/d_2$
    (b) $d_2/(d_2-2)$
    (c) $d_1 \cdot d_2$
    (d) $1$

**Answers.** 1(b), 2(c), 3(b), 4(b), 5(a), 6(b), 7(b), 8(b), 9(b),
10(b).

## Short Answer Questions

1. Explain, in your own words, why a distributional relationship can
   yield a faster generator than a general-purpose method.
2. State the composition method in two steps, using your own notation.
3. Why must the components summed in the convolution method be
   independent?
4. State the definition of the Chi-square distribution used in this
   unit, and explain why it is naturally suited to the convolution
   method.
5. Write down the ratio definition of Student's $t_\nu$ and identify
   the distribution of each of its two building blocks.
6. Write down the ratio definition of $F_{d_1,d_2}$ and identify the
   distribution of each of its two building blocks.
7. Explain why the convolution-based Chi-square generator becomes
   computationally unattractive for large degrees of freedom.
8. Give one example, distinct from those in this unit, of a
   distribution well suited to the composition method.
9. Explain, intuitively, why the $t$ distribution has heavier tails
   than the standard Normal.
10. Why does Theorem 5.2 require a *common* scale parameter across the
    summed Gamma variables?

## Long Answer Questions

1. State and prove Theorem 5.1 (correctness of the composition method)
   for both the discrete and continuous mixture cases.
2. State and prove Theorem 5.2 (Gamma additivity) and explain how it
   is used to justify the Chi-square convolution generator.
3. Derive the density of $\chi^2_1$ as the distribution of $Z^2$ for
   $Z\sim N(0,1)$, showing every differentiation step.
4. Derive the density of Student's $t_\nu$ from its ratio definition,
   showing every integration step.
5. Derive the density of $F_{d_1,d_2}$ from its ratio definition,
   showing every step of the bivariate transformation and integration.
6. Compare the composition, convolution, inverse transformation, and
   acceptance-rejection methods in terms of the information each
   requires about the target distribution.
7. Discuss the role of Chi-square, $t$, and $F$ distributions in
   classical hypothesis testing, and explain why fast exact generators
   for these distributions matter for simulation studies of test
   performance.
8. Explain how the composition method underlies simulation from
   hierarchical Bayesian models, using a concrete two-level example of
   your choosing.
9. Discuss the computational trade-off between the convolution-based
   Chi-square generator and a direct Gamma-based generator, including
   how each behaves as the degrees of freedom parameter grows.
10. Explain, using the tail-comparison figure in this unit, how and
    why the $t_\nu$ distribution converges to the standard Normal
    distribution as $\nu \to \infty$.

## Programming Exercises

1. Implement `generate_mixture()` from first principles (already
   provided) and reproduce the manual calculation of Example 1 by
   fixing the random seed appropriately.
2. Implement `generate_chisq_convolution()` from first principles
   (already provided) and verify, using a Kolmogorov-Smirnov test,
   that a sample of size $10{,}000$ with $\nu=6$ matches `rchisq()`.
3. Implement `generate_t_ratio()` and `generate_f_ratio()` from first
   principles (already provided) and reproduce Examples 3 and 4 by
   fixing the random seed appropriately.
4. Using `system.time()`, compare the speed of
   `generate_chisq_convolution()` against `rchisq()` for
   $\nu \in \{5, 50, 500\}$ and $n=100{,}000$, and discuss the trend.
5. Write an R function implementing the composition method for a
   three-component Normal mixture with user-specified means,
   standard deviations, and weights, and produce a histogram overlaid
   with the theoretical mixture density.
6. Write an R function that simulates a Negative Binomial count as a
   sum of independent Geometric trial counts (the convolution method),
   and verify the result against `rnbinom()` using summary statistics.
7. Empirically verify Theorem 5.2 by simulating sums of independent
   Gamma variables with different shape parameters but a common scale,
   and comparing the sample distribution of the sum to the theoretical
   Gamma distribution using a Q-Q plot.
8. Write a function that simulates an Erlang (integer-shape Gamma)
   waiting time as a sum of independent Exponential interarrival
   times, and use it to estimate the probability that a five-stage
   service process completes within a given time budget.
9. Reproduce the $t$-distribution tail-comparison figure of this unit
   for degrees of freedom $\nu \in \{1,2,5,10,30,100\}$, and comment
   on the rate of convergence to the Normal density.
10. Write an R function that simulates from $F_{d_1,d_2}$ using the
    ratio method and separately using R's built-in `rf()`, and compare
    their empirical densities using overlaid `ggplot2` density curves
    for several choices of $(d_1,d_2)$.

## Mini Projects

1. **Building a compound reliability model.** Model the lifetime of a
   system composed of two sub-populations of components: a "weak"
   sub-population (Exponential with a high failure rate, representing
   early failures) and a "strong" sub-population (Exponential with a
   low failure rate), combined via the composition method with a
   chosen mixing weight. Simulate a large sample of system lifetimes,
   estimate the empirical hazard function, and discuss, with
   reference to your simulation, why a mixture of Exponentials can
   produce a hazard function that decreases over time even though each
   individual sub-population has a constant hazard rate.

2. **Simulation study of classical test statistics.** Using the
   Chi-square, $t$, and $F$ generators developed in this unit, conduct
   a simulation study of the size (Type I error rate) of a one-sample
   $t$-test, a Chi-square goodness-of-fit test, and a one-way ANOVA
   $F$-test, each under its null hypothesis, across several sample
   sizes. Report the empirical rejection rate at the nominal $5\%$
   level for each test and sample size, and discuss any discrepancies
   between the nominal and empirical size.

**Difficulty labels.** Multiple choice, short answer: **Easy**. Long
answer questions 3, 6, 7, 8, programming exercises 1-6: **Medium**.
Long answer questions 1, 2, 4, 5, 9, 10, programming exercises 7-10:
**Medium to Advanced**. Mini projects: **Advanced**.

# References

Casella, G., & Berger, R. L. (2002). *Statistical inference* (2nd
ed.). Duxbury.

Devroye, L. (1986). *Non-uniform random variate generation*.
Springer-Verlag.

Fisher, R. A. (1925). *Statistical methods for research workers*.
Oliver and Boyd.

Pearson, K. (1900). On the criterion that a given system of
deviations from the probable in the case of a correlated system of
variables is such that it can be reasonably supposed to have arisen
from random sampling. *Philosophical Magazine*, 50(302), 157-175.

Rizzo, M. L. (2019). *Statistical computing with R* (2nd ed.). CRC
Press.

Ross, S. M. (2020). *Simulation* (6th ed.). Academic Press.

Student [Gosset, W. S.] (1908). The probable error of a mean.
*Biometrika*, 6(1), 1-25.
