---
title: "Unit 4.1: Resampling Techniques"
author: "Dr. Manoj C. Patil"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
    code_folding: hide
    theme: cosmo
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    echo = TRUE,
    warning = FALSE,
    message = FALSE,
    fig.align = "center",
    fig.width = 7,
    fig.height = 4.5
)

set.seed(123)
library(tidyverse)
library(ggplot2)
library(dplyr)
library(knitr)
library(kableExtra)
library(gridExtra)
library(patchwork)
```

# Learning Objectives

By the end of this unit, a student should be able to

- explain the bootstrap principle and justify why resampling from the
  empirical distribution can approximate the sampling distribution of
  a statistic,
- derive and implement the bootstrap estimates of bias and standard
  error for an arbitrary statistic,
- construct bootstrap confidence intervals using the normal
  approximation and percentile methods, and state their limitations,
- derive and implement the jackknife estimates of bias and standard
  error, and prove the jackknife bias-correction identity for a smooth
  statistic,
- compare the bootstrap and jackknife as complementary resampling
  techniques, and select an appropriate technique for a given
  statistic,
- implement all four procedures in R and validate them by simulation
  against known theoretical benchmarks.

# Prerequisites

Students should be comfortable with

- Units 3.1-3.6 (particularly the generation of pseudo-random samples,
  since resampling relies on the same uniform-generation machinery),
- the concepts of bias, standard error, and sampling distribution from
  a first course in mathematical statistics,
- the law of large numbers and the central limit theorem,
- basic Taylor series expansion,
- R programming, including the `sample()` function and writing
  functions that operate on data vectors.

# Introduction

Classical statistical inference typically requires a formula for the
standard error of an estimator, derived analytically from the
assumed sampling distribution — the formula $s/\sqrt n$ for the
standard error of a sample mean is the most familiar example. For many
statistics of practical interest, however, no such closed-form formula
exists, or exists only under restrictive distributional assumptions
that may not hold in a given application. The sample median, a
trimmed mean, a correlation coefficient, or a ratio of two sample
means are all routinely used in practice, yet none has a standard
error formula as simple and universally applicable as the one for the
mean.

Resampling methods resolve this difficulty by using the computer,
rather than a closed-form formula, to approximate the sampling
distribution of a statistic. The **bootstrap**, introduced by Bradley
Efron in his landmark 1979 paper, treats the observed sample itself as
a stand-in for the population, and approximates the sampling
distribution of a statistic by repeatedly resampling, with
replacement, from the observed data and recomputing the statistic on
each resample. The **jackknife**, developed earlier by Maurice
Quenouille (1949, 1956) for bias correction and popularized by John
Tukey (1958) for variance estimation, takes a complementary approach:
rather than resampling with replacement, it systematically leaves out
one observation at a time and examines how the statistic changes.

The practical importance of these methods is difficult to overstate in
modern statistical computing. Bootstrap confidence intervals are now a
standard tool for regression coefficients, machine learning model
performance metrics, and virtually any statistic for which an
analytical standard error is unavailable or untrustworthy. The
jackknife remains valuable both as a fast, deterministic alternative
to the bootstrap and, historically, as the conceptual precursor from
which the bootstrap itself was developed. This unit develops both
methods in full: their theoretical justification, their exact
mechanics, and their implementation in R.

# Intuition

The central difficulty in estimating the sampling variability of a
statistic is that, in practice, only *one* sample is ever observed —
there is no way to draw many independent samples from the true
population to see how much the statistic varies from sample to
sample. The bootstrap's resolution to this problem is a deceptively
simple substitution: since the observed sample is itself the best
available estimate of the population, *pretend* the observed sample
**is** the population, and draw many new samples from it. Because
resampling from a finite population of $n$ observed values is done
"with replacement" (each new observation drawn uniformly from the $n$
observed values, independently, so the same original observation can
appear multiple times or not at all in a given resample), each
resample looks like a fresh sample of the same size from the empirical
distribution, and the spread of the statistic across many such
resamples imitates the spread that would be seen across many fresh
real samples.

The jackknife takes a more surgical approach to the same underlying
question. Instead of asking "what would happen if I drew an entirely
new random sample," the jackknife asks the narrower question "how
sensitive is my statistic to the removal of any single observation."
By recomputing the statistic $n$ times, once with each observation
left out in turn, the jackknife builds up a picture of how much
influence each individual data point has, and this picture turns out
to be enough, for smooth statistics, to estimate both the bias and the
standard error of the original statistic — using only $n$
recomputations rather than the thousands typically used by the
bootstrap.

```{r intuition-figure, echo=FALSE, fig.cap="A bootstrap resample (top) draws n = 8 values with replacement from the original sample, so some original points appear more than once and others not at all; a jackknife sample (bottom) simply omits one observation at a time."}
set.seed(123)
original_data <- c(3.1, 5.4, 2.2, 7.8, 4.5, 6.1, 3.9, 5.0)
n_demo <- length(original_data)

bootstrap_resample <- sample(original_data, size = n_demo, replace = TRUE)

demo_df <- tibble(
  index = rep(1:n_demo, 2),
  value = c(original_data, bootstrap_resample),
  type = rep(c("Original sample", "One bootstrap resample"), each = n_demo)
)

ggplot(demo_df, aes(x = index, y = value)) +
  geom_point(size = 3, colour = "steelblue") +
  geom_segment(aes(xend = index, y = 0, yend = value), colour = "steelblue", alpha = 0.5) +
  facet_wrap(~type, ncol = 1) +
  labs(title = "Original sample vs. a single bootstrap resample", x = "Position", y = "Value") +
  theme_minimal(base_size = 11)
```

# Formal Definitions

## The Empirical Distribution and the Bootstrap Principle

**Definition.** For an observed sample $x_1,\dots,x_n$, the
**empirical distribution** $\hat F$ places probability mass $1/n$ on
each observed value $x_i$. A **bootstrap sample**
$x_1^{*},\dots,x_n^{*}$ is a sample of size $n$ drawn independently
with replacement from $\hat F$ — equivalently, each $x_i^{*}$ is drawn
uniformly at random from $\{x_1,\dots,x_n\}$.

**The bootstrap principle.** For a statistic $\hat\theta =
T(x_1,\dots,x_n)$ computed from the original sample, the sampling
distribution of $\hat\theta$ under the true (unknown) population $F$
is approximated by the distribution of $\hat\theta^{*} =
T(x_1^{*},\dots,x_n^{*})$ across repeated draws of bootstrap samples
from $\hat F$, since $\hat F$ is itself a consistent estimate of $F$.

## Bootstrap Estimates of Bias and Standard Error

**Definition.** Given $B$ independent bootstrap samples, each yielding
a bootstrap replicate $\hat\theta^{*}_b = T(x_1^{*(b)},\dots,
x_n^{*(b)})$, $b=1,\dots,B$, the **bootstrap estimate of standard
error** is

$$
\widehat{\text{se}}_{\text{boot}} = \sqrt{\frac{1}{B-1}
\sum_{b=1}^{B} \Bigl(\hat\theta^{*}_b - \bar\theta^{*}\Bigr)^2},
\qquad \bar\theta^{*} = \frac1B\sum_{b=1}^B \hat\theta^{*}_b,
\tag{1.1}
$$

and the **bootstrap estimate of bias** is

$$
\widehat{\text{bias}}_{\text{boot}} = \bar\theta^{*} - \hat\theta.
\tag{1.2}
$$

## Jackknife Estimates of Bias and Standard Error

**Definition.** Let $\hat\theta_{(i)} = T(x_1,\dots,x_{i-1},
x_{i+1},\dots,x_n)$ denote the statistic recomputed with the $i$-th
observation deleted, for $i=1,\dots,n$, and let
$\bar\theta_{(\cdot)} = n^{-1}\sum_{i=1}^n \hat\theta_{(i)}$. The
**jackknife estimate of bias** is

$$
\widehat{\text{bias}}_{\text{jack}} = (n-1)\bigl(\bar\theta_{(\cdot)} - \hat\theta\bigr),
\tag{1.3}
$$

and the **jackknife estimate of standard error** is

$$
\widehat{\text{se}}_{\text{jack}} = \sqrt{\frac{n-1}{n}
\sum_{i=1}^{n}\bigl(\hat\theta_{(i)} - \bar\theta_{(\cdot)}\bigr)^2}.
\tag{1.4}
$$

The **jackknife bias-corrected estimator** is

$$
\hat\theta_{\text{jack}} = \hat\theta - \widehat{\text{bias}}_{\text{jack}}
= n\hat\theta - (n-1)\bar\theta_{(\cdot)}.
\tag{1.5}
$$

# Theorems

## Theorem 1.1: Consistency of the Bootstrap Standard Error Estimate

**Statement.** Conditional on the observed data $x_1,\dots,x_n$
(hence conditional on $\hat F$), as the number of bootstrap
replications $B\to\infty$,

$$
\widehat{\text{se}}_{\text{boot}} \;\xrightarrow{\text{a.s.}}\;
\text{se}_{\hat F}(\hat\theta) := \sqrt{\operatorname{Var}_{\hat F}\bigl(T(X_1^{*},\dots,X_n^{*})\bigr)},
$$

the exact standard deviation of $T$ under repeated sampling from
$\hat F$ (the "ideal bootstrap" standard error).

**Assumptions.** $\operatorname{Var}_{\hat F}(\hat\theta^{*}) <
\infty$; the bootstrap replicates $\hat\theta^{*}_1,\dots,
\hat\theta^{*}_B$ are, conditional on the data, independent and
identically distributed draws from the distribution of $T$ under
resampling from $\hat F$ (by construction, since each bootstrap sample
is drawn independently).

**Proof.** Conditional on the data, $\hat\theta^{*}_1,
\dots,\hat\theta^{*}_B$ are i.i.d. with finite variance
$\sigma^2_{\hat F} = \operatorname{Var}_{\hat F}(\hat\theta^{*})$. By
the strong law of large numbers applied to $\hat\theta^{*}_b$ and to
$(\hat\theta^{*}_b)^2$ separately,

$$
\bar\theta^{*} \xrightarrow{\text{a.s.}} E_{\hat F}(\hat\theta^{*}),
\qquad
\frac1B\sum_{b=1}^B (\hat\theta^{*}_b)^2 \xrightarrow{\text{a.s.}} E_{\hat F}\bigl((\hat\theta^{*})^2\bigr),
$$

as $B\to\infty$. The sample variance in (1.1) (with the usual
$B/(B-1)$ finite-sample correction, which tends to $1$) is a
continuous function of these two quantities,
$\widehat{\text{se}}^2_{\text{boot}} \to E_{\hat F}((\hat\theta^{*})^2)
- \bigl(E_{\hat F}(\hat\theta^{*})\bigr)^2 = \sigma^2_{\hat F}$, by the
continuous mapping theorem. Taking square roots (a continuous
function) preserves almost sure convergence, giving the result.
$\blacksquare$

**Remarks.** This theorem addresses only the "Monte Carlo error"
introduced by using a finite number of bootstrap replications $B$; it
says nothing about how close $\text{se}_{\hat F}(\hat\theta)$ itself
is to the true sampling standard error under the unknown population
$F$. That second, statistical (as opposed to computational) source of
error is controlled separately by the consistency of $\hat F$ for $F$
as $n\to\infty$, a different and generally harder question addressed
in the bootstrap consistency literature (Efron, 1979; Bickel &
Freedman, 1981).

**Examples.** Verified by simulation in Section 10.

**Applications.** This theorem justifies the practical rule of thumb
that $B=1000$ to $B=2000$ bootstrap replications are usually
sufficient for standard error estimation (where only the Monte Carlo
error in (1.1) matters), while confidence interval estimation, which
depends on the tails of the bootstrap distribution rather than only
its variance, typically requires a larger $B$ for the same accuracy.

## Theorem 1.2: The Jackknife Removes the Leading Bias Term

**Statement.** Suppose the bias of $\hat\theta_n = T(x_1,\dots,x_n)$,
as an estimator of $\theta$, admits the expansion

$$
E(\hat\theta_n) = \theta + \frac{a}{n} + O\Bigl(\frac{1}{n^2}\Bigr)
\tag{1.6}
$$

for some constant $a$ not depending on $n$. Then the jackknife
bias-corrected estimator (1.5) satisfies

$$
E(\hat\theta_{\text{jack}}) = \theta + O\Bigl(\frac{1}{n^2}\Bigr),
$$

so the jackknife reduces the order of the bias from $O(1/n)$ to
$O(1/n^2)$.

**Assumptions.** The expansion (1.6) holds not only for $\hat\theta_n$
itself but, with the same leading coefficient $a$, for the
delete-one statistics $\hat\theta_{(i)}$ viewed as an instance of the
same estimator computed on a sample of size $n-1$, i.e.
$E(\hat\theta_{(i)}) = \theta + a/(n-1) + O(1/n^2)$ for each $i$; this
holds for smooth statistics under standard regularity conditions
(Efron, 1982).

**Proof.** Taking expectations in (1.5) and using linearity,

$$
E(\hat\theta_{\text{jack}}) = n\,E(\hat\theta_n) - (n-1)\,E(\bar\theta_{(\cdot)}).
$$

By the assumption, $E(\hat\theta_{(i)}) = \theta + a/(n-1) + O(1/n^2)$
for every $i$, so $E(\bar\theta_{(\cdot)}) = \theta + a/(n-1) +
O(1/n^2)$ as well (the average of $n$ terms each with the same
expansion has that same expansion). Substituting both expansions,

$$
\begin{aligned}
E(\hat\theta_{\text{jack}})
&= n\Bigl(\theta + \frac{a}{n} + O\bigl(\tfrac1{n^2}\bigr)\Bigr)
- (n-1)\Bigl(\theta + \frac{a}{n-1} + O\bigl(\tfrac1{n^2}\bigr)\Bigr) \\
&= \bigl(n\theta + a + O(\tfrac1n)\bigr) - \bigl((n-1)\theta + a + O(\tfrac1n)\bigr) \\
&= \theta + O\Bigl(\frac1n\Bigr)\cdot n - O\Bigl(\frac1n\Bigr)\cdot(n-1)
\end{aligned}
$$

where multiplying the $O(1/n^2)$ remainder terms by $n$ and by
$n-1$ respectively leaves an $O(1/n)$ contribution from each; a more
careful accounting (retaining the next term $b/n^2$ in (1.6)
explicitly) shows these two $O(1/n)$-after-multiplication contributions
in fact cancel to leading order, leaving a net remainder of
$O(1/n^2)$ — the full accounting is a routine but notationally heavy
exercise in collecting terms, carried out in full in Efron (1982,
Ch. 2). The leading terms $n\theta$ and $-(n-1)\theta$ combine to
$\theta$, and the leading terms $a$ and $-a$ cancel exactly,
confirming that both the $O(1)$ and the $O(1/n)$ parts of the bias are
eliminated, leaving $E(\hat\theta_{\text{jack}}) = \theta +
O(1/n^2)$. $\blacksquare$

**Remarks.** The exact cancellation of the $a$ coefficient — not just
its order of magnitude — is the key mechanism: the jackknife estimate
of bias, $(n-1)(\bar\theta_{(\cdot)}-\hat\theta_n)$, is constructed
precisely so that it converges to $a/n$, the leading bias term itself,
which is why subtracting it removes exactly that term. For a small
number of classical statistics (including the variance estimator
worked out in Example 2 below), the bias expansion (1.6) is *exact*
with no remainder at all, so the jackknife correction removes the bias
completely rather than merely to a higher order.

**Examples.** A fully exact worked example (no remainder term at all)
is given in Section 12, Example 2.

**Applications.** Bias correction of this kind is routinely used for
ratio estimators in survey sampling, for maximum likelihood variance
estimators, and for correlation and $R^2$-type statistics in
regression, all of which have bias of exactly the form (1.6) to
leading order.

# Mathematical Derivations

## The Jackknife Standard Error Reduces to the Classical Formula for the Mean

For the sample mean, $\hat\theta_n = \bar x = n^{-1}\sum_i x_i$, the
delete-one mean is $\hat\theta_{(i)} = (n\bar x - x_i)/(n-1)$, obtained
by removing $x_i$ from the total and dividing by the remaining $n-1$
observations. First,

$$
\hat\theta_{(i)} - \bar\theta_{(\cdot)}
= \frac{n\bar x - x_i}{n-1} - \bar x
= \frac{n\bar x - x_i - (n-1)\bar x}{n-1}
= \frac{\bar x - x_i}{n-1} = -\frac{x_i - \bar x}{n-1},
\tag{1.7}
$$

using $\bar\theta_{(\cdot)} = \bar x$ exactly (an identity verified
directly: $\sum_i \hat\theta_{(i)} = \sum_i (n\bar x - x_i)/(n-1)
= (n^2\bar x - n\bar x)/(n-1) = n\bar x$, so
$\bar\theta_{(\cdot)}=\bar x$). Substituting (1.7) into the jackknife
standard error formula (1.4),

$$
\widehat{\text{se}}_{\text{jack}}^2 = \frac{n-1}{n}\sum_{i=1}^n
\frac{(x_i-\bar x)^2}{(n-1)^2}
= \frac{1}{n(n-1)}\sum_{i=1}^n (x_i-\bar x)^2 = \frac{s^2}{n},
\tag{1.8}
$$

where $s^2 = (n-1)^{-1}\sum_i (x_i-\bar x)^2$ is the usual sample
variance. Hence $\widehat{\text{se}}_{\text{jack}} = s/\sqrt n$ exactly
reproduces the classical standard error formula for the mean — a
useful sanity check confirming the jackknife formula (1.4) is
correctly normalized.

## Types of Bootstrap Confidence Intervals

**Normal-approximation interval.** Treating $\hat\theta$ as
approximately Normally distributed with standard error
$\widehat{\text{se}}_{\text{boot}}$ gives

$$
\hat\theta \pm z_{1-\alpha/2}\, \widehat{\text{se}}_{\text{boot}}.
\tag{1.9}
$$

This is simple but ignores any skewness in the bootstrap distribution
of $\hat\theta^{*}$.

**Percentile interval.** Using the empirical $\alpha/2$ and
$1-\alpha/2$ quantiles of the bootstrap replicates
$\hat\theta^{*}_1,\dots,\hat\theta^{*}_B$ directly,

$$
\bigl(\hat\theta^{*}_{(\lceil B\alpha/2\rceil)},\;
\hat\theta^{*}_{(\lceil B(1-\alpha/2)\rceil)}\bigr),
\tag{1.10}
$$

where $\hat\theta^{*}_{(k)}$ denotes the $k$-th order statistic of the
bootstrap replicates. The percentile interval automatically adapts to
skewness in the bootstrap distribution and is transformation-respecting
(applying a monotone transformation to $\hat\theta$ and then computing
a percentile interval gives the same interval, transformed, as
computing the percentile interval first and then transforming its
endpoints) — a property the normal-approximation interval lacks.

# Algorithms

## Algorithm 1.1: Bootstrap Estimation of Bias and Standard Error

**Purpose.** Estimate the bias and standard error of an arbitrary
statistic without an analytical formula.

**Input.** Sample $x_1,\dots,x_n$; statistic function $T(\cdot)$;
number of bootstrap replications $B$.

**Output.** $\widehat{\text{bias}}_{\text{boot}}$,
$\widehat{\text{se}}_{\text{boot}}$.

**Step-by-step procedure.**

1. Compute $\hat\theta = T(x_1,\dots,x_n)$.
2. For $b=1,\dots,B$: draw a bootstrap sample of size $n$ with
   replacement from $\{x_1,\dots,x_n\}$; compute
   $\hat\theta^{*}_b = T(\cdot)$ on this resample.
3. Compute $\widehat{\text{bias}}_{\text{boot}}$ via (1.2) and
   $\widehat{\text{se}}_{\text{boot}}$ via (1.1).

**Pseudo-code.**

```
Algorithm: Bootstrap Bias and Standard Error
Input:  x[1..n], T(), B
Output: bias_boot, se_boot

1. theta_hat <- T(x)
2. For b = 1 to B:
       x_star <- sample(x, size = n, replace = TRUE)
       theta_star[b] <- T(x_star)
3. bias_boot <- mean(theta_star) - theta_hat
4. se_boot <- sd(theta_star)
5. Return bias_boot, se_boot
```

**Flowchart description.** Start; compute the statistic on the
original sample; repeat $B$ times: draw a with-replacement resample of
size $n$, recompute the statistic; after the loop, compute the mean
and standard deviation of the $B$ replicates; report bias and standard
error; end.

**Complexity.** $O(Bn)$ for the resampling itself, plus $O(B\cdot
c(T))$ where $c(T)$ is the cost of one evaluation of the statistic.

**Practical considerations.** $B=1000$-$2000$ is generally adequate
for standard error estimation (Theorem 1.1); confidence intervals,
especially percentile intervals, benefit from larger $B$ (often
$B\ge 2000$-$5000$) since they depend on tail quantiles rather than
only the overall spread.

## Algorithm 1.2: Bootstrap Percentile Confidence Interval

**Purpose.** Construct a confidence interval for $\theta$ that adapts
to skewness in the bootstrap distribution.

**Input.** Sample $x_1,\dots,x_n$; statistic function $T(\cdot)$;
number of replications $B$; confidence level $1-\alpha$.

**Output.** Interval $(\hat\theta^{*}_{\text{lower}},
\hat\theta^{*}_{\text{upper}})$.

**Step-by-step procedure.**

1. Generate $B$ bootstrap replicates $\hat\theta^{*}_1,\dots,
   \hat\theta^{*}_B$ as in Algorithm 1.1.
2. Sort the replicates and take the empirical $\alpha/2$ and
   $1-\alpha/2$ quantiles as in (1.10).

**Pseudo-code.**

```
Algorithm: Bootstrap Percentile Interval
Input:  x[1..n], T(), B, alpha
Output: (lower, upper)

1. Generate theta_star[1..B] as in Algorithm 1.1
2. lower  <- quantile(theta_star, alpha / 2)
3. upper  <- quantile(theta_star, 1 - alpha / 2)
4. Return (lower, upper)
```

**Flowchart description.** Start; generate $B$ bootstrap replicates;
sort them; read off the empirical quantiles at the two tail
probabilities; report the resulting interval; end.

**Complexity.** $O(Bn + B\log B)$, the additional $B\log B$ coming
from sorting the replicates.

**Practical considerations.** The percentile interval can perform
poorly (be biased or too narrow) when the bootstrap distribution of
$\hat\theta^{*}$ is itself substantially biased or skewed relative to
$\theta$; the bias-corrected and accelerated (BCa) interval (Efron,
1987) addresses this at the cost of additional computation, and is
recommended over the plain percentile interval whenever high accuracy
is required.

## Algorithm 1.3: Jackknife Estimation of Bias and Standard Error

**Purpose.** Estimate bias and standard error using $n$ deterministic
leave-one-out recomputations rather than random resampling.

**Input.** Sample $x_1,\dots,x_n$; statistic function $T(\cdot)$.

**Output.** $\widehat{\text{bias}}_{\text{jack}}$,
$\widehat{\text{se}}_{\text{jack}}$.

**Step-by-step procedure.**

1. Compute $\hat\theta = T(x_1,\dots,x_n)$.
2. For $i=1,\dots,n$: compute
   $\hat\theta_{(i)} = T(x_1,\dots,x_{i-1},x_{i+1},\dots,x_n)$.
3. Compute $\bar\theta_{(\cdot)}$, then
   $\widehat{\text{bias}}_{\text{jack}}$ via (1.3) and
   $\widehat{\text{se}}_{\text{jack}}$ via (1.4).

**Pseudo-code.**

```
Algorithm: Jackknife Bias and Standard Error
Input:  x[1..n], T()
Output: bias_jack, se_jack

1. theta_hat <- T(x)
2. For i = 1 to n:
       theta_i[i] <- T(x with element i removed)
3. theta_bar <- mean(theta_i)
4. bias_jack <- (n - 1) * (theta_bar - theta_hat)
5. se_jack <- sqrt((n - 1) / n * sum((theta_i - theta_bar)^2))
6. Return bias_jack, se_jack
```

**Flowchart description.** Start; compute the statistic on the full
sample; for each observation in turn, remove it and recompute the
statistic; after all $n$ recomputations, compute their mean and sum of
squared deviations; report bias and standard error; end.

**Complexity.** $O(n\cdot c(T))$, exactly $n$ evaluations of the
statistic, with no random resampling required.

**Practical considerations.** The jackknife can fail for
non-smooth statistics — most notoriously the sample median, for which
the jackknife standard error estimate is inconsistent (it does not
converge to the correct value even as $n\to\infty$); the bootstrap
does not share this defect and is generally preferred for such
statistics.

# R Programming

```{r resampling-functions}
set.seed(123)

#' Bootstrap estimates of bias and standard error
#'
#' Purpose: implements Algorithm 1.1.
#'
#' Arguments:
#'   sample_data          numeric vector, the observed sample
#'   statistic_function   function mapping a numeric vector to a scalar
#'   number_of_replicates number of bootstrap resamples (default 2000)
#'
#' Returns:
#'   a list with the original estimate, bootstrap bias, bootstrap
#'   standard error, and the full vector of bootstrap replicates
#'
#' Example:
#'   bootstrap_estimate(rnorm(30), mean, number_of_replicates = 1000)
bootstrap_estimate <- function(sample_data, statistic_function, number_of_replicates = 2000) {
  sample_size <- length(sample_data)
  original_estimate <- statistic_function(sample_data)

  bootstrap_replicates <- numeric(number_of_replicates)
  for (replicate_index in seq_len(number_of_replicates)) {
    resampled_data <- sample(sample_data, size = sample_size, replace = TRUE)
    bootstrap_replicates[replicate_index] <- statistic_function(resampled_data)
  }

  list(
    estimate = original_estimate,
    bootstrap_bias = mean(bootstrap_replicates) - original_estimate,
    bootstrap_se = sd(bootstrap_replicates),
    replicates = bootstrap_replicates
  )
}

#' Bootstrap percentile confidence interval
#'
#' Purpose: implements Algorithm 1.2.
#'
#' Arguments:
#'   sample_data          numeric vector, the observed sample
#'   statistic_function   function mapping a numeric vector to a scalar
#'   number_of_replicates number of bootstrap resamples (default 2000)
#'   confidence_level     confidence level (default 0.95)
#'
#' Returns:
#'   a named numeric vector with the lower and upper interval endpoints
#'
#' Example:
#'   bootstrap_percentile_ci(rnorm(30), mean)
bootstrap_percentile_ci <- function(sample_data, statistic_function,
                                     number_of_replicates = 2000, confidence_level = 0.95) {
  boot_result <- bootstrap_estimate(sample_data, statistic_function, number_of_replicates)
  alpha <- 1 - confidence_level

  quantile(boot_result$replicates, probs = c(alpha / 2, 1 - alpha / 2))
}

#' Jackknife estimates of bias and standard error
#'
#' Purpose: implements Algorithm 1.3.
#'
#' Arguments:
#'   sample_data          numeric vector, the observed sample
#'   statistic_function   function mapping a numeric vector to a scalar
#'
#' Returns:
#'   a list with the original estimate, jackknife bias, jackknife
#'   standard error, the bias-corrected estimate, and the vector of
#'   leave-one-out replicates
#'
#' Example:
#'   jackknife_estimate(rnorm(30), mean)
jackknife_estimate <- function(sample_data, statistic_function) {
  sample_size <- length(sample_data)
  original_estimate <- statistic_function(sample_data)

  leave_one_out_replicates <- numeric(sample_size)
  for (index in seq_len(sample_size)) {
    leave_one_out_replicates[index] <- statistic_function(sample_data[-index])
  }

  mean_leave_one_out <- mean(leave_one_out_replicates)
  jackknife_bias <- (sample_size - 1) * (mean_leave_one_out - original_estimate)
  jackknife_se <- sqrt(
    (sample_size - 1) / sample_size *
      sum((leave_one_out_replicates - mean_leave_one_out)^2)
  )

  list(
    estimate = original_estimate,
    jackknife_bias = jackknife_bias,
    jackknife_se = jackknife_se,
    bias_corrected_estimate = original_estimate - jackknife_bias,
    replicates = leave_one_out_replicates
  )
}
```

# Simulation Studies

## Bootstrap Standard Error Convergence (Theorem 1.1)

```{r bootstrap-convergence-simulation}
set.seed(123)

fixed_sample <- rnorm(30, mean = 5, sd = 2)
true_theoretical_se <- sd(fixed_sample) / sqrt(length(fixed_sample))

replication_counts <- c(10, 50, 200, 500, 1000, 5000, 20000)
se_estimates <- vapply(
  replication_counts,
  function(B) bootstrap_estimate(fixed_sample, mean, number_of_replicates = B)$bootstrap_se,
  numeric(1)
)

tibble(B = replication_counts, Bootstrap_SE = se_estimates, Classical_SE = true_theoretical_se) %>%
  kable(digits = 4, caption = "Bootstrap SE of the mean converges to the classical formula s/sqrt(n) as B grows")
```

## Jackknife Bias Correction for the Variance Estimator

For $X_1,\dots,X_n$ i.i.d. Normal$(\mu,\sigma^2)$, the plug-in
(divide-by-$n$) variance estimator
$\hat\sigma^2_n = n^{-1}\sum_i (X_i-\bar X)^2$ has the exact bias
$E(\hat\sigma^2_n) - \sigma^2 = -\sigma^2/n$, matching the form of
(1.6) exactly with $a=-\sigma^2$ and no remainder term.

```{r jackknife-variance-bias-simulation}
set.seed(123)

plugin_variance <- function(x) mean((x - mean(x))^2)

true_sigma_squared <- 4
number_of_datasets <- 3000
sample_size_n <- 15

bias_estimates <- replicate(number_of_datasets, {
  simulated_sample <- rnorm(sample_size_n, mean = 0, sd = sqrt(true_sigma_squared))
  jackknife_estimate(simulated_sample, plugin_variance)$jackknife_bias
})

tibble(
  Quantity = c("Theoretical bias (-sigma^2 / n)", "Average jackknife bias estimate"),
  Value = c(-true_sigma_squared / sample_size_n, mean(bias_estimates))
) %>%
  kable(digits = 4, caption = "Jackknife bias estimate vs. the exact theoretical bias of the plug-in variance estimator")
```

## Coverage of the Bootstrap Percentile Confidence Interval

```{r bootstrap-coverage-simulation}
set.seed(123)

number_of_trials <- 500
sample_size_coverage <- 25
true_mean <- 10
coverage_indicator <- logical(number_of_trials)

for (trial in seq_len(number_of_trials)) {
  trial_sample <- rnorm(sample_size_coverage, mean = true_mean, sd = 3)
  interval <- bootstrap_percentile_ci(trial_sample, mean, number_of_replicates = 500)
  coverage_indicator[trial] <- (interval[1] <= true_mean) && (true_mean <= interval[2])
}

tibble(
  Quantity = c("Nominal coverage", "Empirical coverage (500 trials)"),
  Value = c(0.95, mean(coverage_indicator))
) %>%
  kable(digits = 4, caption = "Empirical coverage of the 95% bootstrap percentile interval for the mean")
```

# Graphs

```{r bootstrap-distribution-plot, fig.cap="Bootstrap distribution of the sample mean, with the normal-approximation and percentile 95% intervals marked."}
set.seed(123)

boot_result_plot <- bootstrap_estimate(fixed_sample, mean, number_of_replicates = 3000)
normal_ci <- boot_result_plot$estimate + c(-1, 1) * qnorm(0.975) * boot_result_plot$bootstrap_se
percentile_ci <- quantile(boot_result_plot$replicates, probs = c(0.025, 0.975))

ggplot(tibble(theta_star = boot_result_plot$replicates), aes(x = theta_star)) +
  geom_histogram(aes(y = after_stat(density)), bins = 40, fill = "steelblue", colour = "white") +
  geom_vline(xintercept = normal_ci, colour = "darkred", linetype = "dashed", linewidth = 0.8) +
  geom_vline(xintercept = percentile_ci, colour = "darkgreen", linetype = "dotted", linewidth = 0.8) +
  labs(
    title = "Bootstrap distribution of the sample mean (n = 30)",
    subtitle = "Dashed red = normal-approximation CI; dotted green = percentile CI",
    x = expression(hat(theta)^"*"), y = "Density"
  ) +
  theme_minimal(base_size = 11)
```

```{r jackknife-leave-one-out-plot, fig.cap="Leave-one-out means for the illustrative dataset of Section 12, showing the influence of each observation."}
illustrative_data <- c(5, 7, 9, 12, 14)
jackknife_illustrative <- jackknife_estimate(illustrative_data, mean)

ggplot(
  tibble(observation_removed = illustrative_data, leave_one_out_mean = jackknife_illustrative$replicates),
  aes(x = factor(observation_removed), y = leave_one_out_mean)
) +
  geom_col(fill = "darkorange", width = 0.6) +
  geom_hline(yintercept = jackknife_illustrative$estimate, colour = "black", linetype = "dashed") +
  labs(
    title = "Leave-one-out sample means",
    x = "Observation removed", y = "Mean of remaining 4 observations"
  ) +
  theme_minimal(base_size = 11)
```

# Numerical Examples

**Example 1: Bootstrap and jackknife standard error of the mean.**

*Problem.* For the sample $x = (5, 7, 9, 12, 14)$ ($n=5$), compute the
classical, jackknife, and (conceptually) bootstrap standard errors of
the sample mean.

*Solution.* $\bar x = 47/5 = 9.4$. The sample variance is
$s^2 = 53.2/4 = 13.3$, giving the classical standard error
$s/\sqrt5 = 3.6469/2.2361 = 1.6309$.

For the jackknife, the leave-one-out means are $10.5, 10.0, 9.5,
8.75, 8.25$ (removing $5,7,9,12,14$ respectively), with
$\bar\theta_{(\cdot)} = 9.4$ (exactly $\bar x$, as shown in Section
8.1). The sum of squared deviations of the leave-one-out means from
$9.4$ is $1.1^2+0.6^2+0.1^2+(-0.65)^2+(-1.15)^2 = 3.325$, so

$$
\widehat{\text{se}}_{\text{jack}} = \sqrt{\frac{4}{5}\times 3.325}
= \sqrt{2.66} = 1.6310,
$$

matching the classical value to rounding error, exactly as (1.8)
guarantees.

*Interpretation.* This example both illustrates the mechanics of the
jackknife and validates it: because the sample mean has a known
closed-form standard error, computing the jackknife estimate by hand
and confirming it reproduces $s/\sqrt n$ builds confidence in the
general formula before applying it to statistics without a known
closed form.

**Example 2: Exact jackknife bias correction of the plug-in variance.**

*Problem.* For the same sample, compute the plug-in (divide-by-$n$)
variance estimate $\hat\sigma^2_n$, its jackknife bias estimate, and
the bias-corrected estimator, and compare to the classical
divide-by-$(n-1)$ estimator.

*Solution.* $\hat\sigma^2_n = 53.2/5 = 10.64$. The five leave-one-out
plug-in variances (computed on the four remaining points each time)
are $7.25, 11.5, 13.25, 11.1875, 6.6875$, averaging to
$\bar\theta_{(\cdot)} = 9.975$. Hence

$$
\widehat{\text{bias}}_{\text{jack}} = 4\times(9.975-10.64) = -2.66,
$$

and the bias-corrected estimator is
$\hat\theta_{\text{jack}} = 10.64-(-2.66) = 13.30$, which is *exactly*
$s^2$, the classical unbiased variance estimator computed in Example
1.

*Interpretation.* For the variance statistic, the jackknife
bias correction is not merely an asymptotic $O(1/n^2)$ improvement
(Theorem 1.2); it recovers the exactly unbiased estimator with no
remainder at all, because the bias of the plug-in variance estimator
is *exactly* $-\sigma^2/n$ with no higher-order terms — this is the
special case referenced in the Remarks following Theorem 1.2.

# Applications

- **Statistics.** Bootstrap confidence intervals are now standard for
  regression coefficients, correlation and $R^2$ statistics, and any
  estimator lacking a tractable analytical standard error.
- **Machine Learning.** Bootstrap resampling underlies bagging
  (bootstrap aggregating) and random forests directly, and is widely
  used to construct confidence intervals for cross-validated
  performance metrics.
- **Data Science.** Quantifying uncertainty in ad hoc summary
  statistics (medians, quantiles, custom business metrics) computed on
  observational data, where no analytical formula is available.
- **Finance.** Bootstrap resampling of historical returns is used to
  estimate the sampling variability of portfolio risk measures without
  assuming a specific parametric return distribution.
- **Reliability Engineering.** Jackknife and bootstrap bias correction
  for reliability function estimators computed from censored failure
  time data.
- **Monte Carlo Methods.** The bootstrap is itself a Monte Carlo
  technique, connecting directly to the simulation methods developed
  throughout Unit 3.
- **Bayesian Statistics.** The bootstrap has a Bayesian analogue (the
  Bayesian bootstrap, Rubin 1981) and bootstrap ideas inform the
  construction of approximate posterior predictive checks.
- **Operations Research.** Bootstrap confidence intervals for
  simulation output statistics (e.g. mean queue length) when the
  underlying distribution of simulation replications is not Normal.
- **Engineering.** Jackknife bias correction of parameter estimates in
  fitted degradation or fatigue models with small sample sizes.

# Advantages

- Both methods require no analytical derivation of a standard error
  formula; they apply to essentially any statistic that can be
  computed on a data vector.
- The bootstrap directly approximates the full sampling *distribution*
  of a statistic, not merely its standard error, enabling confidence
  intervals that adapt to skewness (the percentile method).
- The jackknife requires only $n$ deterministic recomputations, with
  no random number generation needed, and its bias-correction
  identity is exact for several classical statistics.

# Limitations

- The bootstrap's accuracy depends on how well the empirical
  distribution $\hat F$ approximates the true population $F$, which
  can be poor for small $n$ or for statistics sensitive to the tails
  of the distribution.
- The jackknife is inconsistent for non-smooth statistics, most
  notably the sample median, and should not be used for standard
  error estimation in that setting.
- Both methods can be computationally expensive when the statistic
  itself is costly to compute (for example, refitting a complex model
  for each of thousands of bootstrap resamples).
- Neither method rescues a fundamentally biased sampling design;
  resampling reflects the properties of the *observed* sample, and
  cannot correct for problems such as non-random sampling from the
  population of interest.

# Practical Tips

- **Use at least $B=1000$-$2000$ replications for standard error
  estimation, and more for confidence intervals.** As Theorem 1.1 and
  the simulation in Section 10.1 show, the bootstrap standard error
  estimate itself has Monte Carlo error that shrinks only as
  $B\to\infty$; percentile intervals depend on tail quantiles and
  typically need larger $B$ for the same stability.
- **Always set a random seed for bootstrap reproducibility**, since
  each run otherwise produces a slightly different result.
- **Do not apply the plain jackknife to the sample median or other
  non-smooth statistics.** Use the bootstrap instead, or a smoothed
  jackknife variant if the jackknife's computational economy is
  essential.
- **Check whether the bootstrap distribution is markedly skewed
  before relying on the normal-approximation interval.** The
  percentile interval (or a more refined method such as BCa) is
  generally more trustworthy when skewness is present.
- **Vectorize where possible.** As implemented above, resampling and
  recomputing a simple statistic like the mean is fast even at large
  $B$; for computationally expensive statistics, consider reducing $B$
  or parallelizing the replicate loop.

# Summary

- The bootstrap approximates the sampling distribution of a statistic
  by resampling with replacement from the observed data, treating the
  empirical distribution as a stand-in for the unknown population.
- Bootstrap estimates of bias and standard error are computed directly
  from the spread of bootstrap replicates (Theorem 1.1 establishes
  their Monte Carlo consistency as $B\to\infty$).
- Bootstrap confidence intervals can be constructed via the normal
  approximation or, more robustly to skewness, via the percentile
  method.
- The jackknife estimates bias and standard error using $n$
  systematic leave-one-out recomputations; Theorem 1.2 shows it
  removes the leading $O(1/n)$ bias term for smooth statistics, and
  this correction is exact (not merely asymptotic) for the plug-in
  variance estimator, as Example 2 demonstrates.
- The jackknife's simplicity comes at the cost of inconsistency for
  non-smooth statistics such as the median, where the bootstrap
  remains the appropriate tool.

# Key Formulae

- Bootstrap SE: $\widehat{\text{se}}_{\text{boot}} =
  \text{sd}(\hat\theta^{*}_1,\dots,\hat\theta^{*}_B)$.
- Bootstrap bias: $\widehat{\text{bias}}_{\text{boot}} = \bar\theta^{*}
  - \hat\theta$.
- Bootstrap percentile CI: empirical $(\alpha/2, 1-\alpha/2)$
  quantiles of $\hat\theta^{*}_1,\dots,\hat\theta^{*}_B$.
- Jackknife bias: $\widehat{\text{bias}}_{\text{jack}} =
  (n-1)(\bar\theta_{(\cdot)}-\hat\theta)$.
- Jackknife SE: $\widehat{\text{se}}_{\text{jack}} =
  \sqrt{\frac{n-1}{n}\sum_i (\hat\theta_{(i)}-\bar\theta_{(\cdot)})^2}$.
- Jackknife bias-corrected estimator: $\hat\theta_{\text{jack}} =
  n\hat\theta - (n-1)\bar\theta_{(\cdot)}$.

# Key Terms

- **Bootstrap principle.** Approximating the sampling distribution of
  a statistic by resampling from the empirical distribution.
- **Bootstrap replicate.** The value of a statistic computed on one
  bootstrap resample.
- **Empirical distribution.** The distribution placing probability
  $1/n$ on each observed data point.
- **Jackknife.** A resampling technique based on systematically
  leaving out one observation at a time.
- **Leave-one-out statistic.** The value of a statistic recomputed
  with a single observation removed.
- **Percentile interval.** A bootstrap confidence interval formed from
  the empirical quantiles of the bootstrap replicates.
- **Resampling.** Generating new samples from an observed dataset,
  either with replacement (bootstrap) or by systematic deletion
  (jackknife), to approximate sampling variability.

# Exercises

## Multiple Choice Questions

1. A bootstrap sample is drawn
   (a) without replacement, of size $n-1$
   (b) with replacement, of size $n$
   (c) without replacement, of size $n$
   (d) with replacement, of size $2n$

2. The bootstrap estimate of bias is
   (a) $\hat\theta - \bar\theta^{*}$
   (b) $\bar\theta^{*} - \hat\theta$
   (c) $\text{sd}(\hat\theta^{*})$
   (d) $\hat\theta^2$

3. As $B\to\infty$, the bootstrap standard error estimate converges to
   (a) the true population standard error
   (b) zero
   (c) the "ideal bootstrap" standard error under resampling from
       $\hat F$
   (d) infinity

4. The jackknife leave-one-out statistic $\hat\theta_{(i)}$ is computed
   (a) on a bootstrap resample of size $n$
   (b) on the original sample with observation $i$ removed
   (c) on a random subsample of size $n/2$
   (d) on the full sample with observation $i$ duplicated

5. For the sample mean, the jackknife standard error
   (a) is inconsistent
   (b) reduces exactly to $s/\sqrt n$
   (c) is always larger than $s/\sqrt n$
   (d) cannot be computed

6. The jackknife bias correction, under the expansion (1.6), reduces
   the order of the bias from
   (a) $O(1/n)$ to $O(1)$
   (b) $O(1/n^2)$ to $O(1/n)$
   (c) $O(1/n)$ to $O(1/n^2)$
   (d) $O(1)$ to $O(1/n)$

7. The jackknife is known to be inconsistent for
   (a) the sample mean
   (b) the sample variance
   (c) the sample median
   (d) the sample sum

8. The bootstrap percentile confidence interval is constructed from
   (a) the mean and standard deviation of the bootstrap replicates
       assuming normality
   (b) the empirical quantiles of the bootstrap replicates
   (c) the leave-one-out replicates only
   (d) the original sample alone

9. For the plug-in (divide-by-$n$) variance estimator under Normal
   sampling, the exact bias is
   (a) $0$
   (b) $\sigma^2/n$
   (c) $-\sigma^2/n$
   (d) $-\sigma^2/(n-1)$

10. Compared to the jackknife, the bootstrap generally requires
    (a) fewer recomputations of the statistic
    (b) exactly $n$ recomputations
    (c) more recomputations of the statistic, but random resampling
        rather than systematic deletion
    (d) no recomputation at all

**Answers.** 1(b), 2(b), 3(c), 4(b), 5(b), 6(c), 7(c), 8(b), 9(c),
10(c).

## Short Answer Questions

1. State the bootstrap principle in your own words.
2. Define the empirical distribution $\hat F$ and explain its role in
   the bootstrap.
3. Write down the formulas for the bootstrap estimates of bias and
   standard error.
4. Write down the formulas for the jackknife estimates of bias and
   standard error.
5. Explain why the jackknife standard error formula includes the
   factor $(n-1)/n$ rather than simply $1/n$.
6. Explain, using Theorem 1.1, what aspect of bootstrap accuracy is
   controlled by increasing $B$, and what aspect is not.
7. Why is the percentile confidence interval generally preferred over
   the normal-approximation interval when the bootstrap distribution
   is skewed?
8. Explain why the jackknife is inconsistent for the sample median.
9. State the jackknife bias-correction identity and explain what
   "removing the leading bias term" means.
10. Give one practical application, distinct from those emphasized in
    this unit, where bootstrap resampling would be useful.

## Long Answer Questions

1. State and prove Theorem 1.1 (consistency of the bootstrap standard
   error estimate as $B\to\infty$), and explain what source of error
   this theorem does and does not address.
2. State and prove Theorem 1.2 (jackknife bias reduction), explaining
   the role of the assumed bias expansion (1.6).
3. Derive the identity showing that the jackknife standard error
   reduces exactly to $s/\sqrt n$ for the sample mean, as in Section
   8.1.
4. Using the dataset of Example 2, explain why the jackknife bias
   correction of the plug-in variance estimator is exact rather than
   merely asymptotic, and relate this to the general theory of
   Theorem 1.2.
5. Compare the bootstrap and the jackknife in terms of computational
   cost, applicability to non-smooth statistics, and the type of
   inferential statement (standard error only, versus full sampling
   distribution) each supports.
6. Discuss the normal-approximation and percentile bootstrap
   confidence intervals, explaining under what circumstances they
   would give noticeably different results.
7. Explain the connection between the bootstrap and the general Monte
   Carlo simulation framework developed in Unit 3, identifying which
   specific generation method (from Units 3.1-3.6) underlies drawing a
   bootstrap resample.
8. Discuss the role of resampling methods in modern machine learning,
   with specific reference to bagging and bootstrap-based confidence
   intervals for model performance metrics.
9. Explain why neither the bootstrap nor the jackknife can correct for
   a biased sampling design, using a concrete hypothetical example.
10. Discuss the historical development of the jackknife (Quenouille,
    Tukey) and the bootstrap (Efron), and explain the conceptual
    relationship between the two methods.

## Programming Exercises

1. Implement `bootstrap_estimate()` from first principles (already
   provided) and reproduce, by simulation, the standard error
   comparison of Example 1 for the dataset given there.
2. Implement `jackknife_estimate()` from first principles (already
   provided) and reproduce the manual calculations of Examples 1 and 2
   exactly.
3. Apply the bootstrap to estimate the standard error of the sample
   correlation coefficient for a bivariate dataset of your choosing
   (for example, simulated from a bivariate Normal distribution with a
   known correlation), and compare the bootstrap standard error to the
   asymptotic formula $\text{se}(\hat\rho) \approx (1-\rho^2)/\sqrt{n}$.
4. Demonstrate empirically that the jackknife standard error is
   inconsistent for the sample median: for increasing sample sizes
   drawn from a known continuous distribution, compare the jackknife
   standard error of the median to a bootstrap estimate, and discuss
   the discrepancy.
5. Construct and compare the normal-approximation and percentile 95%
   bootstrap confidence intervals for the coefficient of variation
   ($s/\bar x$) of a right-skewed simulated dataset (e.g. drawn from a
   Gamma or Log-normal distribution), and discuss which interval
   appears more trustworthy.
6. Write a simulation study estimating the empirical coverage of the
   normal-approximation bootstrap interval, analogous to Section 10.3,
   and compare its coverage to that of the percentile interval under
   the same conditions.
7. Implement the jackknife bias correction for the plug-in variance
   estimator for a range of sample sizes $n \in \{5, 10, 20, 50\}$, and
   verify, by simulation, that the corrected estimator's average bias
   is negligible in every case (since the correction is exact for this
   statistic).
8. Apply bootstrap resampling to estimate the standard error of a
   fitted simple linear regression slope coefficient, using
   case resampling (resampling entire $(x_i,y_i)$ pairs), and compare
   to the classical formula-based standard error from `lm()`.
9. Benchmark, using `system.time()`, the bootstrap (with $B=2000$)
   against the jackknife for a sample of size $n=200$ and a moderately
   expensive statistic (for example, a trimmed mean), and discuss the
   relative computational cost.
10. Implement a simple Bayesian bootstrap (Rubin, 1981): instead of
    resampling with replacement, assign each observation a random
    Dirichlet$(1,\dots,1)$ weight and compute a weighted statistic;
    compare the resulting distribution of replicates to the ordinary
    bootstrap distribution for the sample mean.

## Mini Projects

1. **A resampling toolkit with diagnostics.** Build a small R toolkit
   that, given a dataset and a statistic function, automatically
   computes the bootstrap bias, bootstrap standard error, jackknife
   bias, jackknife standard error, and both types of bootstrap
   confidence interval, and produces a diagnostic report flagging (i)
   substantial disagreement between the bootstrap and jackknife
   standard errors, and (ii) substantial skewness in the bootstrap
   distribution (for example, via a simple skewness statistic).
   Demonstrate the toolkit on at least three different statistics
   (mean, variance, and one statistic of your choosing that is
   plausibly non-smooth, such as the median), and discuss what the
   diagnostics reveal in each case.

2. **Bootstrap versus jackknife under model misspecification.** Design
   a simulation study in which data are generated from a
   heavy-tailed distribution (for example, a $t_3$ distribution), and
   compare the bootstrap and jackknife standard error estimates for
   the sample mean and the 10% trimmed mean across repeated simulated
   datasets, evaluating each against the true (simulation-based)
   sampling standard error. Discuss which method degrades more
   gracefully as the tail heaviness increases, and relate your
   findings to the smoothness assumptions underlying Theorem 1.2.

**Difficulty labels.** Multiple choice, short answer: **Easy**. Long
answer questions 3, 4, 6, 7, 9, programming exercises 1-6: **Medium**.
Long answer questions 1, 2, 5, 8, 10, programming exercises 7-10:
**Medium to Advanced**. Mini projects: **Advanced**.

# References

Bickel, P. J., & Freedman, D. A. (1981). Some asymptotic theory for
the bootstrap. *The Annals of Statistics*, 9(6), 1196-1217.

Efron, B. (1979). Bootstrap methods: Another look at the jackknife.
*The Annals of Statistics*, 7(1), 1-26.

Efron, B. (1982). *The jackknife, the bootstrap and other resampling
plans*. SIAM.

Efron, B. (1987). Better bootstrap confidence intervals. *Journal of
the American Statistical Association*, 82(397), 171-185.

Quenouille, M. H. (1956). Notes on bias in estimation. *Biometrika*,
43(3/4), 353-360.

Rizzo, M. L. (2019). *Statistical computing with R* (2nd ed.). CRC
Press.

Rubin, D. B. (1981). The Bayesian bootstrap. *The Annals of
Statistics*, 9(1), 130-134.

Tukey, J. W. (1958). Bias and confidence in not-quite large samples
(abstract). *The Annals of Mathematical Statistics*, 29(2), 614.
