---
title: "Unit 4.2: Numerical Methods"
author: "Dr. Manoj C. Patil"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
    code_folding: hide
    theme: cosmo
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    echo = TRUE,
    warning = FALSE,
    message = FALSE,
    fig.align = "center",
    fig.width = 7,
    fig.height = 4.5
)

set.seed(123)
library(tidyverse)
library(ggplot2)
library(dplyr)
library(knitr)
library(kableExtra)
library(gridExtra)
library(patchwork)
```

# Learning Objectives

By the end of this unit, a student should be able to

- explain why closed-form solutions to nonlinear equations and
  integrals are the exception rather than the rule in statistical
  practice, motivating numerical methods,
- implement and prove the linear convergence of the bisection method
  for locating a root,
- implement and prove the quadratic convergence of the Newton-Raphson
  method, and extend it to systems of nonlinear equations using the
  Jacobian matrix,
- derive the trapezoidal and Simpson's rule quadrature formulas, their
  error terms, and their convergence rates,
- implement all four methods in R and empirically verify their
  theoretical convergence rates,
- recognize the role of these numerical methods within statistical
  computing, particularly maximum likelihood estimation and numerical
  integration of intractable expectations.

# Prerequisites

Students should be comfortable with

- single-variable and multivariable calculus, including Taylor's
  theorem with remainder and partial derivatives,
- basic linear algebra, including matrix inversion and determinants of
  $2\times2$ matrices,
- the concept of a maximum likelihood estimator and the score equation,
- R programming, including writing iterative loops and simple matrix
  operations.

# Introduction

Much of theoretical statistics is stated in a form that presumes an
equation can be solved, or an integral evaluated, exactly. The maximum
likelihood estimator is defined as the value that sets the score
function to zero; the sampling distribution of a statistic is often
described through an integral of a density; a quantile is defined
implicitly as the value $x$ solving $F(x)=u$. In practice, closed-form
solutions to these problems are the exception. The maximum likelihood
equations for all but the simplest models are nonlinear and admit no
algebraic solution; most integrals of interest in probability and
statistics — even something as basic as the standard Normal
cumulative distribution function — have no elementary antiderivative.
Numerical methods supply the machinery to solve such problems to any
desired accuracy using arithmetic operations alone.

This unit develops the two most fundamental families of numerical
methods used throughout statistical computing. **Root-finding
methods** — the bisection method and the Newton-Raphson method —
locate the value or values of $x$ satisfying $f(x)=0$, and extend
naturally to systems of nonlinear equations arising, for instance,
from multiparameter score equations. **Numerical integration methods**
— the trapezoidal rule and Simpson's rule — approximate a definite
integral using only evaluations of the integrand at a finite set of
points, and form the deterministic counterpart to the Monte Carlo
integration methods developed in Unit 4.3. The bisection method dates
to the ancient method of exhaustion and was formalized in the 19th
century; the Newton-Raphson method traces to Isaac Newton (1669) and
Joseph Raphson (1690); Simpson's rule, though named for Thomas Simpson
(1743), was known earlier to Johannes Kepler. Despite their age, these
methods remain the computational engines inside `optim()`, `nlm()`,
`uniroot()`, and `integrate()` in R, and inside virtually every
statistical software package's maximum likelihood fitting routines.

The practical importance of this material is immediate and pervasive:
fitting a generalized linear model, computing a profile likelihood
confidence interval, or evaluating the normalizing constant of a
posterior density all reduce, at some point, to root-finding or
numerical integration. Understanding how these building blocks work,
and their limitations, is essential to using statistical software
correctly and to diagnosing convergence failures when they occur.

# Intuition

Consider searching for a lost item along a long corridor, where a
helper can only tell you, at any point you ask, whether the item is to
your left or to your right. The most conservative strategy is to
check the midpoint of the corridor, learn which half the item is in,
and repeat on that half — halving the search region at every step.
This is exactly the logic of the **bisection method**: given two
points at which a continuous function $f$ has opposite signs, the
Intermediate Value Theorem guarantees a root lies between them, and
evaluating the midpoint and keeping whichever half still brackets a
sign change halves the uncertainty at every step, guaranteed, but
never faster than that.

A sighted searcher who can also see how steeply the corridor's
brightness is changing could do better: standing at a point and
observing both the reading and its rate of change, they could
extrapolate — assuming the trend continues in a straight line — to
where the reading would hit zero, and jump there directly rather than
merely halving the interval. This is the idea behind the
**Newton-Raphson method**: at the current point $x_n$, approximate the
function by its tangent line, and jump to where that tangent line
crosses zero. When the function is well approximated by its tangent
line near the root — which a smooth function always is, sufficiently
close to a simple root — this extrapolation converges far faster than
repeated halving.

A parallel intuition underlies numerical integration. The exact area
under a curve is unavailable in closed form, but the area under a
*simple* curve — a straight line, or a parabola — passing through a
few known points of the true curve is easy to compute exactly. The
**trapezoidal rule** approximates the area under $f$ on each small
sub-interval by the area under the straight line joining the two
endpoints; **Simpson's rule** does better by fitting a parabola
through three points at a time, exploiting the curvature information
that three points, rather than two, provide.

```{r intuition-figure, echo=FALSE, fig.cap="Newton-Raphson (left) jumps to where the tangent line crosses zero, converging much faster near a smooth root than bisection's repeated halving (right)."}
f_demo <- function(x) x^3 - x - 2
f_prime_demo <- function(x) 3 * x^2 - 1

x_grid <- seq(1, 2.2, length.out = 200)
tangent_at <- 1.5
tangent_line <- f_demo(tangent_at) + f_prime_demo(tangent_at) * (x_grid - tangent_at)
newton_next <- tangent_at - f_demo(tangent_at) / f_prime_demo(tangent_at)

p_newton <- ggplot() +
  geom_line(data = tibble(x = x_grid, y = f_demo(x_grid)), aes(x, y), colour = "steelblue", linewidth = 1) +
  geom_line(data = tibble(x = x_grid, y = tangent_line), aes(x, y), colour = "darkred", linetype = "dashed") +
  geom_point(aes(x = tangent_at, y = f_demo(tangent_at)), size = 2, colour = "black") +
  geom_point(aes(x = newton_next, y = 0), size = 2, colour = "darkgreen") +
  geom_hline(yintercept = 0, colour = "grey60") +
  labs(title = "Newton-Raphson: follow the tangent to zero", x = "x", y = "f(x)") +
  theme_minimal(base_size = 11)

bisect_points <- c(1, 2, 1.5, 1.75, 1.625)
p_bisection <- ggplot() +
  geom_line(data = tibble(x = x_grid, y = f_demo(x_grid)), aes(x, y), colour = "steelblue", linewidth = 1) +
  geom_point(aes(x = bisect_points, y = f_demo(bisect_points)), size = 2, colour = "darkorange") +
  geom_hline(yintercept = 0, colour = "grey60") +
  labs(title = "Bisection: repeatedly halve the bracket", x = "x", y = "f(x)") +
  theme_minimal(base_size = 11)

p_newton + p_bisection
```

# Formal Definitions

## The Root-Finding Problem

**Definition.** Given a continuous function $f:\mathbb{R}\to
\mathbb{R}$, a **root** is a value $r$ such that $f(r)=0$. A root $r$
is **simple** if $f'(r)\ne 0$.

**Definition (bracket).** An interval $[a,b]$ is said to **bracket** a
root of $f$ if $f(a)$ and $f(b)$ have opposite signs, i.e.
$f(a)f(b)<0$; by the Intermediate Value Theorem, a continuous $f$ then
has at least one root in $(a,b)$.

## Systems of Nonlinear Equations

**Definition.** For a vector-valued function
$\mathbf F:\mathbb{R}^p\to\mathbb{R}^p$,
$\mathbf F(\mathbf x) = (F_1(\mathbf x),\dots,F_p(\mathbf x))'$, a
**root** is a vector $\mathbf r$ with $\mathbf F(\mathbf r) =
\mathbf 0$. The **Jacobian matrix** of $\mathbf F$ at $\mathbf x$ is

$$
J(\mathbf x) = \begin{pmatrix}
\dfrac{\partial F_1}{\partial x_1} & \cdots & \dfrac{\partial F_1}{\partial x_p} \\
\vdots & \ddots & \vdots \\
\dfrac{\partial F_p}{\partial x_1} & \cdots & \dfrac{\partial F_p}{\partial x_p}
\end{pmatrix}.
\tag{2.1}
$$

## Numerical Quadrature

**Definition.** A **quadrature formula** approximates a definite
integral by a weighted sum of integrand values at prescribed nodes,

$$
\int_a^b f(x)\,dx \;\approx\; \sum_{i=0}^{m} w_i\, f(x_i),
\tag{2.2}
$$

for nodes $a=x_0<x_1<\cdots<x_m=b$ and weights $w_i$ determined by the
specific rule. A quadrature rule has **degree of precision** $d$ if it
integrates every polynomial of degree $\le d$ exactly, and fails to
integrate some polynomial of degree $d+1$ exactly.

# Theorems

## Theorem 2.1: Linear Convergence of the Bisection Method

**Statement.** Let $f$ be continuous on $[a_0,b_0]$ with
$f(a_0)f(b_0)<0$. The bisection method generates a sequence of
midpoints $c_k$ satisfying

$$
|c_k - r| \le \frac{b_0-a_0}{2^{k+1}}
\tag{2.3}
$$

for some root $r\in(a_0,b_0)$, for every $k=0,1,2,\dots$.

**Assumptions.** $f$ is continuous on $[a_0,b_0]$; only a sign change
is required, not differentiability.

**Proof.** By induction on $k$, the bracket $[a_k,b_k]$ maintained by
the algorithm satisfies $b_k-a_k = (b_0-a_0)/2^k$: initially this holds
for $k=0$, and at each step the bracket width is exactly halved by
construction (the midpoint $c_k=(a_k+b_k)/2$ replaces whichever
endpoint does not change sign with it, halving the width). Since
$f(a_k)f(b_k)<0$ throughout (this sign-change property is preserved by
construction: the algorithm always keeps the sub-interval on which the
sign change persists), the Intermediate Value Theorem guarantees a
root $r_k \in (a_k,b_k)$ at every step; because the brackets are
nested, all $r_k$ coincide with a single root $r \in (a_0,b_0)$ common
to every bracket. Since $c_k$ is the midpoint of $[a_k,b_k]$ and
$r\in[a_k,b_k]$,

$$
|c_k-r| \le \frac{b_k-a_k}{2} = \frac{b_0-a_0}{2^{k+1}},
$$

as claimed. $\blacksquare$

**Remarks.** The bound (2.3) holds unconditionally, with no
smoothness requirement on $f$ beyond continuity; this robustness is
the bisection method's chief advantage, at the cost of convergence
that is always linear (each iteration gains a fixed number of correct
bits, not an increasing number).

**Examples.** A full worked example is given in Section 11.

**Applications.** Bisection is the standard fallback method inside
`uniroot()` in R, used whenever a robust (if slow) root-finder is
needed, for example to invert a monotone cumulative distribution
function numerically.

## Theorem 2.2: Quadratic Convergence of the Newton-Raphson Method

**Statement.** Let $r$ be a simple root of $f$ ($f(r)=0$,
$f'(r)\ne0$), with $f$ twice continuously differentiable in a
neighbourhood of $r$. If the Newton-Raphson iterates
$x_{n+1} = x_n - f(x_n)/f'(x_n)$ are generated from a starting point
$x_0$ sufficiently close to $r$, then the errors $e_n = r-x_n$ satisfy

$$
e_{n+1} = -\frac{f''(\xi_n)}{2f'(x_n)}\, e_n^2
\tag{2.4}
$$

for some $\xi_n$ between $r$ and $x_n$, and consequently there exists
a constant $M$ (depending on bounds for $f''$ and $f'$ near $r$) such
that $|e_{n+1}| \le M\, e_n^2$, i.e. convergence is (at least)
quadratic.

**Assumptions.** $f'(r)\ne 0$ (simple root); $f''$ continuous near
$r$; $x_0$ close enough to $r$ that all iterates remain in a
neighbourhood where $f'\ne0$.

**Proof.** By Taylor's theorem with Lagrange remainder, expanding
$f(r)$ about $x_n$,

$$
0 = f(r) = f(x_n) + f'(x_n)(r-x_n) + \frac{f''(\xi_n)}{2}(r-x_n)^2
$$

for some $\xi_n$ between $x_n$ and $r$. Dividing through by $f'(x_n)$
(non-zero for $x_n$ close enough to $r$, by continuity of $f'$ and
$f'(r)\ne0$),

$$
\frac{f(x_n)}{f'(x_n)} + (r-x_n) = -\frac{f''(\xi_n)}{2f'(x_n)}(r-x_n)^2.
$$

Since $x_{n+1} = x_n - f(x_n)/f'(x_n)$, the term $f(x_n)/f'(x_n)$
equals $x_n - x_{n+1}$; substituting,

$$
(x_n-x_{n+1}) + (r-x_n) = -\frac{f''(\xi_n)}{2f'(x_n)}(r-x_n)^2
\;\Longrightarrow\;
r - x_{n+1} = -\frac{f''(\xi_n)}{2f'(x_n)}(r-x_n)^2,
$$

which is exactly (2.4). Taking $M = \sup |f''|/\inf(2|f'|)$ over a
small neighbourhood of $r$ (finite by continuity of $f'$, $f''$ and
$f'(r)\ne0$) gives $|e_{n+1}| \le M e_n^2$. $\blacksquare$

**Remarks.** Quadratic convergence means that, once the iterates are
close enough to $r$, the number of correct decimal digits
approximately *doubles* at every iteration — a dramatically faster
rate than bisection's fixed linear gain, as the numerical example in
Section 11 illustrates directly. The requirement that $x_0$ be
"sufficiently close" is essential: Newton-Raphson has no global
convergence guarantee analogous to bisection's, and can diverge, cycle,
or converge to an unintended root if started poorly or if $f'(x_n)$
comes close to zero during the iteration.

**Examples.** A full worked example is given in Section 11.

**Applications.** Newton-Raphson (and its generalization, Fisher
scoring, which replaces $f'$ by its expectation) is the standard
algorithm used to solve the score equations of maximum likelihood
estimation for generalized linear models and countless other
parametric statistical models.

# Mathematical Derivations

## Newton-Raphson for Systems of Nonlinear Equations

The scalar Newton-Raphson update $x_{n+1}=x_n-f(x_n)/f'(x_n)$
generalizes to a system $\mathbf F(\mathbf x)=\mathbf 0$ by the same
tangent-line (now tangent-hyperplane) reasoning. Expanding
$\mathbf F$ to first order about $\mathbf x_n$,

$$
\mathbf 0 \approx \mathbf F(\mathbf x_n) + J(\mathbf x_n)(\mathbf x_{n+1}-\mathbf x_n),
$$

and solving for $\mathbf x_{n+1}$ (assuming $J(\mathbf x_n)$ is
invertible) gives the multivariate update

$$
\mathbf x_{n+1} = \mathbf x_n - J(\mathbf x_n)^{-1}\mathbf F(\mathbf x_n).
\tag{2.5}
$$

In practice, (2.5) is implemented not by explicitly inverting
$J(\mathbf x_n)$ (computationally wasteful and numerically less
stable) but by solving the linear system
$J(\mathbf x_n)\,\boldsymbol\delta_n = -\mathbf F(\mathbf x_n)$ for the
step $\boldsymbol\delta_n$, and setting
$\mathbf x_{n+1} = \mathbf x_n + \boldsymbol\delta_n$. Under the
analogous smoothness and non-singularity conditions to the scalar
case, the multivariate iteration also converges quadratically near a
simple root (a root at which $J$ is non-singular).

## Derivation of the Trapezoidal Rule and Its Error

On a single panel $[x_0,x_1]$ of width $h=x_1-x_0$, the trapezoidal
rule replaces $f$ by the straight line $P_1$ interpolating
$(x_0,f(x_0))$ and $(x_1,f(x_1))$, giving

$$
\int_{x_0}^{x_1} f(x)\,dx \approx \int_{x_0}^{x_1} P_1(x)\,dx
= \frac{h}{2}\bigl(f(x_0)+f(x_1)\bigr).
\tag{2.6}
$$

The standard linear interpolation error formula gives
$f(x)-P_1(x) = \tfrac12 f''(\eta(x))(x-x_0)(x-x_1)$ for some
$\eta(x)$ between $x_0$ and $x_1$ (a consequence of Rolle's theorem
applied to the interpolation remainder; see Burden & Faires, 2011,
Thm. 3.3). Since $(x-x_0)(x-x_1)\le 0$ throughout $[x_0,x_1]$, the
weighted mean value theorem for integrals gives, for some
$\xi\in(x_0,x_1)$,

$$
\int_{x_0}^{x_1} \bigl[f(x)-P_1(x)\bigr]dx
= \frac{f''(\xi)}{2}\int_{x_0}^{x_1}(x-x_0)(x-x_1)\,dx
= \frac{f''(\xi)}{2}\left(-\frac{h^3}{6}\right) = -\frac{h^3}{12}f''(\xi),
\tag{2.7}
$$

using the standard result
$\int_{x_0}^{x_1}(x-x_0)(x-x_1)\,dx = -h^3/6$ (verified directly by
substitution $u=x-x_0$: $\int_0^h u(u-h)\,du = \bigl[u^3/3-hu^2/2\bigr]_0^h
= h^3/3-h^3/2=-h^3/6$). Summing (2.7) over $n$ panels of a composite
rule and applying a further mean value argument to the sum of the
per-panel errors (each proportional to $h^3f''(\xi_i)$, summed and
recognized as an $O(h)$-spaced Riemann-type sum) gives the composite
trapezoidal error

$$
E_{\text{trap}} = -\frac{(b-a)h^2}{12} f''(\xi), \qquad \xi\in(a,b),
\tag{2.8}
$$

confirming that the composite trapezoidal rule has error $O(h^2)$.

## Derivation of Simpson's Rule and Its Error

Simpson's rule approximates the integral over a panel $[x_0,x_2]$ of
width $2h$, with midpoint $x_1=x_0+h$, by fitting a quadratic through
the three points. Writing $m=x_1$ and expanding both the exact
integral and the quadrature rule in a Taylor series about $m$ makes
the derivation self-contained. Using
$f(x)=f(m)+f'(m)(x-m)+\tfrac12f''(m)(x-m)^2+\tfrac16f'''(m)(x-m)^3
+\tfrac{1}{24}f''''(m)(x-m)^4+\cdots$ and integrating term by term over
$[m-h,m+h]$ (odd-power terms vanish by symmetry),

$$
\int_{m-h}^{m+h} f(x)\,dx = 2h f(m) + \frac{h^3}{3}f''(m)
+ \frac{h^5}{60}f''''(m) + O(h^7).
\tag{2.9}
$$

For the quadrature side, Simpson's rule is
$S = \tfrac{h}{3}\bigl[f(m-h)+4f(m)+f(m+h)\bigr]$; Taylor-expanding
$f(m-h)$ and $f(m+h)$ and adding (again the odd terms cancel),
$f(m-h)+f(m+h) = 2f(m)+h^2f''(m)+\tfrac{h^4}{12}f''''(m)+O(h^6)$, so

$$
S = \frac{h}{3}\Bigl[6f(m) + h^2 f''(m) + \frac{h^4}{12}f''''(m)\Bigr] + O(h^7)
= 2hf(m) + \frac{h^3}{3}f''(m) + \frac{h^5}{36}f''''(m) + O(h^7).
\tag{2.10}
$$

Subtracting (2.10) from (2.9), the $f(m)$ and $f''(m)$ terms cancel
exactly, leaving

$$
E_{\text{Simpson, panel}} = \Bigl(\frac{1}{60}-\frac{1}{36}\Bigr)h^5 f''''(m) + O(h^7)
= -\frac{h^5}{90}f''''(\xi)
\tag{2.11}
$$

for some $\xi$ near $m$ (replacing $f''''(m)$ by $f''''(\xi)$ via the
mean value theorem to absorb the $O(h^7)$ remainder). Summing over
$n/2$ panels of a composite rule (an even number $n$ of sub-intervals
is required) gives the standard composite Simpson error

$$
E_{\text{Simpson}} = -\frac{(b-a)h^4}{180} f''''(\xi), \qquad \xi\in(a,b),
\tag{2.12}
$$

showing the composite Simpson's rule has error $O(h^4)$ — two full
orders better than the trapezoidal rule's $O(h^2)$ for a comparable
number of function evaluations, at the cost of requiring an even
number of sub-intervals and one additional order of smoothness
($f''''$ rather than $f''$).

# Algorithms

## Algorithm 2.1: Bisection Method

**Purpose.** Locate a root of $f$ within a bracket $[a,b]$, with a
guaranteed, if slow, rate of convergence.

**Input.** Function $f$; bracket $[a,b]$ with $f(a)f(b)<0$; tolerance
$\varepsilon$.

**Output.** An approximate root $c$ with $|c-r| < \varepsilon$.

**Step-by-step procedure.**

1. While $(b-a)/2 > \varepsilon$:
   (a) compute $c=(a+b)/2$;
   (b) if $f(a)f(c) < 0$, set $b=c$; otherwise set $a=c$.
2. Return $c=(a+b)/2$.

**Pseudo-code.**

```
Algorithm: Bisection Method
Input:  f, a, b (with f(a)*f(b) < 0), epsilon
Output: approximate root c

1. While (b - a) / 2 > epsilon:
       c <- (a + b) / 2
       If f(a) * f(c) < 0:
           b <- c
       Else:
           a <- c
2. Return (a + b) / 2
```

**Flowchart description.** Start; check the width of the bracket
against the tolerance; if too wide, compute the midpoint, evaluate the
function there, and replace whichever endpoint shares its sign with
the midpoint; repeat; once the bracket is narrow enough, return the
midpoint; end.

**Complexity.** $O(\log_2((b-a)/\varepsilon))$ iterations, each
costing one function evaluation, by Theorem 2.1.

**Practical considerations.** Bisection requires only a valid initial
bracket, and always converges for a continuous function once one is
found; it is the natural fallback when Newton-Raphson fails to
converge or when derivatives of $f$ are unavailable or unreliable.

## Algorithm 2.2: Newton-Raphson Method

**Purpose.** Locate a root of $f$ rapidly, using derivative
information, near a good starting guess.

**Input.** Function $f$ and its derivative $f'$; starting point $x_0$;
tolerance $\varepsilon$; maximum iterations $N$.

**Output.** An approximate root $x_n$.

**Step-by-step procedure.**

1. For $n=0,1,2,\dots$ until $|x_{n+1}-x_n|<\varepsilon$ or $n=N$:
   compute $x_{n+1} = x_n - f(x_n)/f'(x_n)$.
2. Return the final iterate.

**Pseudo-code.**

```
Algorithm: Newton-Raphson Method
Input:  f, f_prime, x0, epsilon, N
Output: approximate root x

1. x <- x0
2. For n = 1 to N:
       x_new <- x - f(x) / f_prime(x)
       If |x_new - x| < epsilon:
           Return x_new
       x <- x_new
3. Return x  (or report non-convergence)
```

**Flowchart description.** Start; evaluate $f$ and $f'$ at the current
iterate; compute the tangent-line intercept as the new iterate; check
for convergence; if not converged and the iteration budget remains,
repeat; end.

**Complexity.** Under quadratic convergence (Theorem 2.2), the number
of iterations to reach tolerance $\varepsilon$, once in the basin of
quadratic convergence, is $O(\log\log(1/\varepsilon))$ — dramatically
fewer than bisection's $O(\log(1/\varepsilon))$ — but each iteration
requires evaluating both $f$ and $f'$.

**Practical considerations.** Newton-Raphson can fail — diverge,
cycle, or converge to the wrong root — if $x_0$ is poorly chosen or
$f'(x_n)$ is close to zero at some iterate; a common safeguard is to
combine Newton-Raphson with a bisection fallback (as `uniroot()` and
professional root-finders do) to retain global reliability while
benefiting from Newton's local speed.

## Algorithm 2.3: Newton-Raphson for Systems of Nonlinear Equations

**Purpose.** Solve $\mathbf F(\mathbf x)=\mathbf 0$ for a
vector-valued $\mathbf F$, generalizing Algorithm 2.2 using (2.5).

**Input.** Function $\mathbf F$ and its Jacobian $J$; starting vector
$\mathbf x_0$; tolerance $\varepsilon$; maximum iterations $N$.

**Output.** An approximate root $\mathbf x_n$.

**Step-by-step procedure.**

1. For $n=0,1,2,\dots$ until $\|\mathbf x_{n+1}-\mathbf x_n\|<
   \varepsilon$ or $n=N$: solve the linear system
   $J(\mathbf x_n)\boldsymbol\delta_n = -\mathbf F(\mathbf x_n)$ for
   $\boldsymbol\delta_n$; set $\mathbf x_{n+1} = \mathbf x_n +
   \boldsymbol\delta_n$.
2. Return the final iterate.

**Pseudo-code.**

```
Algorithm: Newton-Raphson for Systems
Input:  F, J, x0, epsilon, N
Output: approximate root vector x

1. x <- x0
2. For n = 1 to N:
       delta <- solve( J(x), -F(x) )
       x_new <- x + delta
       If norm(x_new - x) < epsilon:
           Return x_new
       x <- x_new
3. Return x  (or report non-convergence)
```

**Flowchart description.** Start; evaluate $\mathbf F$ and $J$ at the
current vector iterate; solve the resulting linear system for the
step; update the iterate; check convergence in norm; repeat if not
converged; end.

**Complexity.** Each iteration costs $O(p^3)$ for solving the
$p\times p$ linear system (via Gaussian elimination or LU
decomposition), in addition to the cost of evaluating $\mathbf F$ and
$J$; the number of iterations required is again $O(\log\log(1/
\varepsilon))$ near a simple root under the multivariate analogue of
Theorem 2.2.

**Practical considerations.** A singular or near-singular Jacobian at
any iterate causes the linear solve to fail or become numerically
unstable; monitoring the condition number of $J(\mathbf x_n)$, or
adding a small ridge term when necessary, is standard practice in
production implementations (as in the Levenberg-Marquardt
modification used for nonlinear least squares).

## Algorithm 2.4: Composite Trapezoidal and Simpson's Rules

**Purpose.** Approximate $\int_a^b f(x)\,dx$ using evenly spaced
function evaluations.

**Input.** Function $f$; interval $[a,b]$; number of sub-intervals
$n$ (even, for Simpson's rule).

**Output.** Approximate integral value.

**Step-by-step procedure (Trapezoidal).**

1. Set $h=(b-a)/n$ and $x_i = a+ih$, $i=0,\dots,n$.
2. Compute
   $T = \dfrac{h}{2}\Bigl[f(x_0) + 2\sum_{i=1}^{n-1}f(x_i) + f(x_n)\Bigr]$.

**Step-by-step procedure (Simpson's, $n$ even).**

1. Set $h=(b-a)/n$ and $x_i=a+ih$, $i=0,\dots,n$.
2. Compute
   $S = \dfrac{h}{3}\Bigl[f(x_0)+4\!\!\sum_{i\ \text{odd}}\!\!f(x_i)
   +2\!\!\sum_{i\ \text{even},\,0<i<n}\!\!f(x_i)+f(x_n)\Bigr]$.

**Pseudo-code.**

```
Algorithm: Composite Trapezoidal / Simpson's Rule
Input:  f, a, b, n
Output: approximate integral

1. h <- (b - a) / n
2. x[0..n] <- a + i * h  for i = 0..n
3. Trapezoidal:
       T <- h/2 * ( f(x[0]) + 2*sum(f(x[1..n-1])) + f(x[n]) )
4. Simpson (n even):
       S <- h/3 * ( f(x[0]) + 4*sum(f(x[odd i])) + 2*sum(f(x[even i, interior])) + f(x[n]) )
5. Return T (or S)
```

**Flowchart description.** Start; partition $[a,b]$ into $n$ equal
sub-intervals; evaluate $f$ at every node; combine the values with the
prescribed weights (uniform for the endpoints and alternating 4-2-4-2
weights for Simpson's interior nodes); return the weighted sum; end.

**Complexity.** $O(n)$ function evaluations for both rules; Simpson's
rule achieves $O(h^4)$ accuracy against the trapezoidal rule's
$O(h^2)$ for the same $n$ (hence the same computational cost), making
it strictly preferable whenever $f''''$ is well behaved.

**Practical considerations.** Both rules degrade for integrands with
sharp peaks, discontinuities, or singularities within $[a,b]$;
adaptive quadrature (as implemented in R's `integrate()`), which
concentrates evaluation points where the integrand varies most
rapidly, should be preferred for such integrands over a fixed-$n$
composite rule.

# R Programming

```{r numerical-methods-functions}
set.seed(123)

#' Bisection method for root-finding
#'
#' Purpose: implements Algorithm 2.1.
#'
#' Arguments:
#'   target_function   function f for which a root is sought
#'   lower_bound       left endpoint of the bracket
#'   upper_bound       right endpoint of the bracket
#'   tolerance         stopping tolerance on the bracket width
#'
#' Returns:
#'   a list with the approximate root and the iteration history
#'
#' Example:
#'   bisection_method(function(x) x^3 - x - 2, 1, 2, 1e-8)
bisection_method <- function(target_function, lower_bound, upper_bound, tolerance = 1e-8) {
  if (target_function(lower_bound) * target_function(upper_bound) >= 0) {
    stop("The supplied interval does not bracket a root.")
  }

  iteration_history <- numeric(0)
  while ((upper_bound - lower_bound) / 2 > tolerance) {
    midpoint <- (lower_bound + upper_bound) / 2
    iteration_history <- c(iteration_history, midpoint)

    if (target_function(lower_bound) * target_function(midpoint) < 0) {
      upper_bound <- midpoint
    } else {
      lower_bound <- midpoint
    }
  }

  list(root = (lower_bound + upper_bound) / 2, history = iteration_history)
}

#' Newton-Raphson method for root-finding
#'
#' Purpose: implements Algorithm 2.2.
#'
#' Arguments:
#'   target_function     function f
#'   target_derivative   function f'
#'   starting_point      initial guess x0
#'   tolerance           stopping tolerance on successive iterates
#'   max_iterations      maximum number of iterations
#'
#' Returns:
#'   a list with the approximate root and the iteration history
#'
#' Example:
#'   newton_raphson_method(
#'     function(x) x^3 - x - 2, function(x) 3 * x^2 - 1, 1.5, 1e-10
#'   )
newton_raphson_method <- function(target_function, target_derivative, starting_point,
                                   tolerance = 1e-10, max_iterations = 100) {
  current_point <- starting_point
  iteration_history <- current_point

  for (iteration in seq_len(max_iterations)) {
    next_point <- current_point - target_function(current_point) / target_derivative(current_point)
    iteration_history <- c(iteration_history, next_point)

    if (abs(next_point - current_point) < tolerance) {
      return(list(root = next_point, history = iteration_history))
    }
    current_point <- next_point
  }

  list(root = current_point, history = iteration_history)
}

#' Newton-Raphson method for systems of nonlinear equations
#'
#' Purpose: implements Algorithm 2.3.
#'
#' Arguments:
#'   system_function     function returning the vector F(x)
#'   jacobian_function    function returning the Jacobian matrix J(x)
#'   starting_vector      initial guess x0
#'   tolerance            stopping tolerance on the step norm
#'   max_iterations       maximum number of iterations
#'
#' Returns:
#'   a list with the approximate root vector and the iteration history
#'
#' Example:
#'   newton_raphson_system(
#'     function(v) c(v[1]^2 + v[2]^2 - 4, v[1] * v[2] - 1),
#'     function(v) matrix(c(2 * v[1], v[2], 2 * v[2], v[1]), nrow = 2),
#'     c(1.8, 0.6)
#'   )
newton_raphson_system <- function(system_function, jacobian_function, starting_vector,
                                   tolerance = 1e-10, max_iterations = 100) {
  current_vector <- starting_vector
  iteration_history <- list(current_vector)

  for (iteration in seq_len(max_iterations)) {
    step_vector <- solve(jacobian_function(current_vector), -system_function(current_vector))
    next_vector <- current_vector + step_vector
    iteration_history[[iteration + 1]] <- next_vector

    if (sqrt(sum((next_vector - current_vector)^2)) < tolerance) {
      return(list(root = next_vector, history = iteration_history))
    }
    current_vector <- next_vector
  }

  list(root = current_vector, history = iteration_history)
}

#' Composite trapezoidal rule
#'
#' Purpose: implements the trapezoidal part of Algorithm 2.4.
#'
#' Arguments:
#'   integrand_function   function to integrate
#'   lower_limit          lower limit of integration
#'   upper_limit          upper limit of integration
#'   number_of_intervals  number of sub-intervals
#'
#' Returns:
#'   the approximate integral (numeric scalar)
#'
#' Example:
#'   trapezoidal_rule(function(x) x^2, 0, 2, 4)
trapezoidal_rule <- function(integrand_function, lower_limit, upper_limit, number_of_intervals) {
  step_size <- (upper_limit - lower_limit) / number_of_intervals
  grid_points <- lower_limit + (0:number_of_intervals) * step_size
  function_values <- integrand_function(grid_points)

  step_size / 2 * (function_values[1] + 2 * sum(function_values[2:number_of_intervals]) +
    function_values[number_of_intervals + 1])
}

#' Composite Simpson's rule
#'
#' Purpose: implements the Simpson's part of Algorithm 2.4.
#'
#' Arguments:
#'   integrand_function   function to integrate
#'   lower_limit          lower limit of integration
#'   upper_limit          upper limit of integration
#'   number_of_intervals  number of sub-intervals (must be even)
#'
#' Returns:
#'   the approximate integral (numeric scalar)
#'
#' Example:
#'   simpsons_rule(function(x) x^2, 0, 2, 4)
simpsons_rule <- function(integrand_function, lower_limit, upper_limit, number_of_intervals) {
  if (number_of_intervals %% 2 != 0) {
    stop("Simpson's rule requires an even number of sub-intervals.")
  }

  step_size <- (upper_limit - lower_limit) / number_of_intervals
  grid_points <- lower_limit + (0:number_of_intervals) * step_size
  function_values <- integrand_function(grid_points)

  odd_indices <- seq(2, number_of_intervals, by = 2)
  even_indices <- if (number_of_intervals >= 4) {
    seq(3, number_of_intervals - 1, by = 2)
  } else {
    integer(0)
  }

  step_size / 3 * (
    function_values[1] +
      4 * sum(function_values[odd_indices]) +
      2 * sum(function_values[even_indices]) +
      function_values[number_of_intervals + 1]
  )
}
```

# Simulation Studies

## Empirical Convergence Rates: Bisection vs. Newton-Raphson

```{r convergence-rate-comparison}
set.seed(123)

target_f <- function(x) x^3 - x - 2
target_f_prime <- function(x) 3 * x^2 - 1
true_root <- uniroot(target_f, c(1, 2), tol = 1e-14)$root

bisection_result <- bisection_method(target_f, 1, 2, tolerance = 1e-12)
newton_result <- newton_raphson_method(target_f, target_f_prime, 1.5, tolerance = 1e-14)

bisection_errors <- abs(bisection_result$history - true_root)
newton_errors <- abs(newton_result$history - true_root)

tibble(
  Iteration = seq_along(bisection_errors),
  Bisection_Error = bisection_errors
) %>%
  slice(1:8) %>%
  kable(digits = 8, caption = "Bisection error by iteration (halves each step)")

tibble(
  Iteration = seq_along(newton_errors) - 1,
  Newton_Error = newton_errors
) %>%
  kable(digits = 12, caption = "Newton-Raphson error by iteration (squares each step, once close)")
```

## Convergence Order of Trapezoidal and Simpson's Rules

```{r quadrature-convergence}
set.seed(123)

test_integrand <- function(x) exp(-x^2)
true_value <- integrate(test_integrand, 0, 1)$value

interval_counts <- c(2, 4, 8, 16, 32, 64, 128)
trap_errors <- vapply(
  interval_counts, function(n) abs(trapezoidal_rule(test_integrand, 0, 1, n) - true_value), numeric(1)
)
simpson_errors <- vapply(
  interval_counts, function(n) abs(simpsons_rule(test_integrand, 0, 1, n) - true_value), numeric(1)
)

tibble(
  n = interval_counts, Trapezoidal_Error = trap_errors, Simpson_Error = simpson_errors
) %>%
  kable(digits = 10, caption = "Integration error for the integral of exp(-x^2) on [0,1] (true value 0.7468241)")
```

# Graphs

```{r log-log-convergence-plot, fig.cap="Log-log plot of quadrature error against number of sub-intervals: the trapezoidal rule's slope of -2 and Simpson's rule's slope of -4 confirm the O(h^2) and O(h^4) rates of Section 8."}
convergence_df <- tibble(
  n = rep(interval_counts, 2),
  error = c(trap_errors, simpson_errors),
  method = rep(c("Trapezoidal", "Simpson"), each = length(interval_counts))
)

ggplot(convergence_df, aes(x = n, y = error, colour = method)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  scale_x_log10() +
  scale_y_log10() +
  labs(
    title = "Quadrature error vs. number of sub-intervals (log-log scale)",
    x = "Number of sub-intervals (n)", y = "Absolute error"
  ) +
  theme_minimal(base_size = 11)
```

```{r newton_basin_plot, fig.cap="Newton-Raphson iterates converge quadratically to the true root, visualized against the target function."}
x_grid_plot <- seq(1, 2, length.out = 200)

ggplot() +
  geom_line(data = tibble(x = x_grid_plot, y = target_f(x_grid_plot)), aes(x, y), colour = "steelblue") +
  geom_hline(yintercept = 0, colour = "grey60") +
  geom_point(
    data = tibble(x = newton_result$history, y = target_f(newton_result$history)),
    aes(x, y), colour = "darkred", size = 2
  ) +
  geom_path(
    data = tibble(x = newton_result$history, y = target_f(newton_result$history)),
    aes(x, y), colour = "darkred", linetype = "dotted"
  ) +
  labs(title = "Newton-Raphson iterates on f(x) = x^3 - x - 2", x = "x", y = "f(x)") +
  theme_minimal(base_size = 11)
```

# Numerical Examples

**Example 1: Bisection method.**

*Problem.* Find a root of $f(x)=x^3-x-2$ in $[1,2]$, performing three
iterations of bisection by hand.

*Solution.* $f(1)=-2$, $f(2)=4$ (opposite signs, bracket valid).

$$
\begin{aligned}
c_1 &= 1.5, & f(1.5) &= -0.125 \ (\text{same sign as } f(1)) \Rightarrow \text{new bracket } [1.5,2] \\
c_2 &= 1.75, & f(1.75) &= 1.609375 \ (\text{opposite sign}) \Rightarrow \text{new bracket } [1.5,1.75] \\
c_3 &= 1.625, & f(1.625) &= 0.666016 \ (\text{opposite sign}) \Rightarrow \text{new bracket } [1.5,1.625]
\end{aligned}
$$

*Interpretation.* After three iterations the bracket has shrunk from
width $1$ to width $0.125 = 1/8$, exactly matching the halving
predicted by Theorem 2.1; the true root, $r\approx1.521380$, indeed
lies in $[1.5,1.625]$.

**Example 2: Newton-Raphson method.**

*Problem.* Using the same $f$, with $f'(x)=3x^2-1$, perform two
iterations of Newton-Raphson from $x_0=1.5$.

*Solution.*

$$
x_1 = 1.5 - \frac{f(1.5)}{f'(1.5)} = 1.5 - \frac{-0.125}{5.75} = 1.521739,
$$

$$
x_2 = x_1 - \frac{f(x_1)}{f'(x_1)} = 1.521739 - \frac{0.002137}{5.94707} = 1.521380.
$$

*Interpretation.* Two Newton-Raphson iterations already match the true
root $r=1.521380$ to six decimal places, whereas three bisection
iterations (Example 1) had only localized the root to the interval
$[1.5,1.625]$ — a striking illustration of the practical difference
between linear and quadratic convergence.

**Example 3: Newton-Raphson for a system.**

*Problem.* Solve $x^2+y^2=4$, $xy=1$ starting from
$(x_0,y_0)=(1.8,0.6)$, performing one iteration by hand.

*Solution.* $\mathbf F(1.8,0.6) = (1.8^2+0.6^2-4,\ 1.8\times0.6-1) =
(-0.4,\ 0.08)$. The Jacobian is
$J = \begin{pmatrix}2x&2y\\y&x\end{pmatrix}$, so
$J(1.8,0.6)=\begin{pmatrix}3.6&1.2\\0.6&1.8\end{pmatrix}$, with
determinant $3.6(1.8)-1.2(0.6)=5.76$. Solving
$J\boldsymbol\delta=-\mathbf F$,

$$
\boldsymbol\delta = \frac{1}{5.76}\begin{pmatrix}1.8&-1.2\\-0.6&3.6\end{pmatrix}
\begin{pmatrix}0.4\\-0.08\end{pmatrix}
= \frac{1}{5.76}\begin{pmatrix}0.816\\-0.528\end{pmatrix}
= \begin{pmatrix}0.141667\\-0.091667\end{pmatrix},
$$

so $(x_1,y_1) = (1.8+0.141667,\ 0.6-0.091667) = (1.941667,\ 0.508333)$.

*Interpretation.* Checking, $x_1^2+y_1^2=4.0285$ and $x_1y_1=0.9870$,
both markedly closer to the target values $4$ and $1$ than the
starting point; further iterations converge quadratically to the true
solution $(\sqrt{2+\sqrt3},\ 1/\sqrt{2+\sqrt3}) \approx (1.931852,\
0.517638)$.

**Example 4: Trapezoidal and Simpson's rule.**

*Problem.* Approximate $\int_0^2 x^2\,dx$ (true value $8/3=2.666667$)
using $n=4$ sub-intervals ($h=0.5$) with both rules.

*Solution.* Nodes: $0, 0.5, 1, 1.5, 2$ with values $0, 0.25, 1, 2.25,
4$.

$$
T = \frac{0.5}{2}\bigl[0+2(0.25+1+2.25)+4\bigr] = 0.25(11) = 2.75.
$$

$$
S = \frac{0.5}{3}\bigl[0+4(0.25)+2(1)+4(2.25)+4\bigr]
= \frac{0.5}{3}(16) = 2.666667.
$$

*Interpretation.* Simpson's rule reproduces the true value *exactly*,
because $f(x)=x^2$ is a polynomial of degree $2<4$ and Simpson's rule
has degree of precision $3$ (it integrates every cubic exactly, hence
certainly every quadratic); the trapezoidal rule, exact only for
linear integrands, overestimates the true value by $0.0833$, consistent
with the error formula (2.8), since $f''(x)=2>0$ makes the trapezoidal
rule systematically overestimate a convex function.

# Applications

- **Statistics.** Newton-Raphson (or its close relative, Fisher
  scoring) solves the score equations of maximum likelihood
  estimation for logistic regression, Poisson regression, and
  virtually every generalized linear model fit by standard software.
- **Machine Learning.** Newton's method and its quasi-Newton
  variants (BFGS, L-BFGS) are core second-order optimization
  algorithms; the Newton-Raphson system update (2.5) is precisely the
  Newton step used in optimizing a smooth loss function, applied to
  its gradient.
- **Data Science.** Numerical integration is used to compute
  marginal densities, expectations, and normalizing constants of
  fitted probability models that lack closed-form integrals.
- **Finance.** Computing implied volatility from an option's market
  price is a classical one-dimensional root-finding problem, almost
  universally solved by Newton-Raphson or a safeguarded bisection
  hybrid.
- **Reliability Engineering.** Solving for a system's failure time
  or reliability threshold from a nonlinear hazard or reliability
  function often requires root-finding when no closed-form inverse
  exists.
- **Monte Carlo Methods.** Numerical quadrature provides the
  deterministic benchmark against which the Monte Carlo integration
  methods of Unit 4.3 are compared for low-dimensional integrals.
- **Bayesian Statistics.** The Laplace approximation to a posterior
  density requires locating the posterior mode via Newton-Raphson (or
  a quasi-Newton method) and evaluating the Hessian there.
- **Operations Research.** Root-finding is used to solve for
  equilibrium points or break-even conditions in nonlinear resource
  allocation and inventory models.
- **Engineering.** Newton-Raphson for systems is the standard method
  for solving the nonlinear equilibrium equations arising in
  structural and circuit analysis.

# Advantages

- Bisection is completely robust: given a valid bracket, it is
  mathematically guaranteed to converge, requiring only continuity of
  $f$.
- Newton-Raphson converges very rapidly (quadratically) once near a
  simple root, needing very few iterations for high precision.
- Simpson's rule achieves a substantially higher order of accuracy
  than the trapezoidal rule at essentially the same computational cost
  per additional node.

# Limitations

- Bisection requires an initial bracket with a sign change, which may
  not always be easy to locate, particularly for functions with
  multiple close roots or no sign change at a simple root (a root of
  even multiplicity).
- Newton-Raphson requires the derivative $f'$ (or the Jacobian, for
  systems), can diverge or converge to an unintended root from a poor
  starting point, and can fail outright if $f'(x_n)$ is ever zero or
  numerically close to it during the iteration.
- Fixed-$n$ quadrature rules such as trapezoidal and Simpson's rules
  degrade sharply for integrands with singularities, sharp peaks, or
  discontinuities, and provide no built-in error estimate without
  additional (e.g. Richardson extrapolation) machinery.

# Practical Tips

- **Prefer a hybrid strategy in production code.** Combining
  Newton-Raphson's speed with bisection's robustness (falling back to
  a bisection step whenever a Newton step would leave the current
  bracket) is standard practice, as implemented internally by
  professional root-finders including R's `uniroot()`.
- **Always verify convergence, not just iteration count.** Checking
  both $|x_{n+1}-x_n|$ and $|f(x_{n+1})|$ against separate tolerances
  guards against premature termination when the two quantities behave
  differently (as can happen near a root with a very flat or very
  steep derivative).
- **Use Simpson's rule by default over the trapezoidal rule** for
  smooth integrands, since it is essentially free — the same number
  of function evaluations yields two additional orders of accuracy.
- **Use `integrate()`'s adaptive quadrature, not a fixed-$n$ composite
  rule, for integrands with peaks or singularities.** Fixed-node
  rules waste evaluations on flat regions and under-resolve rapidly
  varying regions.
- **When implementing Newton-Raphson for systems, monitor the
  condition number of the Jacobian.** A poorly conditioned or singular
  Jacobian at any iterate is a common, silent cause of numerical
  failure that a simple convergence check may not reveal until many
  wasted iterations have passed.

# Summary

- The bisection method locates a root by repeatedly halving a bracket
  known to contain a sign change; Theorem 2.1 establishes its
  unconditional linear convergence.
- The Newton-Raphson method follows the tangent line to a root and,
  near a simple root, converges quadratically (Theorem 2.2), roughly
  doubling the number of correct digits per iteration.
- Newton-Raphson generalizes to systems of nonlinear equations via the
  Jacobian matrix, replacing division by $f'$ with solving a linear
  system involving $J$.
- The trapezoidal and Simpson's rules approximate a definite integral
  using linear and quadratic local interpolation respectively, with
  error terms $O(h^2)$ and $O(h^4)$ derived explicitly in Section 8,
  and verified empirically by the log-log convergence plot of Section
  10.
- These deterministic numerical methods underlie the practical
  computation of maximum likelihood estimates and intractable
  integrals throughout statistical computing.

# Key Formulae

- Bisection error bound: $|c_k-r| \le (b_0-a_0)/2^{k+1}$.
- Newton-Raphson update: $x_{n+1}=x_n - f(x_n)/f'(x_n)$; quadratic
  error recursion $e_{n+1} = -\dfrac{f''(\xi_n)}{2f'(x_n)}e_n^2$.
- Newton-Raphson for systems: $\mathbf x_{n+1} = \mathbf x_n -
  J(\mathbf x_n)^{-1}\mathbf F(\mathbf x_n)$.
- Trapezoidal rule: $T=\dfrac{h}{2}[f(x_0)+2\sum f(x_i)+f(x_n)]$,
  error $-\dfrac{(b-a)h^2}{12}f''(\xi)$.
- Simpson's rule: $S=\dfrac{h}{3}[f(x_0)+4\sum_{\text{odd}}f(x_i)
  +2\sum_{\text{even}}f(x_i)+f(x_n)]$, error
  $-\dfrac{(b-a)h^4}{180}f''''(\xi)$.

# Key Terms

- **Bisection method.** A bracketing root-finding method that
  repeatedly halves an interval known to contain a sign change.
- **Bracket.** An interval on which a continuous function changes
  sign, guaranteeing a root within it.
- **Degree of precision.** The highest polynomial degree a quadrature
  rule integrates exactly.
- **Jacobian matrix.** The matrix of first partial derivatives of a
  vector-valued function.
- **Newton-Raphson method.** A root-finding method that follows the
  tangent line (or hyperplane) at the current iterate to a new
  estimate.
- **Quadratic convergence.** A convergence rate in which the error at
  each step is proportional to the square of the previous error.
- **Simpson's rule.** A quadrature formula based on local quadratic
  (parabolic) interpolation.
- **Trapezoidal rule.** A quadrature formula based on local linear
  interpolation.

# Exercises

## Multiple Choice Questions

1. The bisection method requires
   (a) a derivative of $f$
   (b) a bracket $[a,b]$ with $f(a)f(b)<0$
   (c) a Hessian matrix
   (d) a Taylor series expansion

2. The bisection method's convergence rate is
   (a) quadratic
   (b) cubic
   (c) linear
   (d) exponential

3. The Newton-Raphson update formula is
   (a) $x_{n+1}=x_n+f(x_n)/f'(x_n)$
   (b) $x_{n+1}=x_n-f(x_n)/f'(x_n)$
   (c) $x_{n+1}=x_n-f'(x_n)/f(x_n)$
   (d) $x_{n+1}=x_n-f(x_n)f'(x_n)$

4. Newton-Raphson achieves quadratic convergence provided
   (a) $f$ is continuous only
   (b) the root is simple and $f''$ is continuous nearby
   (c) $f$ is a polynomial of even degree
   (d) the starting point is exactly the root

5. The Newton-Raphson update for a system of equations replaces
   division by $f'$ with
   (a) multiplication by $J$
   (b) solving a linear system involving the Jacobian $J$
   (c) computing eigenvalues of $J$
   (d) transposing $J$

6. The trapezoidal rule's composite error is of order
   (a) $O(h)$
   (b) $O(h^2)$
   (c) $O(h^3)$
   (d) $O(h^4)$

7. Simpson's rule's composite error is of order
   (a) $O(h^2)$
   (b) $O(h^3)$
   (c) $O(h^4)$
   (d) $O(h^5)$

8. Simpson's rule requires
   (a) an odd number of sub-intervals
   (b) an even number of sub-intervals
   (c) exactly 2 sub-intervals
   (d) no restriction on the number of sub-intervals

9. Simpson's rule integrates every polynomial of degree up to
   (a) 1
   (b) 2
   (c) 3
   (d) 4

10. A common safeguard against Newton-Raphson divergence is to
    (a) always start at $x_0=0$
    (b) combine it with a bisection fallback
    (c) use only the trapezoidal rule instead
    (d) increase the tolerance to a very large value

**Answers.** 1(b), 2(c), 3(b), 4(b), 5(b), 6(b), 7(c), 8(b), 9(c),
10(b).

## Short Answer Questions

1. State the bracketing condition required to start the bisection
   method, and explain why it guarantees a root exists.
2. Explain, in one or two sentences, the geometric idea behind the
   Newton-Raphson method.
3. Write the multivariate Newton-Raphson update formula and identify
   the role of the Jacobian matrix.
4. State the error order of the trapezoidal rule and of Simpson's
   rule, and explain which is more accurate for the same $n$.
5. Explain why Simpson's rule requires an even number of
   sub-intervals.
6. Why can Newton-Raphson fail even when a root exists nearby?
7. Explain the difference between linear and quadratic convergence in
   terms of the number of correct digits gained per iteration.
8. Why is bisection often used as a fallback for Newton-Raphson in
   production software?
9. Give one statistical application of Newton-Raphson distinct from
   those emphasized in this unit.
10. Explain why Simpson's rule integrates $x^2$ exactly, as verified
    in Example 4.

## Long Answer Questions

1. State and prove Theorem 2.1 (linear convergence of bisection).
2. State and prove Theorem 2.2 (quadratic convergence of
   Newton-Raphson), explaining the role of Taylor's theorem in the
   proof.
3. Derive the composite trapezoidal rule error formula, showing every
   step from the linear interpolation error to the final composite
   bound.
4. Derive the single-panel Simpson's rule error formula using the
   Taylor expansion approach of Section 8.2, showing every algebraic
   step.
5. Compare bisection and Newton-Raphson in terms of robustness,
   speed, and information required about $f$, and describe a scenario
   in which each would be preferred.
6. Derive the Newton-Raphson update formula for a system of two
   nonlinear equations from the first-order Taylor expansion of
   $\mathbf F$, explaining each step.
7. Discuss the role of Newton-Raphson (or Fisher scoring) in maximum
   likelihood estimation, explaining how the score equation and the
   observed or expected information relate to $f$ and $f'$ in the
   scalar case.
8. Explain why fixed-$n$ trapezoidal and Simpson's rules can perform
   poorly for integrands with sharp peaks or singularities, and
   describe how adaptive quadrature addresses this limitation.
9. Using the numerical example of Section 11 (Example 4), explain why
   the trapezoidal rule systematically overestimates the integral of a
   convex function, relating your answer to the sign of the error
   term (2.8).
10. Discuss the practical safeguards used in professional root-finding
    software to combine the reliability of bisection with the speed of
    Newton-Raphson.

## Programming Exercises

1. Implement `bisection_method()` and `newton_raphson_method()` from
   first principles (already provided) and reproduce the manual
   iterations of Examples 1 and 2 exactly.
2. Implement `newton_raphson_system()` from first principles (already
   provided) and verify the manual first iteration of Example 3,
   then run the algorithm to convergence and confirm it reaches the
   true solution given in the interpretation of that example.
3. Implement `trapezoidal_rule()` and `simpsons_rule()` from first
   principles (already provided) and reproduce the manual calculation
   of Example 4 exactly.
4. Reproduce the log-log convergence plot of Section 10 for a
   different smooth integrand of your choosing, and verify the fitted
   slopes are close to $-2$ (trapezoidal) and $-4$ (Simpson).
5. Use Newton-Raphson to compute the maximum likelihood estimate of
   the rate parameter of an Exponential distribution from simulated
   data, and verify your answer against the known closed-form MLE
   $1/\bar x$.
6. Use `newton_raphson_system()` to fit, by maximum likelihood, the
   two parameters of a simple two-parameter distribution of your
   choosing (for example, a Gamma distribution) to simulated data,
   comparing your result to R's built-in `optim()` or `fitdistr()`.
7. Write a function that applies Newton-Raphson to invert the standard
   Normal cumulative distribution function (i.e. compute $\Phi^{-1}(u)$
   without using `qnorm()`), and compare its output and speed to
   `qnorm()` for a range of $u$ values.
8. Investigate Newton-Raphson divergence empirically: for
   $f(x)=x^3-2x+2$, show that starting from $x_0=0$ causes the
   iteration to cycle indefinitely, while starting from $x_0=-2$
   converges; explain the difference using a plot of $f$ and its
   tangent lines.
9. Implement Richardson extrapolation for the trapezoidal rule
   (combining the estimates at $n$ and $2n$ sub-intervals to cancel
   the leading error term) and verify empirically that it achieves an
   error order comparable to Simpson's rule.
10. Write a function comparing the number of function evaluations
    required by bisection, Newton-Raphson, and R's built-in
    `uniroot()` to reach a fixed tolerance for a common test function,
    and tabulate the results.

## Mini Projects

1. **A robust hybrid root-finder.** Implement a root-finding function
   that combines bisection and Newton-Raphson: at each step, attempt a
   Newton-Raphson update, but fall back to a bisection step whenever
   the Newton step would leave the current bracket or the derivative
   is too close to zero. Test your hybrid method against plain
   Newton-Raphson on at least three test functions chosen to make
   plain Newton-Raphson struggle (for example, functions with nearby
   inflection points or very flat derivatives near the root), and
   report the iteration counts and reliability of each approach.

2. **Numerical integration for a Bayesian normalizing constant.**
   Consider a posterior density known only up to a normalizing
   constant, $\pi(\theta) \propto L(\theta)\,p(\theta)$, for a
   likelihood $L$ and prior $p$ of your choosing (for a single
   parameter $\theta$). Use the composite Simpson's rule to compute
   the normalizing constant $\int \pi(\theta)\,d\theta$ over a
   suitable range, verify convergence of your answer as $n$ increases,
   and use the normalized posterior to compute the posterior mean via
   a second application of Simpson's rule. Compare your results to a
   Monte Carlo integration estimate (a preview of Unit 4.3).

**Difficulty labels.** Multiple choice, short answer: **Easy**. Long
answer questions 3, 4, 6, 8, 9, programming exercises 1-6: **Medium**.
Long answer questions 1, 2, 5, 7, 10, programming exercises 7-10:
**Medium to Advanced**. Mini projects: **Advanced**.

# References

Burden, R. L., & Faires, J. D. (2011). *Numerical analysis* (9th
ed.). Brooks/Cole.

Kelley, C. T. (2003). *Solving nonlinear equations with Newton's
method*. SIAM.

Nocedal, J., & Wright, S. J. (2006). *Numerical optimization* (2nd
ed.). Springer.

Rizzo, M. L. (2019). *Statistical computing with R* (2nd ed.). CRC
Press.

Ross, S. M. (2020). *Simulation* (6th ed.). Academic Press.
