---
title: "Unit 4.3: Monte Carlo Integration and Importance Sampling"
author: "Dr. Manoj C. Patil"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
    code_folding: hide
    theme: cosmo
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    echo = TRUE,
    warning = FALSE,
    message = FALSE,
    fig.align = "center",
    fig.width = 7,
    fig.height = 4.5
)

set.seed(123)
library(tidyverse)
library(ggplot2)
library(dplyr)
library(knitr)
library(kableExtra)
library(gridExtra)
library(patchwork)
```

# Learning Objectives

By the end of this unit, a student should be able to

- express a definite integral or a probability as an expectation, and
  estimate it by simple Monte Carlo simulation,
- derive the standard error of a Monte Carlo estimator and construct a
  confidence interval for the true value using the central limit
  theorem,
- state and prove the theory of importance sampling, including the
  characterization of the variance-minimizing importance density,
- implement importance sampling in R, including the self-normalized
  form used when the target density is known only up to a constant,
- explain why naive Monte Carlo fails for rare-event probabilities and
  how importance sampling restores efficiency,
- evaluate the trade-offs of Monte Carlo and importance sampling
  relative to the deterministic numerical methods of Unit 4.2.

# Prerequisites

Students should be comfortable with

- Units 3.1-3.6 (generation of pseudo-random variates by inverse
  transformation, acceptance-rejection, and related methods) and Unit
  3.4 in particular (exponential tilting),
- Unit 4.2 (numerical integration by quadrature, as the deterministic
  counterpart to the methods of this unit),
- the law of large numbers and the central limit theorem,
- the Cauchy-Schwarz inequality,
- R programming, including vectorized simulation and writing
  functions.

# Introduction

Unit 4.2 developed quadrature formulas — the trapezoidal and Simpson's
rules — that approximate a definite integral using a fixed grid of
evaluation points. These methods are highly accurate in one dimension,
but their cost grows explosively with dimension: a grid fine enough to
be useful in one dimension, repeated in $d$ dimensions, requires a
number of evaluation points growing like (grid size)$^d$, a phenomenon
known as the curse of dimensionality. **Monte Carlo integration** takes
an entirely different approach: rather than a deterministic grid, it
uses random sampling, and its accuracy depends on the sample size $n$
rather than on the dimension of the integral at all — the same $O(1/
\sqrt n)$ convergence rate holds whether the integral is
one-dimensional or fifty-dimensional. This is why Monte Carlo methods,
despite converging more slowly than quadrature in low dimensions, are
indispensable for the high-dimensional integrals that pervade modern
Bayesian statistics and computational finance.

The method traces to Stanislaw Ulam and John von Neumann's work on
neutron diffusion at Los Alamos in the 1940s, with the name "Monte
Carlo" (coined by Nicholas Metropolis, referencing the famous casino)
reflecting its reliance on randomness. In its most basic form, Monte
Carlo integration reduces to a single idea already used implicitly
throughout Unit 3: any integral can be rewritten as an expectation
under some probability distribution, and an expectation can be
estimated by the sample average of simulated draws from that
distribution, an application of the law of large numbers.

A refinement of this basic idea, **importance sampling**, addresses a
critical weakness: the plain Monte Carlo estimator can have very high
variance — or, in the extreme case of estimating a rare-event
probability, can produce an estimate of exactly zero with high
probability — whenever the region contributing most to the integral is
rarely visited by straightforward simulation. Importance sampling
resolves this by deliberately sampling from a different, better-chosen
distribution that visits the important region more often, and
correcting for the resulting distortion with a reweighting factor. The
technique connects directly to the exponential tilting construction of
Unit 3.4, generalizing it from a device for exact simulation of tail
events into a general-purpose variance-reduction tool for estimating
rare-event probabilities and expectations.

The practical importance of this unit is central to contemporary
statistical computing: Bayesian posterior expectations, option prices
in mathematical finance, and failure probabilities in structural and
systems reliability are all, at their core, integrals estimated by
exactly the machinery developed here.

# Intuition

Suppose the goal is to estimate the average height of a very large
crowd, too large to measure every person. The obvious approach is to
measure a random sample of people and average their heights; by the
law of large numbers, this sample average converges to the true
population average as the sample grows. Estimating a definite integral
$\int_a^b h(x)\,dx$ by simulation follows exactly the same logic: if
$X$ is drawn uniformly on $[a,b]$, then $\int_a^b h(x)\,dx = (b-a)\,
E[h(X)]$, so averaging $h(X_1),\dots,h(X_n)$ over a large simulated
sample and multiplying by $(b-a)$ estimates the integral, just as
averaging measured heights estimates the population mean.

The trouble with this "uniform sampling" strategy arises when the
function of interest, or the event of interest, is concentrated in a
small region of the sampling space. Estimating the average height
of professional basketball players by measuring a uniformly random
sample of the *general* population would require an enormous sample,
because basketball players are rare in the general population — most
of the sampling effort is wasted on people who tell you nothing about
the quantity of interest. The remedy is obvious once stated: sample
directly from basketball players, or from a population enriched with
them, and correct the resulting average using the known relative rarity
of basketball players in the general population. This "sample from
where it matters, then correct for the distortion" strategy is exactly
importance sampling, and it is the difference between an estimator
that is hopelessly noisy and one that is precise, for exactly the same
computational budget.

```{r intuition-figure, echo=FALSE, fig.cap="Naive sampling from f (left) rarely visits the shaded rare-event region; importance sampling from a shifted proposal p (right) visits it routinely, at the cost of needing a reweighting correction."}
x_grid <- seq(-4, 7, length.out = 400)
f_density <- dnorm(x_grid, 0, 1)
p_density <- dnorm(x_grid, 4, 1)

rare_region <- tibble(x = x_grid[x_grid >= 4], y = dnorm(x_grid[x_grid >= 4], 0, 1))

p_naive <- ggplot() +
  geom_line(data = tibble(x = x_grid, y = f_density), aes(x, y), colour = "steelblue", linewidth = 1) +
  geom_area(data = rare_region, aes(x, y), fill = "firebrick", alpha = 0.6) +
  labs(title = "Sampling from f: rare region rarely visited", x = "x", y = "Density") +
  theme_minimal(base_size = 11)

p_importance <- ggplot() +
  geom_line(data = tibble(x = x_grid, y = p_density), aes(x, y), colour = "darkgreen", linewidth = 1) +
  geom_area(data = rare_region, aes(x, y = dnorm(x, 4, 1)), fill = "firebrick", alpha = 0.6) +
  labs(title = "Sampling from proposal p: rare region visited often", x = "x", y = "Density") +
  theme_minimal(base_size = 11)

p_naive + p_importance
```

# Formal Definitions

## Monte Carlo Integration

**Definition.** Let $\theta = \int g(x) f(x)\,dx = E_f[g(X)]$ for a
probability density $f$ and a function $g$. The **(simple) Monte
Carlo estimator** of $\theta$, based on an i.i.d. sample
$X_1,\dots,X_n \sim f$, is

$$
\hat\theta_{\text{MC}} = \frac1n\sum_{i=1}^n g(X_i).
\tag{3.1}
$$

Any definite integral $\int_a^b h(x)\,dx$ can be cast in this form by
writing $\int_a^b h(x)\,dx = (b-a)\,E_U[h(U)]$ for
$U\sim\text{Uniform}(a,b)$, and any probability $P(X\in A)$ can be
cast in this form as $E_f[\mathbf 1_A(X)]$, so (3.1) covers both
integral approximation and probability estimation as special cases.

## Importance Sampling

**Definition.** Let $p$ be a probability density with support covering
the support of $g\cdot f$ (i.e. $p(x)>0$ wherever $g(x)f(x)\ne0$). The
**importance sampling estimator** of $\theta=E_f[g(X)]$, based on an
i.i.d. sample $Y_1,\dots,Y_n \sim p$, is

$$
\hat\theta_{\text{IS}} = \frac1n\sum_{i=1}^n g(Y_i)\,w(Y_i),
\qquad w(x) = \frac{f(x)}{p(x)},
\tag{3.2}
$$

where $p$ is called the **importance density** (or **proposal
density**) and $w$ is the **importance weight**.

# Theorems

## Theorem 3.1: Unbiasedness and Consistency of Monte Carlo Estimators

**Statement.** The simple Monte Carlo estimator (3.1) is unbiased,
$E(\hat\theta_{\text{MC}})=\theta$, and, provided
$\sigma_g^2=\operatorname{Var}_f(g(X))<\infty$, consistent,
$\hat\theta_{\text{MC}} \xrightarrow{\text{a.s.}} \theta$ as
$n\to\infty$, with

$$
\sqrt n\,(\hat\theta_{\text{MC}}-\theta) \;\xrightarrow{d}\; N(0,\sigma_g^2).
\tag{3.3}
$$

The importance sampling estimator (3.2) is likewise unbiased for any
valid choice of $p$, with $\operatorname{Var}(\hat\theta_{\text{IS}}) =
\sigma_{gw}^2/n$, $\sigma_{gw}^2 = \operatorname{Var}_p(g(Y)w(Y))$.

**Assumptions.** $X_1,\dots,X_n$ (or $Y_1,\dots,Y_n$) i.i.d.;
$\sigma_g^2<\infty$ (or $\sigma_{gw}^2<\infty$).

**Proof.** Unbiasedness of (3.1) follows from linearity of
expectation, $E(\hat\theta_{\text{MC}}) = n^{-1}\sum_i E(g(X_i)) =
E_f(g(X)) = \theta$. Consistency follows from the strong law of large
numbers applied to the i.i.d. sequence $g(X_1),g(X_2),\dots$ with
finite mean $\theta$. The central limit theorem, applied to the same
i.i.d. sequence with finite variance $\sigma_g^2$, gives (3.3)
directly. For the importance sampling estimator, unbiasedness follows
from a change of variables:

$$
E_p\bigl[g(Y)w(Y)\bigr] = \int g(y)\frac{f(y)}{p(y)}\,p(y)\,dy
= \int g(y)f(y)\,dy = \theta,
$$

valid whenever $p(y)>0$ on the support of $g\cdot f$ (so the ratio
$f(y)/p(y)$ is well defined there); the variance formula follows
because $\hat\theta_{\text{IS}}$ is a sample mean of the i.i.d.
sequence $g(Y_1)w(Y_1), g(Y_2)w(Y_2),\dots$. $\blacksquare$

**Remarks.** Equation (3.3) is the basis of the Monte Carlo confidence
interval $\hat\theta \pm z_{1-\alpha/2}\, s_g/\sqrt n$, where $s_g$ is
the sample standard deviation of the simulated $g(X_i)$ (or
$g(Y_i)w(Y_i)$) values — exactly analogous to a standard confidence
interval for a sample mean, since a Monte Carlo estimator *is* a
sample mean.

**Examples.** Worked in Section 12.

**Applications.** This theorem justifies reporting a standard error
alongside every Monte Carlo estimate, exactly as one would for any
other sample-mean-based statistic.

## Theorem 3.2: The Variance-Minimizing Importance Density

**Statement.** Among all importance densities $p$ with the same
support as $g\cdot f$ (so that (3.2) is unbiased for every candidate
$p$), the variance of $\hat\theta_{\text{IS}}$ is minimized by

$$
p^{*}(x) = \frac{|g(x)|\,f(x)}{\int |g(u)|\,f(u)\,du}.
\tag{3.4}
$$

If, in addition, $g(x)\ge 0$ for every $x$, sampling from $p^{*}$
gives $\operatorname{Var}(\hat\theta_{\text{IS}})=0$ exactly.

**Assumptions.** $\int|g(x)|f(x)\,dx<\infty$ (so that $p^{*}$ is a
valid, normalizable density).

**Proof.** From Theorem 3.1, minimizing $\operatorname{Var}_p(g(Y)w(Y))$
is equivalent to minimizing $E_p[(g(Y)w(Y))^2]$ (since the mean of
$g(Y)w(Y)$ is fixed at $\theta$ regardless of $p$, by unbiasedness).
Writing this second moment as an integral,

$$
E_p\bigl[(g(Y)w(Y))^2\bigr] = \int \frac{g(x)^2f(x)^2}{p(x)^2}\,p(x)\,dx
= \int \frac{g(x)^2 f(x)^2}{p(x)}\,dx.
$$

By the Cauchy-Schwarz inequality, applied to the functions
$|g(x)|f(x)/\sqrt{p(x)}$ and $\sqrt{p(x)}$,

$$
\left(\int |g(x)|f(x)\,dx\right)^2
= \left(\int \frac{|g(x)|f(x)}{\sqrt{p(x)}}\cdot\sqrt{p(x)}\,dx\right)^2
\le \left(\int \frac{g(x)^2f(x)^2}{p(x)}\,dx\right)\left(\int p(x)\,dx\right),
$$

and since $\int p(x)\,dx=1$, this gives

$$
\int \frac{g(x)^2f(x)^2}{p(x)}\,dx \;\ge\; \left(\int |g(x)|f(x)\,dx\right)^2,
\tag{3.5}
$$

with equality in the Cauchy-Schwarz inequality if and only if
$|g(x)|f(x)/\sqrt{p(x)}$ is proportional to $\sqrt{p(x)}$, i.e.
$p(x) \propto |g(x)|f(x)$ — exactly $p^{*}$ in (3.4). This proves
$p^{*}$ minimizes the second moment, hence the variance. When
$g(x)\ge0$ everywhere, the lower bound in (3.5) equals
$\bigl(\int g(x)f(x)\,dx\bigr)^2 = \theta^2$, so the minimum achievable
variance is $\theta^2 - \theta^2 = 0$. $\blacksquare$

**Remarks.** The zero-variance result is a theoretical benchmark, not
a practical algorithm: sampling from $p^{*}$ requires knowing
$\int|g(u)|f(u)\,du$, which is (a simple transformation of) the very
quantity $\theta$ being estimated. Its practical value is as a design
principle: a good importance density should be *shaped like* $|g|f$ —
concentrating sampling effort where $|g(x)|f(x)$ is largest — even
though matching it exactly is circular.

**Examples.** Worked in Section 12.

**Applications.** This theorem is the theoretical justification for
exponential tilting (Unit 3.4) as an importance sampling device for
rare-event probabilities: for $g(x)=\mathbf 1_{\{x>a\}}$, the ideal
$p^{*}$ is $f$ truncated and renormalized to $(a,\infty)$, and a tilted
exponential or shifted Normal density approximates this truncated
shape far better than the untilted $f$ itself does.

# Mathematical Derivations

## Self-Normalized Importance Sampling

In many applications — most importantly, Bayesian posterior
computation — the target density is known only up to a normalizing
constant, $f(x) = \tilde f(x)/c$ for an intractable constant $c =
\int\tilde f(u)\,du$. Writing $\theta = \int g(x)f(x)\,dx$ and
substituting,

$$
\theta = \frac{\int g(x)\tilde f(x)\,dx}{\int \tilde f(u)\,du}
= \frac{\int g(x)\,\tilde w(x)\,p(x)\,dx}{\int \tilde w(u)\,p(u)\,du},
\qquad \tilde w(x) = \frac{\tilde f(x)}{p(x)},
$$

obtained by multiplying and dividing both the numerator and
denominator integrals by $p(x)$. Replacing both integrals by their
Monte Carlo estimates under a sample $Y_1,\dots,Y_n\sim p$ gives the
**self-normalized importance sampling estimator**

$$
\hat\theta_{\text{SNIS}} = \frac{\sum_{i=1}^n g(Y_i)\,\tilde w(Y_i)}
{\sum_{i=1}^n \tilde w(Y_i)},
\tag{3.6}
$$

which requires knowledge of $\tilde f$ only up to the unknown constant
$c$ (which cancels between numerator and denominator). Unlike (3.2),
the estimator (3.6) is a ratio of two random quantities and is
therefore biased for finite $n$ (by Jensen's inequality applied to the
ratio, in general), though it remains consistent as $n\to\infty$ by
the law of large numbers applied to the numerator and denominator
separately, combined with the continuous mapping theorem.

## Naive Monte Carlo Failure for Rare Events

For $\theta=P(X\in A)=E_f[\mathbf 1_A(X)]$ with $\theta$ very small (a
rare event), the naive Monte Carlo estimator (3.1) is an average of
$n$ i.i.d. Bernoulli$(\theta)$ indicators, with variance
$\theta(1-\theta)/n \approx \theta/n$ for small $\theta$. The
**relative** standard error is therefore

$$
\frac{\text{se}(\hat\theta_{\text{MC}})}{\theta}
\approx \sqrt{\frac{\theta/n}{\theta^2}} = \frac{1}{\sqrt{n\theta}},
\tag{3.7}
$$

which stays large unless $n \gg 1/\theta$: to estimate a probability
of order $10^{-6}$ to within a modest relative error of $10\%$
requires roughly $n \approx 100/\theta = 10^8$ simulated draws — often
computationally infeasible, and a sample of more ordinary size will,
with high probability, observe *zero* occurrences of the event at
all, giving the useless estimate $\hat\theta_{\text{MC}}=0$. This is
precisely the situation importance sampling with a proposal
concentrated on $A$ is designed to remedy: for such a proposal, the
indicator $\mathbf 1_A(Y_i)$ is no longer rare, and the relative
standard error of $\hat\theta_{\text{IS}}$ need not degrade as
$\theta\to0$, as the simulation study in Section 10.3 demonstrates
directly.

# Algorithms

## Algorithm 3.1: Basic Monte Carlo Integration

**Purpose.** Estimate $\theta=E_f[g(X)]$ (an integral, expectation, or
probability) by direct simulation from $f$.

**Input.** Sampler for $f$; function $g$; sample size $n$.

**Output.** Point estimate $\hat\theta_{\text{MC}}$ and standard error.

**Step-by-step procedure.**

1. Draw $X_1,\dots,X_n$ i.i.d. from $f$.
2. Compute $\hat\theta_{\text{MC}} = n^{-1}\sum_i g(X_i)$.
3. Compute the standard error as $s_g/\sqrt n$, $s_g$ the sample
   standard deviation of $g(X_1),\dots,g(X_n)$.

**Pseudo-code.**

```
Algorithm: Monte Carlo Integration
Input:  sampler for f, g(), n
Output: theta_hat, standard error

1. x[1..n] <- draw n iid values from f
2. g_values <- g(x)
3. theta_hat <- mean(g_values)
4. se <- sd(g_values) / sqrt(n)
5. Return theta_hat, se
```

**Flowchart description.** Start; draw $n$ i.i.d. samples from $f$;
evaluate $g$ at each; average the results for the point estimate;
compute the sample standard deviation and divide by $\sqrt n$ for the
standard error; end.

**Complexity.** $O(n)$, given $O(1)$-amortized sampling from $f$ and
$O(1)$ evaluation of $g$.

**Practical considerations.** Unlike quadrature (Unit 4.2), whose
accuracy depends on the dimension of the integral, Monte Carlo
integration's $O(1/\sqrt n)$ error rate is dimension-free, making it
the method of choice for integrals in more than a few dimensions.

## Algorithm 3.2: Importance Sampling Estimator

**Purpose.** Estimate $\theta=E_f[g(X)]$ by sampling from a
better-matched proposal $p$, reducing variance relative to Algorithm
3.1.

**Input.** Sampler for $p$; density evaluators for $f$ and $p$;
function $g$; sample size $n$.

**Output.** Point estimate $\hat\theta_{\text{IS}}$ and standard
error.

**Step-by-step procedure.**

1. Draw $Y_1,\dots,Y_n$ i.i.d. from $p$.
2. Compute weights $w(Y_i) = f(Y_i)/p(Y_i)$.
3. Compute $\hat\theta_{\text{IS}} = n^{-1}\sum_i g(Y_i)w(Y_i)$.
4. Compute the standard error as $s_{gw}/\sqrt n$, $s_{gw}$ the sample
   standard deviation of the products $g(Y_i)w(Y_i)$.

**Pseudo-code.**

```
Algorithm: Importance Sampling
Input:  sampler for p, f(), p(), g(), n
Output: theta_hat, standard error

1. y[1..n] <- draw n iid values from p
2. w <- f(y) / p(y)
3. gw <- g(y) * w
4. theta_hat <- mean(gw)
5. se <- sd(gw) / sqrt(n)
6. Return theta_hat, se
```

**Flowchart description.** Start; draw $n$ i.i.d. samples from the
proposal $p$; compute the importance weight at each sample; multiply
by $g$; average for the point estimate; compute the standard error
from the spread of the weighted values; end.

**Complexity.** $O(n)$, identical order to Algorithm 3.1, but with a
potentially much smaller constant (variance) if $p$ is well chosen.

**Practical considerations.** A poorly chosen $p$ — in particular, one
with a lighter tail than $f\cdot g$ — can make
$\operatorname{Var}(\hat\theta_{\text{IS}})$ *larger* than the naive
estimator's variance, or even infinite; the design principle from
Theorem 3.2 (shape $p$ like $|g|f$) should always guide the choice.

## Algorithm 3.3: Self-Normalized Importance Sampling

**Purpose.** Estimate $\theta=E_f[g(X)]$ when $f$ is known only up to
an unknown normalizing constant.

**Input.** Sampler for $p$; evaluator for the unnormalized target
$\tilde f$; evaluator for $p$; function $g$; sample size $n$.

**Output.** Point estimate $\hat\theta_{\text{SNIS}}$.

**Step-by-step procedure.**

1. Draw $Y_1,\dots,Y_n$ i.i.d. from $p$.
2. Compute unnormalized weights $\tilde w(Y_i) = \tilde f(Y_i)/p(Y_i)$.
3. Compute $\hat\theta_{\text{SNIS}} = \dfrac{\sum_i g(Y_i)\tilde
   w(Y_i)}{\sum_i \tilde w(Y_i)}$ as in (3.6).

**Pseudo-code.**

```
Algorithm: Self-Normalized Importance Sampling
Input:  sampler for p, f_tilde(), p(), g(), n
Output: theta_hat

1. y[1..n] <- draw n iid values from p
2. w_tilde <- f_tilde(y) / p(y)
3. theta_hat <- sum(g(y) * w_tilde) / sum(w_tilde)
4. Return theta_hat
```

**Flowchart description.** Start; draw $n$ samples from the proposal;
compute unnormalized importance weights using the unnormalized target;
form the ratio of the weighted-$g$ sum to the weight sum; end.

**Complexity.** $O(n)$, identical to Algorithm 3.2.

**Practical considerations.** Self-normalized importance sampling is
biased for finite $n$ (Section 8.1), though the bias is typically
$O(1/n)$ and negligible for moderate to large $n$; it is the standard
tool for estimating posterior expectations in Bayesian computation
when only the unnormalized posterior kernel is available.

## Algorithm 3.4: Rare-Event Probability Estimation via Importance Sampling

**Purpose.** Estimate a small probability $\theta=P(X>a)$ efficiently,
avoiding the catastrophic relative-error growth of naive Monte Carlo
described in Section 8.2.

**Input.** Threshold $a$; target density $f$; a shifted or tilted
proposal $p$ concentrating mass beyond $a$ (for example, $N(a,1)$ when
$f$ is standard Normal, or the exponentially tilted density of Unit
3.4); sample size $n$.

**Step-by-step procedure.**

1. Draw $Y_1,\dots,Y_n$ i.i.d. from $p$.
2. Compute $\hat\theta_{\text{IS}} = n^{-1}\sum_i \mathbf 1_{\{Y_i>a\}}
   \, w(Y_i)$, $w(x)=f(x)/p(x)$, using Algorithm 3.2 with
   $g(x)=\mathbf 1_{\{x>a\}}$.

**Pseudo-code.**

```
Algorithm: Rare-Event IS
Input:  a, f(), sampler for tilted proposal p, p(), n
Output: theta_hat, se

1. y[1..n] <- draw n iid values from p
2. w <- f(y) / p(y)
3. indicator <- (y > a)
4. gw <- indicator * w
5. theta_hat <- mean(gw)
6. se <- sd(gw) / sqrt(n)
7. Return theta_hat, se
```

**Flowchart description.** Start; draw samples from a proposal shifted
toward the rare region beyond $a$; weight each by the density ratio;
average the weighted indicators; end.

**Complexity.** $O(n)$, but with $n$ orders of magnitude smaller than
naive Monte Carlo would require for the same relative precision, as
quantified in Section 10.3.

**Practical considerations.** The proposal's tail must not be lighter
than $f$'s tail beyond $a$ (the same requirement noted for
acceptance-rejection envelopes in Unit 3.4), or the importance weights
will have infinite variance; a proposal that is itself a tilted or
shifted version of $f$, as constructed in Unit 3.4, generally
satisfies this requirement by design.

# R Programming

```{r monte-carlo-functions}
set.seed(123)

#' Basic Monte Carlo integration estimator
#'
#' Purpose: implements Algorithm 3.1.
#'
#' Arguments:
#'   sampler_function     zero-argument function returning one draw from f
#'   integrand_function   the function g to average
#'   sample_size          number of Monte Carlo draws
#'
#' Returns:
#'   a list with the point estimate and standard error
#'
#' Example:
#'   monte_carlo_integrate(function() runif(1), function(x) exp(-x), 2000)
monte_carlo_integrate <- function(sampler_function, integrand_function, sample_size) {
  simulated_draws <- vapply(seq_len(sample_size), function(i) sampler_function(), numeric(1))
  g_values <- integrand_function(simulated_draws)

  list(
    estimate = mean(g_values),
    standard_error = sd(g_values) / sqrt(sample_size)
  )
}

#' Importance sampling estimator
#'
#' Purpose: implements Algorithm 3.2.
#'
#' Arguments:
#'   proposal_sampler      zero-argument function returning one draw from p
#'   target_density        function evaluating f(x)
#'   proposal_density       function evaluating p(x)
#'   integrand_function     the function g
#'   sample_size            number of importance samples
#'
#' Returns:
#'   a list with the point estimate, standard error, and importance
#'   weights (useful for diagnosing weight degeneracy)
#'
#' Example:
#'   importance_sampling_estimate(
#'     function() rnorm(1, mean = 3, sd = 1),
#'     function(x) dnorm(x, 0, 1),
#'     function(x) dnorm(x, 3, 1),
#'     function(x) as.numeric(x > 3),
#'     2000
#'   )
importance_sampling_estimate <- function(proposal_sampler, target_density, proposal_density,
                                          integrand_function, sample_size) {
  simulated_draws <- vapply(seq_len(sample_size), function(i) proposal_sampler(), numeric(1))
  importance_weights <- target_density(simulated_draws) / proposal_density(simulated_draws)
  weighted_values <- integrand_function(simulated_draws) * importance_weights

  list(
    estimate = mean(weighted_values),
    standard_error = sd(weighted_values) / sqrt(sample_size),
    weights = importance_weights
  )
}

#' Self-normalized importance sampling estimator
#'
#' Purpose: implements Algorithm 3.3, for a target known only up to a
#' normalizing constant.
#'
#' Arguments:
#'   proposal_sampler        zero-argument function returning one draw from p
#'   unnormalized_target     function evaluating the unnormalized target
#'   proposal_density         function evaluating p(x)
#'   integrand_function       the function g
#'   sample_size              number of importance samples
#'
#' Returns:
#'   the self-normalized point estimate (numeric scalar)
#'
#' Example:
#'   self_normalized_is(
#'     function() rnorm(1, 0, 2),
#'     function(x) exp(-x^2 / 2),
#'     function(x) dnorm(x, 0, 2),
#'     function(x) x^2,
#'     2000
#'   )
self_normalized_is <- function(proposal_sampler, unnormalized_target, proposal_density,
                                integrand_function, sample_size) {
  simulated_draws <- vapply(seq_len(sample_size), function(i) proposal_sampler(), numeric(1))
  unnormalized_weights <- unnormalized_target(simulated_draws) / proposal_density(simulated_draws)

  sum(integrand_function(simulated_draws) * unnormalized_weights) / sum(unnormalized_weights)
}
```

# Simulation Studies

## Monte Carlo Confidence Interval Coverage (Theorem 3.1)

```{r mc-ci-coverage}
set.seed(123)

true_theta <- 1 - exp(-1)
n_trials <- 1000
sample_size_mc <- 500
coverage_hits <- logical(n_trials)

for (trial in seq_len(n_trials)) {
  draws <- runif(sample_size_mc)
  g_vals <- exp(-draws)
  est <- mean(g_vals)
  se <- sd(g_vals) / sqrt(sample_size_mc)
  coverage_hits[trial] <- (true_theta >= est - 1.96 * se) && (true_theta <= est + 1.96 * se)
}

tibble(
  Quantity = c("Nominal coverage", "Empirical coverage (1000 trials)"),
  Value = c(0.95, mean(coverage_hits))
) %>%
  kable(digits = 4, caption = "Coverage of the Monte Carlo 95% CLT-based confidence interval")
```

## Variance Reduction via Importance Sampling (Theorem 3.2)

```{r variance-reduction-simulation}
set.seed(123)

n_is_sim <- 20000

naive_estimate <- monte_carlo_integrate(
  function() rnorm(1, 0, 1), function(x) as.numeric(x > 3), n_is_sim
)

is_estimate <- importance_sampling_estimate(
  function() rnorm(1, mean = 3, sd = 1),
  function(x) dnorm(x, 0, 1),
  function(x) dnorm(x, 3, 1),
  function(x) as.numeric(x > 3),
  n_is_sim
)

true_p3 <- 1 - pnorm(3)

tibble(
  Method = c("Naive Monte Carlo", "Importance Sampling (N(3,1) proposal)"),
  Estimate = c(naive_estimate$estimate, is_estimate$estimate),
  Standard_Error = c(naive_estimate$standard_error, is_estimate$standard_error),
  True_Value = true_p3
) %>%
  kable(digits = 6, caption = "Estimating P(Z > 3): naive Monte Carlo vs. importance sampling, same n")
```

## Rare-Event Simulation at an Extreme Threshold

```{r rare-event-extreme}
set.seed(123)

extreme_threshold <- 4.5
true_p_extreme <- 1 - pnorm(extreme_threshold)

n_extreme <- 20000
naive_extreme <- monte_carlo_integrate(
  function() rnorm(1, 0, 1), function(x) as.numeric(x > extreme_threshold), n_extreme
)
is_extreme <- importance_sampling_estimate(
  function() rnorm(1, mean = extreme_threshold, sd = 1),
  function(x) dnorm(x, 0, 1),
  function(x) dnorm(x, extreme_threshold, 1),
  function(x) as.numeric(x > extreme_threshold),
  n_extreme
)

tibble(
  Method = c("Naive Monte Carlo", "Importance Sampling"),
  Estimate = c(naive_extreme$estimate, is_extreme$estimate),
  Standard_Error = c(naive_extreme$standard_error, is_extreme$standard_error),
  True_Value = true_p_extreme
) %>%
  kable(digits = 8, caption = "Estimating P(Z > 4.5) (true value approx 3.4e-6): naive MC typically observes zero hits")
```

The naive Monte Carlo estimator at this extreme threshold is very
likely to be exactly zero (since $n \times \theta \approx
20000\times 3.4\times10^{-6} \approx 0.07$, so a hit is rare even
across the whole sample), while the importance sampling estimator,
using a proposal centred at the threshold itself, reliably produces an
accurate, low-variance estimate with the same computational budget.

# Graphs

```{r importance-weights-diagnostic, fig.cap="Distribution of importance weights for the P(Z>3) estimation: weights are well-behaved (no extreme outliers), indicating a well-matched proposal."}
ggplot(tibble(weight = is_estimate$weights), aes(x = weight)) +
  geom_histogram(bins = 40, fill = "steelblue", colour = "white") +
  labs(title = "Importance weights, N(3,1) proposal targeting P(Z>3)", x = "Weight w(Y)", y = "Count") +
  theme_minimal(base_size = 11)
```

```{r mc-convergence-plot, fig.cap="Monte Carlo estimate of the integral of exp(-x) on [0,1] converging to the true value as n grows, with the CLT-based 95% band narrowing at rate 1/sqrt(n)."}
set.seed(123)

n_sequence <- round(10^seq(1, 4.3, length.out = 40))
running_draws <- runif(max(n_sequence))
running_g <- exp(-running_draws)
running_mean <- cumsum(running_g) / seq_along(running_g)
running_se <- sqrt(cumsum((running_g - mean(running_g))^2) / seq_along(running_g)^2)

trace_df <- tibble(
  n = seq_along(running_g),
  estimate = running_mean,
  se = running_se
) %>% filter(n %in% n_sequence)

ggplot(trace_df, aes(x = n, y = estimate)) +
  geom_ribbon(aes(ymin = estimate - 1.96 * se, ymax = estimate + 1.96 * se), fill = "steelblue", alpha = 0.3) +
  geom_line(colour = "steelblue") +
  geom_hline(yintercept = 1 - exp(-1), colour = "darkred", linetype = "dashed") +
  scale_x_log10() +
  labs(
    title = "Monte Carlo convergence for the integral of exp(-x) on [0,1]",
    x = "Sample size n (log scale)", y = "Estimate"
  ) +
  theme_minimal(base_size = 11)
```

# Numerical Examples

**Example 1: Basic Monte Carlo integration.**

*Problem.* Estimate $\theta=\int_0^1 e^{-x}\,dx$ (true value
$1-e^{-1}=0.632121$) using the five illustrative points
$u=0.1,0.3,0.5,0.7,0.9$.

*Solution.* $e^{-0.1}=0.904837$, $e^{-0.3}=0.740818$,
$e^{-0.5}=0.606531$, $e^{-0.7}=0.496585$, $e^{-0.9}=0.406570$; their
mean is

$$
\hat\theta = \frac{0.904837+0.740818+0.606531+0.496585+0.406570}{5}
= 0.631068.
$$

*Interpretation.* This mechanical illustration with five evenly
spaced (rather than genuinely random) points already lands close to
the true value $0.632121$ because $e^{-x}$ varies smoothly; a real
Monte Carlo estimate with a random sample of this size would typically
be noisier, which is exactly why Section 10 uses a much larger $n$
and reports a standard error alongside the point estimate.

**Example 2: Monte Carlo probability estimation.**

*Problem.* Estimate $\theta=P(U_1+U_2>1.5)$ for independent
$U_1,U_2\sim\text{Uniform}(0,1)$ (true value, by elementary geometry,
$(2-1.5)^2/2=0.125$) using the five illustrative pairs
$(0.8,0.9),(0.3,0.4),(0.6,0.95),(0.5,0.5),(0.9,0.7)$.

*Solution.* Sums: $1.7,\ 0.7,\ 1.55,\ 1.0,\ 1.6$; indicators (sum
$>1.5$): $1,0,1,0,1$; $\hat\theta = 3/5 = 0.6$.

*Interpretation.* With only $n=5$ points this estimate is far from
the true value $0.125$ — a deliberate illustration of the point made
by (3.7): the relative standard error of a probability estimate is
large unless $n$ is substantially larger than $1/\theta$; the R
simulation of Section 10 with thousands of draws is needed for an
estimate that can be trusted.

**Example 3: Importance sampling weight mechanics.**

*Problem.* For target $f=N(0,1)$, importance proposal $p=N(3,1)$
(used to estimate $P(Z>3)$), compute the importance weight $w(y)$ for
$y=3.2$ and $y=3.5$.

*Solution.* Following the same completing-the-square technique used
for the tilted Normal-tail envelope in Unit 3.4, the weight simplifies
to $w(y)=\varphi(y)/\varphi(y-3) = \exp(4.5-3y)$. For $y=3.2$:
$w=\exp(4.5-9.6)=\exp(-5.1)=0.006097$. For $y=3.5$:
$w=\exp(4.5-10.5)=\exp(-6)=0.002479$.

*Interpretation.* The weight is small precisely because $Y$ values
near $3.2$-$3.5$ are far more probable under the shifted proposal
$N(3,1)$ than under the original target $N(0,1)$; the weight corrects
for this over-representation so that the weighted average remains an
unbiased estimate of the original $P(Z>3)$.

**Example 4: Quantitative variance reduction.**

*Problem.* Using the simulation of Section 10.2 (estimating
$P(Z>3)$, true value $0.0013499$), compare the reported standard
errors of the naive and importance sampling estimators and compute
the variance reduction factor.

*Solution.* From the R output in Section 10.2, the naive Monte Carlo
standard error and the importance sampling standard error can be
compared directly; the ratio of their squares gives the variance
reduction factor achieved by importance sampling for the identical
sample size $n$.

*Interpretation.* The dramatic reduction — typically two to three
orders of magnitude at this threshold, and far more at the more
extreme threshold of Section 10.3 — reflects exactly the mechanism of
Theorem 3.2: the $N(3,1)$ proposal is shaped much more closely like
the ideal $p^{*}\propto \mathbf 1_{\{x>3\}}f(x)$ than the original
$N(0,1)$ target is.

# Applications

- **Statistics.** Monte Carlo estimation of $p$-values and power for
  test statistics whose null or alternative sampling distribution has
  no closed form.
- **Machine Learning.** Estimating expectations under complex,
  high-dimensional posterior or variational distributions, where
  importance sampling (and its sequential extensions) remains a
  building block despite the popularity of MCMC alternatives.
- **Data Science.** Estimating rare-event probabilities in large
  observational datasets (for example, fraud or anomaly rates) using
  reweighted subsampling schemes analogous to importance sampling.
- **Finance.** Monte Carlo option pricing under risk-neutral measures,
  and importance sampling for Value-at-Risk and Expected Shortfall
  estimation, where the risk of interest is, by construction, a rare
  event under the physical measure.
- **Reliability Engineering.** Importance sampling is the standard
  technique for estimating small system failure probabilities in
  structural reliability analysis, where naive simulation would
  require infeasibly many replications.
- **Monte Carlo Methods.** This unit's methods are the direct
  extension of the simulation techniques developed throughout Unit 3
  into the specific task of numerical integration.
- **Bayesian Statistics.** Self-normalized importance sampling
  (Section 8.1) estimates posterior expectations directly from an
  unnormalized posterior kernel, and importance sampling more broadly
  underlies particle filtering and sequential Monte Carlo methods.
- **Operations Research.** Estimating rare congestion, blocking, or
  failure probabilities in queueing and network reliability models via
  importance sampling.
- **Engineering.** Estimating the probability of exceeding a safety
  threshold under random loading, a rare-event problem structurally
  identical to the tail-probability examples of this unit.

# Advantages

- Monte Carlo integration's error rate, $O(1/\sqrt n)$, does not
  depend on the dimension of the integral, unlike quadrature methods,
  making it the practical choice for high-dimensional problems.
- Importance sampling can achieve dramatic variance reduction — often
  several orders of magnitude — for the same computational budget,
  when the proposal is well chosen.
- Self-normalized importance sampling extends the method to the
  common practical situation where the target density's normalizing
  constant is unknown.

# Limitations

- Monte Carlo integration converges more slowly than quadrature in low
  dimensions ($O(1/\sqrt n)$ versus $O(h^4)$ for Simpson's rule, for
  example), so quadrature remains preferable for smooth, low-
  dimensional integrands.
- A poorly chosen importance proposal can *increase* variance relative
  to naive Monte Carlo, or produce importance weights with infinite
  variance if the proposal's tail is lighter than the target's.
- The zero-variance ideal proposal of Theorem 3.2 is generally
  unattainable in practice, since constructing it requires already
  knowing the quantity being estimated.
- Self-normalized importance sampling is biased for finite samples,
  though the bias generally vanishes as $n\to\infty$.

# Practical Tips

- **Always report a standard error alongside a Monte Carlo estimate.**
  Theorem 3.1 makes this straightforward, and an estimate without an
  accompanying measure of its own uncertainty is difficult to
  interpret responsibly.
- **Diagnose importance sampling by inspecting the distribution of
  weights**, as in the histogram of Section 11. A small number of
  extremely large weights (weight degeneracy) signals a poorly matched
  proposal and an unreliable variance estimate, even if the point
  estimate looks plausible.
- **Match the proposal's tail behaviour to the target's**, exactly as
  required for acceptance-rejection envelopes in Unit 3.4; a
  lighter-tailed proposal risks infinite-variance importance weights.
- **For rare-event probabilities, shift or tilt the proposal toward
  the event of interest**, following the design principle of Theorem
  3.2, rather than relying on naive simulation, which becomes
  hopelessly inefficient as the event becomes rarer (Section 8.2).
- **Use self-normalized importance sampling whenever the target
  density's normalizing constant is unavailable**, as is routine in
  Bayesian posterior computation, rather than attempting to compute
  the constant separately.

# Summary

- Monte Carlo integration estimates $\theta=E_f[g(X)]$ by the sample
  average of $g$ evaluated at simulated draws from $f$; Theorem 3.1
  establishes its unbiasedness, consistency, and CLT-based standard
  error, with an error rate independent of dimension.
- Importance sampling estimates the same quantity by sampling from an
  alternative proposal $p$ and reweighting by $f/p$; Theorem 3.2
  characterizes the variance-minimizing proposal
  $p^{*}\propto|g|f$, achieving zero variance in the idealized case
  $g\ge0$.
- Self-normalized importance sampling extends the method to targets
  known only up to a normalizing constant, at the cost of a finite-
  sample bias that vanishes asymptotically.
- Naive Monte Carlo fails catastrophically for rare-event
  probabilities, with relative standard error growing like
  $1/\sqrt{n\theta}$; importance sampling with a proposal shifted or
  tilted toward the rare region (directly extending the exponential
  tilting technique of Unit 3.4) restores efficient estimation, as
  demonstrated quantitatively in this unit's simulation studies.

# Key Formulae

- Monte Carlo estimator: $\hat\theta_{\text{MC}} = n^{-1}\sum_i g(X_i)$,
  $X_i\sim f$.
- Monte Carlo standard error: $s_g/\sqrt n$.
- Importance sampling estimator: $\hat\theta_{\text{IS}} =
  n^{-1}\sum_i g(Y_i)w(Y_i)$, $w(x)=f(x)/p(x)$, $Y_i\sim p$.
- Variance-minimizing proposal: $p^{*}(x) \propto |g(x)|f(x)$.
- Self-normalized IS: $\hat\theta_{\text{SNIS}} = \dfrac{\sum_i
  g(Y_i)\tilde w(Y_i)}{\sum_i \tilde w(Y_i)}$.
- Rare-event relative error: $\text{se}(\hat\theta_{\text{MC}})/\theta
  \approx 1/\sqrt{n\theta}$.

# Key Terms

- **Importance density (proposal).** The density $p$ from which
  importance samples are actually drawn.
- **Importance sampling.** Estimating an expectation by sampling from
  an alternative distribution and reweighting.
- **Importance weight.** The density ratio $w(x)=f(x)/p(x)$ used to
  correct for sampling from $p$ instead of $f$.
- **Monte Carlo integration.** Estimating a definite integral or
  expectation by the sample average of simulated function evaluations.
- **Rare-event simulation.** Estimation of a very small probability,
  for which naive simulation is inefficient.
- **Self-normalized importance sampling.** An importance sampling
  variant usable when the target density is known only up to a
  constant.
- **Weight degeneracy.** The failure mode in which a small number of
  importance weights dominate the estimator, inflating its variance.

# Exercises

## Multiple Choice Questions

1. Monte Carlo integration estimates $\theta=\int g(x)f(x)dx$ as
   (a) the maximum of $g(X_i)$ over a sample
   (b) the sample average of $g(X_i)$, $X_i\sim f$
   (c) the median of $g(X_i)$
   (d) a quadrature sum on a fixed grid

2. The Monte Carlo estimator's standard error decreases with $n$ at
   rate
   (a) $O(1/n)$
   (b) $O(1/\sqrt n)$
   (c) $O(1/n^2)$
   (d) it does not decrease

3. The importance sampling weight is defined as
   (a) $p(x)/f(x)$
   (b) $f(x)/p(x)$
   (c) $f(x)p(x)$
   (d) $g(x)/f(x)$

4. The variance-minimizing importance density is proportional to
   (a) $f(x)$ alone
   (b) $p(x)$ alone
   (c) $|g(x)|f(x)$
   (d) $g(x)^2$

5. When $g(x)\ge0$ everywhere, sampling from the ideal importance
   density $p^{*}$ gives variance
   (a) equal to the naive Monte Carlo variance
   (b) exactly zero
   (c) infinite
   (d) undefined

6. Self-normalized importance sampling is used when
   (a) the proposal is unknown
   (b) the target density is known only up to a normalizing constant
   (c) $g$ is not integrable
   (d) $n$ is very small

7. The relative standard error of a naive Monte Carlo probability
   estimator behaves, for small $\theta$, like
   (a) $\sqrt{\theta/n}$
   (b) $1/\sqrt{n\theta}$
   (c) $n\theta$
   (d) $\theta^2/n$

8. A proposal with a lighter tail than the target can cause importance
   weights to have
   (a) zero variance
   (b) infinite variance
   (c) exactly unit variance
   (d) no effect on variance

9. Self-normalized importance sampling is
   (a) unbiased for every $n$
   (b) biased for finite $n$ but consistent
   (c) always inconsistent
   (d) identical to ordinary importance sampling

10. A histogram of importance weights showing a few extremely large
    values relative to the rest indicates
    (a) an ideal proposal
    (b) weight degeneracy, a poorly matched proposal
    (c) that self-normalization is unnecessary
    (d) that $n$ is too large

**Answers.** 1(b), 2(b), 3(b), 4(c), 5(b), 6(b), 7(b), 8(b), 9(b),
10(b).

## Short Answer Questions

1. Explain how a definite integral $\int_a^b h(x)dx$ can be rewritten
   as an expectation suitable for Monte Carlo estimation.
2. State the standard error formula for the simple Monte Carlo
   estimator and explain its dependence on $n$.
3. Define the importance weight and explain its role in correcting for
   sampling from $p$ rather than $f$.
4. State the variance-minimizing importance density from Theorem 3.2.
5. Explain why the zero-variance ideal importance sampler is not
   directly usable in practice.
6. Explain, using equation (3.7), why naive Monte Carlo performs
   poorly for rare-event probabilities.
7. What is weight degeneracy, and how can it be diagnosed?
8. Explain the difference between ordinary and self-normalized
   importance sampling, and when each is used.
9. Explain the connection between importance sampling and the
   exponential tilting technique of Unit 3.4.
10. Give one application, distinct from those emphasized in this unit,
    where importance sampling would be valuable.

## Long Answer Questions

1. State and prove Theorem 3.1 (unbiasedness and consistency of Monte
   Carlo estimators), including the importance sampling case.
2. State and prove Theorem 3.2 (the variance-minimizing importance
   density), explaining the role of the Cauchy-Schwarz inequality.
3. Derive equation (3.7) (the relative standard error of a naive
   Monte Carlo probability estimator) and explain its implications for
   rare-event simulation.
4. Derive the self-normalized importance sampling estimator (3.6) from
   first principles, explaining why it remains valid when only an
   unnormalized target is available.
5. Compare Monte Carlo integration and the quadrature methods of Unit
   4.2 in terms of convergence rate and dependence on dimension,
   explaining when each should be preferred.
6. Discuss the design principle "shape the proposal like $|g|f$" in
   practical terms, using the shifted-Normal proposal of this unit's
   examples as a concrete illustration.
7. Explain, with reference to Unit 3.4, how exponential tilting can be
   used as an importance sampling proposal for rare-event probability
   estimation, and why its efficiency improves as the event becomes
   rarer.
8. Discuss the risk of importance sampling producing an estimator with
   *higher* variance than naive Monte Carlo, and describe how to guard
   against this in practice.
9. Explain the role of Monte Carlo and importance sampling methods in
   Bayesian computation, distinguishing their use from Markov chain
   Monte Carlo methods.
10. Discuss, using the extreme-threshold simulation of Section 10.3,
    why naive Monte Carlo can return an estimate of exactly zero for a
    rare event, and why this is a more serious failure than merely a
    large standard error.

## Programming Exercises

1. Implement `monte_carlo_integrate()` from first principles (already
   provided) and reproduce the manual calculation of Example 1 with a
   large random sample, comparing the result to the true value.
2. Implement `importance_sampling_estimate()` from first principles
   (already provided) and reproduce the weight calculations of Example
   3 exactly.
3. Implement `self_normalized_is()` from first principles (already
   provided) and use it to estimate $E[X^2]$ for a target density
   known only up to a constant (for example, an unnormalized Normal
   kernel), comparing to the known closed-form answer.
4. Reproduce the variance reduction comparison of Section 10.2 for a
   different threshold (for example, $P(Z>2)$), and report the
   variance reduction factor achieved.
5. Investigate importance sampling failure: use a proposal with a
   *lighter* tail than the target (for example, target $t_3$, proposal
   $N(0,1)$) to estimate $E[X^2]$, and demonstrate empirically that the
   importance weights have very high (or, in the limit, infinite)
   variance, producing an unreliable estimator.
6. Use self-normalized importance sampling to estimate the posterior
   mean of a parameter under a simple Bayesian model (for example, a
   Beta-Binomial model) using only the unnormalized posterior kernel,
   and compare to the known closed-form posterior mean.
7. Write a function that automatically flags weight degeneracy (for
   example, by computing the effective sample size
   $n_{\text{eff}} = (\sum_i w_i)^2/\sum_i w_i^2$) and apply it to both
   a well-matched and a poorly matched importance sampling example.
8. Compare, using `system.time()`, the sample size required by naive
   Monte Carlo versus importance sampling to achieve the same
   standard error when estimating $P(Z>4)$, and report the
   computational savings.
9. Extend the rare-event simulation of Section 10.3 to a
   two-dimensional rare event (for example,
   $P(Z_1>3, Z_2>3)$ for independent standard Normals), using a
   bivariate shifted-Normal proposal.
10. Implement Monte Carlo integration of a genuinely
    high-dimensional integral (for example, the volume of a
    10-dimensional unit ball, with known closed-form value), and
    compare its accuracy and computational cost to what a naive grid-
    based quadrature approach would require, to illustrate the
    dimension-free convergence rate discussed in this unit.

## Mini Projects

1. **Importance sampling for option pricing.** Simulate the price of a
   deep out-of-the-money European call option under the standard
   Black-Scholes model (i.e., the terminal asset price is log-normal),
   where the option payoff is non-zero only in a rare tail event. Using
   naive Monte Carlo and an importance sampling scheme that shifts the
   simulated terminal price distribution toward the option's strike
   price, compare the standard errors of the two estimators for a
   fixed computational budget, and discuss the practical implications
   for pricing deep out-of-the-money options.

2. **A rare-event simulation toolkit with effective sample size
   diagnostics.** Build a small R toolkit implementing naive Monte
   Carlo, importance sampling, and self-normalized importance
   sampling for a class of rare-event probability estimation problems
   of your choosing (for example, tail probabilities of a Gamma or
   Weibull distribution relevant to reliability engineering). Include
   automatic computation of the effective sample size and a diagnostic
   warning when importance weights show signs of degeneracy. Apply
   your toolkit across a range of thresholds, from moderately rare to
   extremely rare, and produce a report documenting how the required
   proposal shift changes with the rarity of the event.

**Difficulty labels.** Multiple choice, short answer: **Easy**. Long
answer questions 3, 4, 6, 8, 9, programming exercises 1-6: **Medium**.
Long answer questions 1, 2, 5, 7, 10, programming exercises 7-10:
**Medium to Advanced**. Mini projects: **Advanced**.

# References

Glynn, P. W., & Iglehart, D. L. (1989). Importance sampling for
stochastic simulations. *Management Science*, 35(11), 1367-1392.

Owen, A. B. (2013). *Monte Carlo theory, methods and examples*.
Self-published.

Rizzo, M. L. (2019). *Statistical computing with R* (2nd ed.). CRC
Press.

Robert, C. P., & Casella, G. (2004). *Monte Carlo statistical
methods* (2nd ed.). Springer.

Ross, S. M. (2020). *Simulation* (6th ed.). Academic Press.
