---
title: "Lecture 3.4: The Acceptance-Rejection Method"
author: "Dr. Manoj C. Patil"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, fig.align = "center")
```

## Motivation

The inverse transformation method of Lecture 3.3 is elegant but has a
real limitation: it needs the quantile function $F_X^{-1}$, and many
important distributions — the Gamma distribution with non-integer
shape, the standard Normal, most Bayesian posterior densities — have
no closed-form quantile function at all. Numerically inverting the
cdf at every single draw is possible but slow and can be delicate.
The acceptance-rejection method sidesteps the problem entirely: it
needs no quantile function whatsoever, only the ability to *evaluate*
the target density (even up to an unknown constant) and to sample
from some other, more convenient distribution. The idea originates
with von Neumann's unpublished 1947 note on Monte Carlo methods for
the Manhattan Project (von Neumann, 1951).

## Intuition

Picture the graph of the target density $f$, sitting entirely beneath
the graph of some scaled *proposal* density $Mg(x)$, where $g$ is easy
to sample from and $M\ge1$ is large enough that $Mg$ never dips below
$f$. Suppose it is possible to scatter points uniformly at random in
the two-dimensional region lying *under* the curve $Mg$. Among these
scattered points, the ones that happen to also fall under the lower
curve $f$ are, geometrically, uniformly scattered under $f$ — and a
point picked uniformly from the region under a density curve has, as
its horizontal coordinate, precisely a random draw from that density.
Points falling above $f$ (but still under $Mg$) are simply thrown
away; the points that survive are exact draws from $f$, with no
approximation involved.

Turning this picture into an algorithm requires a practical way to
scatter points under $Mg$: draw the horizontal coordinate $Y$ from $g$
itself, then draw the vertical coordinate uniformly between $0$ and
$Mg(Y)$. This is equivalent to drawing an independent
$U\sim\text{Uniform}(0,1)$ and accepting the point exactly when
$U\le f(Y)/(Mg(Y))$ — a single numerical comparison, with no
reference to areas or integrals needed while the algorithm is
actually running.

## Setup

**Definition 4.1.** Let $f$ be a target density and $g$ a *proposal*
density from which sampling is easy, with a constant $M\ge1$ such that

$$
f(x)\le Mg(x) \qquad \text{for every } x.
$$

Draw $Y\sim g$ and $U\sim\text{Uniform}(0,1)$ independently; **accept**
$X=Y$ if $U\le f(Y)/(Mg(Y))$, otherwise discard both and repeat.

## Correctness

**Theorem 4.1.** The accepted variate $X$ has density exactly $f$, and
each trial is accepted with probability $1/M$.

*Proof.* Conditional on $Y=y$, the acceptance probability is
$f(y)/(Mg(y))$, a number in $[0,1]$ because of the bound
$f\le Mg$ assumed in Definition 4.1. Averaging this acceptance
probability over the distribution of $Y$ gives the unconditional
probability that any single trial succeeds:

$$
P(\text{accept}) = \int \frac{f(y)}{Mg(y)}\,g(y)\,dy = \frac1M\int f(y)\,dy = \frac1M,
$$

using $\int f(y)\,dy=1$ since $f$ is a density. This proves the second
claim. For the first claim, fix any $x$ and compute the probability
that a trial *both* has $Y\le x$ *and* is accepted:

$$
P(Y\le x,\ \text{accept}) = \int_{-\infty}^x \frac{f(y)}{Mg(y)}\,g(y)\,dy
= \frac{F(x)}{M},
$$

where $F$ is the cdf corresponding to $f$. By the definition of
conditional probability, the cdf of the *accepted* value $X$ is

$$
P(X\le x)=P(Y\le x\mid\text{accept}) = \frac{P(Y\le x,\ \text{accept})}{P(\text{accept})}=\frac{F(x)/M}{1/M}=F(x).
$$

Since this holds for every $x$, $X$ has cdf $F$, i.e. density $f$
exactly — not approximately, and regardless of how loosely or tightly
the envelope $Mg$ was chosen (as long as $f\le Mg$ holds everywhere).
$\blacksquare$

**Corollary.** Because each trial is accepted independently with
probability $1/M$, the number of trials needed to obtain one accepted
value follows a Geometric$(1/M)$ distribution, with mean exactly $M$.
This is the reason the constant $M$ is worth minimizing: it is
literally the average number of candidates that must be generated and
discarded for every one that is kept.

**Worked example.** Let $f(x)=6x(1-x)$ on $(0,1)$ (a Beta$(2,2)$
density, with maximum value $1.5$ at $x=0.5$), and take $g$ to be the
Uniform$(0,1)$ density with $M=1.5$. For a candidate $y=0.9$,
$f(0.9)=6(0.9)(0.1)=0.54$ and $Mg(0.9)=1.5$, so the acceptance ratio
is $0.54/1.5=0.36$. A draw $u=0.5$ exceeds $0.36$, so this candidate
is **rejected**; a draw $u=0.2$ does not exceed $0.36$, so the same
candidate would be **accepted**. Note that $y=0.9$ lies far from the
density's peak at $0.5$, which is exactly why its acceptance ratio of
$0.36$ is well below $1$ — candidates near the peak of $f$ are
accepted far more readily than candidates in the tails.

```{r ar-plot, fig.width=5, fig.height=3.2, fig.cap="Acceptance-rejection for f(x) = 6x(1-x) using a Uniform(0,1) envelope, M = 1.5; accepted points (red) fall under f, rejected points (grey) lie above it."}
set.seed(1)
n <- 400
y <- runif(n); v <- runif(n, 0, 1.5)
f <- 6 * y * (1 - y)
acc <- v <= f
x <- seq(0, 1, length.out = 200)
plot(x, 6 * x * (1 - x), type = "l", col = "steelblue", lwd = 2,
     ylim = c(0, 1.5), xlab = "x", ylab = "height")
abline(h = 1.5, col = "darkorange", lwd = 2)
points(y[!acc], v[!acc], pch = 16, cex = 0.4, col = "grey70")
points(y[acc], v[acc], pch = 16, cex = 0.4, col = "firebrick")
```

## Choice of $M$ and Exponential Tilting

By the corollary above, the average computational cost per accepted
value is exactly $M$, so the natural design goal is to make $M$ as
small as possible. The smallest valid constant is
$M^{*}=\sup_x f(x)/g(x)$; achieving a finite $M^{*}$ at all requires
$g$'s support and tails to be at least as generous as $f$'s — for
instance, a Normal proposal can never bound a Cauchy target, because
the Cauchy density decays far more slowly than the Normal density does
in the tails, so the ratio $f(x)/g(x)$ grows without bound.

**Exponential tilting** is a systematic way to build a good proposal
when the region of interest is a distribution's tail. It chooses a
proposal from the family $g_\theta(x)\propto e^{\theta x}h(x)$, for
some convenient base density $h$, and tunes the "tilt" $\theta$ so
that $g_\theta$ concentrates its mass exactly where the target region
lies. As a concrete instance, to sample the standard Normal tail
$\{Z>a\}$ efficiently for large $a$, take $h$ to be the Exponential
density shifted to start at $a$, with rate $\theta=a$; a short
calculation (completing the square in the exponent of the Normal
density) shows the resulting bounding constant is

$$
M^{*}=\frac{\varphi(a)}{a\,Q(a)}, \qquad Q(a)=P(Z>a),
$$

and by the classical Mills' ratio approximation $Q(a)\sim\varphi(a)/a$
for large $a$, this constant satisfies $M^{*}\to1$ as $a\to\infty$.
In words: the tilted method becomes *more* efficient the further out
into the tail one simulates, which is the exact opposite of naive
sampling from the untilted Normal distribution and discarding values
below the threshold — an approach whose acceptance probability $Q(a)$
shrinks toward zero as $a$ grows, making it hopelessly wasteful for
exactly the extreme, rare events that are often of the greatest
practical interest.

**Numerical illustration.** At threshold $a=2$, $\varphi(2)=0.05399$
and $Q(2)=P(Z>2)=0.02275$, giving
$M^{*}=0.05399/(2\times0.02275)=1.187$, so the tilted method accepts a
candidate with probability $1/M^{*}\approx0.843$. The naive method —
drawing an ordinary standard Normal value and keeping it only if it
exceeds $2$ — accepts with probability $Q(2)=0.02275$ directly. The
tilted method is therefore roughly $0.843/0.02275\approx37$ times more
efficient at this threshold alone, and the gap widens further as $a$
increases.

## References

von Neumann, J. (1951). Various techniques used in connection with
random digits. In *Monte Carlo Method* (National Bureau of Standards
Applied Mathematics Series 12, pp. 36–38). U.S. Government Printing
Office.
