---
title: "Lecture 3.1: Theory of Random Number Generation"
author: "Dr. Manoj C. Patil"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, fig.align = "center")
```

## Motivation

Every simulation — a Monte Carlo integral, a bootstrap resample, a
randomized algorithm — ultimately needs a long supply of numbers that
*behave* like independent draws from the Uniform$(0,1)$ distribution.
A computer cannot produce numbers that are genuinely random (its
arithmetic is entirely deterministic), so instead it produces numbers
that are deterministic but designed to be statistically
indistinguishable from random ones. Understanding exactly how this is
done, and exactly where it can go wrong, is the purpose of this
lecture.

## Intuition

Imagine a clock face with $m$ marks on it. Starting at some mark
$X_0$, suppose a rule says: multiply the current mark by $a$, add
$c$, and read off the remainder after dividing by $m$ — that
remainder becomes the next mark. Even though this rule is completely
mechanical, the sequence of marks visited can look erratic, jumping
around the clock face in no obvious pattern, *provided* $a$, $c$, and
$m$ are chosen well. If they are chosen badly, the sequence might
visit only a handful of marks before repeating itself, which would be
disastrous for a simulation expecting thousands of "independent"
values. The mathematics below answers, precisely, which choices of
$a,c,m$ guarantee that *every* mark on the clock face is visited
exactly once before the sequence repeats.

## Pseudo-Random Numbers

**Definition 1.1.** A *pseudo-random number generator* (PRNG) is a
deterministic recurrence $X_{n+1} = g(X_n)$ on a finite state space
$S = \{0,1,\dots,m-1\}$, together with a seed $X_0 \in S$ and an
output map $U_n = X_n/m \in [0,1)$, intended to imitate independent
Uniform$(0,1)$ draws. Because $S$ has only $m$ elements, the sequence
$(X_n)$ must eventually repeat itself (among any $m+1$ terms, two must
coincide); the smallest $P$ with $X_{n+P}=X_n$ for all large $n$ is
called the *period*. If $P=m$, every possible state is visited once
per cycle, and the generator is said to have **full period** — the
best a deterministic recurrence on $m$ states can achieve.

## The Linear Congruential Generator

**Definition 1.2.** The *linear congruential generator* (LCG) is the
clock-arithmetic rule described above,

$$
X_{n+1} = (aX_n+c) \bmod m,
$$

with multiplier $a$, increment $c$, and modulus $m$. Setting $c=0$
gives the *multiplicative* generator (MCG), historically the first
version proposed (Lehmer, 1949); keeping $c\ne0$ gives the *mixed*
generator, which — as the next theorem shows — can achieve full
period on *any* modulus, not only a prime one.

The following theorem is the central result of this lecture: it
converts the vague hope "the sequence should look random" into three
checkable arithmetic conditions on $a$, $c$, and $m$.

**Theorem 1.1 (Hull–Dobell, 1962).** The LCG has full period $m$ for
every seed if and only if

1. $\gcd(c,m)=1$;
2. $a\equiv 1 \pmod p$ for every prime $p\mid m$;
3. $a\equiv 1 \pmod 4$ if $4\mid m$.

*Proof.* The first step is to write down the state after $n$ steps
explicitly, rather than one step at a time. Unrolling the recurrence
$X_1=(aX_0+c)\bmod m$, $X_2=(aX_1+c)\bmod m,\dots$ and substituting
repeatedly gives the closed form

$$
X_n = \bigl(a^nX_0 + cS(n)\bigr)\bmod m, \qquad S(n)=\sum_{i=0}^{n-1}a^i,
$$

which can be checked by induction: it holds for $n=1$, and if it
holds for $n$ then $X_{n+1}=(aX_n+c)\bmod m =
\bigl(a^{n+1}X_0+c(aS(n)+1)\bigr)\bmod m$, and $aS(n)+1=S(n+1)$ by
direct expansion of the geometric sum.

Full period means that $n=m$ is the *smallest* positive integer for
which $X_n$ returns to $X_0$, and this must hold *simultaneously for
every possible seed* $X_0$ — not just for one lucky starting value.
Comparing the closed form above at two different seeds (or, equally,
comparing the coefficient of $X_0$ and the constant term separately)
shows that "$X_n\equiv X_0\pmod m$ for every $X_0$" is equivalent to
the pair of conditions $a^n\equiv1\pmod m$ and $cS(n)\equiv0\pmod m$
holding together, and full period asks that $n=m$ be the smallest $n$
for which both hold.

Now write $b=a-1$. Conditions (2) and (3) say exactly that $p\mid b$
for every prime $p$ dividing $m$, and that $4\mid b$ whenever $4\mid
m$. A classical number-theoretic fact (the "lifting-the-exponent"
lemma; see Knuth, 1997, §3.2.1.2) says that these two divisibility
requirements on $b$ are *precisely* what is needed to guarantee, for
every prime power $p^{e}$ exactly dividing $m$, that $p^{e}\mid S(n)$
as soon as $p^{e}\mid n$ — and no earlier. Piecing the prime-power
factors of $m$ back together with the Chinese Remainder Theorem then
shows $a^m\equiv1\pmod m$ and $m\mid S(m)$, with no smaller $n$
achieving both. Condition (1), $\gcd(c,m)=1$, is exactly what allows
$cS(n)\equiv0\pmod m$ to be replaced by the condition $m\mid S(n)$
regardless of the specific value of $c$. Putting these pieces
together shows the smallest valid $n$ is exactly $m$.

For the converse, if any one of the three conditions fails for some
prime power factor of $m$, the map $X\mapsto(aX+c)\bmod m$, restricted
to that factor, fails to be a single cycle through all its residues;
since the overall cycle structure on $\mathbb{Z}_m$ is built from the
cycle structures on its prime-power factors (again via the Chinese
Remainder Theorem), a shorter cycle on any one factor forces a period
shorter than $m$ overall. $\blacksquare$

**Worked example.** Take $a=5$, $c=3$, $m=16$, seed $X_0=7$. Since
$16=2^4$, the only prime factor is $p=2$; check $a-1=4$, which is
divisible by $2$ (condition 2) and by $4$ (condition 3, since $4\mid
16$); also $\gcd(3,16)=1$ (condition 1). All three conditions hold, so
the theorem guarantees full period $16$. Iterating by hand,

$$
X_1=(5\cdot7+3)\bmod16=38\bmod16=6,\quad
X_2=(5\cdot6+3)\bmod16=33\bmod16=1,
$$
$$
X_3=(5\cdot1+3)\bmod16=8,\qquad X_4=(5\cdot8+3)\bmod16=11,\qquad X_5=(5\cdot11+3)\bmod16=10,
$$

and continuing this process visits all $16$ residues $0,\dots,15$
exactly once before the sequence returns to $X_0=7$, exactly as
promised.

**Lemma 1.2.** If $m$ is prime, the MCG $X_{n+1}=aX_n \bmod m$ (seed
$X_0\ne0$) has period equal to the multiplicative order of $a$ modulo
$m$; this attains its maximum, $m-1$, exactly when $a$ is a
**primitive root** mod $m$.

*Proof.* Because $m$ is prime, the nonzero residues
$\{1,\dots,m-1\}$ form a group under multiplication modulo $m$ — the
group $(\mathbb Z/m\mathbb Z)^{*}$, of order $m-1$. The orbit
$X_n=a^nX_0$ returns to its starting value $X_0$ exactly when
$a^n\equiv1\pmod m$, that is, exactly when $n$ is a multiple of the
smallest such exponent, called the multiplicative order
$\operatorname{ord}(a)$. Lagrange's theorem for finite groups says
$\operatorname{ord}(a)$ must divide the group's order $m-1$, so the
period can never exceed $m-1$; it equals $m-1$ exactly when $a$'s
powers sweep out the *entire* group before repeating, which is the
definition of a primitive root. $\blacksquare$

**Worked example.** With $m=31$ (prime) and $a=3$, computing
successive powers of $3$ modulo $31$ — $3,9,27,19,26,16,17,20,29,25,
13,8,24,10,30,28,22,4,12,5,15,14,11,2,6,18,23,7,21,1$ — shows that $1$
first reappears only at the $30$-th power. Since $30=m-1$, this
confirms $3$ is a primitive root modulo $31$, and the corresponding
MCG has the maximal possible period, $30$.

## Why Full Period Is Not the Whole Story

**Remark.** Full period only guarantees that every state is visited
once per cycle; it says nothing about whether *consecutive* outputs
look independent of one another when viewed several at a time. The
generator RANDU ($a=65539$, $c=0$, $m=2^{31}$), once used widely on
IBM mainframes, illustrates this dramatically. Because
$a=2^{16}+3$, squaring gives $a^2=2^{32}+6\cdot2^{16}+9$, and since
$2^{32}\equiv0\pmod{2^{31}}$, this reduces to the exact identity
$a^2\equiv6a-9\pmod{2^{31}}$. Multiplying through by $X_n$ (recall the
generator is $X_{n+1}=aX_n\bmod2^{31}$, so $X_{n+2}=a^2X_n\bmod
2^{31}$) gives

$$
X_{n+2}\equiv 6X_{n+1}-9X_n \pmod{2^{31}}.
$$

**Numerical check.** With $a=65539$, direct computation gives
$a^2\bmod2^{31}=393225$ and $6a-9=6(65539)-9=393225$ — the two sides
agree exactly, confirming the identity is not an approximation but an
exact arithmetic fact about this specific multiplier.

In words: every *third* value in the sequence is an exact linear
combination of the two before it. Geometrically, this forces every
triple of consecutive outputs $(U_n,U_{n+1},U_{n+2})$, plotted as a
point in the unit cube, to lie on one of only $15$ parallel planes,
rather than filling the cube — a severe defect that is completely
invisible if one only looks at pairs of consecutive values, as the
figure below shows.

```{r randu-plot, fig.width=6, fig.height=3, fig.cap="Successive pairs from RANDU (left) versus a well-chosen generator (right); the planar defect of RANDU only becomes visible in three dimensions."}
mcg <- function(seed, a, m, n) {
  x <- numeric(n); cur <- seed
  for (i in 1:n) { cur <- (a * cur) %% m; x[i] <- cur }
  x / m
}
n <- 2000
randu <- mcg(1, 65539, 2^31, n)
good <- runif(n)
par(mfrow = c(1, 2), mar = c(4, 4, 2, 1))
plot(randu[-n], randu[-1], pch = 16, cex = 0.3, col = "firebrick",
     xlab = expression(U[n]), ylab = expression(U[n+1]), main = "RANDU")
plot(good[-n], good[-1], pch = 16, cex = 0.3, col = "steelblue",
     xlab = expression(U[n]), ylab = expression(U[n+1]), main = "runif()")
```

The lesson is that the Hull–Dobell theorem and Lemma 1.2 certify only
the *arithmetic* property of full period; whether the output also
*looks* statistically random — the subject of the next lecture — must
be checked separately.

## References

Hull, T. E., & Dobell, A. R. (1962). Random number generators. *SIAM
Review*, 4(3), 230–254.

Knuth, D. E. (1997). *The art of computer programming, Volume 2:
Seminumerical algorithms* (3rd ed.). Addison-Wesley.

Lehmer, D. H. (1949). Mathematical methods in large-scale computing
units. *Proceedings of the Second Symposium on Large-Scale Digital
Calculating Machinery*, 141–146.
