---
title: "Lecture 3.2: Testing of Random Number Generators"
author: "Dr. Manoj C. Patil"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, fig.align = "center")
```

## Motivation and Intuition

Lecture 3.1 showed that a generator can achieve the maximal possible
period yet still hide serious defects, as RANDU's collapse onto
parallel planes demonstrated. Since the arithmetic structure of a
generator does not, by itself, certify statistical quality, the only
remaining option is to examine the generator's *output* directly and
ask whether it is statistically distinguishable from a genuine
Uniform$(0,1)$ sample. Two quite different things can go wrong. First,
the individual values might not be spread evenly over $[0,1]$ — this
is a failure of **uniformity**. Second, even if each value looks fine
in isolation, the *order* in which they appear might not be random —
this is a failure of **independence**, and it is exactly the kind of
defect that trapped RANDU. This lecture develops one representative
test of each kind in full, together with brief descriptions of three
further independence tests.

## The Testing Problem

Given output $U_1,\dots,U_n$ of a generator, every test in this
lecture addresses the same joint null hypothesis:

$$
H_0:\quad U_1,\dots,U_n \overset{\text{iid}}{\sim}\text{Uniform}(0,1).
$$

Rejecting $H_0$ signals that the output can be statistically
distinguished from an ideal uniform sample; because the alternative is
left completely unspecified, no single test can reveal *which* aspect
of $H_0$ has failed, which is exactly why a battery of complementary
tests — some sensitive to shape, others to sequence — is used in
practice rather than any one test alone.

## Uniformity: The Chi-square Test

The most direct way to check whether values are spread evenly is to
bin them and compare the observed counts to what evenness would
predict. Partition $[0,1]$ into $k$ equal bins with observed counts
$O_i$. Under $H_0$, each of the $n$ values falls into each bin with
probability $1/k$, so $(O_1,\dots,O_k)\sim\text{Multinomial}(n;1/k,
\dots,1/k)$, and the discrepancy between observed and expected counts
is summarized by

$$
\chi^2=\sum_{i=1}^k\frac{(O_i-n/k)^2}{n/k}.
$$

**Theorem 2.1.** Under $H_0$, $\chi^2 \xrightarrow{d} \chi^2_{k-1}$ as
$n\to\infty$ with $k$ fixed.

*Proof.* By the multivariate central limit theorem applied to the
multinomial counts, the vector of standardized deviations converges,
$n^{-1/2}(O_i-n/k)_{i=1}^k \xrightarrow{d} N_k(0,\Sigma)$, with
covariance matrix $\Sigma=\tfrac1k I_k - \tfrac1{k^2}\mathbf{11}'$.
This matrix is *singular* — it annihilates the all-ones vector
$\mathbf 1$, since $\Sigma\mathbf1 = \tfrac1k\mathbf1 -
\tfrac1{k^2}\cdot k\cdot\mathbf1=\mathbf 0$ — and this singularity is
not an accident: it reflects the fact that the counts $O_i$ are not
free to vary independently, since they must always sum to the fixed
total $n$. Losing one "degree of freedom" to this constraint is
exactly why the limiting chi-square distribution below has $k-1$, not
$k$, degrees of freedom.

Writing $\chi^2$ as the quadratic form
$n^{-1}(O-n\mathbf p)'D^{-1}(O-n\mathbf p)$, with $D=\text{diag}(\mathbf
p)$ and $\mathbf p=(1/k,\dots,1/k)'$, one can check directly that the
matrix $D^{-1}\Sigma$ is idempotent (that is, $(D^{-1}\Sigma)^2 =
D^{-1}\Sigma$) with trace exactly $k-1$. A standard fact about
quadratic forms in a (possibly singular) multivariate Normal vector
states that if $V\sim N_k(0,\Sigma)$ and $D^{-1}\Sigma$ is idempotent
of rank $r$, then $V'D^{-1}V\sim\chi^2_r$; applying this fact to the
limiting Normal vector above, with $r=k-1$, gives the stated
convergence. $\blacksquare$

**Worked example.** Suppose $n=40$ generated values are placed into
$k=4$ equal bins with observed counts $O=(8,12,11,9)$. The expected
count per bin is $E=40/4=10$, so

$$
\chi^2=\frac{(8-10)^2}{10}+\frac{(12-10)^2}{10}+\frac{(11-10)^2}{10}+\frac{(9-10)^2}{10}=\frac{4+4+1+1}{10}=1.0.
$$

With $3$ degrees of freedom, the $5\%$ critical value is $7.815$;
since $1.0$ is far below this threshold, there is no evidence against
uniformity — the observed counts are well within the spread expected
by chance alone.

## Uniformity: The Kolmogorov–Smirnov Test

A drawback of the chi-square test is that it requires an arbitrary
choice of bins, and can be insensitive to smooth departures from
uniformity that happen not to line up with the bin edges. The
Kolmogorov–Smirnov test avoids binning entirely by comparing the
*entire* empirical distribution function to the target directly. With
empirical cdf $F_n(x)=n^{-1}\sum_i\mathbf1\{U_i\le x\}$, the statistic

$$
D_n=\sup_x|F_n(x)-x|
$$

measures the single largest vertical gap between the step function
$F_n$ and the straight line $F(x)=x$. Because $F_n$ only changes value
at the observed order statistics $U_{(1)}\le\cdots\le U_{(n)}$, the
supremum must be attained at one of these jump points (approached
either from the left or from the right), giving the computable form

$$
D_n = \max_{1\le i\le n}\max\Bigl\{\tfrac{i}{n}-U_{(i)},\ U_{(i)}-\tfrac{i-1}{n}\Bigr\}.
$$

Kolmogorov (1933) showed that $\sqrt n\,D_n$ converges to a specific
limiting distribution under $H_0$; the argument requires the theory of
weak convergence of the empirical process to a Brownian bridge, a
substantially deeper tool than anything used elsewhere in this
lecture, and the proof is omitted.

**Worked example.** For the small ordered sample $u_{(1)},\dots,
u_{(5)}=0.10,0.35,0.42,0.68,0.90$, compute both quantities at each
$i$:

| $i$ | $u_{(i)}$ | $i/5-u_{(i)}$ | $u_{(i)}-(i-1)/5$ |
|---|---|---|---|
| 1 | 0.10 | 0.10 | 0.10 |
| 2 | 0.35 | 0.05 | 0.15 |
| 3 | 0.42 | 0.18 | 0.02 |
| 4 | 0.68 | 0.12 | 0.08 |
| 5 | 0.90 | 0.10 | 0.10 |

The largest entry anywhere in the table is $0.18$, attained at $i=3$,
so $D_5=0.18$: the third order statistic falls short of its "expected"
position $3/5=0.6$ by this amount.

## Independence: The Runs Test

Uniformity tests say nothing about order. A **run** is a maximal
increasing or decreasing block in the sequence; intuitively, if the
values are independent, the sequence should switch between rising and
falling reasonably often, but not with perfect, suspicious regularity.

**Theorem 2.2.** Under $H_0$ (with a continuous distribution, so ties
have probability zero), the number of runs $R$ satisfies
$E(R)=(2n-1)/3$.

*Proof.* Call an interior position $i\in\{2,\dots,n-1\}$ a *turning
point* if $U_i$ is a local maximum or local minimum relative to its
two neighbours $U_{i-1}$ and $U_{i+1}$. Because $U_{i-1},U_i,U_{i+1}$
are three continuous, exchangeable random variables, all $3!=6$
orderings of their relative sizes are equally likely; exactly $4$ of
these $6$ orderings place $U_i$ either largest or smallest of the
three (i.e. at an extreme), so $P(\text{turning point at }i)=4/6=2/3$.
Summing this probability over the $n-2$ interior positions, the
expected number of turning points is $E(T)=(n-2)\cdot\tfrac23$.
Finally, every turning point marks the boundary between one run and
the next, so the number of runs is always exactly one more than the
number of turning points, $R=T+1$. Combining,

$$
E(R)=E(T)+1=\frac{2n-4}{3}+1=\frac{2n-1}{3}.
$$
$\blacksquare$

A more elaborate version of the same turning-point argument, tracking
how nearby turning-point indicators are correlated with one another,
gives $\operatorname{Var}(R)=(16n-29)/90$; the standardized statistic
$Z=(R-E(R))/\sqrt{\operatorname{Var}(R)}$ is then approximately
$N(0,1)$ for large $n$, allowing a formal test.

**Worked example.** For the ten values
$0.51,0.62,0.40,0.85,0.33,0.71,0.29,0.58,0.44,0.77$, the signs of
consecutive differences are $+,-,+,-,+,-,+,-,+$: the sequence switches
direction at *every single step*, so every sign is its own run,
giving $R=9$. With $n=10$, $E(R)=19/3=6.33$ and
$\operatorname{Var}(R)=131/90=1.456$, so

$$
Z=\frac{9-6.33}{\sqrt{1.456}}=2.21.
$$

Since $|Z|=2.21>1.96$, this is significant at the $5\%$ level: the
sequence alternates far more regularly than genuine randomness would
produce (recall that under $H_0$ a turning point occurs with
probability $2/3$ at each step, not with certainty), so this pattern
is exactly the kind of excessive negative dependence the runs test is
designed to catch.

```{r runs-plot, fig.width=5, fig.height=3.2, fig.cap="Simulated distribution of the number of runs (n = 50) against its Normal approximation from Theorem 2.2."}
count_runs <- function(u) {
  s <- sign(diff(u)); s <- s[s != 0]
  1 + sum(s[-1] != s[-length(s)])
}
set.seed(1)
n <- 50
r <- replicate(4000, count_runs(runif(n)))
hist(r, breaks = seq(min(r) - 0.5, max(r) + 0.5, 1), freq = FALSE,
     col = "steelblue", border = "white", main = "", xlab = "Number of runs R")
curve(dnorm(x, mean = (2 * n - 1) / 3, sd = sqrt((16 * n - 29) / 90)),
      add = TRUE, col = "darkred", lwd = 2)
```

## Three Further Independence Tests

The gap, poker, and serial correlation tests below are classical, all
traceable to Knuth's (1997) standard treatment of statistical tests
for random sequences.

- **Gap test.** Fix an interval $[\alpha,\beta]\subset[0,1]$ of
  probability $p=\beta-\alpha$. Under $H_0$, the number of values
  skipped between one visit to $[\alpha,\beta]$ and the next is
  Geometric$(p)$ — this is simply the waiting-time distribution for a
  Bernoulli$(p)$ "success" of landing in the interval. A chi-square
  test comparing observed gap-length frequencies to this geometric law
  reveals clustering (too many short gaps) or repulsion (too many long
  gaps) relative to independence.

  *Numerical illustration.* With $[\alpha,\beta]=[0,0.5)$, so $p=0.5$,
  suppose a sequence hits this interval at positions $3$, $6$, and
  $12$. The first gap skips the $2$ values at positions $4,5$, and the
  second gap skips the $5$ values at positions $7,\dots,11$. Under
  $H_0$, $P(\text{gap}=2)=(0.5)^2(0.5)=0.125$ and
  $P(\text{gap}=5)=(0.5)^5(0.5)=0.015625$; a full test would compare
  many such observed gap lengths, across a long sequence, to these
  geometric probabilities using a chi-square statistic.

- **Poker test.** Group $m$ consecutive decimal digits of each
  generated value into a "hand" and classify it by its repeated-digit
  pattern, exactly as one classifies a poker hand by rank. Under
  $H_0$, elementary counting gives fixed probabilities for each
  pattern (e.g. for $m=5$ digits: all different $0.3024$, one pair
  $0.5040$, two pairs $0.1080$, three of a kind $0.0720$, full house
  $0.0090$, four of a kind $0.0045$, five of a kind $0.0001$), and a
  chi-square test compares the observed hand frequencies to these.

  *Numerical illustration.* The value $u=0.42517$ has digits
  $4,2,5,1,7$ — all different, the most likely hand, with probability
  $0.3024$ under $H_0$. The value $u=0.33218$ has digits $3,3,2,1,8$ —
  the digit $3$ repeats once, giving a "one pair" hand, with
  probability $0.5040$ under $H_0$; a full poker test would classify
  many such hands this way and compare the resulting frequency table
  to these theoretical probabilities.

- **Serial correlation.** The lag-$k$ sample autocorrelation $r_k$
  measures the linear association between values $k$ steps apart;
  under $H_0$ it satisfies $r_k\sqrt{n-k}\xrightarrow{d}N(0,1)$, so
  values of $r_k$ far from zero (relative to $1/\sqrt{n-k}$) indicate
  that successive (or $k$-apart) outputs are not behaving
  independently.

  *Numerical illustration.* For the eight values $0.2, 0.5, 0.3, 0.8,
  0.4, 0.6, 0.1, 0.9$ (mean $\bar u=0.475$), the lag-$1$
  autocorrelation works out to $r_1=-0.634$, and
  $Z=r_1\sqrt{8-1}=-1.68$. Since $|Z|=1.68<1.96$, this particular
  short sequence does not give significant evidence of lag-$1$
  dependence at the $5\%$ level, even though $r_1=-0.634$ looks large
  numerically — with only $n=8$ values, sample correlations are highly
  variable under $H_0$, and the formal $Z$-test, not the raw
  correlation, is what should be reported.

Each of these targets a different notion of "structure," which is
precisely why applying several of them together, as in the batteries
used in practice, gives far more reliable protection than relying on
any single test.

## References

Kolmogorov, A. N. (1933). Sulla determinazione empirica di una legge
di distribuzione. *Giornale dell'Istituto Italiano degli Attuari*, 4,
83–91.

Knuth, D. E. (1997). *The art of computer programming, Volume 2:
Seminumerical algorithms* (3rd ed.). Addison-Wesley.
